### 2020-06-15 notes/notes.tex
June 15 notes

%        File: notes.tex
%     Created: Mon Jun 15 09:00 AM 2020 E
%

\documentclass[english]{article}
\usepackage[latin9]{inputenc}
\usepackage[letterpaper]{geometry}
\geometry{verbose}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage{subfigure}
\usepackage[english]{babel}
\usepackage{braket}
\usepackage{hyperref}
\newcommand{\bibliographytext}{References}

\begin{document}
\title{Research Notes}
\author{Stuart Thomas and Chris Monahan}
\date{Last updated \today}
\maketitle

\section{Monday, June 15}
\begin{enumerate}
    \item Preliminary meeting
    \item We will begin with $\phi^4$ term due to lower energy bound.
    \item Beginning with code in Python, switch to C/C++ if necessary.
\end{enumerate}

\end{document}

### 2020-06-16 notes/notes.tex
Tuesday, June 16 meeting Add references

@@
 
+\section{Tuesday, June 16}
+\begin{enumerate}
+
+    \item Preliminary concepts to understand: 
+        \begin{enumerate}
+            \item scalar field theory on lattice
+            \item Markov chains and Monte Carlo
+            \item the gradient flow
+            \item $O(n)$ symmetry
+        \end{enumerate}
+
+    \item General components to research, executed in parallel
+        \begin{enumerate}    
+            \item Reading
+            \item Mathematical analysis
+            \item Writing code
+            \item Present (writing, plot generation,\dots)
+        \end{enumerate}
+
+    \item Things to get out of dissertation \cite{Schaich2006}
+        \begin{enumerate}
+            \item Markov chain
+            \item Cluster algorithms
+
+            \begin{enumerate}
+                \item Metropolis is a method for narrowing possibilities by accepting only some changes.
+It can get stuck in a local mininum, loss of ergodicity.
+We solve these with cluster algoriths.
+Wolff grows clusters probabilistically and flips, while Swenson and Yang identifies clusters and flips them probabilistically.
+
+                \item Near a phase transition, correlation length grows and changes become less likely to be accepted: need clusters.
+Clusters dont work far from the phase transition.
+This is manifested as a sequence of a few metropolis steps and a cluster step.
+            \end{enumerate}
+        \end{enumerate}
+    \item Research plan
+        \begin{enumerate}
+            \item Start with 2D $\phi^4$
+
+            \begin{enumerate}
+                \item Set up lattice with sign flip for reflection
+                \item Use Markov chain Monte Carlo to simulate.
+                \item Measure, magnetization and suseptibility, Binder cumulant
+            \end{enumerate}
+            \item Transition to 3D, then maybe transition to C/C++.
+
+            \item Implement the gradient flow
+
+            \item move to 2-3d nonlinear sigma model.
+
+            \item Motivation: the nonlinear sigma model works for QCD given the asymtotic freedom. We may also want to explore topology.
+
+        \end{enumerate}
+\nocite{*}
+\bibliographystyle{plain}
+\bibliography{notes}
+
+\end{enumerate}
+
 \end{document}

### 2020-06-17 code/phi4.py
Initial Monte Carlo Code

@@ -1,0 +1,179 @@
+from random import randint, choices
+from colorsys import hls_to_rgb
+from copy import copy as copy
+from matplotlib import pyplot as plt
+import numpy as np
+import time
+from functools import lru_cache
+from matplotlib import pyplot as plt
+
+class Lattice(object):
+
+    rel_coords = [(-1,0),(0,-1),(1,0),(0,1)]
+
+    def __init__(self, dim):
+        self.dim = dim
+        self.size = dim**2
+        self.data = choices([True,False], k=self.size)
+
+    def __repr__(self):
+        return repr(self.data)
+
+    def __iter__(self):
+        return iter(self.data)
+
+    def __getitem__(self, i):
+        return self.data[i]
+
+    def __setitem__(self, i, val):
+        self.data[i] = val
+
+    def random(self):
+        return randint(0,self.size-1)
+
+    def coords(self):
+        return range(self.size)
+
+    def neighbors(self, coord):
+        d = self.dim
+        x = coord % d
+        y = coord // d
+
+        for i,j in Lattice.rel_coords:
+            if 0 < (x+i) + d*(y+j) < self.size:
+                yield (x+i) + d*(y+j)
+
+    def site_hamiltonian(self, site):
+        for neighbor in self.neighbors(site):
+                if self[site] ^ self[neighbor]:
+                    return 1
+                else:
+                    return -1
+
+    def hamiltonian(self):
+        energy = 0
+        J = 1
+        for coord in range(self.size):
+            energy += self.site_hamiltonian(coord)
+        return energy
+
+    def show(self, figsize=(6,6)):
+        M = np.transpose(np.resize( np.array(self.data), (self.dim, self.dim) ))
+        r = np.abs(M)
+        arg = np.angle(M)
+
+        # This code is generalized for complex fields
+        h = (arg + np.pi)  / (2 * np.pi) + 0.5
+        l = 1.0 - 1.0/(1.0 + r**1.0)
+        s = 0.8
+
+        c = np.vectorize(hls_to_rgb) (h,l,s) # --> tuple
+        c = np.array(c)  # -->  array of (3,n,m) shape, but need (n,m,3)
+        c = c.swapaxes(0,2)
+        plt.figure(figsize=figsize)
+        plt.imshow(c)
+        plt.show()
+
+    def magnetization(self):
+        return sum(self.data) / self.size
+
+
+
+class RandomWalk(object):
+    def __init__(self, lat, beta):
+        self.lattice = lat
[…]

### 2020-06-17 code/phi4.py
Wed June 17 notes

@@ -1,7 +1,6 @@
 from random import randint, choices
 from colorsys import hls_to_rgb
 from copy import copy as copy
-from matplotlib import pyplot as plt
 import numpy as np
 import time
 from functools import lru_cache
@@ -75,8 +74,7 @@ class Lattice(object):
         plt.show()
 
     def magnetization(self):
-        return sum(self.data) / self.size
-
+        return sum([1 if phi==True else -1 for phi in self]) / self.size
 
 
 class RandomWalk(object):
@@ -177,3 +175,4 @@ if __name__=="__main__":
     plt.show()
 
     print(l.hamiltonian())
+

### 2020-06-17 code/phi4.py
optimize energy calculation, add susceptibility

@@ -1,19 +1,25 @@
-from random import randint, choices
+from random import randint, choices, random
 from colorsys import hls_to_rgb
 from copy import copy as copy
 import numpy as np
 import time
 from functools import lru_cache
 from matplotlib import pyplot as plt
+import sys
 
 class Lattice(object):
 
     rel_coords = [(-1,0),(0,-1),(1,0),(0,1)]
 
-    def __init__(self, dim):
+    def __init__(self, dim, beta, J):
+        self.beta = beta
         self.dim = dim
         self.size = dim**2
         self.data = choices([True,False], k=self.size)
+        self.J = J
+        self.energy = 0
+        for coord in range(self.size):
+            self.energy += self.site_hamiltonian(coord)
 
     def __repr__(self):
         return repr(self.data)
@@ -26,6 +32,7 @@ class Lattice(object):
 
     def __setitem__(self, i, val):
         self.data[i] = val
+        self.update_energy(i)
 
     def random(self):
         return randint(0,self.size-1)
@@ -39,22 +46,22 @@ class Lattice(object):
         y = coord // d
 
         for i,j in Lattice.rel_coords:
-            if 0 < (x+i) + d*(y+j) < self.size:
+            if 0 <= x+i < self.dim and 0 <= y+j < self.dim:
                 yield (x+i) + d*(y+j)
 
     def site_hamiltonian(self, site):
+        e = 0
         for neighbor in self.neighbors(site):
                 if self[site] ^ self[neighbor]:
-                    return 1
+                    e +=  1
                 else:
-                    return -1
+                    e += -1
 
-    def hamiltonian(self):
-        energy = 0
-        J = 1
-        for coord in range(self.size):
-            energy += self.site_hamiltonian(coord)
-        return energy
+        return self.J * e
+
+    def update_energy(self, *sites):
+        for site in sites:
+            self.energy += 4 * self.site_hamiltonian(site)
 
     def show(self, figsize=(6,6)):
         M = np.transpose(np.resize( np.array(self.data), (self.dim, self.dim) ))
@@ -76,39 +83,42 @@ class Lattice(object):
     def magnetization(self):
         return sum([1 if phi==True else -1 for phi in self]) / self.size
 
+    def susceptibility(self):
+        m = self.magnetization()
+        m2 = 1
+        return self.size * self.beta * (m2 - m**2)
 
[…]

### 2020-06-19 code/phi4.py
Transition to phi4 model

@@ -1,25 +1,47 @@
-from random import randint, choices, random
+from random import randint, choices, random, seed
 from colorsys import hls_to_rgb
-from copy import copy as copy
+from copy import copy, deepcopy
 import numpy as np
 import time
 from functools import lru_cache
 from matplotlib import pyplot as plt
 import sys
+import imageio
 
 class Lattice(object):
 
     rel_coords = [(-1,0),(0,-1),(1,0),(0,1)]
 
-    def __init__(self, dim, beta, J):
-        self.beta = beta
-        self.dim = dim
-        self.size = dim**2
-        self.data = choices([True,False], k=self.size)
-        self.J = J
+    def __init__(self, m, l, data=None, dim=None):
+        self._construct( m, l, data, dim)
+
+    def _construct(self, m, l, data, dim):
+        self.m = m
+        self.l = l
+
+        if data:
+            self.data = data
+            self.dim = int(np.sqrt(len(data)))
+            self.size = len(data)
+
+        elif dim:
+            self.dim = dim
+            self.size = dim**2
+            self.data = [self._rand_init_val() for _ in range(self.size)]
+
+        else:
+            raise Exception("dim and data cannot both be None")
+
+        self.sites = list(range(self.size))
         self.energy = 0
         for coord in range(self.size):
             self.energy += self.site_hamiltonian(coord)
+
+        self.neighbors = lru_cache(maxsize=self.size) (self.neighbors)
+
+    def _rand_init_val(self):
+        # random value in [-1.5, 1.5)
+        return random() * 3 - 1.5 # EDIT
 
     def __repr__(self):
         return repr(self.data)
@@ -34,6 +56,12 @@ class Lattice(object):
         self.data[i] = val
         self.update_energy(i)
 
+    def __getstate__(self):
+        return self.m, self.l, self.data
+
+    def __setstate__(self, state):
+        self._construct(*state, None)
+
     def random(self):
         return randint(0,self.size-1)
 
@@ -44,26 +72,26 @@ class Lattice(object):
         d = self.dim
         x = coord % d
         y = coord // d
-
+        ns = []
         for i,j in Lattice.rel_coords:
-            if 0 <= x+i < self.dim and 0 <= y+j < self.dim:
-                yield (x+i) + d*(y+j)
+            ns.append( (x+i) % self.dim + d*((y+j) % self.dim) )
[…]

### 2020-06-19 code/phi4.py
a few typos

@@ -83,7 +83,7 @@ class Lattice(object):
         for neighbor in self.neighbors(site):
             e -= phi * self[neighbor]
 
-        e += 0.5 * self.m**2 * phi**2 + 0.25 * self.l * phi**4 # based on 7.16, QUESTION
+        e += (4 + 0.5 * self.m**2) * phi**2 + 0.25 * self.l * phi**4 # based on 7.16, QUESTION
 
         return e
 
@@ -131,7 +131,6 @@ class RandomWalk(object):
         self.lattice[i] =-self.lattice[i]
 
     def change(self, i):
-        self.flip(i)
         self.lattice[i] += 3 * random() - 1.5
 
     def metropolis(self):
@@ -209,7 +208,7 @@ class GifProducer(object):
 
 if __name__=="__main__":
     seed(a=14231)
-    L = 256
+    L = 128
     m = 1
     lam = 1
     wolff = False
@@ -230,14 +229,14 @@ if __name__=="__main__":
     record_rate = l.size
     thermalization = 200
     iter_index = []
-    gp = GifProducer()
+    # gp = GifProducer()
     def record_state(lat):
         iter_index.append(sweep_count)
         susceptibilities.append(l.susceptibility())
         binder_cums.append(l.binder_cumulant())
         magnetizations.append(abs(l.magnetization()))
         energies.append(l.energy/l.size)
-        gp.save_lat(l)
+        # gp.save_lat(l)
 
     record_state(l)
     timeout = 100
@@ -266,7 +265,7 @@ if __name__=="__main__":
     exec_time = time.time() - start
     print(f"Done in {exec_time}")
 
-    gp.save("test.gif")
+    # gp.save("test.gif")
 
     fig, (ax1, ax2, ax3, ax4) = plt.subplots(4,1, figsize=(18,12), sharex = True)
     ax1.plot(iter_index, magnetizations)
@@ -280,8 +279,8 @@ if __name__=="__main__":
     ax4.set_ylabel("Binder Cumulant")
     ax4.set_xlabel("Sweep")
     # ax3.set_ylim((-0.05,1.05))
-    ax1.set_title(f"Monte Carlo Simulation of Ising Model using Metropolis and {'Wolff' if wolff else 'Swenson-Wang'} Algorithms, $L={L}$, $\\lambda={lam}$, $\\mu_0^2={m}$, $t={exec_time:.1f}s$")
[…]

### 2020-06-19 notes/notes.tex
Wed June 17 notes
June 19 notes

@@
         \item Start with 2D $\phi^4$
-
         \begin{enumerate}
@@
         \item Transition to 3D, then maybe transition to C/C++.
-
         \item Implement the gradient flow
-
         \item move to 2-3d nonlinear sigma model. 
-
         \item Motivation: the nonlinear sigma model works for QCD given the asymtotic freedom. We may also want to explore topology.
@@
     \end{enumerate}
+\end{enumerate}
+
+\section{Wednesday, June 17}
+\begin{enumerate}
+    \item Code tips:
+    \begin{enumerate}
+        \item Try Swendsen-Wang algorithm in addition to Wolff
+        \item Print out time taken
+        \item Optimize Hamiltonian
+        \item Save every tenth measurement or store configurations to calculate path integral. Exclude thermaliztion (first 200)
+        \item Write in terms of sweeps, not iterations
+        \item Parallelize (look into checkerboard algorithm)
+        \item Implement Binder cumulant and suseptibility
+        \item Store every few states
+        \item Look into multigrid algorithm
+
+    \end{enumerate}
+
+    \item Reading on Monte Carlo Markov chain and cluster algorithms.
+\end{enumerate}
+
+\section{Friday, June 19}
+    \begin{enumerate}
+        \item Looking over code
+        \begin{enumerate}
+            \item Might be too slow, move to C/C++ eventually?
+            \item Why is the energy increasing with metropolis algorithm?
+            \item Shift to using action instead of Hamiltonian.
+            \item Transition from Broken Phase, look at $\mu_0^2$ term.
+            \item Profile code for possible optimizations.
+        \end{enumerate}
+    \end{enumerate}
+
 \nocite{*}
@@
 \bibliography{notes}
-
-\end{enumerate}
-
 \end{document}

### 2020-06-19 code/phi4.py
optimize

@@ -2,15 +2,23 @@ from random import randint, choices, random, seed
 from colorsys import hls_to_rgb
 from copy import copy, deepcopy
 import numpy as np
-import time
+from time import time, sleep
 from functools import lru_cache
 from matplotlib import pyplot as plt
 import sys
 import imageio
+import timeit
+from math import exp, sqrt
+
+if 'line_profiler' not in globals():
+    def profile(f):
+        return f
+
+
 
 class Lattice(object):
 
-    rel_coords = [(-1,0),(0,-1),(1,0),(0,1)]
+    rel_coords = ((-1,0),(0,-1),(1,0),(0,1))
 
     def __init__(self, m, l, data=None, dim=None):
         self._construct( m, l, data, dim)
@@ -19,9 +27,12 @@ class Lattice(object):
         self.m = m
         self.l = l
 
+        self.__redefined_mass = 4 + 1/2 * self.m
+        self.__quarter_lambda = 1/4 * self.l
+
         if data:
             self.data = data
-            self.dim = int(np.sqrt(len(data)))
+            self.dim = int(sqrt(len(data)))
             self.size = len(data)
 
         elif dim:
@@ -33,15 +44,14 @@ class Lattice(object):
             raise Exception("dim and data cannot both be None")
 
         self.sites = list(range(self.size))
-        self.energy = 0
-        for coord in range(self.size):
-            self.energy += self.site_hamiltonian(coord)
-
+        self.__lagrangian_cache = [self.lagrangian(i) for i in self.sites]
+        self.__action = sum(self.__lagrangian_cache)
         self.neighbors = lru_cache(maxsize=self.size) (self.neighbors)
+        self.__previous_state = None
 
     def _rand_init_val(self):
         # random value in [-1.5, 1.5)
-        return random() * 3 - 1.5 # EDIT
+        return random() * 3 - 1.5
 
     def __repr__(self):
         return repr(self.data)
@@ -52,15 +62,38 @@ class Lattice(object):
     def __getitem__(self, i):
         return self.data[i]
 
-    def __setitem__(self, i, val):
-        self.data[i] = val
-        self.update_energy(i)
+    def __setitem__(self, i, new_phi):
+        phi = self[i]
+        self.__previous_state = {
+                'i' : i,
+                'phi' : phi,
+                'lagrangian' : self.__lagrangian_cache[i],
[…]

### 2020-06-19 code/test.py

@@ -1,0 +1,22 @@
+from timeit import timeit
+from matplotlib import pyplot as plt
+import math
+import numpy as np
+
+times = list(range(1,50))
+reassign = []
+no_reassign = []
+def numpy():
+    np.exp(-1000.0)
+
+def math():
+    math.exp(-1000.0)
+
+for t in times:
+    reassign.append( timeit(math, globals=dict( t=t)))
+    no_reassign.append( timeit(numpy, globals=dict(t=t)))
+
+plt.plot(times,reassign, label="math")
+plt.plot(times,no_reassign, label="numpy")
+plt.legend()
+plt.show()

### 2020-06-22 notes/notes.tex
June 22 Notes; gitignore plots

@@
             \item Transition from Broken Phase, look at $\mu_0^2$ term.
-            \item Profile code for possible optimizations.
+            \item Profile code for possible optimizations. % http://users.physik.fu-berlin.de/~kleinert/kleiner_reb8/psfiles/phi4.pdf
+        \end{enumerate}
+    \end{enumerate}
+
+ \section{Monday, June 22}
+    \begin{enumerate}
+        \item Coding
+        \begin{enumerate}
+            \item Add plots to \texttt{.gitignore}.
+            \item Try parallelizing code, using either multigrid or checkerboard algorithm
+        \end{enumerate}
+
+        \item Reading
+        \begin{enumerate}
+            \item Start to focus more on understanding the theory behind research.
+            \item Read dissertation \cite{Schaich2006} Chap. 6.5 and 6.6, take notes on questions.
+            \item Newman \cite{Newman1999} (Main textbook for Monte Carlo in Statistical Physics)
+        \end{enumerate}
+
+        \item Just some things to remember
+        \begin{enumerate}
+            \item Correlation functions correlate values in statistical systems and relate to propagators in QFT.
+            \item The problem of renormalization: $a\rightarrow0$ leads to unbounded correlation function.
+As real physical lengths are measured in terms of the lattice constant, these sometimes tend to infinity.
         \end{enumerate}

### 2020-06-22 code/phi4.py

@@ -254,7 +254,7 @@ if __name__=="__main__":
     L = 248
     m = 1
     lam = 1
-    wolff = True
+    wolff = False
     l = Lattice(dim=L, m=m, l=lam)
 
     rw = RandomWalk(l)
@@ -324,7 +324,7 @@ if __name__=="__main__":
     ax4.set_xlabel("Sweep")
     # ax3.set_ylim((-0.05,1.05))
     ax1.set_title(f"Monte Carlo Simulation of $\phi^4$ Model using Metropolis and {'Wolff' if wolff else 'Swenson-Wang'} Algorithms, $L={L}$, $\\lambda={lam}$, $\\mu_0^2={m**2}$, $t={exec_time:.1f}s$")
-    plt.savefig('plots/temp.png')
+    plt.savefig('plots/phi_248_sw.png')
     # plt.show()
 
 

### 2020-07-07 notes/notes.tex
Wednesday notes
tuesday, june 30 notes
convert to MPI

@@
 
+ \section{Wednesday, June 24}
+    \begin{enumerate}
+
+        \item Code
+        \begin{enumerate}
+            \item Parallelize
+            \item Transition to numpy
+        \end{enumerate}
+        \item Theory Question
+        \begin{enumerate}
+            \item renormalize and regularlize: what do they mean?
+            \item Look at LePage
+        \end{enumerate}
+
+    \end{enumerate}
+
+ \section{Tuesday, June 30}
+    \begin{enumerate}
+        \item Code
+        \begin{enumerate}
+            \item Try \texttt{mpi4py}.
+            \item Move to 3D (this may decrease parallel overhead)
+        \end{enumerate}
+
+        \item Reading
+        \begin{enumerate}
+            \item Continue Reading Collins, others. 
+        \end{enumerate}
+    \end{enumerate}
+
+ \section{Thursday, July 2}
+    \begin{enumerate}
+        \item Coding
+            \begin{enumerate}
+                \item Continue implementing MPI
+                \item Parallelization may be more apparent in 3D
+            \end{enumerate}
+        \item Reading
+            \begin{enumerate }
+                \item Dirac fermions, represented by 4D spinor field (spin up/down, electron/positron)    
+                \item Look at Tong (Chap. 4)
+                \item Charge (See Tong, Noether's Theorem)
+            \end{enumerate}
+
+        \item Next week back: start gradient flow on linear phi4 model
+
+    \end{enumerate}
+
 \nocite{*}

### 2020-07-07 code/keys.py

@@ -1,0 +1,4 @@
+WHITE = 0
+BLACK = 1
+WOLFF = 0
+SWENDSEN_WANG = 1

### 2020-07-07 code/lattice.py

@@ -1,0 +1,155 @@
+from functools import lru_cache
+from itertools import product
+from colorsys import hls_to_rgb
+from random import randrange
+
+from matplotlib import pyplot as plt
+import numpy as np
+from mpi4py import MPI
+
+from keys import *
+
+COMM = MPI.COMM_WORLD
+RANK = COMM.Get_rank()
+SIZE = COMM.Get_size()
+
+REL_COORDS = ((-1,0),(0,-1),(1,0),(0,1))
+
+
+class Lattice(object):
+
+    def __init__(self, m, l, data=None, dim=None):
+        self._construct( m, l, data, dim)
+
+
+    def _construct(self, m, l, data, dim):
+        self.m = m
+        self.l = l
+
+        self.__redefined_mass = 4 + 1/2 * self.m
+        self.__quarter_lambda = 1/4 * self.l
+
+        if data is not None:
+            self.data = data
+            self.dim = data.shape[0]
+            self.size = data.size
+
+        elif dim is not None:
+            self.dim = dim
+            self.size = dim**2
+            if RANK==0:
+                self.data = self.rand_dist(np.random.rand(self.dim, self.dim)) # Hot start
+            else:
+                self.data = np.empty((self.dim, self.dim))
+
+        else:
+            raise Exception("dim and data cannot both be None")
+
+        self.__neighbors = lru_cache(maxsize=self.size) (self.__neighbors)
+        self.sites = list(product(range(self.dim), repeat=2))
+        # populate cache
+        # for s in self.sites:
+            # self.__neighbors(s)
+        if RANK==0:
+            self.calculate_action()
+        else:
+            self.action=0
+
+        self.mpi_assignment = {}
+        if RANK>0:
+            global site_arrays
+            for color in [WHITE, BLACK]:
+                all_sites = list(self.checker_iter(color))
+                split_sites = np.array_split(all_sites, SIZE-1)
+                self.mpi_assignment[color] = split_sites[RANK-1]
+
+    @staticmethod
+    def rand_dist(r):
+        return 3 * r - 1.5
+
+    @staticmethod
+    def static_lagrangian(phi, neighbor_phis, redef_mass, quarter_lam):
+        return -phi * sum(neighbor_phis) + redef_mass * phi**2 + quarter_lam * phi**4
+
+    def calculate_action(self):
+        self.action = sum(self.lagrangian(c) for c in self.sites)
+
[…]

### 2020-07-07 code/phi4.py

@@ -1,330 +1,99 @@
-from random import randint, choices, random, seed
-from colorsys import hls_to_rgb
-from copy import copy, deepcopy
-import numpy as np
+# from random import randint, choices, random, seed, randrange
+# from copy import copy, deepcopy
+# import numpy as np
 from time import time, sleep
-from functools import lru_cache
-from matplotlib import pyplot as plt
+# from functools import lru_cache
+# from itertools import product
 import sys
-import imageio
-import timeit
-from math import exp, sqrt
+# import imageio
+# import timeit
+# from math import exp, sqrt
+# import inspect
+# import tempfile
+# import multiprocessing as mp
+# import pickle
+import os
+from getopt import getopt
+# import tables
+# from dataclasses import dataclass
+# from typing import Callable
+# import dill
+# from matplotlib import pyplot as plt
+from mpi4py import MPI
+# from datetime import datetime
+
+from lattice import Lattice
+from random_walk import RandomWalk
+from keys import *
+from recorder import Recorder
+
+
+COMM = MPI.COMM_WORLD
+RANK = COMM.Get_rank()
+SIZE = COMM.Get_size()
+
 
 if 'line_profiler' not in globals():
     def profile(f):
         return f
 
+MPI_RES_CHUNKSIZE = 0
 
+def main():
+    os.environ['OMP_NUM_THREADS'] = '1'
+    os.environ['NUMEXPR_NUM_THREADS'] = '1'
+    os.environ['MKL_NUM_THREADS'] = '1'
 
-class Lattice(object):
+    opts, args = getopt(sys.argv[1:],'pL:s:S')
+    dopts = dict(opts)
 
-    rel_coords = ((-1,0),(0,-1),(1,0),(0,1))
+    if '-s' in dopts:
+        seed = int(dopts['-s'])
+        seed(a=seed)
+        np.random.seed(seed)
 
-    def __init__(self, m, l, data=None, dim=None):
-        self._construct( m, l, data, dim)
-
-    def _construct(self, m, l, data, dim):
-        self.m = m
-        self.l = l
-
-        self.__redefined_mass = 4 + 1/2 * self.m
-        self.__quarter_lambda = 1/4 * self.l
-
-        if data:
-            self.data = data
-            self.dim = int(sqrt(len(data)))
-            self.size = len(data)
-
-        elif dim:
-            self.dim = dim
-            self.size = dim**2
-            self.data = [self._rand_init_val() for _ in range(self.size)]
-
-        else:
[…]

### 2020-07-07 code/phi4_serial.py

@@ -1,0 +1,361 @@
+from random import randint, choices, random, seed, randrange
+from colorsys import hls_to_rgb
+from copy import copy, deepcopy
+import numpy as np
+from time import time, sleep
+from functools import lru_cache
+from matplotlib import pyplot as plt
+from functools import partial
+import sys
+import imageio
+import timeit
+from math import exp, sqrt
+import multiprocessing as mp
+import pickle
+import inspect
+
+if 'line_profiler' not in globals():
+    def profile(f):
+        return f
+
+
+class Lattice(object):
+
+    rel_coords = ((-1,0),(0,-1),(1,0),(0,1))
+
+    def __init__(self, m, l, data=None, cache=None, dim=None):
+        self._construct( m, l, data, cache, dim)
+
+    def _construct(self, m, l, data, cache, dim):
+        self.m = m
+        self.l = l
+
+        self.__redefined_mass = 4 + 1/2 * self.m
+        self.__quarter_lambda = 1/4 * self.l
+
+        if data:
+            self.data = data
+            self.dim = int(sqrt(len(data)))
+            self.size = len(data)
+
+        elif dim:
+            self.dim = dim
+            self.size = dim**2
+            self.data = [self._rand_init_val() for _ in range(self.size)]
+
+        else:
+            raise Exception("dim and data cannot both be None")
+
+        self.sites = list(range(self.size))
+        if cache is None:
+            self.__lagrangian_cache = [self.lagrangian(i) for i in self.sites]
+        else:
+            self.__lagrangian_cache = cache
+        self.__action = sum(self.__lagrangian_cache)
+        self.neighbors = lru_cache(maxsize=self.size) (self.neighbors)
+        self.__previous_state = None
+
+    def full_action(self):
+        self.__lagrangian_cache = [self.lagrangian(i) for i in self.sites]
+        self.__action = sum(self.__lagrangian_cache)
+        return self.__action
+
+
+    def _rand_init_val(self):
+        # random value in [-1.5, 1.5)
+        return random() * 3 - 1.5
+
+    def __repr__(self):
+        return repr(self.data)
+
+    def __iter__(self):
+        return iter(self.data)
+
+    def __getitem__(self, i):
+        return self.data[i]
+    @profile
+    def __setitem__(self, i, new_phi):
+        phi = self[i]
+        self.__previous_state = {
+                'i' : i,
[…]

### 2020-07-07 code/random_walk.py

@@ -1,0 +1,129 @@
+from random import random
+from math import exp
+
+import numpy as np
+from mpi4py import MPI
+
+COMM = MPI.COMM_WORLD
+RANK = COMM.Get_rank()
+SIZE = COMM.Get_size()
+
+if 'line_profiler' not in globals():
+    def profile(f):
+        return f
+
+
+class RandomWalk(object):
+    def __init__(self, lattice):
+        self.lat = lattice
+        self.mpi_chunksize = lattice.size ** 2 // 2 // SIZE // 2
+
+    @profile
+    def checker_met(self, site, color):
+        neighbor_phis = tuple(self.lat.data[n] for n in self.lat.neighbors(site))
+        old_action = self.lat.action
+        old_L = self.lat.lagrangian(site)
+        dphi = self.lat.rand_dist(random())
+        self.lat.data[site] += dphi
+        new_L = self.lat.lagrangian(site)
+        dS = new_L - old_L
+        for nphi in neighbor_phis:
+            dS -= nphi * dphi
+
+        if dS < 0 or random() <= exp(-dS): # EDIT
+            return dphi, dS
+        else:
+            self.lat.data[site] -= dphi
+            self.lat.action = old_action
+            return None
+
+
+    @profile
+    def half_checkerboard(self, color):
+        COMM.Bcast(self.lat.data, root=0)
+        self.lat.action = COMM.bcast(self.lat.action, root=0)
+
+        # One node compiles changes from other nodes
+        if RANK==0:
+            running_nodes = SIZE - 1
+            while running_nodes > 0:
+                data = COMM.recv()
+                xs = data['xs']
+                ys = data['ys']
+                dphis = data['dphis']
+                dS = data['dS']
+
+                self.lat.data[xs, ys] += dphis
+                self.lat.action += dS
+                if data['done']:
+                    running_nodes -= 1
+
+        else:
+            size = 0
+            xs = []
+            ys = []
+            dphis = []
+            tot_dS = 0
+            for site in self.lat.mpi_assignment[color]:
+                res = self.checker_met(tuple(site), color)
+                if res is not None:
+                    dphi, dS = res
+                    xs.append(site[0])
+                    ys.append(site[1])
+                    dphis.append(dphi)
+                    tot_dS += dS
+                    size += 1
+
+                    if size % self.mpi_chunksize == 0:
[…]

### 2020-07-07 code/recorder.py

@@ -1,0 +1,66 @@
+from mpi4py import MPI
+from matplotlib import pyplot as plt
+import imageio
+
+COMM = MPI.COMM_WORLD
+RANK = COMM.Get_rank()
+
+class Recorder(object):
+    def __init__(self):
+        if RANK==0:
+            self.gp = GifProducer()
+            self.iter_index = []
+            self.magnetizations = []
+            self.energies = []
+            self.susceptibilities = []
+            self.binder_cums = []
+            self.states = []
+            self.record_count = 0
+            self.thermalization = 50
+
+    def save(self, l):
+        if RANK==0:
+            self.iter_index.append(self.record_count)
+            self.susceptibilities.append(l.susceptibility())
+            self.binder_cums.append(l.binder_cumulant())
+            self.magnetizations.append(abs(l.magnetization()))
+            l.calculate_action()
+            self.energies.append(l.action/l.size)
+            self.gp.save_lat(l)
+            self.record_count += 1
+
+    def save_gif(self, fname):
+        if RANK==0:
+            self.gp.save(fname)
+
+    def plot(self, L, m, lam, exec_time, show):
+        if RANK==0:
+            x = self.iter_index
+            fig, (ax1, ax2, ax3, ax4) = plt.subplots(4,1, figsize=(18,12), sharex = True)
+            ax1.plot(x, self.magnetizations)
+            ax2.plot(x, self.energies)
+            ax3.plot(x, self.susceptibilities)
+            ax4.plot(x, self.binder_cums)
+
+            ax1.set_ylabel("Magnetization")
+            ax2.set_ylabel("Total Action")
+            ax3.set_ylabel("Susceptibility")
+            ax4.set_ylabel("Binder Cumulant")
+            ax4.set_xlabel("Sweep")
+            # ax3.set_ylim((-0.05,1.05))
+            ax1.set_title(f"Monte Carlo Simulation of $\phi^4$ Model using Metropolis and Wolff Algorithms, $L={L}$, $\\lambda={lam}$, $\\mu_0^2={m**2}$, $t={exec_time:.1f}s$")
+            plt.savefig('plots/temp_parallel.png')
+            if show:
+                plt.show()
+
+class GifProducer(object):
+    def __init__(self):
+        self.frames = []
+
+    def save_lat(self, lat):
+        im = lat.show(show=False)
+        self.frames.append(im.make_image("AGG")[0])
+        plt.close()
+
+    def save(self, fn, fps=3):
+        imageio.mimwrite(fn, self.frames, fps=fps)

### 2020-07-07 code/test.py

@@ -1,22 +1,0 @@
-from timeit import timeit
-from matplotlib import pyplot as plt
-import math
-import numpy as np
-
-times = list(range(1,50))
-reassign = []
-no_reassign = []
-def numpy():
-    np.exp(-1000.0)
-
-def math():
-    math.exp(-1000.0)
-
-for t in times:
-    reassign.append( timeit(math, globals=dict( t=t)))
-    no_reassign.append( timeit(numpy, globals=dict(t=t)))
-
-plt.plot(times,reassign, label="math")
-plt.plot(times,no_reassign, label="numpy")
-plt.legend()
-plt.show()

### 2021-03-01 code/cpp_code/constants.h
make C++ code compatible with vortex cluster

@@ -2,4 +2,4 @@
 #define MASTER 0
 #define PROG_CHAR "#"
 #define D 2
-#define MAXERROR 0.01
+#define MAXERROR 0.005

### 2021-03-01 code/cpp_code/lattice.cpp

@@ -2,6 +2,9 @@
 
 #include <iostream>
 #include "lattice.h"
+#include <algorithm>
+
+using namespace std;
 
 int Lattice2D::L;
 double Lattice2D::beta;

### 2021-03-01 code/cpp_code/lattice.h

@@ -4,6 +4,7 @@
 
 #include "phi.h"
 #include <vector>
+#include <tuple>
 
 typedef tuple<int,int,int,int> Plaquette;
 

### 2021-03-01 code/cpp_code/main.cpp

@@ -52,6 +52,7 @@ int main(int argc, char *argv[]) {
     int measurements = root["measurements"].As<int>();
     int thermalization = root["thermalization"].As<int>();
     int record_rate = root["record_rate"].As<int>();
+    int cluster_rate = 5;
 
     vector<double> taus;
     auto iter =  root["taus"].Begin();
@@ -97,13 +98,14 @@ int main(int argc, char *argv[]) {
     Recorder recorder(observables);
     recorder.reserve(measurements);
 
-    const sweep_args args {
-        .sweeps = sweeps,
-        .thermalization = thermalization,
-        .record_rate = record_rate,
-        .ts = taus,
-        .progress = progress,
-        .cluster_algorithm = ClusterAlgorithm::WOLFF
+    const sweep_args args = {
+        sweeps,
+        thermalization,
+        record_rate,
+        ClusterAlgorithm::WOLFF,
+        taus,
+        cluster_rate,
+        progress
     };
 
     sweeper.full_sweep(&recorder, args);

### 2021-03-01 code/cpp_code/progress.cpp

@@ -6,19 +6,18 @@ class progress_bar
 {
     static const auto overhead = sizeof " [100%]";
 
-    ostream& os;
+    ostream os;
     const size_t bar_width;
     string message;
     const string full_bar;
 
  public:
-    progress_bar(ostream& os, size_t line_width,
+    progress_bar(ostream& stream, size_t line_width,
                  string message_, const char symbol = '.')
-        : os{os},
-          bar_width{line_width - overhead},
+    :     bar_width{line_width - overhead},
           message{move(message_)},
-
-          full_bar{string(bar_width, symbol) + string(bar_width, ' ')}
+          full_bar{string(bar_width, symbol) + string(bar_width, ' ')},
+          os{stream.rdbuf()}
     {
         if (message.size()+1 >= bar_width || message.find('\n') != message.npos) {
             os << message << '\n';

### 2021-03-01 code/cpp_code/sweep.cpp

@@ -99,9 +99,7 @@ Sweeper::Sweeper (int DIM, MPI_Comm c) :
     }
 
     lat.action = full_action(); 
-
     int offset = process_Rank *  sites_per_node;
-
 }
 
 Sweeper::~Sweeper() {

### 2021-03-01 code/cpp_code/sweep.h

@@ -45,13 +45,14 @@ class Recorder {
 
 
 struct sweep_args {
+
     int sweeps;
     int thermalization;
     int record_rate;
-    ClusterAlgorithm cluster_algorithm = WOLFF; 
+    ClusterAlgorithm cluster_algorithm; 
     vector<double> ts;
-    int cluster_rate = 5;
-    bool progress = true;
+    int cluster_rate;
+    bool progress;
 };
 
 class Sweeper {

### 2021-03-01 code/run.sh

@@ -1,15 +1,29 @@
 #!/bin/tcsh
-#PBS -N phi4_bc
-#PBS -l walltime=72:00:00
+#PBS -N gradientflow
+#PBS -l walltime=48:00:00
 #PBS -m abef
 #PBS -j oe
 #PBS -M snthomas01@email.wm.edu
 #PBS -l nodes=1:vortex:ppn=12
 
-#if ( ! (-e $PBS_O_WORKDIR) ) then
-    #cd $PBS_O_WORKDIR
 
 
-set N=12
-mpiexec -n $N python gradient_flow2.py 48 >& log.out
+set N=1
+setenv OPENMPI_HOME /usr/local/intel64/broadwell/intel-2017/openmpi-3.1.4-ib
+setenv LD_LIBRARY_PATH /usr/local/intel64/broadwell/intel-2017/openmpi-3.1.4-ib/lib:/usr/local/intel-2017/compilers_and_libraries_2017.0.098/linux/compiler/lib/intel64_lin:/usr/local/intel-2017/compilers_and_libraries_2017.0.098/linux/compiler/lib/intel64:/usr/local/intel-2017/compilers_and_libraries_2017.0.098/linux/mkl/lib/intel64:/usr/local/intel-2017/compilers_and_libraries_2017.0.098/linux/tbb/lib/intel64/gcc4.4:/usr/local/intel-2017/debugger_2017/iga/lib:/usr/local/intel-2017/debugger_2017/libipt/intel64/lib:/usr/local/intel-2017/compilers_and_libraries_2017.0.098/linux/daal/lib/intel64_lin:/usr/local/intel-2017/compilers_and_libraries_2017.0.098/linux/daal/../tbb/lib/intel64_lin/gcc4.4:/usr/local/gcc-5.2.0/lib:/usr/local/gcc-5.2.0/lib64:/lib64:/usr/lib64:/usr/local/torque-6.1.1.1/lib:/usr/local/anaconda3-2020.02/lib
+
+if ($?PBS_O_WORKDIR) then
+    cd $PBS_O_WORKDIR
+else
+    echo 'Running on head node'
+endif
+
+foreach input (inputs/*)
+    set output="outputs/`basename $input yml`csv"
+    #nohup mpiexec -n $N ./cpp_code/bin/sweep -i $input -o $output -q >! log.out &
+    ./cpp_code/bin/sweep -i $input -o $output -q >& log.out &
+end
[…]

### 2021-03-03 final-report/main.tex

\documentclass[12pt]{report}
\usepackage{amsmath}
\usepackage{graphicx}
\setlength{\oddsidemargin}{0.5in}
\setlength{\evensidemargin}{0.5in}
\setlength{\textwidth}{6.0in}
\setlength{\topmargin}{.75in}
\setlength{\textheight}{8.6in}

% Modified by S. Aubin for the class of 2019 Physics theses

%
% This is a LaTeX template for an undergraduate honors or
%   senior thesis in physics at William & Mary. Created from
%   several theses from previous students of D.S. Armstrong,
%   pared down to bare bones.
%
%   you will need this present file (thesis_template.tex)
%   as well as the file  figure1.eps  to make the whole thing
%   (the latter is just an encapsulated postscript figure)
%     For simplicity, we don't use bibtex here; experts can choose
%   to use that more powerful way of dealing with citations.
%
%   To compile it, do:
% > pdflatex thesis_template.tex
% > pdflatex thesis_template.tex  (yes, you need to do it twice to
%                                  fill in the list of tables, figures
%                                  and references).

\begin{document}
\setlength{\baselineskip}{24pt}
\begin{titlepage}
\LARGE
\begin{center}
{\bf A Wonderful Piece of Scientific Research}\\[2.3cm]

\normalsize
A thesis submitted in partial fulfillment of the requirement \\
for the degree of Bachelor of Science with Honors in \\
Physics from the College of William and Mary in Virginia,\\[0.5cm]
by\\[0.5cm]
Stuart Thomas \\[2.5cm]
\end{center}
\normalsize
\begin{flushright}
%\hfill Accepted for Honors \\[.5cm]
\hfill \hrulefill \\
\hfill \hfill Advisor: Prof. Christopher J. Monahan\\[.6cm]
\hfill \hrulefill \\Prof. Todd Averett \\[.6cm]
\hfill \hrulefill \\Prof. Andreas Stathopoulos \\[.6cm]
\end{flushright}
\begin{center}
Williamsburg, Virginia\\
May 2021
\end{center}
\end{titlepage}

\setlength{\topmargin}{0.0in}

\pagenumbering{roman}
\tableofcontents
\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
\paragraph
\indent I would like to thank lots of people, and this is where
I will do it...

\listoffigures
\addcontentsline{toc}{chapter}{List of Figures}
\listoftables
\addcontentsline{toc}{chapter}{List of Tables}

\begin{abstract}
\setcounter{page}{5}
\addcontentsline{toc}{chapter}{Abstract}
\paragraph
\indent An experiment was conducted at the Thomas Jefferson National
Accelerator Facility (JLab) to study something very interesting and
unusual....

The abstract should present a synopsis of all the main results of the research,
and should be self-contained. Generally, there should not be citations
in an abstract.

\end{abstract}

\chapter{Introduction}
\pagenumbering{arabic}
I have provided some suggested Chapter titles here, but you should use what makes sense for your research.

The introduction should present the primary motivation (i.e. big picture motivation, what big science question(s) your research is trying to answer or assist in answering) for the research and the results that you are presenting. You should also provide background information on your research (what has been done before) and also mention what some of the applications of your research are. You can also discuss some of the specifics of the exact thing you are working on.

\section{The Goal of the Experiment/Research}
\paragraph
\indent One can have sections of chapters and sub-sections as well (sub-sub-sections are possible, but are probably overkill for your purposes.

\chapter{Theory}
\paragraph
\indent Lots of theory...

You can introduce the basic concepts and theory that are necessary for understanding your research objective, methods, and results.

\section{Quantum Chromodynamics}\label{QCD} % label for referring to this section

\indent This is a subsection of the Theory section. Quantum chromodynamics
is a wonderful theory... here is a citation~\cite{Perkins} from a textbook, and here is another~\cite{IsgurKumano} from a journal article.

\subsubsection{Meson Spectrum in QCD}

Here is a numbered equation, taken from the thesis of A. Dubanowitz~\cite{UPV}:
%
\begin{eqnarray}
\frac{\phi\rightarrow a_0\gamma}{\phi\rightarrow f_0\gamma} = 0 .
\end{eqnarray}\label{eq:a0_decay_equation} %for referring to equation
%
We now introduce another equation, where we show that we need to define
every symbol in an equation. The parity-violating electron-proton
scattering asymmetry $A_{ep}$ can be expressed as
%
\begin{equation}
A_{ep} / A_0 = Q_{W}^{p}+Q^{2}F^{p}(Q^{2},\theta) \;\;\;\;\;  ,
\label{eq:asymmetry}
\end{equation}
%
where
$A_0 = -G_{F}Q^{2} / \left( {4\pi\alpha\sqrt{2}} \, \right)$,
$G_F$ is the Fermi constant, and $\alpha$ the fine structure constant.
$Q^2$ is the four-momentum transfer squared.
The second term, $Q^2 F^p$, contains the nucleon structure defined in terms of electromagnetic, neutral-weak, and axial form factors, which are encoded in the
hadronic form factor $F^p$.

Notice that equations are parts of sentences,
and should be punctuated as such. Read a real article from a refereed journal
(such as Physical Review or Physical Review Letters to see how this should
be done). Short, unnumbered equations can even appear
inline in a sentence, as was done in the preceding paragraph.

Every table you include must be referred to in the body of the text,
they can't just stand on their own. For example, we show the
dominant decay branching ratios of the $\phi$ meson in Table~\ref{tab:phidecays}.
You can refer to a section or chapter of your thesis, for example, we can look
back to Sec.~\ref{QCD} for a description of Quantum Chromodynamics.

\begin{table}[!thb]
\begin{center}
\begin{tabular}{|l|r|}
\hline
Decay Modes &\\ $\phi \rightarrow$ & Branching Ratio \\ \hline\hline
$K^+K^-$ & $(49.1\pm 0.8)\%$ \\ \hline
$K_{L}^{0}K_{S}^{0}$ & $(34.1\pm 0.6)\%$ \\ \hline
$\rho \pi + \pi^+ \pi^- \pi^0$ & $(15.5\pm 0.7)\%$ \\ \hline
$\omega \gamma$ & $<5\%$ \\ \hline
$f_0(980)\gamma$ & $<1\times 10^{-4}$ \\ \hline
$a_0(980)\gamma$ & $<5\times 10^{-3}$ \\ \hline
\end{tabular}
\end{center}
\caption[Main decays of the $\phi$ meson.] % title of the table for List of Tables
        {Main decays of the $\phi$ meson and their measured branching ratios. Notice that the caption is
        not the same as the much shorter title for the table, which appears in the List of Tables. You can see how to provide both the full caption and the title of the table by looking at the .tex source code.} % caption to the table
\label{tab:phidecays} % for referring to the table
\end{table}

A similar statement applies for figures. For example, we show a single event display of
a typical radiative decay in Fig.~\ref{fig:NeatFigure}. Figures and tables should not appear
earlier than they are first referenced in the body of the text, if at all possible.

Here I refer back to Eq.~\ref{eq:a0_decay_equation}; notice the use of
an abbreviation for the word ``equation''. Equation~\ref{eq:asymmetry} shows
how we refer to an equation at the start of a sentence.

\begin{figure*}[!tb]
\begin{center}
%\includegraphics[width=0.8\textwidth]{figure1.pdf} % set the width of the figure to 80% of the text width
\end{center}
\caption[A neat figure] %(this is the name of the figure, for the Table of Figures, and should be short and sweet).
{\label{fig:NeatFigure} % this is a label with which to refer to the figure
  This is a very nifty figure, which should be described in detail in this caption. Remember to
  describe all relevant aspects: what is plotted against what, if it is a graph, what the relevant features and
  symbols are, {\em etc.} . Your reader should be able to read the caption and understand everything of relevance in the figure. Figure captions should be in
  complete, properly punctuated sentences.
  % this is the body of the caption to the figure
}
\end{figure*}

\chapter{Experimental Technique}

\chapter{Results}

\chapter{Conclusion/Outlook}

\appendix
\chapter{My Great Computer Program}
You may wish to include some material in appendices, such as this one.
An appendix should have material that is not required to be read by
the average reader of the thesis, but which may be useful for
some readers, and to document important material for the future - {\em i.e.}
the next folks to carry and expand your work. You can include things like
additional data, code, instructions for use of some piece of equipment or
software, etc.

\section{Code sample}
\paragraph
\indent The following is the C code which .....
\setlength{\baselineskip}{12pt}
{\footnotesize \begin{verbatim}
/***********************************************************/

/*   Ima G. Physicist   */

int makeBSDHits(itape_header_t *event, time_list_t *timeList)
{
#define nBSDChannels 48  /*corresponds to the number of scintillators of the BSD*/
  adc_values_t *bsd=NULL;
  tdc_values_t *tbsd=NULL;
  int i,j,k,m;
  int size;

  int adcValue[nBSDChannels];
  int ntdchits[nBSDChannels];

  struct bsdtimes{
    int ntimes;
    int tdctimes[16];
  };

  struct bsdtimes tdcValue[nBSDChannels];

  for(i=0; i < nBSDChannels; i++){
    tdcValue[i].ntimes=0;
    for(j=0; j < 16; j++){
      tdcValue[i].tdctimes[j]=0;
    }
  }

  return(0);
}
\end{verbatim}}

\begin{flushleft}
\bibliographystyle{unsrt}
\begin{thebibliography}{99.}

\bibitem{Perkins} Perkins, Donald H.  Introduction to High Energy
Physics.  3rd ed.  Menlo Park: Addison-Wesley, 1987.

\bibitem{IsgurKumano} F. Close, N. Isgur and S. Kumano,
Nucl. Phys. {\bf B389}, 513 (1993).

\bibitem{UPV} Alex Dubanowitz, Senior project thesis, College of
William and Mary, 1998. (unpublished).

\end{thebibliography}
\end{flushleft}

\end{document}

### 2021-03-04 code/cpp_code/constants.h
return tolerance to normal level

@@ -2,4 +2,4 @@
 #define MASTER 0
 #define PROG_CHAR "#"
 #define D 2
-#define MAXERROR 0.005
+#define MAXERROR 0.01

### 2021-03-08 final-report.latex/main.tex
move final report, update notebook

--- 2021-03-03 final-report/main.tex
@@
 \paragraph
-\indent An experiment was conducted at the Thomas Jefferson National
-Accelerator Facility (JLab) to study something very interesting and
-unusual....
-
-The abstract should present a synopsis of all the main results of the research,
-and should be self-contained. Generally, there should not be citations
-in an abstract.
+\indent \textit{Abstract}
 
@@
 \pagenumbering{arabic}
-I have provided some suggested Chapter titles here, but you should use what makes sense for your research.
 
-The introduction should present the primary motivation (i.e.
-big picture motivation, what big science question(s) your research is trying to answer or assist in answering) for the research and the results that you are presenting.
-You should also provide background information on your research (what has been done before) and also mention what some of the applications of your research are.
-You can also discuss some of the specifics of the exact thing you are working on.
-
-\section{The Goal of the Experiment/Research}
-\paragraph
-\indent One can have sections of chapters and sub-sections as well (sub-sub-sections are possible, but are probably overkill for your purposes.
+%\section{Main Experimental Goal}
 
 \chapter{Theory}
-\paragraph
-\indent Lots of theory...
+\section{Gradient Flow}
+\section{Field Theories on the Lattice}
+\section{Non-linear $\sigma$ model}
 
-You can introduce the basic concepts and theory that are necessary for understanding your research objective, methods, and results.
-
-\section{Quantum Chromodynamics}\label{QCD} % label for referring to this section
-
-\indent This is a subsection of the Theory section. Quantum chromodynamics
-is a wonderful theory... here is a citation~\cite{Perkins} from a textbook, and here is another~\cite{IsgurKumano} from a journal article.
-
-\subsubsection{Meson Spectrum in QCD}
-
-Here is a numbered equation, taken from the thesis of A. Dubanowitz~\cite{UPV}:
-%
-\begin{eqnarray}
-\frac{\phi\rightarrow a_0\gamma}{\phi\rightarrow f_0\gamma} = 0 .
-\end{eqnarray}\label{eq:a0_decay_equation} %for referring to equation
-%
-We now introduce another equation, where we show that we need to define
-every symbol in an equation. The parity-violating electron-proton
-scattering asymmetry $A_{ep}$ can be expressed as
-%
-\begin{equation}
-A_{ep} / A_0 = Q_{W}^{p}+Q^{2}F^{p}(Q^{2},\theta) \;\;\;\;\;  ,
-\label{eq:asymmetry}
-\end{equation}
-%
-where
-$A_0 = -G_{F}Q^{2} / \left( {4\pi\alpha\sqrt{2}} \, \right)$,
-$G_F$ is the Fermi constant, and $\alpha$ the fine structure constant.
-$Q^2$ is the four-momentum transfer squared.
-The second term, $Q^2 F^p$, contains the nucleon structure defined in terms of electromagnetic, neutral-weak, and axial form factors, which are encoded in the
-hadronic form factor $F^p$.
-
-Notice that equations are parts of sentences,
-and should be punctuated as such. Read a real article from a refereed journal
-(such as Physical Review or Physical Review Letters to see how this should
-be done). Short, unnumbered equations can even appear
-inline in a sentence, as was done in the preceding paragraph.
-
-Every table you include must be referred to in the body of the text,
-they can't just stand on their own. For example, we show the
-dominant decay branching ratios of the $\phi$ meson in Table~\ref{tab:phidecays}.
-You can refer to a section or chapter of your thesis, for example, we can look
-back to Sec.~\ref{QCD} for a description of Quantum Chromodynamics.
-
-\begin{table}[!thb]
-\begin{center}
-\begin{tabular}{|l|r|}
-\hline
-Decay Modes &\\ $\phi \rightarrow$ & Branching Ratio \\ \hline\hline
-$K^+K^-$ & $(49.1\pm 0.8)\%$ \\ \hline
-$K_{L}^{0}K_{S}^{0}$ & $(34.1\pm 0.6)\%$ \\ \hline
-$\rho \pi + \pi^+ \pi^- \pi^0$ & $(15.5\pm 0.7)\%$ \\ \hline
-$\omega \gamma$ & $<5\%$ \\ \hline
-$f_0(980)\gamma$ & $<1\times 10^{-4}$ \\ \hline
-$a_0(980)\gamma$ & $<5\times 10^{-3}$ \\ \hline
-\end{tabular}
-\end{center}
-\caption[Main decays of the $\phi$ meson.] % title of the table for List of Tables
-        {Main decays of the $\phi$ meson and their measured branching ratios. Notice that the caption is
-        not the same as the much shorter title for the table, which appears in the List of Tables.
-You can see how to provide both the full caption and the title of the table by looking at the .tex source code.} % caption to the table
-\label{tab:phidecays} % for referring to the table
-\end{table}
-
-A similar statement applies for figures. For example, we show a single event display of
-a typical radiative decay in Fig.~\ref{fig:NeatFigure}. Figures and tables should not appear
-earlier than they are first referenced in the body of the text, if at all possible.
-
-Here I refer back to Eq.~\ref{eq:a0_decay_equation}; notice the use of
-an abbreviation for the word ``equation''. Equation~\ref{eq:asymmetry} shows
-how we refer to an equation at the start of a sentence.
-
-\begin{figure*}[!tb]
-\begin{center}
-%\includegraphics[width=0.8\textwidth]{figure1.pdf} % set the width of the figure to 80% of the text width
-\end{center}
-\caption[A neat figure] %(this is the name of the figure, for the Table of Figures, and should be short and sweet).
-{\label{fig:NeatFigure} % this is a label with which to refer to the figure
-  This is a very nifty figure, which should be described in detail in this caption. Remember to
-  describe all relevant aspects: what is plotted against what, if it is a graph, what the relevant features and
-  symbols are, {\em etc.} . Your reader should be able to read the caption and understand everything of relevance in the figure. Figure captions should be in
-  complete, properly punctuated sentences.
-  % this is the body of the caption to the figure
-}
-\end{figure*}
-
-\chapter{Experimental Technique}
+\chapter{Methods}
+\section{Markov Chain Monte Carlo}
+\subsection{Metropolis Algorithm}
+\subsection{Wolff Cluster Algorithm}
+\subsection{Autocorrelation}
 
 \chapter{Results}
+\chapter{Statistical Analysis}
 
-\chapter{Conclusion/Outlook}
+\chapter{Conclusion \& Outlook}
 
@@
 \chapter{My Great Computer Program}
-You may wish to include some material in appendices, such as this one.
-An appendix should have material that is not required to be read by
-the average reader of the thesis, but which may be useful for
-some readers, and to document important material for the future - {\em i.e.}
-the next folks to carry and expand your work. You can include things like
-additional data, code, instructions for use of some piece of equipment or
-software, etc.
 
 \section{Code sample}
-\paragraph
-\indent The following is the C code which .....
-\setlength{\baselineskip}{12pt}
-{\footnotesize \begin{verbatim}
-/***********************************************************/
-
-/*   Ima G. Physicist   */
-
-int makeBSDHits(itape_header_t *event, time_list_t *timeList)
-{
-#define nBSDChannels 48  /*corresponds to the number of scintillators of the BSD*/
-  adc_values_t *bsd=NULL;
-  tdc_values_t *tbsd=NULL;
-  int i,j,k,m;
-  int size;
-
-  int adcValue[nBSDChannels];
-  int ntdchits[nBSDChannels];
-
-  struct bsdtimes{
-    int ntimes;
-    int tdctimes[16];
-  };
-
-  struct bsdtimes tdcValue[nBSDChannels];
-
-  for(i=0; i < nBSDChannels; i++){
-    tdcValue[i].ntimes=0;
-    for(j=0; j < 16; j++){
-      tdcValue[i].tdctimes[j]=0;
-    }
-  }
-
-  return(0);
-}
-\end{verbatim}}
-
-\begin{flushleft}
 \bibliographystyle{unsrt}
-\begin{thebibliography}{99.}
-
-\bibitem{Perkins} Perkins, Donald H.  Introduction to High Energy
-Physics.  3rd ed.  Menlo Park: Addison-Wesley, 1987.
-
-\bibitem{IsgurKumano} F. Close, N. Isgur and S. Kumano,
-Nucl. Phys. {\bf B389}, 513 (1993).
-
-\bibitem{UPV} Alex Dubanowitz, Senior project thesis, College of
-William and Mary, 1998. (unpublished).
-
-\end{thebibliography}
-\end{flushleft}
 

### 2021-03-10 final-report.latex/main.tex
final-report: include topology, fix Wolff cluster, write gradient flow for NSM

@@
 \usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{hyperref}
 \usepackage{graphicx}
@@
 \section{Gradient Flow}
+To remove ultraviolet fluctuations, we introduce a momentum cutoff, effectively reducing the predictive power of our theory by ignoring very high momentum states.
+In a gauge theory, such as QCD, a hard cutoff violates gauge invariance.
+Therefore, renormalization techniques must be more nuanced in the reduction of high-frequency divergences.
+
+One such technique is ``smearing,'' a local averaging of a gauge field in order to remove diverging fluctuations \cite{solbrig2007}.
+In this study, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time''.
+The flow time parameterizes the smearing; an evolution in flow time corresponds to suppressing ultraviolet divergences.
+In a 2D $\phi^4$ scalar field theory, given by the action
+\begin{equation}
+    S_\phi [\phi] = \frac{1}{2}\int d^2x\left[(\partial \phi)^2+m^2\phi^2+\frac{\lambda}{2}\right],
+\end{equation}
+we can define the evolution in flow time as
+\begin{equation}
+    \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
+\end{equation}
+where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime and $\tau$ is the flow time.
+Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
+When solved exactly, this function forms a Gaussian which forces operators to converge.
+Therefore, in all orders of pertubation theory, any correlation function in the gauge field that has been evolved in flow time becomes immune to these divergences\cite{makino2015a}.
+
+Generally, we can choose any flow time equation that drives the field towards a classical minimum.
+Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields.
+In the nonlinear $\sigma$ model, it has been shown that an appropriate manifestation of the flow time can resolve divergencies \cite{makino2015a}.
+Therefore, it is possible to define a different flow equation for this model as well.
+Our goal is to understand this renormalizability in the nonlinear $\sigma$ model using the gradient flow.
+Specifically, we seek to define vacuum condensates, or nonpertubative vacuum expectation values.
+
+To explore this technique, we first plan to produce numerical calculations using a nonlinear $\sigma$ model.
+Using a Monte Carlo method in 1+1 dimensions, we can calculate the correlation functions of quantum field theories.
+With more computing power, expanding this simulation to 2+1 dimensions would provide an analysis more relevant to quantum chromodynamics.
+Following this numerical calculation, we seek to find and understand an implementation of the gradient flow in this model using analytic methods.
+This understanding can then be verified by the numerical solution.
+
 \section{Field Theories on the Lattice}
@@
 
+The nonlinear $\sigma$ model is a prototypical theory for many physical phenomena, including applications in string theory.
+As a simple nonperturbative model, it provides an ideal starting point for lattice QCD studies.
+Specifically, the nonlinear $\sigma$ model exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
+Furthermore, it has an exact application to Heisenberg ferromagnets in condensed matter.
+The theory is defined by the Euclidean action
+\begin{equation}
+    S_E = \frac{1}{2g^2} \int d^Dx \partial^\mu \vec\phi \cdot \partial_\mu \vec\phi
+\end{equation}
+subject to the constraint that $\vec\phi\cdot\vec\phi = 1$. In this report, we primarily consider $O(3)$, where $\vec\phi$ is a $3$ dimensional vector.
+
+Following \cite{bietenholz2018}, we can define the gradient flow in this model via the differential equation
+\begin{equation}
+    \label{eq:nsm_gradflow}
+    \partial_\tau \vec\rho (\tau,x) = P(\tau,x) \nabla^2 \vec\rho(\tau,x),
+\end{equation}
+where $\nabla^2$ is the Laplacian operator in Euclidean space and $P(\tau,x)$ is a $N \times N$ matrix defined as
+\begin{equation*}
+    P(\tau, x) = 1 - \vec\rho(\tau,x) \vec\rho(\tau,x)^T.
+\end{equation*}
+We solve this equation numerically using the boundary condition $\vec\rho(0,x) = \phi(x)$.
+
+In order to use this theory on the lattice, we discretize the action. We replace the derivatives with a simple difference and impose periodic boundary conditions on the system.
+
+\section{Topological Charge}
+
+For every configuration of the non-linear sigma model, there is an associated integer quantity known as the topological charge.
+By viewing the field as a mapping from periodic real space to the sphere $S^{N-1}$, we can can associate each configuration with a member of the homotopy group $\Pi_2(S^2) = \mathbb{Z}$.
+This integer value is the topological charge of such a configuration.
+
 \chapter{Methods}
-\section{Markov Chain Monte Carlo}
+\label{sec:methods}
+Our study of the gradient flow in the nonlinear $\sigma$ model consists of a computational part and an analytical part.
+We begin by outlining a numerical Monte Carlo method to simulate the lattice in 2 and 3 dimensions.
+We verify our program with the well-studied $\phi^4$ scalar field theory.
+We then generalize our model to a vector field to simulate the non-linear sigma model.
+This simulation system provides data on vacuum states with which we study the gradient flow's effect on topology.
+
+We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories.
+
+\section{Monte Carlo Simulations}
+\label{sec:mc}
+
+We implement a Markov Chain Monte Carlo method following Schaich's thesis \cite{schaich2006}.
+This implementation utilizes a ``random walk,'' i.e.
+a set of random steps through phase space, to determine statistical values such as correlation functions across the lattice.
+By the definition of the Markov chain, the probability of adoption of each state, and therefore its inclusion in the Monte Carlo calculation, depends only on the current state and the proposed state.
+This probability is denoted as $P(\mu\rightarrow\nu)$ where $\mu$ and $\nu$ are the existing and proposed lattice configurations respectively.
+We use a combination of the Metropolis and Wolff algorithms to determine this value.
+
 \subsection{Metropolis Algorithm}
+We primarily use the Metropolis algorithm for the calculation of new Markov chain configurations.
+We begin with a so-called ``hot start,'' where each field value at each lattice site is randomly selected.
+Then we propose a new value for a single lattice point, which is accepted with a probability
+\begin{equation}
+    P(\mu\rightarrow\nu) = \begin{cases} 
+        e^{-(S_\nu - S_\mu)} & S_\nu > S_\mu \\
+        1 & \mathrm{otherwise} \\
+   \end{cases}
+\end{equation}
+where $S_\nu$ and $S_\mu$ are the lattice actions of configuration before and after the change.
+This process is performed for each point on the lattice, making up a ``sweep.'' Repeating this sweep many times pushes the lattice toward the action minimum.
+
 \subsection{Wolff Cluster Algorithm}
-\subsection{Autocorrelation}
+
+Though the Metropolis algorithm will slowly find the absolute minimum of the theory, the presence of local minima can greatly prolong the convergence.
+Both the $\phi^4$ model and the nonlinear $\sigma$ model feature ``kinetic'' terms with gradients of $\phi$.
+Therefore, the presence of large similarly-valued regions in the lattice can lead to a local minimum.
+The Wolff algorithm helps reduce the presence of these clusters through two steps: identifying a cluster and flipping it along some arbitrary vector.
+In the case of $\phi^4$ theory, this flipping takes the form of a simple sign change.
+In the non-linear sigma model we choose a random unit vector $\vec r$ and consider the projection of the field on this vector.
+When the cluster flips, each site is flipped along this direction.
+\footnote{citation needed}.
+To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site.
+In the non-linear $\sigma$ model, this value is derived as
+\begin{equation}
+    P_{add} = 1-e^{- \beta \vec\phi_a\vec\phi_b}.
+\end{equation}
+ % TODO: derive this
+
+\subsection{Checkerboard algorithm}
+
+In order to parallelize the Metropolis algorithm, we use a checkerboard algorithm.
+Since the Lagrangian density at each site does not depend on any sites diagonal to it, the lattice can be split into ``white'' sites and ``black'' sites, like the tiles on a checkerboard.
+Each white site is independent of every other white site and likewise with black sites.
+Therefore, we can split the sites of each color into separate parallel processing nodes and independently run the Metropolis algorithm, ensuring that no site affects the Lagrangian density at any other site.
+We use this method to parallelize the code through the Message Passing Interface (MPI).
+
+\subsection{Autocorrelation times and thermalization}
+One important aspect is the correlation between different states of the Markov chain.
+Ideally, each sample of the lattice would be completely independent, but this is not the case.
+Therefore, we must sweep the lattice sufficiently to produce an effective Monte Carlo result.
+We use Wolff's automatic windowing procedure \cite{wolff2007} and the magnetic susceptibility $\chi_m$ to estimate the autocorrelation and determine an appropriate number of sweep between measurements for each theory.
+Furthermore, the initial values of our Markov chain will be highly correlated with the initial random start.
+Therefore, we also wait a set number of sweeps before taking any measurements so that the lattice can approach a minimum of the action.
 
@@
 \bibliographystyle{unsrt}
+\bibliography{library}
 

### 2021-03-15 final-report.latex/main.tex
add discretized gradient flow

@@
 \setlength{\textheight}{8.6in}
+
+\newcommand{\e}{\vec e}
 
@@
     \label{eq:nsm_gradflow}
-    \partial_\tau \vec\rho (\tau,x) = P(\tau,x) \nabla^2 \vec\rho(\tau,x),
+    \partial_\tau \vec\rho (\tau,x) = \left( 1 - \vec\rho(\tau,x) \vec\rho(\tau,x)^T \right) \nabla^2 \vec\rho(\tau,x),
 \end{equation}
-where $\nabla^2$ is the Laplacian operator in Euclidean space and $P(\tau,x)$ is a $N \times N$ matrix defined as
-\begin{equation*}
-    P(\tau, x) = 1 - \vec\rho(\tau,x) \vec\rho(\tau,x)^T.
-\end{equation*}
-We solve this equation numerically using the boundary condition $\vec\rho(0,x) = \phi(x)$.
+where $\nabla^2$ is the Laplacian operator in Euclidean space. We solve this equation numerically using the boundary condition $\vec\rho(0,x) = \phi(x)$.
 
-In order to use this theory on the lattice, we discretize the action. We replace the derivatives with a simple difference and impose periodic boundary conditions on the system.
+In order to use this theory on the lattice, we discretize the action.
+We replace the derivatives with a simple difference and impose periodic boundary conditions on the system.
+For a more in-depth discussion, see Section~\ref{sec:grad_flow}.
 
@@
 \begin{equation}
-    P_{add} = 1-e^{- \beta \vec\phi_a\vec\phi_b}.
+    P_{add} = 1-e^{- \beta \vec e_a\vec e_b}.
 \end{equation}
@@
 
+\section{Gradient Flow}
+\label{sec:gradflow_disc}
+On the lattice, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes 
+\begin{equation}
+    \label{eq:nsm_gradflow_disc}
+    \partial_\tau \e (\tau)_x = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \nabla^2 \e(\tau)_x,
+\end{equation}
+ where the Laplace operator $\nabla^2$ is defined as
+\begin{equation*}
+    \nabla^2 \e(t)_x = \sum^d_{\mu=1} \left( \e(t)_{x+\hat \mu} - \e(t)_{x-\hat \mu}\right) - 2 d \e(t)_x.
+\end{equation*}
+
+In order to calculate the gradient flow, we numerically solve the differential equation using a fourth-order Runga-Kutta approximation.
+We use an adaptive step size algorithm to dramatically increase efficiency.
+
 \chapter{Results}

### 2021-03-22 final-report.latex/main.tex
expand Runge Kutta description

@@
 
-In order to calculate the gradient flow, we numerically solve the differential equation using a fourth-order Runga-Kutta approximation.
-We use an adaptive step size algorithm to dramatically increase efficiency.
+\subsection{Runge-Kutta Algorithm}
+In order to calculate the gradient flow, we numerically solve the differential equation using a fourth-order Runga-Kutta approximation. This algorithm refines the Euler method 
+\begin{equation*}
+    \e(\tau+h)_x \approx \e(\tau)_x + h f(\e(\tau)_x).
+\end{equation*}
+where $f(\e)$ is defined for convenience as 
+\begin{equation}
+    f(\e)=\partial_\tau \e (\tau)_x  = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \nabla^2 \e(\tau)_x,
+\end{equation}
+following from Eq.~\ref{eq:nsm_gradflow_disc}. To the fourth order, this approximation becomes 
+%
+\begin{align}
+    \label{eq:rungekutta}
+    k_1 &= h f(\e(\tau)_x) \\ 
+    k_2 &= h f(\e(\tau)_x + \frac{k_1}{2}) \\ 
+    k_3 &= h f(\e(\tau)_x + \frac{k_2}{2}) \\ 
+    k_4 &= h f(\e(\tau)_x + k_3) \\ 
+    \e(\tau+h)_x &= \e(\tau)_x + \frac{k_1}{6} + \frac{k_2}{3} + \frac{k_3}{3} + \frac{k_4}{6} + O(h^5).
+\end{align}
+This method is usually superior to Euler's method and the midpoint method \cite{vetterling1992}.
+
+To increase the efficiency of this algorithm, we implement the step-doubling algorithm to adaptively adjust $h$.
+If the error of a Runge-Kutta step is greater than the tolerance, the same step is repeated with half the step size.
+Alternatively, if the error is less than half of the tolerance, the step size is doubled for the next calculation.
+Finally, if the step size is greater than the distance to the next measurement, that distance is used as the step size, using the normal value afterwards.
+Otherwise, the algorithm proceeds with the consistent step size.
 

### 2021-03-29 code/cpp_code/observables.h
xi2 calculation, Bietenholz fig 7

@@ -33,14 +33,51 @@ namespace observables {
         public:
             double operator()(const Lattice2D& lat) const {
                 double val = 0;
-                const Phi phi0 = lat[0];
-                //for (const Phi& phi : as_const(lat)) {
-                for (auto it = lat.cbegin(); it!=lat.cend(); ++it) {
-                    val += (*it)*phi0;
+
+                for (auto itx = lat.cbegin(); itx!=lat.cend(); ++itx) {
+                    for (auto ity = lat.cbegin(); ity!=lat.cend(); ++ity) {
+                        val += (*itx)*(*ity);
+                    }
                 }
                 return val;
             };
             const string name() const { return "chi_m"; };
+    };
+
+    //class C : public BaseObservable {
+        //public:
+            //double operator()(const Lattice2D& lat) const {
+                //double val = 0;
+                //const Phi phi0 = lat[0];
+                ////for (const Phi& phi : as_const(lat)) {
+                //for (auto it = lat.cbegin(); it!=lat.cend(); ++it) {
+                    //val += (*it)*phi0;
+                //}
+                //return val;
+            //};
+            //const string name() const { return "chi_m"; };
+    //};
+
+    class F : public BaseObservable {
+        public:
+            double operator()(const Lattice2D& lat) const {
+                double val = 0;
+
+                int x=0;
+                int y=0;
+
+                //for (const Phi& phi : as_const(lat)) {
+                for (auto itx = lat.cbegin(); itx!=lat.cend(); ++itx) {
+                    for (auto ity = lat.cbegin(); ity!=lat.cend(); ++ity) {
+                        val += (*itx)*(*ity) * cos( 2pi_L * (x - y));
+                        x++;
+                        if (x==lat.L) x=0; // This ensures that x is the Euclidean space dimension
+                    }
+                    if (y==lat.L) y=0;
+                }
+                return val;
+            };
+            const string name() const { return "F"; };
     };
 
     class Q : public BaseObservable {

### 2021-03-29 final-report.latex/main.tex
add introduction, flesh out theory

@@
 
+\pagenumbering{arabic}
 \chapter{Introduction}
-\pagenumbering{arabic}
+
+Quantum field theory (QFT) is an extraordinarily successful model which can be applied to a range of physical phenomena.
+Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental framework of particle physics.
+In condensed matter physics, effective field theories can model emergent effects such as phonons and quasiparticles.
+Compared to experiment, QFT is remarkably accurate, famously matching the experimental value for the electron $g$-factor to eight significant figures.\cite{weisskopf1981}
+
+However, this power comes at a cost: the study of quantum fields is rife with infinities.
+A na\"ive treatment of quantum field theory produces divergent values for physical quantities, a clearly impossible result.
+Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the so-called \textit{renormalization group}.
+However, the technique fails with perturbatively nonrenormalizable theories.
+
+One such example is the \textit{non-linear sigma model} which provides a prototypical theory in both condensed matter and particle physics.
+In solid-state systems, this model describes Heisenberg ferromagnets.
+In nuclear physics, the non-linear sigma model acts as a prototype for quantum chromodynamics, exhibiting characteristic features such as a mass gap and asymptotic freedom.
+% citation needed
+
+In this study, we specifically consider the $O(3)$ nonlinear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time).
+This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action.
+
+Since the non-linear sigma model cannot be renormalized perturbatively, we cannot study these topological effects with normal perturbative techniques.
+An alternative solution is placing the field on a discretized lattice, a technique originally used for quantum chromodynamics.
+In this scenario, field configurations become computationally calculable.
+This process introduces an nonphysical length scale $a$, the lattice spacing.
+Therefore, we expect any physical result to converge in the continuum limit, i.e.
+when $a\rightarrow 0$.
+However, this is not always the case as observables mix on the lattice, leading to divergences.
+As an example, states of definite angular momentum mix when discretized, a clear violation of the angular momentum commutation relations.
+
+The gradient flow is a technique to remove these divergences.
+By dampening high-momentum fluctuations, the gradient flow reduces power-divergent mixing and make observables finite on the lattice.\cite{monahan2016} To explore this technique, we first plan to produce numerical calculations using a non-linear sigma model.
+Using a Monte Carlo method in 1+1 dimensions, we can calculate the topological properties of quantum field theories.
+
+\section{Conventions}
+
+Throughout this paper, we use natural units, i.e. $\hbar = 1$ and $c=1$.
 
@@
 \chapter{Theory}
+
+This thesis incorporates two main bodies of knowledge: quantum field theory and statistical simulation.
+Through the path integral formulation of quantum field theory, we are able to describe the physics of the former with the established mathematics of the latter.
+
+\section{Quantum Field Theory}
+
+In this section we outline a rough description of quantum field theory.
+A full introduction is beyond the scope of this paper, however we do assume knowledge of nonrelativistic quantum mechanics and classical field theory.
+
+The fundamental hypothesis of quantum field theory (QFT) describes particles as discrete packets of energy on a quantum field.
+But what is a quantum field?
+Like in classical mechanics, a field is a function of spacetime with some mathematical object assigned to each point in space and time.
+In the case of the electric field, this object is a three-dimensional vector, while the electric potential is a scalar field.
+Classical and quantum fields have Lagrangians which define how they evolve in time.
+What differentiates a quantum field from a classical field is superposition: where classical fields have a definite configuration, quantum fields exist in a superposition of all possible configurations.
+It is possible---though nontrivial and outside the scope of this description--- to motivate the appearance of discrete particles from this superposition.
+
+This general description allows QFT to easily incorporate special relativity.
+By ensuring that the Lagrangian of a theory is invariant under Lorentz transformations, we can ensure that the theory is physical.
+% is this enough?
+
+Though most introductions to QFT use the ``second quantization,'' we will use a alternative formulation known as the path integral formulation (for a similar introduction see Zee's textbook \cite{zee2010}).
+
+\subsection{Path Integral Formulation}
+
+As stated above, we can model a quantum field theory as a superposition of all possible classical fields.
+Like single-particle quantum mechanics, each configuration has a probability amplitude.
+To measure expectation values of observables, we simply take an average over all configurations weighted by this complex amplitude.
+We can formalize this notion using the fundamental formula\footnote{Since this study concerns vacua, we do not include a source term.}
+\begin{equation}
+    \label{eq:pathintegral}
+    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; e^{iS[\phi]}
+\end{equation}
+% not sure if the semicolons is good form
+where $\langle \hat O \rangle$ is the expectation value of an arbitrary operator $\hat O$; $Z$ is a normalization constant; $S$ is the action functional, defined from the field's Lagrangian; and $\int \mathcal{D}\phi$ represents the eponymous path integral.
+Though it is possible to define this integral rigorously, for our purposes we can equate it to a sum over all possible configurations.
+This form generalizes the classical principle of least action, a relationship that Richard Feynman pedagogically explains in his lectures on physics \cite{feynman1963a}.
+
+At first glance, Eq.~\ref{eq:pathintegral} is remarkably similar to the statistics of the canonical ensemble.
+Through this similarity, we will be able to use mathematical tools from statistical mechanics to study quantum field theories.
+However, the factor of $i$ in the exponent currently prohibits us from making this jump.
+To remedy this issue, we perform a ``Wick rotation'' which shifts spacetime into Euclidean coordinates.
+In normal spacetime, defined by the Minkowski metric, the Lorentz-invariant distance is given as
+\begin{equation}
+    s = x^2_0 - x^2_1- x^2_3- x^2_3
+\end{equation}
+where $x_0=ct$ and $\vec{x} = (x_1, x_2, x_3)^T$. By redefining the time coordinate $x_0$ of a spacetime point $x$ to be $x_4=ix_0$, we find that the quantity
+\begin{equation}
+    s_E = x^2_1+ x^2_3+ x^2_3 + x^2_4,
+\end{equation}
+is invariant, which is representative of a four-dimensional Euclidean space. Furthermore, we find that
+\begin{align}
+    d^4x_E &= d^3\vec{x}dx_4 \\
+    &= i d^3\vec{x}dx_0 \\
+    &= i d^4x. \label{eq:wickdifferential}
+\end{align}
+
+We can use this transformation to redefine the Lagrangian $\mathcal{L}$ in Euclidean space as $\mathcal{L}_E$.
+Since a Lorentz-invariant Lagrangian must include only even powers and derivatives of $x$, the Euclidean Lagrangian is necessarily real.
+Subsequently, we can define a Euclidean action based on the differential in Eq.~\ref{eq:wickdifferential}:
+\begin{align}
+    S_E &= \int d^4x_E \mathcal{L}_E \\
+    &= i \int d^4x \mathcal{L}_E,
+\end{align}
+allowing us to redefine the path integral as 
+\begin{equation}
+    \label{eq:pathintegraleuclidean}
+    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O (\phi)\; e^{-S_E[\phi]}.
+\end{equation}
+
+By replacing the Minkowski action $S$ with a Euclidean action $S_E$, we have transformed the amplitude $e^{iS}$ to a statistical Boltzmann factor $e^{-S_E}$.
+
+\subsection{Ultraviolet Divergences}
+
+\section{Markov Chain Monte Carlo}
+
+To accomplish a statistical analysis of quantum fields, we use a Monte Carlo simulation.
+This method amounts to producing a large number of configurations and calculating statistics on the sample.
+A brute-force calculation over all possible configurations---as Eq.~\ref{eq:pathintegraleuclidean} suggests--- is clearly infinite and computationally infeasible.
+However, the exponential nature of the Boltzmann factor dictates that only configurations near the action minimum contribute to observable statistics.
+Therefore, by selecting a sample of configurations near this minimum, we are able to extract meaningful results with a finite computation.
+
+\subsection{The Markov Chain}
+The Markov chain is an effective method to identity these low-action states.
+Essentially, we begin with a random configuration and then make small adjustments, gradually lowering the action.
+Starting with a field configuration $\phi_a$, we propose a new field $\phi_b$.
+The probability of accepting this change, thereby adding the configuration to the chain, is given by the function
+\[P(\phi_a \rightarrow \phi_b).\]
+
+There are four requirements that this function must obey to produce a Boltzmann distribution of samples:
+\begin{enumerate}
+    \item $P(\phi_a \rightarrow \phi_b)$ must depend only on the configurations $\phi_a$ and $\phi_b$.
+    \item The probability must be properly normalized, i.e. $\sum_{\phi} P(\phi_a \rightarrow \phi) = 1$.
+    \item Every configuration must be reachable in a finite number of steps. In other words, the chain must be ergodic.
+    \item In order to reach equilibrium, the chain must be reversible.
+In other words, the probability of a $\phi_a\rightarrow\phi_b$ transition must be equal to the probability of a $\phi_b\rightarrow\phi_a$ transition.
+Mathematically, this condition takes the form of \textit{detailed balance equations}:
+\begin{equation}
+    \label{eq:detailedbalance}
+    P(\phi_a)\,P(\phi_a\rightarrow\phi_b) = P(\phi_b)\,P(\phi_b\rightarrow\phi_a),
+\end{equation}
+where $P(\phi)$ is the probability of a system existing in state $\phi$.
+% this needs citation
+\end{enumerate}
+
+This final condition will allow us to explicitly define the transition probability using the action. From the Boltzmann distribution, we know
+\begin{equation}
+    P(\phi) = \frac{1}{Z} e^{-S_E[\phi]}.
+\end{equation}
+Therefore, be rearranging Eq.~\ref{eq:detailedbalance}, we find
+\begin{equation}
+    \frac{P(\phi_a\rightarrow\phi_b)}{P(\phi_b\rightarrow\phi_a)} = e^{-(S_E[\phi_b] - S_E[\phi_a])}.
+\end{equation}
+
+This formula will provide the explicit probabilities of the Metropolis and Wolff algorithms.
+
 \section{Gradient Flow}
-To remove ultraviolet fluctuations, we introduce a momentum cutoff, effectively reducing the predictive power of our theory by ignoring very high momentum states.
-In a gauge theory, such as QCD, a hard cutoff violates gauge invariance.
-Therefore, renormalization techniques must be more nuanced in the reduction of high-frequency divergences.
+To remove ultraviolet fluctuations, we introduce a momentum cutoff which reduces ultraviolet divergences.
+In a gauge theory, such as QCD, a hard cutoff violates gauge invariance, so more nuanced techniques are required.
 
@@
 In a 2D $\phi^4$ scalar field theory, given by the action
+
 \begin{equation}
@@
 Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields.
-In the nonlinear $\sigma$ model, it has been shown that an appropriate manifestation of the flow time can resolve divergencies \cite{makino2015a}.
+In the non-linear sigma model, it has been shown that an appropriate manifestation of the flow time can resolve divergencies \cite{makino2015a}.
 Therefore, it is possible to define a different flow equation for this model as well.
-Our goal is to understand this renormalizability in the nonlinear $\sigma$ model using the gradient flow.
-Specifically, we seek to define vacuum condensates, or nonpertubative vacuum expectation values.
+Our goal is to understand this renormalizability in the non-linear sigma model using the gradient flow.
 
-To explore this technique, we first plan to produce numerical calculations using a nonlinear $\sigma$ model.
-Using a Monte Carlo method in 1+1 dimensions, we can calculate the correlation functions of quantum field theories.
-With more computing power, expanding this simulation to 2+1 dimensions would provide an analysis more relevant to quantum chromodynamics.
-Following this numerical calculation, we seek to find and understand an implementation of the gradient flow in this model using analytic methods.
-This understanding can then be verified by the numerical solution.
+\section{Non-linear sigma model}
 
-\section{Field Theories on the Lattice}
-\section{Non-linear $\sigma$ model}
-
-The nonlinear $\sigma$ model is a prototypical theory for many physical phenomena, including applications in string theory.
+The non-linear sigma model is a prototypical theory for many physical phenomena, including applications in string theory.
 As a simple nonperturbative model, it provides an ideal starting point for lattice QCD studies.
-Specifically, the nonlinear $\sigma$ model exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
+Specifically, the non-linear sigma model exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
 Furthermore, it has an exact application to Heisenberg ferromagnets in condensed matter.
@@
 
-For every configuration of the non-linear sigma model, there is an associated integer quantity known as the topological charge.
-By viewing the field as a mapping from periodic real space to the sphere $S^{N-1}$, we can can associate each configuration with a member of the homotopy group $\Pi_2(S^2) = \mathbb{Z}$.
-This integer value is the topological charge of such a configuration.
+The $O(3)$ non-linear sigma model model features topological properties originating from two properties: 
+\begin{enumerate}
+    \item At $x\rightarrow\infty$, the field must become uniform since the Lagrangian must vanish.
+This allows us to model $x\rightarrow\infty$ as a single point on the field, forming a Riemann sphere in three dimensions.
+    \item The elements of the $O(3)$ non-linear sigma model are three-dimensional unit vectors, thereby existing on a three dimensional unit sphere. 
+\end{enumerate}
+
+With these two properties, we can view the field as a continuous mapping between two 3D spheres, denoted as $S^2$, and associate an integer number of wrappings to each mapping from $S^2$ to $S^2$.
+We can envision a tangible metaphor for this wrapping with a balloon and a baseball: by simply inserting the baseball into the balloon, we have established a mapping from every point on the balloon to every point on the baseball.
+We can create an equally valid map by twisting the balloon's mouth and wrapping the baseball again.
+In a purely mathematical world, we perform this process an infinite number of times, thereby associating every possible mapping with an integer.
+The group of integers is known as the \textit{homotopy group} of the non-linear sigma model.
+We associate every field configuration with an element of this group, known as the \textit{topological charge}, which we denote as $Q$.
 
@@
 \label{sec:methods}
-Our study of the gradient flow in the nonlinear $\sigma$ model consists of a computational part and an analytical part.
+Our study of the gradient flow in the non-linear sigma model consists of a computational part and an analytical part.
 We begin by outlining a numerical Monte Carlo method to simulate the lattice in 2 and 3 dimensions.
@@
 Though the Metropolis algorithm will slowly find the absolute minimum of the theory, the presence of local minima can greatly prolong the convergence.
-Both the $\phi^4$ model and the nonlinear $\sigma$ model feature ``kinetic'' terms with gradients of $\phi$.
+Both the $\phi^4$ model and the non-linear sigma model feature ``kinetic'' terms with gradients of $\phi$.
 Therefore, the presence of large similarly-valued regions in the lattice can lead to a local minimum.
@@
 To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site.
-In the non-linear $\sigma$ model, this value is derived as
+In the non-linear sigma model, this value is derived as
 \begin{equation}

### 2021-03-31 final-report.latex/main.tex
Flesh out introduction, add some phi4 results, small fixes

@@
 \usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subcaption}
 \setlength{\oddsidemargin}{0.5in}
@@
 \newcommand{\e}{\vec e}
+\newcommand{\citeneeded}{}
 
@@
 
-Quantum field theory (QFT) is an extraordinarily successful model which can be applied to a range of physical phenomena.
-Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental framework of particle physics.
-In condensed matter physics, effective field theories can model emergent effects such as phonons and quasiparticles.
+Quantum field theory (QFT) is an extraordinarily successful framework which can be applied to a range of physical phenomena.
+Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics.
+In condensed matter physics, effective field theories model emergent effects such as phonons and quasiparticles.
 Compared to experiment, QFT is remarkably accurate, famously matching the experimental value for the electron $g$-factor to eight significant figures.\cite{weisskopf1981}
@@
 
-One such example is the \textit{non-linear sigma model} which provides a prototypical theory in both condensed matter and particle physics.
-In solid-state systems, this model describes Heisenberg ferromagnets.
-In nuclear physics, the non-linear sigma model acts as a prototype for quantum chromodynamics, exhibiting characteristic features such as a mass gap and asymptotic freedom.
-% citation needed
+One such example is the \textit{non-linear sigma model}, a prototypical theory in both condensed matter and particle physics.
+In solid-state systems, this model describes Heisenberg ferromagnets and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), exhibiting characteristic features such as a mass gap and asymptotic freedom.
+\citeneeded
 
-In this study, we specifically consider the $O(3)$ nonlinear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time).
+In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time).
 This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action.
@@
 
-The gradient flow is a technique to remove these divergences.
-By dampening high-momentum fluctuations, the gradient flow reduces power-divergent mixing and make observables finite on the lattice.\cite{monahan2016} To explore this technique, we first plan to produce numerical calculations using a non-linear sigma model.
-Using a Monte Carlo method in 1+1 dimensions, we can calculate the topological properties of quantum field theories.
+The gradient flow is a technique designed to remove these divergences.
+By dampening high-momentum fluctuations, the gradient flow reduces power-divergent mixing and make observables finite on the lattice.\cite{monahan2016} In quantum chromodynamics, previous studies have verified the ability of the gradient flow to make observables finite\citeneeded.
+Due to its success in QCD, there has been interest in using the gradient flow to finitize the topological susceptibility in the 1+1 $O(3)$ non-linear sigma model.\cite{bietenholz2018}.
+
+%goals here
+
+\section{Method Overview}
+
+To numerically study the topological qualities of the non-linear sigma model, we first implement a Markov Chain Monte Carlo simulation.
+We initially construct a proof-of-concept Python program that models the simpler $\phi^4$ model (see Sec.~\ref{sec:phi4}.
+After comparing with existing literature, we transition to a C++ simulation for efficiency, implementing the non-linear sigma model in larger lattices.
+Since the gradient flow has no exact solution in the non-linear sigma model, we implement a numerical solution using a fourth-order Runge-Kutta approximation.
+By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.
 
 \section{Conventions}
-
-Throughout this paper, we use natural units, i.e. $\hbar = 1$ and $c=1$.
+\begin{itemize}
+    \item Throughout this paper, we use natural units, i.e. $\hbar = 1$ and $c=1$.
+\end{itemize}
 
@@
 
+\subsection{$\phi^4$ model}
+\label{sec:phi4}
+
+The simplest interacting field theory is known as the $\phi^4$ model. This theory describes a spin-0 boson and consists of a real scalar field given by the action
+\begin{equation}
+    \label{eq:phi4 action}
+    S[\phi] = \int d^D x \left[ \frac{1}{2}\partial^\mu \phi \partial_\mu\phi - \frac{1}{2} m_0^2 \phi^2 - \frac{\lambda}{4!}\phi^4\right].
+\end{equation}
+The first two terms describe a free relativistic particle of mass $m_0$ while the last term describes an interaction with strength $\lambda$.
+Per Einstein summation notation, there is an implicit sum over spacetime dimensions $\mu\in\{0,1,2,3\}$, indexing the derivative vectors\footnote{This canonical representation of the kinetic term $\frac{1}{2}\partial^\mu\phi\partial_\mu\phi$ is equivalent to $\frac{1}{2}\dot\phi^2-\frac{1}{2}\left(\nabla \phi\right)^2$.}
+\begin{align*}
+    \partial^\mu &= \left( \frac{\partial}{\partial t}, \frac{\partial}{\partial x},\frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) \\
+    \partial_\mu &= \left( \frac{\partial}{\partial t}, -\frac{\partial}{\partial x}, -\frac{\partial}{\partial y}, -\frac{\partial}{\partial z} \right).
+\end{align*}
+
 \subsection{Ultraviolet Divergences}
@@
 \section{Gradient Flow}
+\label{sec:gradflow}
 To remove ultraviolet fluctuations, we introduce a momentum cutoff which reduces ultraviolet divergences.
 In a gauge theory, such as QCD, a hard cutoff violates gauge invariance, so more nuanced techniques are required.
+One such technique is ``smearing,'' a local averaging designed to remove diverging fluctuations \cite{solbrig2007}.
+In this study, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.
 
-One such technique is ``smearing,'' a local averaging of a gauge field in order to remove diverging fluctuations \cite{solbrig2007}.
-In this study, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time''.
-The flow time parameterizes the smearing; an evolution in flow time corresponds to suppressing ultraviolet divergences.
-In a 2D $\phi^4$ scalar field theory, given by the action
-
-\begin{equation}
-    S_\phi [\phi] = \frac{1}{2}\int d^2x\left[(\partial \phi)^2+m^2\phi^2+\frac{\lambda}{2}\right],
-\end{equation}
-we can define the evolution in flow time as
+Specifically, the gradient flow pushes field configuration toward classical minima of the action.
+Additionally, renormalized correlation functions remain renormalized at nonzero flow time.\cite{luscher2013} In 2D $\phi^4$ scalar field theory, the gradient flow is defined by the differential equation
 \begin{equation}
@@
 \end{equation}
-where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime and $\tau$ is the flow time.
+where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime\footnote{$\partial^2 = \frac{\partial}{\partial t} - \nabla^2$} and $\tau$ is the flow time.
 Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
-When solved exactly, this function forms a Gaussian which forces operators to converge.
-Therefore, in all orders of pertubation theory, any correlation function in the gauge field that has been evolved in flow time becomes immune to these divergences\cite{makino2015a}.
+In the $\phi^4$ theory, we can solve this equation exactly to find
+\begin{equation}
+    \rho(\tau, x) = \frac{1}{4 \pi \tau} \int d^2 y e^{-(x-y)^2/4\tau} \phi(y).
+\end{equation}
+This function forms a Gaussian, smoothly dampening high-momentum modes and removing ultraviolet divergences from evolved correlation functions.\cite{makino2015a}
 
@@
 Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields.
-In the non-linear sigma model, it has been shown that an appropriate manifestation of the flow time can resolve divergencies \cite{makino2015a}.
+In the non-linear sigma model, it has been shown that an appropriate manifestation of the flow time can resolve divergences \cite{makino2015a}.
 Therefore, it is possible to define a different flow equation for this model as well.
-Our goal is to understand this renormalizability in the non-linear sigma model using the gradient flow.
 
@@
 \end{equation}
-subject to the constraint that $\vec\phi\cdot\vec\phi = 1$. In this report, we primarily consider $O(3)$, where $\vec\phi$ is a $3$ dimensional vector.
+subject to the constraint that $\vec\phi\cdot\vec\phi = 1$. In this report, we primarily consider $O(3)$, where $\vec\phi$ is a three dimensional vector.
 
@@
 We replace the derivatives with a simple difference and impose periodic boundary conditions on the system.
-For a more in-depth discussion, see Section~\ref{sec:grad_flow}.
+For a more in-depth discussion, see Section~\ref{sec:gradflow}.
 
@@
 Our study of the gradient flow in the non-linear sigma model consists of a computational part and an analytical part.
-We begin by outlining a numerical Monte Carlo method to simulate the lattice in 2 and 3 dimensions.
+We begin by outlining a numerical Monte Carlo method to simulate the lattice in two and three dimensions.
 We verify our program with the well-studied $\phi^4$ scalar field theory.
@@
 In the non-linear sigma model we choose a random unit vector $\vec r$ and consider the projection of the field on this vector.
-When the cluster flips, each site is flipped along this direction.
-\footnote{citation needed}.
-To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site.
+When the cluster flips, each site is flipped along this direction.\citeneeded To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site.
 In the non-linear sigma model, this value is derived as
@@
     \label{eq:rungekutta}
-    k_1 &= h f(\e(\tau)_x) \\ 
-    k_2 &= h f(\e(\tau)_x + \frac{k_1}{2}) \\ 
-    k_3 &= h f(\e(\tau)_x + \frac{k_2}{2}) \\ 
-    k_4 &= h f(\e(\tau)_x + k_3) \\ 
+    k_1 &= h f\left(\e\left(\tau\right)_x\right) \\ 
+    k_2 &= h f\left(\e\left(\tau\right)_x + \frac{k_1}{2}\right) \\ 
+    k_3 &= h f\left(\e\left(\tau\right)_x + \frac{k_2}{2}\right) \\ 
+    k_4 &= h f\left(\e\left(\tau\right)_x + k_3\right) \\ 
     \e(\tau+h)_x &= \e(\tau)_x + \frac{k_1}{6} + \frac{k_2}{3} + \frac{k_3}{3} + \frac{k_4}{6} + O(h^5).
@@
 \chapter{Results}
+\section{$\phi^4$ model}
+
+We initially implemented the $\phi^4$ model using the Monte Carlo method described in Sec.~\ref{sec:methods} to verify the results of our system.
+According to previous studies (\cite{monahan2016}, \cite{schaich2006}), the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ for $\lambda = 0.5$.
+We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
+
+\begin{figure}[h]
+  \centering
+      \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
+      \caption{The lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$.
+$N=32$, $\lambda=0.5$.
+The lattice was thermalized from a hot start for 1000 sweeps.
+Afterwards, 100 measurements were taken with 100 sweeps between each.
+The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
+  \label{fig:phi4}
+\end{figure}
+
+To improve efficiency, we transitioned from a Python based simulation to a C++ code.
+The comparison between these results are found in Fig.~\ref{fig:compare}, using the same values from Fig.~\ref{fig:phi4}.
+
+\begin{figure}[h]
+  \centering
+      \includegraphics[width=0.7\textwidth]{imgs/compare.png}
+      \caption{Comparison between Python and C++ code, using the lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$.
+$N=32$, $\lambda=0.5$.
+The lattice was thermalized from a hot start for 1000 sweeps.
+Afterwards, 100 measurements were taken with 100 sweeps between each.
+The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
+  \label{fig:compare}
+\end{figure}
+
+Furthermore, we implemented the gradient flow in this theory. Fig.~\ref{fig:flow} demonstrates the effect of the gradient flow on scalar lattice, specifically the dampening of high frequency modes.
+\begin{figure}[h]
+  \centering
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+        \includegraphics[width=\textwidth]{imgs/0.png}
+        \caption{$\tau=0$}
+      \end{subfigure}
+      \hfill
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+        \includegraphics[width=\textwidth]{imgs/0_5.png}
+        \caption{$\tau=0.5$}
+      \end{subfigure}
+      \hfill
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+        \includegraphics[width=\textwidth]{imgs/2.png}
+        \caption{$\tau=2$}
+      \end{subfigure}
+      \hfill
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+            \includegraphics[width=\textwidth]{imgs/10.png}
+            \caption{$\tau=10$}
+      \end{subfigure}
+      \hfill
+      \caption{\label{fig:flow} Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
+
+\end{figure}
+
 \chapter{Statistical Analysis}
@@
 \chapter{Conclusion \& Outlook}
+
+% MPI is unnecessary in this case
 

### 2021-04-02 final-report.latex/main.tex
typo
flesh out some parts

@@
 Quantum field theory (QFT) is an extraordinarily successful framework which can be applied to a range of physical phenomena.
-Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics.
+Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying subatomic particles.
 In condensed matter physics, effective field theories model emergent effects such as phonons and quasiparticles.
@@
 To numerically study the topological qualities of the non-linear sigma model, we first implement a Markov Chain Monte Carlo simulation.
-We initially construct a proof-of-concept Python program that models the simpler $\phi^4$ model (see Sec.~\ref{sec:phi4}.
+We initially construct a proof-of-concept Python program that models the simpler $\phi^4$ model (see Sec.~\ref{sec:phi4}).
 After comparing with existing literature, we transition to a C++ simulation for efficiency, implementing the non-linear sigma model in larger lattices.
@@
     \item Throughout this paper, we use natural units, i.e. $\hbar = 1$ and $c=1$.
+    \item We use Einstein summation notation, an implicit sum over repeated spacetime indices. For example, if $x^\mu$ is a spacetime four-vector and $x_\mu$ is its covariant form, the term
+        \begin{align*}
+            x^\mu x_\mu &= \sum^4_{\mu=0} x^\mu x_\mu \\
+            &= x_0^2-x_1^2-x_2^2-x_3^2.
+        \end{align*}
 \end{itemize}
@@
 \subsection{Path Integral Formulation}
+\label{sec:pathintegral}
 
@@
 Though it is possible to define this integral rigorously, for our purposes we can equate it to a sum over all possible configurations.
-This form generalizes the classical principle of least action, a relationship that Richard Feynman pedagogically explains in his lectures on physics \cite{feynman1963a}.
+This form is the quantum analog of the classical principle of least action and reduces to such for large values of the action.
+For a more pedagogical explanation, see Richard Feynman's lectures on physics.\cite{feynman1963a}
 
@@
 \begin{equation}
-    s = x^2_0 - x^2_1- x^2_3- x^2_3
+    s^2 = x^2_0 - x^2_1- x^2_3- x^2_3
 \end{equation}
-where $x_0=ct$ and $\vec{x} = (x_1, x_2, x_3)^T$. By redefining the time coordinate $x_0$ of a spacetime point $x$ to be $x_4=ix_0$, we find that the quantity
+where $x_0=ct$ and $\vec{x} = (x_1, x_2, x_3)^T$. By redefining the time coordinate of a spacetime point $x$ to be $x_4=ix_0$, we find that the quantity
 \begin{equation}
-    s_E = x^2_1+ x^2_3+ x^2_3 + x^2_4,
+    s_E^2 = x^2_1+ x^2_3+ x^2_3 + x^2_4,
 \end{equation}
-is invariant, which is representative of a four-dimensional Euclidean space. Furthermore, we find that
+is equivalently Lorentz invariant, which is representative of a four-dimensional Euclidean space. Furthermore, we find that
 \begin{align}
@@
 
-We can use this transformation to redefine the Lagrangian $\mathcal{L}$ in Euclidean space as $\mathcal{L}_E$.
-Since a Lorentz-invariant Lagrangian must include only even powers and derivatives of $x$, the Euclidean Lagrangian is necessarily real.
+We can use this transformation to redefine the Lagrangian $\mathcal{L}$ in Euclidean space as $\mathcal{L}_E$, replacing all $x_0$ with $ix_0$.
+Since a Lorentz-invariant Lagrangian must include only even powers and derivatives of $x$, the Euclidean Lagrangian remains real.
 Subsequently, we can define a Euclidean action based on the differential in Eq.~\ref{eq:wickdifferential}:
@@
 By replacing the Minkowski action $S$ with a Euclidean action $S_E$, we have transformed the amplitude $e^{iS}$ to a statistical Boltzmann factor $e^{-S_E}$.
+This new form will allow us to use statistical techniques to simulate quantum fields.
 
@@
 The first two terms describe a free relativistic particle of mass $m_0$ while the last term describes an interaction with strength $\lambda$.
-Per Einstein summation notation, there is an implicit sum over spacetime dimensions $\mu\in\{0,1,2,3\}$, indexing the derivative vectors\footnote{This canonical representation of the kinetic term $\frac{1}{2}\partial^\mu\phi\partial_\mu\phi$ is equivalent to $\frac{1}{2}\dot\phi^2-\frac{1}{2}\left(\nabla \phi\right)^2$.}
+Note that we have generalized the action for $D$ dimensions.
+Per Einstein summation notation, there is an implicit sum over spacetime dimensions $\mu\in\{0,1,2,3\}$ indexing the derivative vectors\footnote{This canonical representation of the kinetic term $\frac{1}{2}\partial^\mu\phi\partial_\mu\phi$ is equivalent to $\frac{1}{2}\dot\phi^2-\frac{1}{2}\left(\nabla \phi\right)^2$.}
 \begin{align*}
@@
 
+This field features spontaneous symmetry breaking at a critical value of $m_0^2$.
+In practice this property causes the field to spontaneously align, similarly to spins aligning in a ferromagnet.
+The name ``symmetry breaking'' refers to the transformation $\phi\rightarrow-\phi$, which changes the values of observables in the aligned regime but not the disordered regime.
+
 \subsection{Ultraviolet Divergences}
+
+In Section~\rec{sec:pathintegral},  we defined a fundamental equation of quantum fields using an ``path integral'' which encompasses an uncountably infinite configuration space.
+Likewise in perturbation theory, infinite integrals over momenta are commonplace.
+Unfortunately, these integrals often diverge, even when calculating physical values.
+
+The accepted remedy is unintuitive.
+Essentially, we adopt infinite values for the parameters of the Lagragian ($m_0^2$ and $\lambda$ in $\phi^4$ theory).
+In practice, this technique is involved and consists of two steps: regularization and renormalization.
+
+Regularization is a process which introduces a new parameter into calculations this can take multiple forms
 
@@
 Furthermore, we implemented the gradient flow in this theory. Fig.~\ref{fig:flow} demonstrates the effect of the gradient flow on scalar lattice, specifically the dampening of high frequency modes.
+
 \begin{figure}[h]

### 2021-04-02 final-report.latex/main.tex
small updates

@@
 \end{equation}
-where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime\footnote{$\partial^2 = \frac{\partial}{\partial t} - \nabla^2$} and $\tau$ is the flow time.
+where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime\footnote{Explicitly, $\partial^2 = \frac{\partial^2}{\partial t^2} + \nabla^2$} and $\tau$ is the flow time.
 Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
-In the $\phi^4$ theory, we can solve this equation exactly to find
+In the $\phi^4$ theory, we can solve this equation exactly to find \cite{monahan2016}
 \begin{equation}
@@
 \begin{equation}
-    P(\mu\rightarrow\nu) = \begin{cases} 
+    P(\phi_a\rightarrow\phi_b) = \begin{cases} 
         e^{-(S_\nu - S_\mu)} & S_\nu > S_\mu \\

### 2021-04-02 code/phi4_serial.py
small updates

@@ -162,6 +162,12 @@ class Lattice(object):
         c = np.vectorize(hls_to_rgb) (h,l,s) # --> tuple
         c = np.array(c)  # -->  array of (3,n,m) shape, but need (n,m,3)
         c = c.swapaxes(0,2)
+
+        beta = 6
+        c[:,:,0] = 1/(1+np.exp(-beta*np.real(M)))
+        c[:,:,1] = 1/(1+np.exp(-beta*np.real(M)))
+        c[:,:,2] = 1/(1+np.exp(-beta*np.real(M)))
+
         plt.figure(figsize=figsize)
         im = plt.imshow(c)
         if show:
@@ -277,16 +283,16 @@ class GifProducer(object):
 if __name__=="__main__":
     # seed(a=14231)
     L = int(sys.argv[1])
-    m = 0
-    lam = 0
+    m = -0.9
+    lam = 0.5
     ncores = 6
-    timeout = 1
+    timeout = 1000
     wolff = True
     parallel = 'parallel' in sys.argv
     l = Lattice(dim=L, m=m, l=lam)
 
     rw = RandomWalk(l)
-    wolff_rate = 1
+    wolff_rate = 10
     stop_acceptance = 0.0
 
     counter = 0
@@ -326,11 +332,12 @@ if __name__=="__main__":
             rw.full_sweep()
 
             sweep_count += 1
-            record_state(l)
 
 
         if wolff:
-            #rw.wolff()
+            record_state(l)
+            rw.wolff()
+            record_state(l)
             pass
         else:
             rw.swendsen_wang()

### 2021-04-02 final-report.latex/main.tex
unknown

@@
 Quantum field theory (QFT) is an extraordinarily successful framework which can be applied to a range of physical phenomena.
-Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying subatomic particles.
+Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics.
 In condensed matter physics, effective field theories model emergent effects such as phonons and quasiparticles.
@@
 
-Regularization is a process which introduces a new parameter into calculations this can take multiple forms
+Regularization is a process which introduces a new parameter into calculations. One example is a momentum cutoff. This technique transforms infinite momentum integral integrals as 
+\begin{equation*}
+    \int_0^\infty dk \rightarrow \int_0^\Lambda dk,
+\end{equation*}
+introducing $\Lambda$ as a regularization parameter.
+
+\subsection{Fields on the lattice}
 
@@
       \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
-      \caption{The lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$.
-$N=32$, $\lambda=0.5$.
-The lattice was thermalized from a hot start for 1000 sweeps.
-Afterwards, 100 measurements were taken with 100 sweeps between each.
-The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
-  \label{fig:phi4}
-\end{figure}
-
-To improve efficiency, we transitioned from a Python based simulation to a C++ code.
-The comparison between these results are found in Fig.~\ref{fig:compare}, using the same values from Fig.~\ref{fig:phi4}.
-
-\begin{figure}[h]
-  \centering
-      \includegraphics[width=0.7\textwidth]{imgs/compare.png}
-      \caption{Comparison between Python and C++ code, using the lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$.
-$N=32$, $\lambda=0.5$.
-The lattice was thermalized from a hot start for 1000 sweeps.
-Afterwards, 100 measurements were taken with 100 sweeps between each.
-The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
-  \label{fig:compare}
-\end{figure}
-
-Furthermore, we implemented the gradient flow in this theory. Fig.~\ref{fig:flow} demonstrates the effect of the gradient flow on scalar lattice, specifically the dampening of high frequency modes.
-
-\begin{figure}[h]
-  \centering
       \begin{subfigure}[b]{0.2\textwidth}\centering

### 2021-04-05 final-report.latex/main2.tex
add discretization procedure, slight reorganization, add to Wolff derivation

--- 2021-04-02 final-report.latex/main.tex
@@
 
-\subsection{Fields on the lattice}
-
 \section{Markov Chain Monte Carlo}
@@
 When the cluster flips, each site is flipped along this direction.\citeneeded To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site.
-In the non-linear sigma model, this value is derived as
+
+We can derive this equation using the detailed balance equation (Eq.~\ref{eq:detailedbalance}). In the non-linear sigma model, this value is derived as 
+    \label{eq:detailedbalance}
 \begin{equation}
@@
  % TODO: derive this
+\label{eq:detailedbalance}
 
@@
 \section{$\phi^4$ model}
-
 We initially implemented the $\phi^4$ model using the Monte Carlo method described in Sec.~\ref{sec:methods} to verify the results of our system.

### 2021-04-08 final-report.latex/main.tex
add topological charge derivation

@@
 \usepackage{hyperref}
+\usepackage{tikz}
 \usepackage{graphicx}
@@
 \newcommand{\e}{\vec e}
-\newcommand{\citeneeded}{}
+\newcommand{\citeneeded}{\footnote{\textit{citation needed}}}
+\newcommand{\half}{\frac{1}{2}}
 
@@
 \begin{center}
-{\bf A Wonderful Piece of Scientific Research}\\[2.3cm]
+    {\bf Topology of the $O(3)$ non-linear sigma model under the gradient flow}\\[2.3cm]
 
@@
     \label{eq:phi4 action}
-    S[\phi] = \int d^D x \left[ \frac{1}{2}\partial^\mu \phi \partial_\mu\phi - \frac{1}{2} m_0^2 \phi^2 - \frac{\lambda}{4!}\phi^4\right].
+    S[\phi] = \int d^D x \left[ \frac{1}{2}\partial^\mu \phi \partial_\mu\phi - \frac{1}{2} m_0^2 \phi^2 - \frac{\lambda}{4}\phi^4\right].
 \end{equation}
@@
 \end{align*}
+Following Sec.~\ref{sec:pathintegral}, we calculate the Euclidean action in 1+1 dimensions as 
+\begin{equation}
+    \label{eq:phi4 euclidean action}
+    S_E[\phi] = \int d^2 x_E \left[\frac{1}{2}\left(\partial_t \phi\right) + \frac{1}{2} \left(\partial_x \phi \right) + \frac{1}{2} m_0^2 \phi^2 + \frac{\lambda}{4}\phi^4\right].
+\end{equation}
+where $\partial_t$ and $\partial_x$ are the two derivatives in 1+1 Euclidean spacetime.
 
@@
 The name ``symmetry breaking'' refers to the transformation $\phi\rightarrow-\phi$, which changes the values of observables in the aligned regime but not the disordered regime.
+These two phases are known as the ``broken'' and ``symmetric'' phases and their transition is well understood.
 
-\subsection{Ultraviolet Divergences}
+\subsection{Non-linear sigma model}
 
-In Section~\rec{sec:pathintegral},  we defined a fundamental equation of quantum fields using an ``path integral'' which encompasses an uncountably infinite configuration space.
-Likewise in perturbation theory, infinite integrals over momenta are commonplace.
-Unfortunately, these integrals often diverge, even when calculating physical values.
+The non-linear sigma model (NLSM) is a prototypical theory for many physical phenomena, including applications in string theory.
+As a simple nonperturbative model, it provides an ideal starting point for lattice QCD studies.
+Specifically, the NLSM exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
+Furthermore, it has an exact application to Heisenberg ferromagnets in condensed matter.
 
-The accepted remedy is unintuitive.
-Essentially, we adopt infinite values for the parameters of the Lagragian ($m_0^2$ and $\lambda$ in $\phi^4$ theory).
-In practice, this technique is involved and consists of two steps: regularization and renormalization.
+Unlike the $phi^4$ model, which consists of a real value at each point in spacetime, the $O(3)$ NLSM consists of a 3D unit vector at each point.
+For this reason, every transformation of the field must be ``norm preserving''.
+Per its name, the non-linear sigma model features a global symmetry under the 3D orthogonal group $O(3)$.
+In other words, the theory is invariant under rotating all vectors the same way.
+To differentiate it from the $\phi^4$ model, we denote the NLSM field as $\e(x)$.
 
-Regularization is a process which introduces a new parameter into calculations. One example is a momentum cutoff. This technique transforms infinite momentum integral integrals as 
-\begin{equation*}
-    \int_0^\infty dk \rightarrow \int_0^\Lambda dk,
-\end{equation*}
-introducing $\Lambda$ as a regularization parameter.
+The theory is defined by the Euclidean action 
+\begin{equation}
+    \label{eq:nlsm euclidean action}
+    S_E = \frac{\beta}{2} \int d^Dx \partial^\mu \e \cdot \partial_\mu \e
+\end{equation}
+subject to the constraint that $\e\cdot\e = 1$. Here, $\beta$ is the inverse coupling.
 
-\subsection{Fields on the lattice}
+\subsection{NLSM $\theta$ term}
+
+In order to make the non-linear sigma model topologically nontrivial, we introduce a $\theta$ term into the action such that 
+\begin{equation}
+    S[\e] \rightarrow S[\e] - i \theta Q[\phi],
+\end{equation}
+where $Q$ is the topological charge (see Sec.~\ref{sec:topological charge}).
 
@@
 
-\section{Gradient Flow}
-\label{sec:gradflow}
-To remove ultraviolet fluctuations, we introduce a momentum cutoff which reduces ultraviolet divergences.
-In a gauge theory, such as QCD, a hard cutoff violates gauge invariance, so more nuanced techniques are required.
-One such technique is ``smearing,'' a local averaging designed to remove diverging fluctuations \cite{solbrig2007}.
-In this study, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.
+\section{Observables}
 
-Specifically, the gradient flow pushes field configuration toward classical minima of the action.
-Additionally, renormalized correlation functions remain renormalized at nonzero flow time.\cite{luscher2013} In 2D $\phi^4$ scalar field theory, the gradient flow is defined by the differential equation
+To extract physics from Monte Carlo simulations, we define a set of ``observables''.
+These quantities manifest as expectation values of operators, calculated using the Euclidean path integral formula (Eq.~\ref{eq:pathintegraleuclidean}).
+We can classify these observables into two categories: primary and secondary observables.
+Primary observables are calculated as expectation values of global operators while secondary observables are derived from these quantities.
+
+\subsection{Primary Observables}
+Each primary observable is defined on each configuration independently, meaning they do not encode ensemble statistics of the Markov Chain.\footnote{One slight exception is the magnetic susceptibility, which can be defined both ways.} In the $\phi^4$ model, we can develop an intuition around these quantities by visualizing the symmetric and broken phases.
+Fig~\ref{fig:primary observables} and Tab.~\ref{tab:primary observables} show examples of these quantities in three different configurations: one in the broken phase, one in the symmetric phase, and one at the transition.
+
+There are two potential points of confusion here.
+The first lies in the definition of ``broken`` phase.
+Though the symmetric phase more closely resembles a pane of broken glass, it leaves $\phi\rightarrow-\phi$ symmetry \textit{un}-broken, thereby giving the title ``broken'' to the more visually uniform configuration.
+An additional potential pitfall is the distinction between the \textit{lattice average} and the \textit{ensemble average}.
+The first is a mean over all lattice sites while the second is a mean over all configurations in the Markov Chain.
+
+\begin{figure}[h]
+  \centering
+      \begin{subfigure}[b]{0.3\textwidth}\centering
+        \includegraphics[width=0.6\textwidth]{imgs/broken.png}
+        \caption{broken phase}
+      \end{subfigure}%
+      \hfill
+      \begin{subfigure}[b]{0.3\textwidth}\centering
+        \includegraphics[width=0.6\textwidth]{imgs/transition.png}
+        \caption{transition}
+      \end{subfigure}%
+      \hfill
+      \begin{subfigure}[b]{0.3\textwidth}\centering
+        \includegraphics[width=0.6\textwidth]{imgs/symmetric.png}
+        \caption{symmetric phase}
+      \end{subfigure}
+      \hfill
+      \caption{\label{fig:primary observables} Visualization of broken phase, symmetric phase and transition. Simulation run on $64\times64$ lattice, plotted after 1000 sweep thermalization.}
+\end{figure}
+
+\begin{table}[h]
+    \begin{center}
+    \begin{tabular}{ c|| c | c | c}
+        & broken & transition & symmetric \\ \hline
+        $|\bar\phi|$ & 0.56 & 0.07 & 0.02 \\ 
+        $\chi_m$ & 1274.05& 22.37 & 1.98\\  
+        $\S_E/L^2$ & 0.29 & 0.40 & 0.44
+    \end{tabular}
+    \end{center}
+    \caption{\label{tab:primary observables} Primary observables with average action per site corresponding to the particular configurations in Fig.~\ref{fig:primary observables}.}
+\end{table}
+
+\subsubsection{Average Magnetization}
+The average magnetization quantifies the total alignment of the field.
+In the $\phi^4$ theory, this value is a scalar and indicates the spontaneous symmetry breaking of the field.
+In the NLSM, this value is a vector and also represents a spontaneous symmetry breaking, though this quality is outside the scope of this study.
+In both theories, a value of zero indicates the symmetric phase while a nonzero value indicates broken symmetry.
+In the NLSM, a magnitude of one represents total alignment.
+
+Mathematically, we can define this quantity in the $\phi^4$ model as 
 \begin{equation}
-    \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
+    |\bar\phi| \equiv \frac{1}{L^2}\sum_i^{L^2} \phi(x_i).
 \end{equation}
-where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime\footnote{Explicitly, $\partial^2 = \frac{\partial^2}{\partial t^2} + \nabla^2$} and $\tau$ is the flow time.
-Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
-In the $\phi^4$ theory, we can solve this equation exactly to find \cite{monahan2016}
+In the NLSM, the equivalent value is defined as 
 \begin{equation}
-    \rho(\tau, x) = \frac{1}{4 \pi \tau} \int d^2 y e^{-(x-y)^2/4\tau} \phi(y).
+    |\bar\e| \equiv \frac{1}{L^2}\sum_i^{L^2} \e(x_i).
 \end{equation}
-This function forms a Gaussian, smoothly dampening high-momentum modes and removing ultraviolet divergences from evolved correlation functions.\cite{makino2015a}
 
-Generally, we can choose any flow time equation that drives the field towards a classical minimum.
-Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields.
-In the non-linear sigma model, it has been shown that an appropriate manifestation of the flow time can resolve divergences \cite{makino2015a}.
-Therefore, it is possible to define a different flow equation for this model as well.
+\subsubsection{Magnetic Susceptibility}
 
-\section{Non-linear sigma model}
+\subsection{Secondary Observables}
+\subsubsection{Binder Cumulant}
+\subsection{Bimodality}
 
-The non-linear sigma model is a prototypical theory for many physical phenomena, including applications in string theory.
-As a simple nonperturbative model, it provides an ideal starting point for lattice QCD studies.
-Specifically, the non-linear sigma model exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
-Furthermore, it has an exact application to Heisenberg ferromagnets in condensed matter.
-The theory is defined by the Euclidean action
-\begin{equation}
-    S_E = \frac{1}{2g^2} \int d^Dx \partial^\mu \vec\phi \cdot \partial_\mu \vec\phi
-\end{equation}
-subject to the constraint that $\vec\phi\cdot\vec\phi = 1$. In this report, we primarily consider $O(3)$, where $\vec\phi$ is a three dimensional vector.
-
-Following \cite{bietenholz2018}, we can define the gradient flow in this model via the differential equation
-\begin{equation}
-    \label{eq:nsm_gradflow}
-    \partial_\tau \vec\rho (\tau,x) = \left( 1 - \vec\rho(\tau,x) \vec\rho(\tau,x)^T \right) \nabla^2 \vec\rho(\tau,x),
-\end{equation}
-where $\nabla^2$ is the Laplacian operator in Euclidean space. We solve this equation numerically using the boundary condition $\vec\rho(0,x) = \phi(x)$.
-
-In order to use this theory on the lattice, we discretize the action.
-We replace the derivatives with a simple difference and impose periodic boundary conditions on the system.
-For a more in-depth discussion, see Section~\ref{sec:gradflow}.
-
-\section{Topological Charge}
-
+\subsection{Jackknife Method}
+\section{Topological Observables}
+\label{sec:topological charge}
 The $O(3)$ non-linear sigma model model features topological properties originating from two properties: 
@@
 
+In practice, this quantity is nontrivial to calculate.
+To begin, we define a local topological charge density $q$, defined for each square of adjacent lattice points.
+This square, known as a \textit{plaquette}, is denoted $x^*$.
+The global charge $Q$ is the sum of all local charges:
+\begin{equation}
+    Q \equiv \sum_{x^*} q(x^*).
+\end{equation}
+As a function of $x^*$, the charge density is a function of the field on the plaquette vertices, an idea visualized in Fig.~\ref{fig:plaquette}.
+\begin{figure}
+    \centering
+    \begin{tikzpicture}
+        \draw[step=4cm,gray,thin] (-2,-2) grid (6,6);
+        \draw[dotted] (0,0) -- (4,4);
+        \draw[thick](0,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_1$} -- 
+                    (0,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_2$} -- 
+                    (4,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_3$} -- 
+                    (4,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_4$} -- cycle ;
+                    \draw[] (2,2) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north east]{$x^*$};
+
+        \draw[-stealth] (0.2,1) -- (0.2,3);
+        \draw[-stealth] (1,3.8) -- (3,3.8);
+        \draw[-stealth] (3,3.5) -- (1,1.5);
+
+        \draw[-stealth] (1,0.5) -- (3,2.5);
+        \draw[-stealth] (3.8,3) -- (3.8,1);
+        \draw[-stealth] (3,0.2) -- (1,0.2);
+    \end{tikzpicture}
+    \caption{\label{fig:plaquette} Visualization of plaquette $x^*$.
+Dotted line separated plaquette into two signed areas which are used to define the topological charge density $q(x^*)$.
+Arrows represent order of signed area.}
+\end{figure}
+In the NLSM, the field at each of these vertices is a point on the sphere $S^2$.
+Therefore, there is a signed area $A$ on the sphere associated with each triplet of points (the sign changes with odd permutations of the ordering).
+We follow the derivation in \cite{berg1981} and split the plaquette into two triangles, as shown in Fig.~\ref{fig:plaquette}, with the ordering determining the sign.
+The topological charge density is defined using this signed area as
+\begin{equation}
+    q(x^*) = \frac{1}{4\pi} \bigg[A\Big(\e(x_1), \e(x_3), \e(x_3)\Big) + A\Big(\e(x_1), \e(x_3), \e(x_4)\Big) \bigg].
+\end{equation}
+This sign is defined if $A\neq 0, 2\pi$, or in other words, as long as the three points on the sphere are distinct and do not form a hemisphere.
+In numerical calculations, these points can be ignored.
+Therefore, we impose that the signed area is defined on the smallest spherical triangle, or mathematically
+\begin{equation}
+    -2\pi < A < 2\pi.
+\end{equation}
+Following \cite{berg1981}, this yields an expression for the signed area
+\begin{equation}
+    A(\e_1, \e_2, \e_1) = 2 \:\mathrm{arg}\Big(1+\e_1\cdot\e_2 + \e_2\cdot\e_3 + \e_3\cdot\e_1 + i\e_1 \cdot (\e_2\times\e_3)\Big).
+\end{equation}
+Under periodic boundary conditions, these triangles on the sphere necessarily wrap $S^2$ an integer number of times, ensuring $Q$ is an integer.
+
+Following this quantity, we can define a topological susceptibility $\chi_t$
+\begin{equation}
+\chi_t \equiv \frac{1}{L^2} \Big( \langle Q^2 \rangle - \langle Q \rangle^2 \Big).
+\end{equation}
+
+\section{Ultraviolet Divergences}
+
+In Section~\ref{sec:pathintegral},  we defined a fundamental equation of quantum fields using a ``path integral'' which encompasses an uncountably infinite configuration space.
+However, we said nothing of the integral's convergence.
+In fact, many fundamental processes in QFT have divergent amplitudes, yielding nonsensical results.
+The most type of divergence stems from high-momentum states, giving them the name ``ultraviolet divergences''.
+The remedy to this catastrophe is unintuitive.
+Essentially, we adopt infinite values for the parameters of the Lagrangian ($m_0^2$ and $\lambda$ in $\phi^4$ theory).
+Since neither of these two quantities is ever measured directly, we do not have to assume that their values are finite.
+In practice, this technique is involved and consists of two steps: regularization and renormalization.
+
+\subsection{Regularization: fields on the lattice}
+Regularization is a process which introduces a new parameter into calculations. One example is a momentum cutoff. This technique transforms infinite momentum integrals as follows:  
+\begin{equation*}
+    \int_0^\infty dk \rightarrow \int_0^\Lambda dk,
+\end{equation*}
+introducing $\Lambda$ as a regularization parameter.
+This process makes results $\Lambda$-dependent, but finite.
+Another example is dimensional regularization, which calculates results in terms of the spacetime dimension $d$ and analytically continues this parameter into the real numbers.
+
+In this study, we employ lattice regularization.
+This process discretizes the field, modeling the field $\phi(x)$ as a lattice $\phi_i$ where $i$ indexes lattice sites.
+The inherent parameter in this case is the lattice spacing $a$ which measures the width of each lattice chunk.
+
+\subsection{Renormalization}
+After regularization, we redefine the Lagrangian parameters in terms of the regularization parameter, following a handful of boundary conditions.
+In this study, we assert that $L/\xi$ remains constant, where $L$ is the side length of the system and $\xi$ is the coherence length.
+In perturbation theory, this process is arduous and includes the introduction of counter-terms into the Lagrangian.
+In the case of the non-linear sigma model, it is impossible using counterterms\citeneeded but can be performed numerically.
+In this study, we use predetermined values from \cite{bietenholz2018}.
+
+At this point, we can calculate observables as functions of regularization parameters.
+To achieve physical values, we take the limit as the regularization parameters approach their physical values.
+With a momentum cutoff, we take $\Lambda \rightarrow \infty$ and with dimensional regularization we usually take $d\rightarrow 4$.
+With lattice regularization, we approach the continuum, taking the lattice spacing $a\rightarrow 0$.
+
+At this point, we have surely eliminated all divergences, right?
+Unfortunately, this is not always the case.
+The topological susceptibility $\chi_t$ is one such value that diverges in the continuum limit.
+As we decrease the width of each lattice site, high frequency modes become more significant, leading to an ultraviolet divergence in the operator.
+
+\subsection{The Gradient Flow}
+\label{sec:gradflow}
+To remove this ultraviolet divergence, we adopt a technique is ``smearing'', a local averaging of the field.\cite{solbrig2007} Specifically, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.
+
+Specifically, the gradient flow pushes field configuration toward classical minima of the action.
+Additionally, renormalized correlation functions remain renormalized at nonzero flow time.\cite{luscher2013} In 2D $\phi^4$ scalar field theory, the gradient flow is defined by the differential equation
+\begin{equation}
+    \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
+\end{equation}
+where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime and $\tau$ is the flow time.
+Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
+In the $\phi^4$ theory, we can solve this equation exactly to find \cite{monahan2016}
+\begin{equation}
+    \rho(\tau, x) = \frac{1}{4 \pi \tau} \int d^2 y e^{-(x-y)^2/4\tau} \phi(y).
+\end{equation}
+This function forms a Gaussian, smoothly dampening high-momentum modes and removing ultraviolet divergences from evolved correlation functions.\cite{makino2015a} We can visualize this by plotting the $\phi$ field, shown in Fig.~\ref{fig:flow}.
+These plots demonstrate the reduction of high momentum modes.
+
+%TODO: make this in black and white
+\begin{figure}[h]
+  \centering
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+        \includegraphics[width=\textwidth]{imgs/0.png}
+        \caption{$\tau=0$}
+      \end{subfigure}%
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+        \includegraphics[width=\textwidth]{imgs/0_5.png}
+        \caption{$\tau=0.5$}
+      \end{subfigure}%
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+        \includegraphics[width=\textwidth]{imgs/2.png}
+        \caption{$\tau=2$}
+      \end{subfigure}%
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+            \includegraphics[width=\textwidth]{imgs/10.png}
+            \caption{$\tau=10$}
+      \end{subfigure}%
+      \caption{\label{fig:flow} Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
+\end{figure}
+
+Generally, we can choose any flow time equation that drives the field towards a classical minimum.
+Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields.
+In the non-linear sigma model, it has been shown that an appropriate manifestation of the flow time can resolve divergences \cite{makino2015a}.
+Therefore, it is possible to define a different flow equation for this model as well.
+Following \cite{bietenholz2018}, we can define the gradient flow in this model via the differential equation
+\begin{equation}
+    \label{eq:nsm_gradflow}
+    \partial_\tau \vec\rho (\tau,x) = \left( 1 - \vec\rho(\tau,x) \vec\rho(\tau,x)^T \right) \partial^2 \vec\rho(\tau,x),
+\end{equation}
+where $\partial^2$ is the Laplacian operator in Euclidean space\footnote{Explicitly, $\partial^2 = \frac{\partial^2}{\partial t^2} + \nabla^2$}.
+We solve this equation numerically using the boundary condition $\vec\rho(0,x) = \e(x)$.
+
 \chapter{Methods}
@@
 We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories. 
+\section{Fields on the Lattice}
+To implement the lattice regularization technique, we must redefine the field action in terms of discrete positions, a process known as ``discretization''.
+We transition from $x$ as a continuous vector in $\mathbb{R}^2$ to $x_{i,j}$ where
+\begin{equation}
+    x_{i,j} = ia \hat{\mu}_t + j a \hat{\mu}_x.
+\end{equation}
+Here, $a$ is the lattice constant and $\hat{\mu}_0$ and $\hat{\mu}_1$ are unit vectors.
+This change effectively shifts the domain of the field from $\mathbb{R}^2$, which is uncountably infinite, to $\mathbb{Z}^2$, which is countably infinite.
+To achieve a finite domain, we impose periodic boundary conditions, such that
+\begin{equation}
+    \phi\left(x_{i+L,j}\right) = \phi\left(x_{i,j+L}\right) = \phi\left(x_{i,j}\right)
+\end{equation}
+where $L$ is the side length (in units of the lattice constant $a$) of the system. In this study we focus solely on square geometries and thus the side length $L$ is unambiguous.
+
+In the $\phi^4$ model, we can specify a discrete action using the Euclidean action from Eq. \ref{eq:phi4 euclidean action}. We begin by redefining the derivative operator as a difference:
+\begin{align}
+    \partial_\mu \phi = \frac{\phi\left(x + a \hat{\mu}\right) - \phi(x)}{a}
+\end{align}
+We can then define the kinetic term
+\begin{align}
+    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow& \frac{1}{2a^2}\left[ \left(\phi(x+a \hat{t}) -\phi(x)\right)^2 + \left(\phi(x+a \hat{x}) -\phi(x)\right)^2\right] \\
+    \begin{split} \rightarrow& \frac{1}{2a^2}\left[ \phi^2(x+a\hat{t}) + \phi^2(x+a\hat{x}) + 2\phi^2(x)\right.\\ &\qquad \left.- 2\phi(x+a\hat{t})\phi(x) - 2\phi(x+a\hat{x})\phi(x) \right]\hfilneg\end{split}
+\end{align}
+Since we will eventually sum over all sites $x$, the periodic boundary conditions imply that an overall shift in $x$ does not effect the final action.
+Therefore, we can combine the first two terms with the third term to produce
+\begin{equation}
+    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow \frac{1}{a^2}\left[ 2\phi^2(x) - \phi(x+a\hat{t})\phi(x) - \phi(x+a\hat{x})\phi(x) \right]
+\end{equation}
+Unlike the kinetic term, the mass and interaction terms remain unchanged under the discretization procedure. The only remaining change is a shift from an integral to a sum. This takes the form
+\begin{equation}
+    \int dtdx \rightarrow a^2 \sum_i
+\end{equation}
+such that the final discretized action becomes
+\begin{equation}
+    \label{eq:phi4 discretized action}
+    S_{\mathrm{lat}}[\phi] = \sum_i \left[-\phi(x_i + a\hat{t}) - \phi(x_i + a\hat{x}) + \left(2+\half m_0^2\right)\phi^2(x_i) + \frac{1}{4}\lambda \phi^4(x_i)\right]
+\end{equation}
+
+Likewise, we can discretize the non-linear sigma model. In this case, the derivative term becomes 
+\begin{equation}
+    \half \left(\partial_t\e\right)^2 + \half \left(\partial_x\e\right)^2 \rightarrow \frac{1}{a^2}\left[ 2 - \e(x+a\hat{t})\cdot\e(x) - \e(x+a\hat{x})\cdot\e(x) \right].
+\end{equation}
+Note that we have used the identity $\e\cdot\e = 1$. Inserting this into equation\label{eq:nlsm euclidean action} yields the discretized action
+\begin{equation}
+    \label{eq:nlsm discretized action}
+    S_\mathrm{lat}[\e] = \sum_i \left[ 2 - \e(x+a\hat{t})\cdot\e(x) - \e(x+a\hat{x})\cdot\e(x) \right].
+\end{equation}
+
+Finally, we redefine the gradient flow in on the lattice.
+Since the gradient flow is solved exactly in the $\phi^4$ model, we rely on a Fast Fourier Transform.
+\textit{TODO: explicit expression for this} In the NLSM, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes
+\begin{equation}
+    \label{eq:nsm_gradflow_disc}
+    \partial_\tau \e (\tau, x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T\right) \partial^2 \e(\tau,x),
+\end{equation}
+ where the Laplacian operator $\partial^2$ is defined as
+\begin{equation*}
+    \partial^2 \e(\tau,x) = \e(\tau, x+a \hat{t}) + \e(\tau,x-a\hat t) + \e(\tau, x+a \hat{x}) + \e(\tau,x-a\hat x) - 2 d \e(t)_x.
+\end{equation*}
+%TODO: derive the laplacian operator
 
@@
 \label{sec:mc}
-
 We implement a Markov Chain Monte Carlo method following Schaich's thesis \cite{schaich2006}.
@@
     P(\phi_a\rightarrow\phi_b) = \begin{cases} 
-        e^{-(S_\nu - S_\mu)} & S_\nu > S_\mu \\
+        e^{-(S[\phi_b] - S_[\phi_a])} & S[\phi_b] < S_[\phi_a] \\
         1 & \mathrm{otherwise} \\
@@
 \end{equation}
-where $S_\nu$ and $S_\mu$ are the lattice actions of configuration before and after the change.
+where $\phi_a$ is the initial configuration and $\phi_b$ is the proposed configuration.
 This process is performed for each point on the lattice, making up a ``sweep.'' Repeating this sweep many times pushes the lattice toward the action minimum.
@@
 When the cluster flips, each site is flipped along this direction.\citeneeded To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site.
-In the non-linear sigma model, this value is derived as
+
+We define a cluster $C$ as a continuous set of aligned sites which is generated probabilistically.
+In $\phi^4$ theory, we define ``aligned'' as having the same sign.
+In the non-linear sigma model, we choose a random unit vector $\vec{r}$ and use it to project the field vector.
+The sign of this quantity defines ``aligned'' in the $O(3)$ NLSM.
+In order to find the probability $p_i$ of adding a site $x_i$ to the cluster $C$, we use the detailed balance equation (Eq.~\ref{eq:detailedbalance}) which relates the Markov chain probability $P(\phi_a \rightarrow \phi_b)$ to the actions of each configuration.
+In the context of the Wolff cluster algorithm, we will notate this probability as $P\left(\phi \rightarrow f_C\left(\phi\right)\right)$ and express it as
 \begin{equation}
-    P_{add} = 1-e^{- \beta \vec e_a\vec e_b}.
+    P\left(\phi \rightarrow f_C\left(\phi\right)\right) = \left( \prod_{i \in C} p_i \right) \left( \prod_{j \in N} \left(1 - p_j \right)\right)
+\end{equation}
+where $N$ is the set of aligned neighboring sites.
+Since $p_i=0$ for sites that are not aligned with the cluster, they are not included in this expression.
+Now if we consider the probability of flipping the cluster back, i.e.
+$P\left(f_C\left(\phi\right) \rightarrow \phi\right)$, we achieve a similar expression:
+\begin{equation}
+    P\left(f_C\left(\phi\right) \rightarrow \phi  \right) = \left( \prod_{i \in C} p_i \right) \left( \prod_{j \in M} \left(1 - p_j \right)\right)
+\end{equation}
+where $M$ is the set of initially unaligned sites (they become aligned after the first flip).
+
+Using the detailed balance equations, we can now relate these quantities with the actions $S[\phi]$ and $S[f_C(\phi)]$:
+
+\begin{equation}
+    \frac{ \prod_{j \in N} \left(1 - p_j \right) }{ \prod_{j \in M} \left(1 - p_j \right) } = e^{-\left(S[f_c(\phi)] - S[\phi]\right)}.
+\end{equation}
+
+\textit{TODO: add end of derivation}
+
+\begin{equation}
+    p_i = 1-e^{- \beta \vec e_a \cdot \vec e_b}.
 \end{equation}
  % TODO: derive this
+
+Fig.~\ref{fig:wolff} shows a real demonstration of this process.
+We can see a large cluster of negative field values becoming positive.
+Note that periodic boundary conditions apply, so the small island of black at the bottom is actually part of the larger cluster.
+Furthermore, this visualization demonstrates the probabilistic nature of the Wolff algorithm.
+Since states are added probabilistically, there are some small holes in the cluster.
+These will be removed by following Metropolis sweeps.
+\begin{figure}[h]
+  \centering
+      \begin{subfigure}[b]{0.5\textwidth}\centering
+        \includegraphics[width=0.6\textwidth]{imgs/wolffa.png}
+        \caption{before cluster flip}
+      \end{subfigure}%
+      \hfill
+      \begin{subfigure}[b]{0.5\textwidth}\centering
+        \includegraphics[width=0.6\textwidth]{imgs/wolffb.png}
+        \caption{after cluster flip}
+      \end{subfigure}
+      \hfill
+      \caption{\label{fig:wolff} An example of the Wolff cluster algorithm in the $\phi^4$ model.
+White represents negative values of $\phi$ while black represents positive.
+$\lambda=0.5$, $m_0^2=-0.9$}
+
+\end{figure}
 
@@
 
-\section{Gradient Flow}
-\label{sec:gradflow_disc}
-On the lattice, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes 
-\begin{equation}
-    \label{eq:nsm_gradflow_disc}
-    \partial_\tau \e (\tau)_x = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \nabla^2 \e(\tau)_x,
-\end{equation}
- where the Laplace operator $\nabla^2$ is defined as
-\begin{equation*}
-    \nabla^2 \e(t)_x = \sum^d_{\mu=1} \left( \e(t)_{x+\hat \mu} - \e(t)_{x-\hat \mu}\right) - 2 d \e(t)_x.
-\end{equation*}
-
 \subsection{Runge-Kutta Algorithm}
@@
 \begin{equation}
-    f(\e)=\partial_\tau \e (\tau)_x  = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \nabla^2 \e(\tau)_x,
+    f(\e)=\partial_\tau \e (\tau)_x  = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \partial^2 \e(\tau)_x,
 \end{equation}
@@
 We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
-
 \begin{figure}[h]
@@
       \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=\textwidth]{imgs/0.png}
-        \caption{$\tau=0$}
-      \end{subfigure}
-      \hfill
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=\textwidth]{imgs/0_5.png}
-        \caption{$\tau=0.5$}
-      \end{subfigure}
-      \hfill
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=\textwidth]{imgs/2.png}
-        \caption{$\tau=2$}
-      \end{subfigure}
-      \hfill
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-            \includegraphics[width=\textwidth]{imgs/10.png}
-            \caption{$\tau=10$}
-      \end{subfigure}
-      \hfill
-      \caption{\label{fig:flow} Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
+      \caption{\label{fig:phi4} Simulation of the phase transition in the $\phi^4$ model: Magnetization $\langle |\bar\phi|\rangle$, magnetic susceptibility $\chi_m$, Binder cumulant $U$ and bimodality $B$ for various values of the mass squared $m_0^2$.}
 

### 2021-04-08 final-report.latex/main.tex
add sphere visualization

@@
 \usepackage{tikz}
+\usetikzlibrary{arrows.meta}
 \usepackage{graphicx}
@@
 \usepackage{subcaption}
+\usepackage{listings}
 \setlength{\oddsidemargin}{0.5in}
@@
 \newcommand{\half}{\frac{1}{2}}
+\newcommand{\codeexample}[1]{
+    \section{\texttt{#1}}
+    \lstinputlisting[language=C++]{../code/cpp_code/#1}
+}
 
@@
 \end{equation}
-Therefore, be rearranging Eq.~\ref{eq:detailedbalance}, we find
+Therefore, by rearranging Eq.~\ref{eq:detailedbalance}, we find
 \begin{equation}
-    \frac{P(\phi_a\rightarrow\phi_b)}{P(\phi_b\rightarrow\phi_a)} = e^{-(S_E[\phi_b] - S_E[\phi_a])}.
+    \label{eq:detailedbalance2}
+    \frac{P(\phi_a\rightarrow\phi_b)}{P(\phi_b\rightarrow\phi_a)} = e^{S_E[\phi_a] - S_E[\phi_b]}.
 \end{equation}
@@
         \includegraphics[width=0.6\textwidth]{imgs/broken.png}
-        \caption{broken phase}
+        \caption{broken phase ($m_0^2=-1.0$)}
       \end{subfigure}%
@@
         \includegraphics[width=0.6\textwidth]{imgs/transition.png}
-        \caption{transition}
+        \caption{transition ($m_0^2=-0.7$)}
       \end{subfigure}%
+
       \hfill
@@
         \includegraphics[width=0.6\textwidth]{imgs/symmetric.png}
-        \caption{symmetric phase}
+        \caption{symmetric phase ($m_0^2=-0.4$)}
       \end{subfigure}
@@
 In the NLSM, the field at each of these vertices is a point on the sphere $S^2$.
-Therefore, there is a signed area $A$ on the sphere associated with each triplet of points (the sign changes with odd permutations of the ordering).
+Therefore, there is a signed area $A$ on the sphere associated with each triplet of points, as shown in Fig.~\ref{fig:sphere} (the sign changes with odd permutations of the ordering).
 We follow the derivation in \cite{berg1981} and split the plaquette into two triangles, as shown in Fig.~\ref{fig:plaquette}, with the ordering determining the sign.
@@
 \end{equation}
+
 Following \cite{berg1981}, this yields an expression for the signed area
@@
 Under periodic boundary conditions, these triangles on the sphere necessarily wrap $S^2$ an integer number of times, ensuring $Q$ is an integer.
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}
+    \pgfdeclarelayer{nodelayer}
+    \pgfdeclarelayer{edgelayer}
+    \pgfdeclareradialshading{sphere4}{\pgfpoint{-0.2cm}{0.5cm}}% 
+        {rgb(0cm)=(1,1,1);
+        rgb(1cm)=(0.5,0.5,0.5); rgb(1.05cm)=(1,1,1)}
+        %rgb(0.7cm)=(0.1,0.1,0.1); rgb(1cm)=(0.5,0.05,0); rgb(1.05cm)=(1,1,1)}
+    \pgfsetlayers{nodelayer,edgelayer}
+    \tikzstyle{label}=[fill=none, draw=none, shape=circle]
+    \tikzstyle{point}=[inner sep=0pt, minimum size=0.2cm,fill=black, draw, shape=circle]
+    %\tikzstyle{interior line}=[{Stealth[scale=1.5]}-,dotted,thick]
+    \tikzstyle{interior line}=[dotted,thick]
 
+    \tikzstyle{triangle}=[thick]
+    \tikzstyle{arrow}=[->, thick]
+	\begin{pgfonlayer}{nodelayer}
+		\node [style=point] (4) at (0.5, 2.5) {};
+		\node [style=point] (5) at (0.55, 0.625) {};
+		\node [style=point] (6) at (2.25, 0.75) {};
+		\node (7) at (0, 0) {};
+		\node (8) at (1.25, 1.375) {};
+		\node (9) at (2.425, 2.525) {};
+	\end{pgfonlayer}
+	\begin{pgfonlayer}{edgelayer}
+        \draw (0,0) circle (3cm);
+        %\shade[inner color=white,outer color=lightgray] (0,0) circle (3cm);
+        \shade[shading=sphere4] (0,0) circle (3cm);
+		\draw [bend left=15,style=triangle] (4.center) to (6.center);
+		\draw [bend left=355,style=triangle] (6.center) to (5.center);
+		\draw [bend right=5,style=triangle] (5.center) to (4.center);
+		\draw [style=interior line] (4.center) to (7.center);
+		\draw [style=interior line] (5.center) to (7.center);
+        \draw [style=interior line] (6.center) to (7.center);
+        \draw [fill=gray!80] (4.center) to [bend left=15] (6.center) to [bend left=355] (5.center) to [bend right=5] cycle;
+
+		%\draw [style=arrow] (8.center) to (9.center);
+
+        \node [style=label, anchor=north] at (2.25, 0.75) {\small $\e(x_1)$};
+        \node [style=label, anchor=north] at (0.55, 0.625) {\small $\e(x_2)$};
+        \node [style=label, anchor=east] at (0.5, 2.5) {\small $\e(x_3)$};
+		\node [style=point] at (4){};
+		\node [style=point] at (5){};
+		\node [style=point] at (6){};
+
+        \node [style=label] at (8) {\small $A$};
+	\end{pgfonlayer}
+\end{tikzpicture}
+\caption{\label{fig:sphere} Visualization of signed area $A$ on the sphere $S^2$ traced out by field at points $x_1$, $x_2$ and $x_3$.}
+\end{figure}
 Following this quantity, we can define a topological susceptibility $\chi_t$
@@
     \label{eq:phi4 discretized action}
-    S_{\mathrm{lat}}[\phi] = \sum_i \left[-\phi(x_i + a\hat{t}) - \phi(x_i + a\hat{x}) + \left(2+\half m_0^2\right)\phi^2(x_i) + \frac{1}{4}\lambda \phi^4(x_i)\right]
+    S_{\mathrm{lat}}[\phi] = \sum_i \left[-\phi(x_i + a\hat{t})\phi(x_i) - \phi(x_i + a\hat{x})\phi(x_i) + \left(2+\half m_0^2\right)\phi^2(x_i) + \frac{1}{4}\lambda \phi^4(x_i)\right]
 \end{equation}
@@
 The sign of this quantity defines ``aligned'' in the $O(3)$ NLSM.
-In order to find the probability $p_i$ of adding a site $x_i$ to the cluster $C$, we use the detailed balance equation (Eq.~\ref{eq:detailedbalance}) which relates the Markov chain probability $P(\phi_a \rightarrow \phi_b)$ to the actions of each configuration.
+In order to find the probability $p_i$ of adding a site $x_i$ to the cluster $C$, we use the detailed balance equation (Eq.~\ref{eq:detailedbalance2}) which relates the Markov chain probability $P(\phi_a \rightarrow \phi_b)$ to the actions of each configuration.
 In the context of the Wolff cluster algorithm, we will notate this probability as $P\left(\phi \rightarrow f_C\left(\phi\right)\right)$ and express it as
@@
 Using the detailed balance equations, we can now relate these quantities with the actions $S[\phi]$ and $S[f_C(\phi)]$:
-
 \begin{equation}
-    \frac{ \prod_{j \in N} \left(1 - p_j \right) }{ \prod_{j \in M} \left(1 - p_j \right) } = e^{-\left(S[f_c(\phi)] - S[\phi]\right)}.
+    \frac{ \prod_{j \in N} \left(1 - p_j \right) }{ \prod_{j \in M} \left(1 - p_j \right) } = e^{S[\phi] - S[f_C(\phi)]}.
 \end{equation}
-
-\textit{TODO: add end of derivation}
+To simplify the exponent, we follow the discretized $\phi^4$ action in Eq.~\ref{eq:phi4 discretized action}.
+We see that the mass and interaction terms are invariant under a sign change of $\phi$ and therefore disappear in the difference.
+In fact, the only change in the action results from the neighboring sites $N\cup M$.
+The first two terms of Eq.~\ref{eq:phi4 discretized action} can be viewed as a sum over all pairs of neighboring lattice points $x_a$ and $x_b$, such that a change
+\begin{align}
+    \phi(x_a) &\rightarrow -\phi(x_a) \\
+    \phi(x_b) &\rightarrow \phi(x_b)
+\end{align}
+implies
+\begin{equation*}
+    -\phi(x_b)\phi(x_a) \rightarrow \phi(x_b)\phi(x_a),
+\end{equation*}
+leading to a term -$2\phi(x_a)\phi(x_b)$ in the exponent. Likewise,
 
@@
 \appendix
-\chapter{My Great Computer Program}
+\chapter{C++ NLSM Monte Carlo Program}
+\lstset{language=C++,
+                basicstyle=\ttfamily,
+                keywordstyle=\color{blue}\ttfamily,
+                stringstyle=\color{red}\ttfamily,
+                commentstyle=\color{green}\ttfamily,
+                morecomment=[l][\color{magenta}]{\#}
+}
 
-\section{Code sample}
 \bibliographystyle{unsrt}

### 2021-04-08 final-report.latex/main.tex
add code to appendix

@@
 \lstset{language=C++,
-                basicstyle=\ttfamily,
+                basicstyle=\small\ttfamily,
                 keywordstyle=\color{blue}\ttfamily,
@@
                 commentstyle=\color{green}\ttfamily,
-                morecomment=[l][\color{magenta}]{\#}
+                morecomment=[l][\color{magenta}]{\#},
+                breaklines=true,
+                postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}
 }
+
+\codeexample{sweep.h}
+\codeexample{sweep.cpp}
+\codeexample{lattice.h}
+\codeexample{lattice.cpp}
+\codeexample{phi.h}
+\codeexample{phi.cpp}
+\codeexample{observables.h}
+\codeexample{constants.h}
 

### 2021-04-08 final-report.latex/main.tex
make gradient flow plot black & white

@@
 \begin{figure}[h]
-  \centering
+    \begin{center}
       \begin{subfigure}[b]{0.3\textwidth}\centering
         \includegraphics[width=0.6\textwidth]{imgs/broken.png}
-        \caption{broken phase ($m_0^2=-1.0$)}
+        \caption{broken phase \\($m_0^2=-1.0$)}
       \end{subfigure}%
@@
         \includegraphics[width=0.6\textwidth]{imgs/transition.png}
-        \caption{transition ($m_0^2=-0.7$)}
+        \caption{transition \\($m_0^2=-0.7$)}
       \end{subfigure}%
-
       \hfill
@@
         \includegraphics[width=0.6\textwidth]{imgs/symmetric.png}
-        \caption{symmetric phase ($m_0^2=-0.4$)}
+        \caption{symmetric phase \\($m_0^2=-0.4$)}
       \end{subfigure}
@@
       \caption{\label{fig:primary observables} Visualization of broken phase, symmetric phase and transition. Simulation run on $64\times64$ lattice, plotted after 1000 sweep thermalization.}
+  \end{center}
 \end{figure}
@@
         $|\bar\phi|$ & 0.56 & 0.07 & 0.02 \\ 
-        $\chi_m$ & 1274.05& 22.37 & 1.98\\  
-        $\S_E/L^2$ & 0.29 & 0.40 & 0.44
+        $S_E/L^2$ & 0.29 & 0.40 & 0.44
     \end{tabular}
     \end{center}
-    \caption{\label{tab:primary observables} Primary observables with average action per site corresponding to the particular configurations in Fig.~\ref{fig:primary observables}.}
+    \caption{\label{tab:primary observables} Average magnetization with average action per site corresponding to the particular configurations in Fig.~\ref{fig:primary observables}.}
 \end{table}
@@
 
-%TODO: make this in black and white
 \begin{figure}[h]
@@
       \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=\textwidth]{imgs/0.png}
+        \includegraphics[width=0.9\textwidth]{imgs/gf0.png}
         \caption{$\tau=0$}
@@
       \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=\textwidth]{imgs/0_5.png}
-        \caption{$\tau=0.5$}
+        \includegraphics[width=0.9\textwidth]{imgs/gf1.png}
+        \caption{$\tau=0.001$}
       \end{subfigure}%
       \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=\textwidth]{imgs/2.png}
-        \caption{$\tau=2$}
+        \includegraphics[width=0.9\textwidth]{imgs/gf2.png}
+        \caption{$\tau=0.01$}
       \end{subfigure}%
       \begin{subfigure}[b]{0.2\textwidth}\centering
-            \includegraphics[width=\textwidth]{imgs/10.png}
-            \caption{$\tau=10$}
+        \includegraphics[width=0.9\textwidth]{imgs/gf3.png}
+        \caption{$\tau=0.1$}
       \end{subfigure}%
@@
       \caption{\label{fig:phi4} Simulation of the phase transition in the $\phi^4$ model: Magnetization $\langle |\bar\phi|\rangle$, magnetic susceptibility $\chi_m$, Binder cumulant $U$ and bimodality $B$ for various values of the mass squared $m_0^2$.}
-
 \end{figure}
@@
 \lstset{language=C++,
-                basicstyle=\small\ttfamily,
+                basicstyle=\scriptsize\ttfamily,
                 keywordstyle=\color{blue}\ttfamily,

### 2021-04-09 final-report.latex/main.tex
fix Wolff algorithm

@@
 Therefore, the presence of large similarly-valued regions in the lattice can lead to a local minimum.
-The Wolff algorithm helps reduce the presence of these clusters through two steps: identifying a cluster and flipping it along some arbitrary vector.
+One method of removing these clusters involves identifying all clusters on the lattice and probabilistically flipping each, a technique known as the Swendsen-Wang algorithm\cite{swendsen1987}.
+
+A more efficient approach is the Wolff algorithm\cite{wolff1989}, which \textit{grows} one cluster probabilistically and flips it unconditionally.
 In the case of $\phi^4$ theory, this flipping takes the form of a simple sign change.
 In the non-linear sigma model we choose a random unit vector $\vec r$ and consider the projection of the field on this vector.
-When the cluster flips, each site is flipped along this direction.\citeneeded To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site.
+When the cluster flips, each site is flipped along this direction.\citeneeded To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site, growing the cluster from a single, randomly selected ``seed''.
+Starting with the seed, the probability of adding each neighboring site is given by the source site $x$ and the proposed site $x'$.
+Wolff defines this probability for arbitrary sigma models as
+\begin{equation}
+    \label{eq:wolff padd}
+    P_{add}(\e(x),\e(x')) = \begin{cases} 
+        1 - e^{2\beta [\vec{r} \cdot \e(x)][\vec{r} \cdot \e(x')]} & \mathrm{sgn}[\vec{r}\cdot\e(x)]=\mathrm{sgn}[\vec{r}\cdot\e(x')]\\
+        0 & \mathrm{otherwise} \\
+   \end{cases}
+\end{equation}
+This expression is designed to preserve the detailed balance equations.
+We can demonstrate this quality, and motivate an equivalent expression for the $\phi^4$ model, by considering the probability $P(\phi\rightarrow f_C(\phi))$ of flipping some cluster $C$.
+Generally,
+\begin{equation}
+    P\big(\phi \rightarrow f_C\left(\phi\right)\big) \propto \prod_{\langle x,x'\rangle \in \partial C}  \Big[1 - P_{add}\big(\e(x),\e(x')\big)\Big]
+\end{equation}
+where $\partial C$ is the set of pairs of sites on the boundary of $C$.
+Since $P_{add}=0$ for unaligned sites, these pairs contribute nothing to the value.
+We can also find the probability $P(f_C(\phi) \rightarrow \phi)$ with the same expression:
+\begin{equation}
+P\big(f_C(\phi)\rightarrow \phi \big) \propto \prod_{\langle x,x'\rangle \in \partial C}  \Big[1 - P_{add}\big(\e(x),R\;\e(x')\big)\Big],
+\end{equation}
+where the matrix $R$ which is a reflection matrix along the vector $\vec{r}$.
 
-We define a cluster $C$ as a continuous set of aligned sites which is generated probabilistically.
-In $\phi^4$ theory, we define ``aligned'' as having the same sign.
-In the non-linear sigma model, we choose a random unit vector $\vec{r}$ and use it to project the field vector.
-The sign of this quantity defines ``aligned'' in the $O(3)$ NLSM.
-In order to find the probability $p_i$ of adding a site $x_i$ to the cluster $C$, we use the detailed balance equation (Eq.~\ref{eq:detailedbalance2}) which relates the Markov chain probability $P(\phi_a \rightarrow \phi_b)$ to the actions of each configuration.
-In the context of the Wolff cluster algorithm, we will notate this probability as $P\left(\phi \rightarrow f_C\left(\phi\right)\right)$ and express it as
+From the discretized action of the NLSM model (Eq.~\ref{eq:nlsm discretized action}) and the detailed balance equation (Eq.~\ref{eq:detailedbalance2}), we derive
 \begin{equation}
-    P\left(\phi \rightarrow f_C\left(\phi\right)\right) = \left( \prod_{i \in C} p_i \right) \left( \prod_{j \in N} \left(1 - p_j \right)\right)
+    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\e(x),\e(x')\big)}{ 1 - P_{add}\big(\e(x),R\;\e(x')\big)}  = \mathrm{exp}\left\{\beta\sum_{\langle x,x'\rangle \in \partial C} \e(x)\cdot[R-1]\e(x')\right\}.
 \end{equation}
-where $N$ is the set of aligned neighboring sites.
-Since $p_i=0$ for sites that are not aligned with the cluster, they are not included in this expression.
-Now if we consider the probability of flipping the cluster back, i.e.
-$P\left(f_C\left(\phi\right) \rightarrow \phi\right)$, we achieve a similar expression:
+Note that all the pairs within and outside the cluster cancel in the fraction on the left and the difference on the right. Using the definition of the reflection matrix
 \begin{equation}
-    P\left(f_C\left(\phi\right) \rightarrow \phi  \right) = \left( \prod_{i \in C} p_i \right) \left( \prod_{j \in M} \left(1 - p_j \right)\right)
+    R\;\e = \e - 2(\e\cdot\vec{r})\e,
 \end{equation}
-where $M$ is the set of initially unaligned sites (they become aligned after the first flip).
+we can simplify the equation to be 
+\begin{equation}
+    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\e(x),\e(x')\big)}{ 1 - P_{add}\big(\e(x),R\;\e(x')\big)}  = \prod_{\langle x,x'\rangle \in \partial C}\mathrm{exp}\big\{2\beta[r\cdot\e(x)][r\cdot\e(x')]\big\}.
+\end{equation}
+By plugging in Eq.~\ref{eq:wolff padd}, it is clear to see this equation is satisfied.
 
-Using the detailed balance equations, we can now relate these quantities with the actions $S[\phi]$ and $S[f_C(\phi)]$:
+Using this same reasoning, we can deduce an expression for $P_{add}$ in the $\phi^4$ model.
+Since this model is one dimensional, the reflection matrix $R$ is equivalent to $-1$.
+Adapting the NLSM detailed balance equation for the $\phi$ field, we find
 \begin{equation}
-    \frac{ \prod_{j \in N} \left(1 - p_j \right) }{ \prod_{j \in M} \left(1 - p_j \right) } = e^{S[\phi] - S[f_C(\phi)]}.
+    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\phi(x),\phi(x')\big)}{ 1 - P_{add}\big(\phi(x),-\phi(x')\big)}  = \prod_{\langle x,x'\rangle \in \partial C}\mathrm{exp}\{-2 \phi(x)\phi(x')\}.
 \end{equation}
-To simplify the exponent, we follow the discretized $\phi^4$ action in Eq.~\ref{eq:phi4 discretized action}.
-We see that the mass and interaction terms are invariant under a sign change of $\phi$ and therefore disappear in the difference.
-In fact, the only change in the action results from the neighboring sites $N\cup M$.
-The first two terms of Eq.~\ref{eq:phi4 discretized action} can be viewed as a sum over all pairs of neighboring lattice points $x_a$ and $x_b$, such that a change
-\begin{align}
-    \phi(x_a) &\rightarrow -\phi(x_a) \\
-    \phi(x_b) &\rightarrow \phi(x_b)
-\end{align}
-implies
-\begin{equation*}
-    -\phi(x_b)\phi(x_a) \rightarrow \phi(x_b)\phi(x_a),
-\end{equation*}
-leading to a term -$2\phi(x_a)\phi(x_b)$ in the exponent. Likewise,
-
+This equation is satisfied by the ansatz
 \begin{equation}
-    p_i = 1-e^{- \beta \vec e_a \cdot \vec e_b}.
+    \label{eq:wolff padd}
+    P_{add}(\phi(x),\phi(x')) = \begin{cases} 
+        1 - e^{-2\phi(x)\phi(x')]} & \mathrm{sgn}[\phi(x)]=\mathrm{sgn}[\phi(x')]\\
+        0 & \mathrm{otherwise} \\
+   \end{cases}.
 \end{equation}
- % TODO: derive this
+We will use this expression in the computational implementation of this algorithm.
 
@@
 
+\section{Topological charge with a $\theta$ term}
+
 \chapter{Results}

### 2021-04-09 final-report.latex/main.tex
add some result figures

@@
 
-Unlike the $phi^4$ model, which consists of a real value at each point in spacetime, the $O(3)$ NLSM consists of a 3D unit vector at each point.
+Unlike the $\phi^4$ model, which consists of a real value at each point in spacetime, the $O(3)$ NLSM consists of a 3D unit vector at each point.
 For this reason, every transformation of the field must be ``norm preserving''.
 Per its name, the non-linear sigma model features a global symmetry under the 3D orthogonal group $O(3)$.
-In other words, the theory is invariant under rotating all vectors the same way.
+In other words, the theory is invariant under rotating all vectors.
 To differentiate it from the $\phi^4$ model, we denote the NLSM field as $\e(x)$.
@@
 subject to the constraint that $\e\cdot\e = 1$. Here, $\beta$ is the inverse coupling.
-
-\subsection{NLSM $\theta$ term}
-
-In order to make the non-linear sigma model topologically nontrivial, we introduce a $\theta$ term into the action such that 
-\begin{equation}
-    S[\e] \rightarrow S[\e] - i \theta Q[\phi],
-\end{equation}
-where $Q$ is the topological charge (see Sec.~\ref{sec:topological charge}).
 
@@
     \begin{center}
+    {\renewcommand{\arraystretch}{1.2} %
     \begin{tabular}{ c|| c | c | c}
@@
         $S_E/L^2$ & 0.29 & 0.40 & 0.44
-    \end{tabular}
+    \end{tabular}}
+
     \end{center}
@@
 \end{figure}
+
 Following this quantity, we can define a topological susceptibility $\chi_t$
@@
 \end{equation}
+
+\subsection{NLSM $\theta$ term}
+
+In the nonlinear sigma model, we expect $\langle Q \rangle=0$. By introducing a $\theta$ term into the action,
+\begin{equation}
+    S[\e] \rightarrow S[\e] - i \theta Q[\phi],
+\end{equation}
+we can construct a topologically nontrivial model (that is, $\langle Q \rangle \neq 0$).
 
@@
       \end{subfigure}%
-      \caption{\label{fig:flow} Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
+      \caption{\label{fig:flow} Effect of flow time evolution on a random lattice in the symmetric phase. White represents positive values of $\phi$ while black represents negative.}
 \end{figure}
@@
       \caption{\label{fig:wolff} An example of the Wolff cluster algorithm in the $\phi^4$ model.
-White represents negative values of $\phi$ while black represents positive.
+White represents positive values of $\phi$ while black represents negative.
 $\lambda=0.5$, $m_0^2=-0.9$}
@@
 
+\begin{figure}[h]
+    \centering
+      \includegraphics[width=\textwidth]{imgs/theta.png}
+      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$.}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+      \includegraphics[width=\textwidth]{imgs/bietenholz.png}
+      \caption{\label{fig:bietenholz} }
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+      \includegraphics[width=0.6\textwidth]{imgs/hist.png}
+      \caption{\label{fig:hist} Histogram of topological charge values $Q$ for trivial NLSM. $L=404$, 10,000 measurements, measurements very 50 sweps, 1,000 sweep thermalization, $\tau=0$}
+\end{figure}
+
 \chapter{Statistical Analysis}

### 2021-04-09 final-report.latex/main.tex
finish observables

@@
 \subsubsection{Average Magnetization}
+\label{sec:avg mag}
 The average magnetization quantifies the total alignment of the field.
-In the $\phi^4$ theory, this value is a scalar and indicates the spontaneous symmetry breaking of the field.
-In the NLSM, this value is a vector and also represents a spontaneous symmetry breaking, though this quality is outside the scope of this study.
 In both theories, a value of zero indicates the symmetric phase while a nonzero value indicates broken symmetry.
@@
 
-Mathematically, we can define this quantity in the $\phi^4$ model as 
+Due to the $\phi\rightarrow-\phi$ symmetry of the $\phi^4$ model, the ensemble mean of the average magnetization $\langle \phi \rangle$ is $0$.
+Likewise, the $O(3)$ symmetry in the NLSM enforces $\langle \e \rangle$.
+To measure the alignment, we therefore use the magnitude of this quantity, defined in the $\phi^4$ model as
 \begin{equation}
-    |\bar\phi| \equiv \frac{1}{L^2}\sum_i^{L^2} \phi(x_i).
+|\bar\phi| \equiv \frac{1}{L^2}\left| \sum_i^{L^2} \phi(x_i)\right|.
 \end{equation}
-In the NLSM, the equivalent value is defined as 
+and in the NLSM as
 \begin{equation}
-    |\bar\e| \equiv \frac{1}{L^2}\sum_i^{L^2} \e(x_i).
+    |\bar\e| \equiv \frac{1}{L^2}\left|\sum_i^{L^2} \e(x_i)\right|.
 \end{equation}
 
+In the symmetric phase, both $\langle \bar\e \rangle$ and $\langle \bar\phi \rangle = 0$ while in the broken phase they are both nonzero.
+
 \subsubsection{Magnetic Susceptibility}
+Despite being the prototypical indicator of a phase transition, the magnitude of the magnetization is not always clear on a discretized lattice.
+For this reason, we consider other observables such as the magnetic susceptibility.
+Generally, this value is defined as
+\begin{equation}
+    \chi_m \equiv L^2 \big(\langle {\bar e}^2 \rangle - \langle \bar e \rangle^2\big)
+\end{equation}
+in the NLSM (the $\phi^2$ expression is nearly identical).
+Manifestly, this expression appears as a secondary observable since it is not defined for each lattice configuration.
+However, due to the rotation invariance of both fields, the second term disappears (see Sec.~\ref{sec:avg mag}), yielding
+\begin{align}
+    \chi_m &= L^2 \left\langle \left( \sum_x e(x) \right)^2\right\rangle  \\
+           &= L^2 \left\langle \sum_{x,y}\e(x)\cdot \e(y) \right\rangle. 
+\end{align}
+Due to translational invariance from the periodic boundary conditions, this expression becomes
+\begin{equation}
+    \chi_m = \left\langle \sum_x \e(0)\cdot\e(x) \right\rangle
+\end{equation}
+for the NLSM. Following the same logic, we derive 
+\begin{equation}
+    \chi_m = \left\langle \sum_x \phi(0)\phi(x) \right\rangle
+\end{equation}
+for the $\phi^4$ model.
+
+\subsubsection{Internal Energy}
+The internal energy is defined as \cite{berg1981}
+\begin{equation}
+    E = \frac{2}{\beta L^2} \langle S \rangle
+\end{equation}
+in the NLSM.
 
 \subsection{Secondary Observables}
+Unlike primary observables, secondary observables are defined for each \textit{ensemble}, not each configuration.
+To further ensure an accurate perspective of the phase transition in the $\phi^4$ model, we incorporate two indicators: the Binder cumulant and the bimodality.
+Since we do not measure these observables on the NLSM, we do not include their analogous expressions here.
+
 \subsubsection{Binder Cumulant}
+We define the Binder cumulant $U$ as \cite{landau2000}
+\begin{equation}
+    U \equiv 1-\frac{\langle \bar{\phi^4} \rangle}{3\langle \bar{\phi^2}\rangle^2}.
+\end{equation}
+Similar to the magnitude of the average magnetization, this formula yields $0$ in the symmetric phase and a nonzero value in the broken phase.
+This nonzero value i $2/3$ for the Binder cumulant.
+The advantage of this metric is a sharper transition at the critical $m_0^2$.
+
 \subsection{Bimodality}
+The final phase transition indicator we use is the bimodality.
+In the symmetric phase, the average magnetization $\bar\phi$ centers around $0$ while in the broken phase, these values cluster around two peaks.
+Qualitatively, this metric measures the separation of these peaks.
 
+We begin by measuring $\bar\phi$ for each configuration.
+We separate each value into an odd number number of bins, ensuring that there is a bin centered at $\bar\phi=0$.
+We then calculate the number of configurations $n_0$ in the center bin and the number of configurations $n_{max}$ in the fullest bin.
+The bimodality is then calculated as
+\begin{equation}
+    B = 1 - \frac{n_0}{n_{max}}.
+\end{equation}
+When the configurations are centered around $\bar\phi=0$, i.e.
+the symmetric phase, this value is $0$.
+When the configurations are aligned such that $\bar\phi\neq0$, i.e.
+the broken phase, this value becomes $1$.
 \subsection{Jackknife Method}
+Though it simple to measure the uncertainties associated with primary observables, secondary observables make this process more complicated.
+While we could propagate the uncertainty of the Binder cumulant, such a process is not clear for the bimodality.
+Therefore, we utilize a method known Jackknife resampling.
+
+We begin by calculating some observable $O$ on an ensemble of $N$ configurations.
+Then, for each configuration $i$, we calculate the same observable but exclude said configuration.
+This leaves us with a set of $N$ observables $O_i$.
+Assuming independent measurements, we can calculate the variance of $O$ as
+\begin{equation}
+    \mathrm{Var}(O) = \sum_{i} \left(C_i - C\right)^2.
+\end{equation}
+We use this formula to calculate all uncertainties in this study.
 \section{Topological Observables}
@@
 \end{equation}
+In the trivial case, $\langle Q \rangle$ disappears and   
+\begin{equation}
+    \chi_t = \frac{1}{L^2} \sum_{x,y} \langle q(x)q(y)\rangle.
+\end{equation}
+Assuming periodic boundary conditions and therefore translational symmetry, this expression simplifies to 
+\begin{equation}
+    \chi_t = \frac{1}{L^2} \sum_{x} \langle q(x)q(0)\rangle.
+\end{equation}
+
+On the lattice, this quantity is known to diverge in the continuum limit.\cite{bietenholz2018}
 
@@
 \end{equation}
-Note that we have used the identity $\e\cdot\e = 1$. Inserting this into equation\label{eq:nlsm euclidean action} yields the discretized action
+Note that we have used the identity $\e\cdot\e = 1$. Inserting this into Eq.~\ref{eq:nlsm euclidean action} yields the discretized action
 \begin{equation}
@@
 \begin{equation}
-    \label{eq:wolff padd}
+    \label{eq:phi4 wolff padd}
     P_{add}(\e(x),\e(x')) = \begin{cases} 
@@
 \end{equation}
-By plugging in Eq.~\ref{eq:wolff padd}, it is clear to see this equation is satisfied.
+By plugging in Eq.~\ref{eq:phi4 wolff padd}, it is clear to see this equation is satisfied.
 
@@
 \begin{equation}
-    \label{eq:wolff padd}
+    \label{eq:nlsm wolff padd}
     P_{add}(\phi(x),\phi(x')) = \begin{cases} 
@@
 \chapter{Results}
-\section{$\phi^4$ model}
+\section{$\phi^4$ model results}
 
@@
 
+\section{Non-linear sigma model results}
+After confirming the phase transition 
+\begin{figure}[h]
+    \centering
+      \includegraphics[width=\textwidth]{imgs/internal_energy.png}
+      \caption{\label{fig:bergluscher}}
+\end{figure}
+
 \begin{figure}[h]
@@
       \includegraphics[width=\textwidth]{imgs/theta.png}
-      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$.}
+      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$. Note the different scaling of the $y$-axis.}
 \end{figure}
@@
       \includegraphics[width=\textwidth]{imgs/bietenholz.png}
-      \caption{\label{fig:bietenholz} }
+      \caption{\label{fig:bietenholz} $\langle Q^2 \rangle=\chi_t L^2$ as a  }
 \end{figure}
@@
 
-\chapter{Statistical Analysis}
+%\chapter{Statistical Analysis}
 

### 2021-04-09 final-report.latex/main.tex
add theta term derivation, autocorrelation time and thermalization

@@
 This probability is denoted as $P(\mu\rightarrow\nu)$ where $\mu$ and $\nu$ are the existing and proposed lattice configurations respectively.
-We use a combination of the Metropolis and Wolff algorithms to determine this value.
 
@@
 
+\subsection{Thermalization}
+Since we begin with a random lattice, the first few configurations of our Monte Carlo simulation will be far from the lattice minimum.
+We plot the action as a function of metropolis sweeps in Fig.~\ref{fig:therm}.
+\begin{figure}[h]
+  \centering
+      \begin{subfigure}[b]{0.5\textwidth}\centering
+        \includegraphics[width=0.6\textwidth]{imgs/therm24.png}
+        \caption{$L=24$}
+      \end{subfigure}%
+      \hfill
+      \begin{subfigure}[b]{0.5\textwidth}\centering
+        \includegraphics[width=0.6\textwidth]{imgs/therm404.png}
+        \caption{$L=404$}
+      \end{subfigure}
+      \hfill
+      \caption{\label{fig:therm} Plots of the action as a function of Monte Carlo time, starting with a random NLSM lattice.}
+\end{figure}
+Based on this plot, we determine that 1000 sweeps will give sufficient time for the system to reach the classical action minimum. We use this value for the remainder of this study.
 \subsection{Autocorrelation times and thermalization}
@@
 Ideally, each sample of the lattice would be completely independent, but this is not the case.
-Therefore, we must sweep the lattice sufficiently to produce an effective Monte Carlo result.
-We use Wolff's automatic windowing procedure \cite{wolff2007} and the magnetic susceptibility $\chi_m$ to estimate the autocorrelation and determine an appropriate number of sweep between measurements for each theory.
-Furthermore, the initial values of our Markov chain will be highly correlated with the initial random start.
-Therefore, we also wait a set number of sweeps before taking any measurements so that the lattice can approach a minimum of the action.
+Since each configuration is based on previous configurations, there is a correlation between each pair of members in the Markov chain, decreasing exponentially based on the number of steps between.
+Explicitly, the autocorrelation scales as \[e^{t/\tau_{int}}\] where $t$ is the number of steps between configurations and $\tau_{int}$ is the autocorrelation time.\footnote{Though it is called a time, $\tau_{int}$ is in units of Markov Chain steps.} When performing simulations of the lattice, the number of sweeps between measurements should be much larger than $\tau_{int}$.
+
+    We use Wolff's automatic windowing procedure \cite{wolff2007} and the magnetic susceptibility $\chi_m$ to estimate the autocorrelation.
+Using Wolff's public MatLab code\footnote{\url{https://www.physik.hu-berlin.de/de/com/ALPHAsoft}}, we estimate the autocorrelation time for $L=24$ and $L=404$ lattices.
+This algorithm identifies the optimal window size with which to calculate the autocorrelation time.
+We perform this process on $L=24$ and $L=404$ lattices using a thermalization of $1000$ sweeps and $500$ total measurements.
+Note that this calculation included a Wolff cluster algorithm every $5$ sweeps.
+The result of this calculation is shown in Fig.~\ref{fig:tauint}.
+\begin{figure}[h]
+  \centering
+      \begin{subfigure}[b]{0.5\textwidth}\centering
+        \includegraphics[width=0.6\textwidth]{imgs/tauint24.png}
+        \caption{$L=24$, $\tau_{int}=4.43$}
+      \end{subfigure}%
+      \hfill
+      \begin{subfigure}[b]{0.5\textwidth}\centering
+        \includegraphics[width=0.6\textwidth]{imgs/tauint404.png}
+        \caption{$L=404$, $\tau_{int}=12.81$}
+      \end{subfigure}
+      \hfill
+      \caption{\label{fig:tauint} Plots of automatic windowing procedure used to calculate $\tau_{int}$ for the NLSM model.}
+\end{figure}
+
+Based on these two values for $\tau_{int}$, we decide to measure every $50$ sweeps for each simulation. This value will ensure that each measurement is effectively independent.
 
@@
 \section{Topological charge with a $\theta$ term}
+In Sec.~\ref{sec:topological charge}, we discussed the introduction of a $\theta$ term into the action.
+This change made the theory ``topological'', pushing $\langle Q \rangle$ away from zero.
+In order to calculate $\langle Q \rangle$ as a function of $Q$, we consider the path integral
+\begin{align}
+    \langle Q \rangle_\theta &=\int \mathcal{D}\e\:Q[\e]e^{-S[\e]+i\theta Q[\e]} \\
+                             &=\int \mathcal{D}\e\:\left( Q[\e]e^{i\theta Q[\e]} \right) e^{-S[\e]} \\
+                             &=\langle Q e^{i \theta Q} \rangle_{\theta=0}.
+\end{align}
+Therefore, we can calculate $\langle Q \rangle_\theta$ for arbitrary $\theta$ using the same simulation framework as the $\theta=0$ case.
+
+We can also relate this function to the topological susceptibility. By expanding the exponent as a Taylor series around $\theta=0$, we find that  
+\begin{equation}
+    \langle Q \rangle_\theta = \langle Q \rangle_{\theta=0} + i \theta \langle Q^2 \rangle_{\theta=0} + O(\theta^2),
+\end{equation}
+such that 
+\begin{align}
+    -i \frac{\partial}{\partial \theta} \langle Q \rangle_\theta &= \langle Q^2 \rangle_{\theta=0}\\
+                                                              &= \chi_t L^2.
+\end{align}
+We expect to see the slope of this function at zero increase for large lattice sizes.
 

### 2021-04-09 final-report.latex/main.tex
results

@@
 \paragraph
-\indent I would like to thank lots of people, and this is where
-I will do it...
+\indent I would like to thank my advisor, Prof.
+Christopher Monahan.
+His constant helpfulness has made this research possible and enjoyable.
+I would also like to thank the entire faculty of the William \& Mary Physics Department.
+The amount of knowledge and wisdom they have bestowed upon me has been beyond my imagination.
+
+I also acknowledge the help the High Performance Computing group at William \& Mary, whose cluster performed many of these simulations.
+And finally, I would like to acknowledge the financial help of the 1693 Scholars Program which funded the summer portion of this research.
 
@@
 \paragraph
-\indent \textit{Abstract}
+\indent The $O(3)$ non-linear model features topological effects such as instantons but cannot be perturbatively renormalized.
+Other nonperturbative theories such as QCD have motivated the development of lattice quantum field theory, where fields are simulated as discrete arrays and extrapolated to the continuum limit.
+Unfortunately, operators such as the topological susceptibility diverge in this limit.
+The gradient flow is a strong contender to remove these infinites, having prior success in QCD.
+We confirm a previous result\cite{bietenholz2018} that the gradient flow does not make the topological susceptibility infinite.
+We also study the nontrivial field theory by introducing a $\theta$ term into the action and plot the topological charge as a function of $\theta$.
 
@@
 We associate every field configuration with an element of this group, known as the \textit{topological charge}, which we denote as $Q$.
+Configurations with $|Q|=1$ are known as instantons.
 
@@
 \end{figure}
+In the trivial NLSM action (when $\theta=0$), the values of $Q$ are roughly Gaussian around $Q=0$, as shown in Fig.~\ref{fig:hist}.
+\begin{figure}[h]
+    \centering
+      \includegraphics[width=0.6\textwidth]{imgs/hist.png}
+      \caption{\label{fig:hist} Histogram of topological charge values $Q$ for trivial NLSM. $L=404$, 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization, $\tau=0$}
+\end{figure}
 
@@
       \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/therm24.png}
+        \includegraphics[width=0.9\textwidth]{imgs/therm24.png}
         \caption{$L=24$}
@@
       \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/therm404.png}
+        \includegraphics[width=0.9\textwidth]{imgs/therm404.png}
         \caption{$L=404$}
@@
 Based on this plot, we determine that 1000 sweeps will give sufficient time for the system to reach the classical action minimum. We use this value for the remainder of this study.
-\subsection{Autocorrelation times and thermalization}
+\subsection{Autocorrelation}
 One important aspect is the correlation between different states of the Markov chain.
@@
 Since each configuration is based on previous configurations, there is a correlation between each pair of members in the Markov chain, decreasing exponentially based on the number of steps between.
-Explicitly, the autocorrelation scales as \[e^{t/\tau_{int}}\] where $t$ is the number of steps between configurations and $\tau_{int}$ is the autocorrelation time.\footnote{Though it is called a time, $\tau_{int}$ is in units of Markov Chain steps.} When performing simulations of the lattice, the number of sweeps between measurements should be much larger than $\tau_{int}$.
+Explicitly, the autocorrelation scales as $e^{t/\tau_{int}}$ where $t$ is the number of steps between configurations and $\tau_{int}$ is the autocorrelation time.\footnote{Though it is called a time, $\tau_{int}$ is in units of Markov Chain steps.} When performing simulations of the lattice, the number of sweeps between measurements should be much larger than $\tau_{int}$.
 
@@
       \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/tauint24.png}
+        \includegraphics[width=0.9\textwidth]{imgs/tauint24.png}
         \caption{$L=24$, $\tau_{int}=4.43$}
@@
       \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/tauint404.png}
+        \includegraphics[width=0.9\textwidth]{imgs/tauint404.png}
         \caption{$L=404$, $\tau_{int}=12.81$}
@@
 \section{Topological charge with a $\theta$ term}
+\label{sec:topotheta}
 In Sec.~\ref{sec:topological charge}, we discussed the introduction of a $\theta$ term into the action.
@@
 \begin{align}
-    -i \frac{\partial}{\partial \theta} \langle Q \rangle_\theta &= \langle Q^2 \rangle_{\theta=0}\\
+    -i \frac{\partial}{\partial \theta}\Big|_{\theta=0} \langle Q \rangle_\theta &= \langle Q^2 \rangle_{\theta=0}\\
                                                               &= \chi_t L^2.
 \end{align}
-We expect to see the slope of this function at zero increase for large lattice sizes.
+Since $\chi_t$ diverges at $\theta=0$, we expect the plot of $\langle Q\rangle_\theta$ to approach a vertical line in the in continuum limit.
 
@@
 
-We initially implemented the $\phi^4$ model using the Monte Carlo method described in Sec.~\ref{sec:methods} to verify the results of our system.
-According to previous studies (\cite{monahan2016}, \cite{schaich2006}), the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ for $\lambda = 0.5$.
-We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
+We initially implement the $\phi^4$ model to verify the results of our system.
+According to previous studies (\cite{monahan2016, schaich2006}, the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ for $\lambda = 0.5$.
+We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the magnetic susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
 \begin{figure}[h]
@@
       \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
-      \caption{\label{fig:phi4} Simulation of the phase transition in the $\phi^4$ model: Magnetization $\langle |\bar\phi|\rangle$, magnetic susceptibility $\chi_m$, Binder cumulant $U$ and bimodality $B$ for various values of the mass squared $m_0^2$.}
-\end{figure}
+      \caption{\label{fig:phi4} The lattice average $|\langle\bar\phi\rangle|$, the lattice suscpetibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$.
+$L=32$, $\lambda=0.5$.
+The lattice was thermalized from a hot start for 1000 sweeps.
+Afterwards, 100 measurements were taken with 100 sweeps between each.
+The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
+\end{figure}
+This figure confirms a phase transition near $m_0^2=-0.72$ and supports the accurancy of our model.
 
 \section{Non-linear sigma model results}
-After confirming the phase transition 
+After confirming the phase transition in the $\phi^4$ model, we turn to the NLSM.
+In order to confirm the accuracy of the model, we compare results with existing literature.
+We first compare our results to Berg \& L\"uscher \cite{berg1981}, specifically measuring the internal energy and magnetic susceptibility.
+
+\subsection{Comparison with Existing Literature}
+Following \cite{berg1981}, we approximate the internal energy in the strong ($\beta<1$) and weak ($\beta>2$) regimes as 
+\begin{equation}
+E \approx \begin{cases} 
+    4-4y-8y^3-\frac{48}{5}y^5& \beta<1 \\
+    \frac{2}{\beta} + \frac{4\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2. 
+\end{cases}
+\end{equation}
+We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{bergluscher}.
 \begin{figure}[h]
@@
 \end{figure}
-
+These two charts show a high degree of agreement with the current literature.
+
+We also seek to confirm the results from Bietenholz et al.\cite{bietenholz2018}.
+Specifically, that the topological susceptibility $\chi_t$ diverges in the continuum limit even at finite flow time.
+Since $\chi_t$ is in units of inverse distance squared, we multiply by $\xi_2^2$, the square of the second moment correlation length, to achieve a scale-invariant value $\chi_t\xi_2^2$.
+Additionally, we use a parameter $t_0$ to scale the flow time such that $t_0\sim L^2$.
+In our Monte Carlo simulation, we adapt the same values as \cite{bietenholz2018} for $\xi_2$.
+
+To begin the comparison, we plot $\chi_t\xi_2^2$ as a function of flow time $\tau$, shown in Fig.~\ref{bietenholz}.
+\begin{figure}[h]
+    \centering
+      \includegraphics[width=\textwidth]{imgs/bietenholz.png}
+      \caption{\label{fig:bietenholz} $\langle Q^2 \rangle=\chi_t L^2$ as a  function of flow time $\tau$.
+Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization.}
+\end{figure}
+We find that the flow time effectively decreases the topological susceptibility by dampening high-momentum modes.
+To analyze the divergence of $\chi_t$ in the continuum limit, we plot $\chi_t \xi_2^2$ as a function of lattice size $L$.
+We perform this simulation at flow time $\tau=5t_0$.
+\begin{figure}[h]
+    \centering
+      \includegraphics[width=\textwidth]{imgs/divergence.png}
+      \caption{\label{fig:bietenholz} $\chi_t\xi_2^2$ as a function of $L$ for flow time $\tau=5t_0$.
+The data is fit with both a log and power fit.
+Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization.}
+\end{figure}
+The data is fit with two options: a log fit
+\begin{equation}
+    \chi_t \xi_2^2 = A \mathrm{log}(B L) + C
+\end{equation}
+and a power fit
+\begin{equation}
+    \chi_t \xi_2^2 = A B^L + C.
+\end{equation}
+The data fits these functions with $\chi^2/DOF$ of $31.3$ and $761.6$ respectively.
+Both of these function diverge as $L\rightarrow \intfy$, indicating that the topological susceptibility does as well.
+This is still a slower divergence than the linear divergence featured in the $\tau=0$ case.
+
+\subsection{Topological charge when $\theta\neq 0$}
+Following the method explained in Sec.~\ref{sec:topotheta}, we calculate the imaginary part of $\langle Q \rangle$ for arbitrary $\theta$.
+We perform this calculation for four values of the flow time $\tau$, shown in Fig~\ref{fig:theta}.
 \begin{figure}[h]
@@
       \includegraphics[width=\textwidth]{imgs/theta.png}
-      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$. Note the different scaling of the $y$-axis.}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=\textwidth]{imgs/bietenholz.png}
-      \caption{\label{fig:bietenholz} $\langle Q^2 \rangle=\chi_t L^2$ as a  }
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=0.6\textwidth]{imgs/hist.png}
-      \caption{\label{fig:hist} Histogram of topological charge values $Q$ for trivial NLSM. $L=404$, 10,000 measurements, measurements very 50 sweps, 1,000 sweep thermalization, $\tau=0$}
-\end{figure}
+      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$.
+Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization.
+Note the different scaling of the $y$-axis.}
+\end{figure}
+These plots demonstrate the divergence of the continuum limit in the $\tau=0$ and the flowed regimes.
+In the $\tau=0$ case, the slope increases sharply, reflecting the linear divergence of $\chi_t$.
+However in the flowed regime, this divergence is much slower, reflecting the logarithmic or power-law relationship shown in Fig.~\ref{fig:divergence}.
 
@@
 
+Using a Monte Carlo simulation, we have analyzed two main quantum field theories: the $\phi^4$ model and the $O(3)$ non-linear sigma model.
+In the former, we verified a phase transition at $m_0^2=-0.72$.
 % MPI is unnecessary in this case
@@
 \appendix
+
 \chapter{C++ NLSM Monte Carlo Program}

### 2021-04-09 final-report.latex/main.tex
conclusions

@@
 \begin{equation}
+    \label{eq:analytic_energy}
 E \approx \begin{cases} 
@@
 \end{equation}
-We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{bergluscher}.
+We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{fig:bergluscher}.
 \begin{figure}[h]
     \centering
-      \includegraphics[width=\textwidth]{imgs/internal_energy.png}
-      \caption{\label{fig:bergluscher}}
+      \includegraphics[width=0.6\textwidth]{imgs/internal_energy.png}
+      \caption{\label{fig:bergluscher}Comparison with \cite{berg1981}.
+First panel: internal energy compared with analytic  energy (Eq.~\ref{eq:analytic_energy}).
+Second panel: magnetic susceptibility compared with literature values.}
 \end{figure}
@@
 The data fits these functions with $\chi^2/DOF$ of $31.3$ and $761.6$ respectively.
-Both of these function diverge as $L\rightarrow \intfy$, indicating that the topological susceptibility does as well.
+Both of these function diverge as $L\rightarrow \infty$, indicating that the topological susceptibility does as well.
 This is still a slower divergence than the linear divergence featured in the $\tau=0$ case.
@@
 In the $\tau=0$ case, the slope increases sharply, reflecting the linear divergence of $\chi_t$.
-However in the flowed regime, this divergence is much slower, reflecting the logarithmic or power-law relationship shown in Fig.~\ref{fig:divergence}.
+However in the flowed regime, this divergence is much slower, reflecting the logarithmic or power-law relationship shown in Fig.~\ref{fig:bietenholz}.
 
@@
 Using a Monte Carlo simulation, we have analyzed two main quantum field theories: the $\phi^4$ model and the $O(3)$ non-linear sigma model.
-In the former, we verified a phase transition at $m_0^2=-0.72$.
+To perform the calculation, we run a series of Metropolis sweeps interspersed with Wolff cluster algorithms, creating a Markov chain of samples.
+We use these field configurations to calculate operators using a Wick-rotated path integral.
+
+In the $\phi^4$ model, we verified a phase transition at $m_0^2=-0.72$ using the magnitude of the average magnetization, the magnetic susceptibility, the Binder cumulant and the bimodality.
+This process verified the effectiveness of the Monte Carlo computational system.
+We then generalized our system to the $O(3)$ non-linear sigma model, confirming our calculations with known results.
+We calculated the topological susceptibility $\chi_t$ to analyze its divergence in the continuum limit, confirming that this is indeed the case even at finite flow time.
+Specifically, the gradient flow reduces the linear divergence of the susceptibility to either a power-law relationship or a logarithmic relationship with $\chi^2/DOF$ of $31.3$ and $761.6$ respectively.
+
+Finally, we produced a plot demonstrating the relationship between the topological charge and the $\theta$ in the topological action.
+This plot demonstrates qualitatively the different divergences of the topological susceptibility, which is equivalent to the slope at $\theta=0$.
+In the $\tau=0$ case, this slope became vertical rapidly for higher lattice sizes while this process occurred more slowly at nonzero flow time.
+
+Though the logarithmic and power-law functions visually fit the susceptibility data well, the $\chi_2$ values are high.
+This imprecision can be attributed to errors that are too small.
+Future work could include more independent simulations which may reduce the variance between points or increase the statistical errors.
+Furthermore, more and larger lattices may help reduce the $\chi^2$ value.
+
+Procedurally, we found that the MPI parallelization and checkerboard algorithm was unnecessary in this calculation.
+The computation time of the Runge-Kutta algorithm for computing the gradient flow far outweighed that of the Monte Carlo simulation.
+A possible improvement could be a parallelization of the Runge-Kutta algorithm or a more accurate approximation to the gradient flow.
+
+Generally, this study has implications in both condensed matter and nuclear physics as the NLSM is used in both.
+Though the calculation of the $\chi_t$ divergence confirms existing literature, the relationship between the topological charge and $\theta$ in the flow time was previously unexplored.
+
+Other future work may include explorations of higher order non-linear sigma models or other topological effective field theories.
+The latter of these models has applications in topological quantum computing and other technologies.
+
 % MPI is unnecessary in this case

### 2021-04-19 final-report.latex/conclusion.tex
split into chapter files

--- 2021-04-09 final-report.latex/main.tex
@@
-\documentclass[12pt]{report}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{hyperref}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta}
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{listings}
-\setlength{\oddsidemargin}{0.5in}
-\setlength{\evensidemargin}{0.5in}
-\setlength{\textwidth}{6.0in}
-\setlength{\topmargin}{.75in}
-\setlength{\textheight}{8.6in}
+Using a Monte Carlo simulation, we have analyzed two main quantum field theories: the $\phi^4$ model and the $O(3)$ non-linear sigma model.
+To perform the calculation, we run a series of Metropolis sweeps interspersed with Wolff cluster algorithms, creating a Markov chain of samples.
+We use these field configurations to calculate operators using a Wick-rotated path integral.
 
-\newcommand{\e}{\vec e}
-\newcommand{\citeneeded}{\footnote{\textit{citation needed}}}
-\newcommand{\half}{\frac{1}{2}}
-\newcommand{\codeexample}[1]{
-    \section{\texttt{#1}}
-    \lstinputlisting[language=C++]{../code/cpp_code/#1}
-}
+In the $\phi^4$ model, we verified a phase transition at $m_0^2=-0.72$ using the magnitude of the average magnetization, the magnetic susceptibility, the Binder cumulant and the bimodality.
+This process verified the effectiveness of the Monte Carlo computational system.
+We then generalized our system to the $O(3)$ non-linear sigma model, confirming our calculations with known results.
+We calculated the topological susceptibility $\chi_t$ to analyze its divergence in the continuum limit, confirming that this is indeed the case even at finite flow time.
+Specifically, the gradient flow reduces the linear divergence of the susceptibility to either a power-law relationship or a logarithmic relationship with $\chi^2/DOF$ of $31.3$ and $761.6$ respectively.
 
-% Modified by S. Aubin for the class of 2019 Physics theses
+Finally, we produced a plot demonstrating the relationship between the topological charge and the $\theta$ in the topological action.
+This plot demonstrates qualitatively the different divergences of the topological susceptibility, which is equivalent to the slope at $\theta=0$.
+In the $\tau=0$ case, this slope became vertical rapidly for higher lattice sizes while this process occurred more slowly at nonzero flow time.
 
-%
-% This is a LaTeX template for an undergraduate honors or
-%   senior thesis in physics at William & Mary. Created from
-%   several theses from previous students of D.S. Armstrong,
-%   pared down to bare bones.
-%
-%   you will need this present file (thesis_template.tex)
-%   as well as the file  figure1.eps  to make the whole thing
-%   (the latter is just an encapsulated postscript figure)
-%     For simplicity, we don't use bibtex here; experts can choose
-%   to use that more powerful way of dealing with citations.
-%
-%   To compile it, do:
-% > pdflatex thesis_template.tex
-% > pdflatex thesis_template.tex  (yes, you need to do it twice to
-%                                  fill in the list of tables, figures
-%                                  and references).
+Though the logarithmic and power-law functions visually fit the susceptibility data well, the $\chi_2$ values are high.
+This imprecision can be attributed to errors that are too small.
+Future work could include more independent simulations which may reduce the variance between points or increase the statistical errors.
+Furthermore, more and larger lattices may help reduce the $\chi^2$ value.
 
-\begin{document}
-\setlength{\baselineskip}{24pt}
-\begin{titlepage}
-\LARGE
-\begin{center}
-    {\bf Topology of the $O(3)$ non-linear sigma model under the gradient flow}\\[2.3cm]
+Procedurally, we found that the MPI parallelization and checkerboard algorithm was unnecessary in this calculation.
+The computation time of the Runge-Kutta algorithm for computing the gradient flow far outweighed that of the Monte Carlo simulation.
+A possible improvement could be a parallelization of the Runge-Kutta algorithm or a more accurate approximation to the gradient flow.
 
-\normalsize
-A thesis submitted in partial fulfillment of the requirement \\
-for the degree of Bachelor of Science with Honors in \\
-Physics from the College of William and Mary in Virginia,\\[0.5cm]
-by\\[0.5cm]
-Stuart Thomas \\[2.5cm]
-\end{center}
-\normalsize
-\begin{flushright}
-%\hfill Accepted for Honors \\[.5cm]
-\hfill \hrulefill \\
-\hfill \hfill Advisor: Prof. Christopher J. Monahan\\[.6cm]
-\hfill \hrulefill \\Prof. Todd Averett \\[.6cm]
-\hfill \hrulefill \\Prof. Andreas Stathopoulos \\[.6cm]
-\end{flushright}
-\begin{center}
-Williamsburg, Virginia\\
-May 2021
-\end{center}
-\end{titlepage}
+Generally, this study has implications in both condensed matter and nuclear physics as the NLSM is used in both.
+Though the calculation of the $\chi_t$ divergence confirms existing literature, the relationship between the topological charge and $\theta$ in the flow time was previously unexplored.
 
-\setlength{\topmargin}{0.0in}
-
-\pagenumbering{roman}
-\tableofcontents
-\chapter*{Acknowledgments}
-\addcontentsline{toc}{chapter}{Acknowledgments}
-\paragraph
-\indent I would like to thank lots of people, and this is where
-I will do it...
-
-\listoffigures
-\addcontentsline{toc}{chapter}{List of Figures}
-\listoftables
-\addcontentsline{toc}{chapter}{List of Tables}
-
-\begin{abstract}
-\setcounter{page}{5}
-\addcontentsline{toc}{chapter}{Abstract}
-\paragraph
-\indent \textit{Abstract}
-
-\end{abstract}
-
-\pagenumbering{arabic}
-\chapter{Introduction}
-
-Quantum field theory (QFT) is an extraordinarily successful framework which can be applied to a range of physical phenomena.
-Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics.
-In condensed matter physics, effective field theories model emergent effects such as phonons and quasiparticles.
-Compared to experiment, QFT is remarkably accurate, famously matching the experimental value for the electron $g$-factor to eight significant figures.\cite{weisskopf1981}
-
-However, this power comes at a cost: the study of quantum fields is rife with infinities.
-A na\"ive treatment of quantum field theory produces divergent values for physical quantities, a clearly impossible result.
-Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the so-called \textit{renormalization group}.
-However, the technique fails with perturbatively nonrenormalizable theories.
-
-One such example is the \textit{non-linear sigma model}, a prototypical theory in both condensed matter and particle physics.
-In solid-state systems, this model describes Heisenberg ferromagnets and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), exhibiting characteristic features such as a mass gap and asymptotic freedom.
-\citeneeded
-
-In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time).
-This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action.
-
-Since the non-linear sigma model cannot be renormalized perturbatively, we cannot study these topological effects with normal perturbative techniques.
-An alternative solution is placing the field on a discretized lattice, a technique originally used for quantum chromodynamics.
-In this scenario, field configurations become computationally calculable.
-This process introduces an nonphysical length scale $a$, the lattice spacing.
-Therefore, we expect any physical result to converge in the continuum limit, i.e.
-when $a\rightarrow 0$.
-However, this is not always the case as observables mix on the lattice, leading to divergences.
-As an example, states of definite angular momentum mix when discretized, a clear violation of the angular momentum commutation relations.
-
-The gradient flow is a technique designed to remove these divergences.
-By dampening high-momentum fluctuations, the gradient flow reduces power-divergent mixing and make observables finite on the lattice.\cite{monahan2016} In quantum chromodynamics, previous studies have verified the ability of the gradient flow to make observables finite\citeneeded.
-Due to its success in QCD, there has been interest in using the gradient flow to finitize the topological susceptibility in the 1+1 $O(3)$ non-linear sigma model.\cite{bietenholz2018}.
-
-%goals here
-
-\section{Method Overview}
-
-To numerically study the topological qualities of the non-linear sigma model, we first implement a Markov Chain Monte Carlo simulation.
-We initially construct a proof-of-concept Python program that models the simpler $\phi^4$ model (see Sec.~\ref{sec:phi4}).
-After comparing with existing literature, we transition to a C++ simulation for efficiency, implementing the non-linear sigma model in larger lattices.
-Since the gradient flow has no exact solution in the non-linear sigma model, we implement a numerical solution using a fourth-order Runge-Kutta approximation.
-By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.
-
-\section{Conventions}
-\begin{itemize}
-    \item Throughout this paper, we use natural units, i.e. $\hbar = 1$ and $c=1$.
-    \item We use Einstein summation notation, an implicit sum over repeated spacetime indices. For example, if $x^\mu$ is a spacetime four-vector and $x_\mu$ is its covariant form, the term
-        \begin{align*}
-            x^\mu x_\mu &= \sum^4_{\mu=0} x^\mu x_\mu \\
-            &= x_0^2-x_1^2-x_2^2-x_3^2.
-        \end{align*}
-\end{itemize}
-
-%\section{Main Experimental Goal}
-
-\chapter{Theory}
-
-This thesis incorporates two main bodies of knowledge: quantum field theory and statistical simulation.
-Through the path integral formulation of quantum field theory, we are able to describe the physics of the former with the established mathematics of the latter.
-
-\section{Quantum Field Theory}
-
-In this section we outline a rough description of quantum field theory.
-A full introduction is beyond the scope of this paper, however we do assume knowledge of nonrelativistic quantum mechanics and classical field theory.
-
-The fundamental hypothesis of quantum field theory (QFT) describes particles as discrete packets of energy on a quantum field.
-But what is a quantum field?
-Like in classical mechanics, a field is a function of spacetime with some mathematical object assigned to each point in space and time.
-In the case of the electric field, this object is a three-dimensional vector, while the electric potential is a scalar field.
-Classical and quantum fields have Lagrangians which define how they evolve in time.
-What differentiates a quantum field from a classical field is superposition: where classical fields have a definite configuration, quantum fields exist in a superposition of all possible configurations.
-It is possible---though nontrivial and outside the scope of this description--- to motivate the appearance of discrete particles from this superposition.
-
-This general description allows QFT to easily incorporate special relativity.
-By ensuring that the Lagrangian of a theory is invariant under Lorentz transformations, we can ensure that the theory is physical.
-% is this enough?
-
-Though most introductions to QFT use the ``second quantization,'' we will use a alternative formulation known as the path integral formulation (for a similar introduction see Zee's textbook \cite{zee2010}).
-
-\subsection{Path Integral Formulation}
-\label{sec:pathintegral}
-
-As stated above, we can model a quantum field theory as a superposition of all possible classical fields.
-Like single-particle quantum mechanics, each configuration has a probability amplitude.
-To measure expectation values of observables, we simply take an average over all configurations weighted by this complex amplitude.
-We can formalize this notion using the fundamental formula\footnote{Since this study concerns vacua, we do not include a source term.}
-\begin{equation}
-    \label{eq:pathintegral}
-    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; e^{iS[\phi]}
-\end{equation}
-% not sure if the semicolons is good form
-where $\langle \hat O \rangle$ is the expectation value of an arbitrary operator $\hat O$; $Z$ is a normalization constant; $S$ is the action functional, defined from the field's Lagrangian; and $\int \mathcal{D}\phi$ represents the eponymous path integral.
-Though it is possible to define this integral rigorously, for our purposes we can equate it to a sum over all possible configurations.
-This form is the quantum analog of the classical principle of least action and reduces to such for large values of the action.
-For a more pedagogical explanation, see Richard Feynman's lectures on physics.\cite{feynman1963a}
-
-At first glance, Eq.~\ref{eq:pathintegral} is remarkably similar to the statistics of the canonical ensemble.
-Through this similarity, we will be able to use mathematical tools from statistical mechanics to study quantum field theories.
-However, the factor of $i$ in the exponent currently prohibits us from making this jump.
-To remedy this issue, we perform a ``Wick rotation'' which shifts spacetime into Euclidean coordinates.
-In normal spacetime, defined by the Minkowski metric, the Lorentz-invariant distance is given as
-\begin{equation}
-    s^2 = x^2_0 - x^2_1- x^2_3- x^2_3
-\end{equation}
-where $x_0=ct$ and $\vec{x} = (x_1, x_2, x_3)^T$. By redefining the time coordinate of a spacetime point $x$ to be $x_4=ix_0$, we find that the quantity
-\begin{equation}
-    s_E^2 = x^2_1+ x^2_3+ x^2_3 + x^2_4,
-\end{equation}
-is equivalently Lorentz invariant, which is representative of a four-dimensional Euclidean space. Furthermore, we find that
-\begin{align}
-    d^4x_E &= d^3\vec{x}dx_4 \\
-    &= i d^3\vec{x}dx_0 \\
-    &= i d^4x. \label{eq:wickdifferential}
-\end{align}
-
-We can use this transformation to redefine the Lagrangian $\mathcal{L}$ in Euclidean space as $\mathcal{L}_E$, replacing all $x_0$ with $ix_0$.
-Since a Lorentz-invariant Lagrangian must include only even powers and derivatives of $x$, the Euclidean Lagrangian remains real.
-Subsequently, we can define a Euclidean action based on the differential in Eq.~\ref{eq:wickdifferential}:
-\begin{align}
-    S_E &= \int d^4x_E \mathcal{L}_E \\
-    &= i \int d^4x \mathcal{L}_E,
-\end{align}
-allowing us to redefine the path integral as 
-\begin{equation}
-    \label{eq:pathintegraleuclidean}
-    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O (\phi)\; e^{-S_E[\phi]}.
-\end{equation}
-
-By replacing the Minkowski action $S$ with a Euclidean action $S_E$, we have transformed the amplitude $e^{iS}$ to a statistical Boltzmann factor $e^{-S_E}$.
-This new form will allow us to use statistical techniques to simulate quantum fields.
-
-\subsection{$\phi^4$ model}
-\label{sec:phi4}
-
-The simplest interacting field theory is known as the $\phi^4$ model. This theory describes a spin-0 boson and consists of a real scalar field given by the action
-\begin{equation}
-    \label{eq:phi4 action}
-    S[\phi] = \int d^D x \left[ \frac{1}{2}\partial^\mu \phi \partial_\mu\phi - \frac{1}{2} m_0^2 \phi^2 - \frac{\lambda}{4}\phi^4\right].
-\end{equation}
-The first two terms describe a free relativistic particle of mass $m_0$ while the last term describes an interaction with strength $\lambda$.
-Note that we have generalized the action for $D$ dimensions.
-Per Einstein summation notation, there is an implicit sum over spacetime dimensions $\mu\in\{0,1,2,3\}$ indexing the derivative vectors\footnote{This canonical representation of the kinetic term $\frac{1}{2}\partial^\mu\phi\partial_\mu\phi$ is equivalent to $\frac{1}{2}\dot\phi^2-\frac{1}{2}\left(\nabla \phi\right)^2$.}
-\begin{align*}
-    \partial^\mu &= \left( \frac{\partial}{\partial t}, \frac{\partial}{\partial x},\frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) \\
-    \partial_\mu &= \left( \frac{\partial}{\partial t}, -\frac{\partial}{\partial x}, -\frac{\partial}{\partial y}, -\frac{\partial}{\partial z} \right).
-\end{align*}
-Following Sec.~\ref{sec:pathintegral}, we calculate the Euclidean action in 1+1 dimensions as 
-\begin{equation}
-    \label{eq:phi4 euclidean action}
-    S_E[\phi] = \int d^2 x_E \left[\frac{1}{2}\left(\partial_t \phi\right) + \frac{1}{2} \left(\partial_x \phi \right) + \frac{1}{2} m_0^2 \phi^2 + \frac{\lambda}{4}\phi^4\right].
-\end{equation}
-where $\partial_t$ and $\partial_x$ are the two derivatives in 1+1 Euclidean spacetime.
-
-This field features spontaneous symmetry breaking at a critical value of $m_0^2$.
-In practice this property causes the field to spontaneously align, similarly to spins aligning in a ferromagnet.
-The name ``symmetry breaking'' refers to the transformation $\phi\rightarrow-\phi$, which changes the values of observables in the aligned regime but not the disordered regime.
-These two phases are known as the ``broken'' and ``symmetric'' phases and their transition is well understood.
-
-\subsection{Non-linear sigma model}
-
-The non-linear sigma model (NLSM) is a prototypical theory for many physical phenomena, including applications in string theory.
-As a simple nonperturbative model, it provides an ideal starting point for lattice QCD studies.
-Specifically, the NLSM exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
-Furthermore, it has an exact application to Heisenberg ferromagnets in condensed matter.
-
-Unlike the $\phi^4$ model, which consists of a real value at each point in spacetime, the $O(3)$ NLSM consists of a 3D unit vector at each point.
-For this reason, every transformation of the field must be ``norm preserving''.
-Per its name, the non-linear sigma model features a global symmetry under the 3D orthogonal group $O(3)$.
-In other words, the theory is invariant under rotating all vectors.
-To differentiate it from the $\phi^4$ model, we denote the NLSM field as $\e(x)$.
-
-The theory is defined by the Euclidean action 
-\begin{equation}
-    \label{eq:nlsm euclidean action}
-    S_E = \frac{\beta}{2} \int d^Dx \partial^\mu \e \cdot \partial_\mu \e
-\end{equation}
-subject to the constraint that $\e\cdot\e = 1$. Here, $\beta$ is the inverse coupling.
-
-\section{Markov Chain Monte Carlo}
-
-To accomplish a statistical analysis of quantum fields, we use a Monte Carlo simulation.
-This method amounts to producing a large number of configurations and calculating statistics on the sample.
-A brute-force calculation over all possible configurations---as Eq.~\ref{eq:pathintegraleuclidean} suggests--- is clearly infinite and computationally infeasible.
-However, the exponential nature of the Boltzmann factor dictates that only configurations near the action minimum contribute to observable statistics.
-Therefore, by selecting a sample of configurations near this minimum, we are able to extract meaningful results with a finite computation.
-
-\subsection{The Markov Chain}
-The Markov chain is an effective method to identity these low-action states.
-Essentially, we begin with a random configuration and then make small adjustments, gradually lowering the action.
-Starting with a field configuration $\phi_a$, we propose a new field $\phi_b$.
-The probability of accepting this change, thereby adding the configuration to the chain, is given by the function
-\[P(\phi_a \rightarrow \phi_b).\]
-
-There are four requirements that this function must obey to produce a Boltzmann distribution of samples:
-\begin{enumerate}
-    \item $P(\phi_a \rightarrow \phi_b)$ must depend only on the configurations $\phi_a$ and $\phi_b$.
-    \item The probability must be properly normalized, i.e. $\sum_{\phi} P(\phi_a \rightarrow \phi) = 1$.
-    \item Every configuration must be reachable in a finite number of steps. In other words, the chain must be ergodic.
-    \item In order to reach equilibrium, the chain must be reversible.
-In other words, the probability of a $\phi_a\rightarrow\phi_b$ transition must be equal to the probability of a $\phi_b\rightarrow\phi_a$ transition.
-Mathematically, this condition takes the form of \textit{detailed balance equations}:
-\begin{equation}
-    \label{eq:detailedbalance}
-    P(\phi_a)\,P(\phi_a\rightarrow\phi_b) = P(\phi_b)\,P(\phi_b\rightarrow\phi_a),
-\end{equation}
-where $P(\phi)$ is the probability of a system existing in state $\phi$.
-% this needs citation
-\end{enumerate}
-
-This final condition will allow us to explicitly define the transition probability using the action. From the Boltzmann distribution, we know
-\begin{equation}
-    P(\phi) = \frac{1}{Z} e^{-S_E[\phi]}.
-\end{equation}
-Therefore, by rearranging Eq.~\ref{eq:detailedbalance}, we find
-\begin{equation}
-    \label{eq:detailedbalance2}
-    \frac{P(\phi_a\rightarrow\phi_b)}{P(\phi_b\rightarrow\phi_a)} = e^{S_E[\phi_a] - S_E[\phi_b]}.
-\end{equation}
-
-This formula will provide the explicit probabilities of the Metropolis and Wolff algorithms.
-
-\section{Observables}
-
-To extract physics from Monte Carlo simulations, we define a set of ``observables''.
-These quantities manifest as expectation values of operators, calculated using the Euclidean path integral formula (Eq.~\ref{eq:pathintegraleuclidean}).
-We can classify these observables into two categories: primary and secondary observables.
-Primary observables are calculated as expectation values of global operators while secondary observables are derived from these quantities.
-
-\subsection{Primary Observables}
-Each primary observable is defined on each configuration independently, meaning they do not encode ensemble statistics of the Markov Chain.\footnote{One slight exception is the magnetic susceptibility, which can be defined both ways.} In the $\phi^4$ model, we can develop an intuition around these quantities by visualizing the symmetric and broken phases.
-Fig~\ref{fig:primary observables} and Tab.~\ref{tab:primary observables} show examples of these quantities in three different configurations: one in the broken phase, one in the symmetric phase, and one at the transition.
-
-There are two potential points of confusion here.
-The first lies in the definition of ``broken`` phase.
-Though the symmetric phase more closely resembles a pane of broken glass, it leaves $\phi\rightarrow-\phi$ symmetry \textit{un}-broken, thereby giving the title ``broken'' to the more visually uniform configuration.
-An additional potential pitfall is the distinction between the \textit{lattice average} and the \textit{ensemble average}.
-The first is a mean over all lattice sites while the second is a mean over all configurations in the Markov Chain.
-
-\begin{figure}[h]
-    \begin{center}
-      \begin{subfigure}[b]{0.3\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/broken.png}
-        \caption{broken phase \\($m_0^2=-1.0$)}
-      \end{subfigure}%
-      \hfill
-      \begin{subfigure}[b]{0.3\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/transition.png}
-        \caption{transition \\($m_0^2=-0.7$)}
-      \end{subfigure}%
-      \hfill
-      \begin{subfigure}[b]{0.3\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/symmetric.png}
-        \caption{symmetric phase \\($m_0^2=-0.4$)}
-      \end{subfigure}
-      \hfill
-      \caption{\label{fig:primary observables} Visualization of broken phase, symmetric phase and transition. Simulation run on $64\times64$ lattice, plotted after 1000 sweep thermalization.}
-  \end{center}
-\end{figure}
-
-\begin{table}[h]
-    \begin{center}
-    {\renewcommand{\arraystretch}{1.2} %
-    \begin{tabular}{ c|| c | c | c}
-        & broken & transition & symmetric \\ \hline
-        $|\bar\phi|$ & 0.56 & 0.07 & 0.02 \\ 
-        $S_E/L^2$ & 0.29 & 0.40 & 0.44
-    \end{tabular}}
-
-    \end{center}
-    \caption{\label{tab:primary observables} Average magnetization with average action per site corresponding to the particular configurations in Fig.~\ref{fig:primary observables}.}
-\end{table}
-
-\subsubsection{Average Magnetization}
-\label{sec:avg mag}
-The average magnetization quantifies the total alignment of the field.
-In both theories, a value of zero indicates the symmetric phase while a nonzero value indicates broken symmetry.
-In the NLSM, a magnitude of one represents total alignment.
-
-Due to the $\phi\rightarrow-\phi$ symmetry of the $\phi^4$ model, the ensemble mean of the average magnetization $\langle \phi \rangle$ is $0$.
-Likewise, the $O(3)$ symmetry in the NLSM enforces $\langle \e \rangle$.
-To measure the alignment, we therefore use the magnitude of this quantity, defined in the $\phi^4$ model as
-\begin{equation}
-|\bar\phi| \equiv \frac{1}{L^2}\left| \sum_i^{L^2} \phi(x_i)\right|.
-\end{equation}
-and in the NLSM as
-\begin{equation}
-    |\bar\e| \equiv \frac{1}{L^2}\left|\sum_i^{L^2} \e(x_i)\right|.
-\end{equation}
-
-In the symmetric phase, both $\langle \bar\e \rangle$ and $\langle \bar\phi \rangle = 0$ while in the broken phase they are both nonzero.
-
-\subsubsection{Magnetic Susceptibility}
-Despite being the prototypical indicator of a phase transition, the magnitude of the magnetization is not always clear on a discretized lattice.
-For this reason, we consider other observables such as the magnetic susceptibility.
-Generally, this value is defined as
-\begin{equation}
-    \chi_m \equiv L^2 \big(\langle {\bar e}^2 \rangle - \langle \bar e \rangle^2\big)
-\end{equation}
-in the NLSM (the $\phi^2$ expression is nearly identical).
-Manifestly, this expression appears as a secondary observable since it is not defined for each lattice configuration.
-However, due to the rotation invariance of both fields, the second term disappears (see Sec.~\ref{sec:avg mag}), yielding
-\begin{align}
-    \chi_m &= L^2 \left\langle \left( \sum_x e(x) \right)^2\right\rangle  \\
-           &= L^2 \left\langle \sum_{x,y}\e(x)\cdot \e(y) \right\rangle. 
-\end{align}
-Due to translational invariance from the periodic boundary conditions, this expression becomes
-\begin{equation}
-    \chi_m = \left\langle \sum_x \e(0)\cdot\e(x) \right\rangle
-\end{equation}
-for the NLSM. Following the same logic, we derive 
-\begin{equation}
-    \chi_m = \left\langle \sum_x \phi(0)\phi(x) \right\rangle
-\end{equation}
-for the $\phi^4$ model.
-
-\subsubsection{Internal Energy}
-The internal energy is defined as \cite{berg1981}
-\begin{equation}
-    E = \frac{2}{\beta L^2} \langle S \rangle
-\end{equation}
-in the NLSM.
-
-\subsection{Secondary Observables}
-Unlike primary observables, secondary observables are defined for each \textit{ensemble}, not each configuration.
-To further ensure an accurate perspective of the phase transition in the $\phi^4$ model, we incorporate two indicators: the Binder cumulant and the bimodality.
-Since we do not measure these observables on the NLSM, we do not include their analogous expressions here.
-
-\subsubsection{Binder Cumulant}
-We define the Binder cumulant $U$ as \cite{landau2000}
-\begin{equation}
-    U \equiv 1-\frac{\langle \bar{\phi^4} \rangle}{3\langle \bar{\phi^2}\rangle^2}.
-\end{equation}
-Similar to the magnitude of the average magnetization, this formula yields $0$ in the symmetric phase and a nonzero value in the broken phase.
-This nonzero value i $2/3$ for the Binder cumulant.
-The advantage of this metric is a sharper transition at the critical $m_0^2$.
-
-\subsection{Bimodality}
-The final phase transition indicator we use is the bimodality.
-In the symmetric phase, the average magnetization $\bar\phi$ centers around $0$ while in the broken phase, these values cluster around two peaks.
-Qualitatively, this metric measures the separation of these peaks.
-
-We begin by measuring $\bar\phi$ for each configuration.
-We separate each value into an odd number number of bins, ensuring that there is a bin centered at $\bar\phi=0$.
-We then calculate the number of configurations $n_0$ in the center bin and the number of configurations $n_{max}$ in the fullest bin.
-The bimodality is then calculated as
-\begin{equation}
-    B = 1 - \frac{n_0}{n_{max}}.
-\end{equation}
-When the configurations are centered around $\bar\phi=0$, i.e.
-the symmetric phase, this value is $0$.
-When the configurations are aligned such that $\bar\phi\neq0$, i.e.
-the broken phase, this value becomes $1$.
-\subsection{Jackknife Method}
-Though it simple to measure the uncertainties associated with primary observables, secondary observables make this process more complicated.
-While we could propagate the uncertainty of the Binder cumulant, such a process is not clear for the bimodality.
-Therefore, we utilize a method known Jackknife resampling.
-
-We begin by calculating some observable $O$ on an ensemble of $N$ configurations.
-Then, for each configuration $i$, we calculate the same observable but exclude said configuration.
-This leaves us with a set of $N$ observables $O_i$.
-Assuming independent measurements, we can calculate the variance of $O$ as
-\begin{equation}
-    \mathrm{Var}(O) = \sum_{i} \left(C_i - C\right)^2.
-\end{equation}
-We use this formula to calculate all uncertainties in this study.
-\section{Topological Observables}
-\label{sec:topological charge}
-The $O(3)$ non-linear sigma model model features topological properties originating from two properties: 
-\begin{enumerate}
-    \item At $x\rightarrow\infty$, the field must become uniform since the Lagrangian must vanish.
-This allows us to model $x\rightarrow\infty$ as a single point on the field, forming a Riemann sphere in three dimensions.
-    \item The elements of the $O(3)$ non-linear sigma model are three-dimensional unit vectors, thereby existing on a three dimensional unit sphere. 
-\end{enumerate}
-
-With these two properties, we can view the field as a continuous mapping between two 3D spheres, denoted as $S^2$, and associate an integer number of wrappings to each mapping from $S^2$ to $S^2$.
-We can envision a tangible metaphor for this wrapping with a balloon and a baseball: by simply inserting the baseball into the balloon, we have established a mapping from every point on the balloon to every point on the baseball.
-We can create an equally valid map by twisting the balloon's mouth and wrapping the baseball again.
-In a purely mathematical world, we perform this process an infinite number of times, thereby associating every possible mapping with an integer.
-The group of integers is known as the \textit{homotopy group} of the non-linear sigma model.
-We associate every field configuration with an element of this group, known as the \textit{topological charge}, which we denote as $Q$.
-
-In practice, this quantity is nontrivial to calculate.
-To begin, we define a local topological charge density $q$, defined for each square of adjacent lattice points.
-This square, known as a \textit{plaquette}, is denoted $x^*$.
-The global charge $Q$ is the sum of all local charges:
-\begin{equation}
-    Q \equiv \sum_{x^*} q(x^*).
-\end{equation}
-As a function of $x^*$, the charge density is a function of the field on the plaquette vertices, an idea visualized in Fig.~\ref{fig:plaquette}.
-\begin{figure}
-    \centering
-    \begin{tikzpicture}
-        \draw[step=4cm,gray,thin] (-2,-2) grid (6,6);
-        \draw[dotted] (0,0) -- (4,4);
-        \draw[thick](0,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_1$} -- 
-                    (0,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_2$} -- 
-                    (4,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_3$} -- 
-                    (4,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_4$} -- cycle ;
-                    \draw[] (2,2) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north east]{$x^*$};
-
-        \draw[-stealth] (0.2,1) -- (0.2,3);
-        \draw[-stealth] (1,3.8) -- (3,3.8);
-        \draw[-stealth] (3,3.5) -- (1,1.5);
-
-        \draw[-stealth] (1,0.5) -- (3,2.5);
-        \draw[-stealth] (3.8,3) -- (3.8,1);
-        \draw[-stealth] (3,0.2) -- (1,0.2);
-    \end{tikzpicture}
-    \caption{\label{fig:plaquette} Visualization of plaquette $x^*$.
-Dotted line separated plaquette into two signed areas which are used to define the topological charge density $q(x^*)$.
-Arrows represent order of signed area.}
-\end{figure}
-In the NLSM, the field at each of these vertices is a point on the sphere $S^2$.
-Therefore, there is a signed area $A$ on the sphere associated with each triplet of points, as shown in Fig.~\ref{fig:sphere} (the sign changes with odd permutations of the ordering).
-We follow the derivation in \cite{berg1981} and split the plaquette into two triangles, as shown in Fig.~\ref{fig:plaquette}, with the ordering determining the sign.
-The topological charge density is defined using this signed area as
-\begin{equation}
-    q(x^*) = \frac{1}{4\pi} \bigg[A\Big(\e(x_1), \e(x_3), \e(x_3)\Big) + A\Big(\e(x_1), \e(x_3), \e(x_4)\Big) \bigg].
-\end{equation}
-This sign is defined if $A\neq 0, 2\pi$, or in other words, as long as the three points on the sphere are distinct and do not form a hemisphere.
-In numerical calculations, these points can be ignored.
-Therefore, we impose that the signed area is defined on the smallest spherical triangle, or mathematically
-\begin{equation}
-    -2\pi < A < 2\pi.
-\end{equation}
-
-Following \cite{berg1981}, this yields an expression for the signed area
-\begin{equation}
-    A(\e_1, \e_2, \e_1) = 2 \:\mathrm{arg}\Big(1+\e_1\cdot\e_2 + \e_2\cdot\e_3 + \e_3\cdot\e_1 + i\e_1 \cdot (\e_2\times\e_3)\Big).
-\end{equation}
-Under periodic boundary conditions, these triangles on the sphere necessarily wrap $S^2$ an integer number of times, ensuring $Q$ is an integer.
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}
-    \pgfdeclarelayer{nodelayer}
-    \pgfdeclarelayer{edgelayer}
-    \pgfdeclareradialshading{sphere4}{\pgfpoint{-0.2cm}{0.5cm}}% 
-        {rgb(0cm)=(1,1,1);
-        rgb(1cm)=(0.5,0.5,0.5); rgb(1.05cm)=(1,1,1)}
-        %rgb(0.7cm)=(0.1,0.1,0.1); rgb(1cm)=(0.5,0.05,0); rgb(1.05cm)=(1,1,1)}
-    \pgfsetlayers{nodelayer,edgelayer}
-    \tikzstyle{label}=[fill=none, draw=none, shape=circle]
-    \tikzstyle{point}=[inner sep=0pt, minimum size=0.2cm,fill=black, draw, shape=circle]
-    %\tikzstyle{interior line}=[{Stealth[scale=1.5]}-,dotted,thick]
-    \tikzstyle{interior line}=[dotted,thick]
-
-    \tikzstyle{triangle}=[thick]
-    \tikzstyle{arrow}=[->, thick]
-	\begin{pgfonlayer}{nodelayer}
-		\node [style=point] (4) at (0.5, 2.5) {};
-		\node [style=point] (5) at (0.55, 0.625) {};
-		\node [style=point] (6) at (2.25, 0.75) {};
-		\node (7) at (0, 0) {};
-		\node (8) at (1.25, 1.375) {};
-		\node (9) at (2.425, 2.525) {};
-	\end{pgfonlayer}
-	\begin{pgfonlayer}{edgelayer}
-        \draw (0,0) circle (3cm);
-        %\shade[inner color=white,outer color=lightgray] (0,0) circle (3cm);
-        \shade[shading=sphere4] (0,0) circle (3cm);
-		\draw [bend left=15,style=triangle] (4.center) to (6.center);
-		\draw [bend left=355,style=triangle] (6.center) to (5.center);
-		\draw [bend right=5,style=triangle] (5.center) to (4.center);
-		\draw [style=interior line] (4.center) to (7.center);
-		\draw [style=interior line] (5.center) to (7.center);
-        \draw [style=interior line] (6.center) to (7.center);
-        \draw [fill=gray!80] (4.center) to [bend left=15] (6.center) to [bend left=355] (5.center) to [bend right=5] cycle;
-
-		%\draw [style=arrow] (8.center) to (9.center);
-
-        \node [style=label, anchor=north] at (2.25, 0.75) {\small $\e(x_1)$};
-        \node [style=label, anchor=north] at (0.55, 0.625) {\small $\e(x_2)$};
-        \node [style=label, anchor=east] at (0.5, 2.5) {\small $\e(x_3)$};
-		\node [style=point] at (4){};
-		\node [style=point] at (5){};
-		\node [style=point] at (6){};
-
-        \node [style=label] at (8) {\small $A$};
-	\end{pgfonlayer}
-\end{tikzpicture}
-\caption{\label{fig:sphere} Visualization of signed area $A$ on the sphere $S^2$ traced out by field at points $x_1$, $x_2$ and $x_3$.}
-\end{figure}
-
-Following this quantity, we can define a topological susceptibility $\chi_t$
-\begin{equation}
-\chi_t \equiv \frac{1}{L^2} \Big( \langle Q^2 \rangle - \langle Q \rangle^2 \Big).
-\end{equation}
-In the trivial case, $\langle Q \rangle$ disappears and   
-\begin{equation}
-    \chi_t = \frac{1}{L^2} \sum_{x,y} \langle q(x)q(y)\rangle.
-\end{equation}
-Assuming periodic boundary conditions and therefore translational symmetry, this expression simplifies to 
-\begin{equation}
-    \chi_t = \frac{1}{L^2} \sum_{x} \langle q(x)q(0)\rangle.
-\end{equation}
-
-On the lattice, this quantity is known to diverge in the continuum limit.\cite{bietenholz2018}
-
-\subsection{NLSM $\theta$ term}
-
-In the nonlinear sigma model, we expect $\langle Q \rangle=0$. By introducing a $\theta$ term into the action,
-\begin{equation}
-    S[\e] \rightarrow S[\e] - i \theta Q[\phi],
-\end{equation}
-we can construct a topologically nontrivial model (that is, $\langle Q \rangle \neq 0$).
-
-\section{Ultraviolet Divergences}
-
-In Section~\ref{sec:pathintegral},  we defined a fundamental equation of quantum fields using a ``path integral'' which encompasses an uncountably infinite configuration space.
-However, we said nothing of the integral's convergence.
-In fact, many fundamental processes in QFT have divergent amplitudes, yielding nonsensical results.
-The most type of divergence stems from high-momentum states, giving them the name ``ultraviolet divergences''.
-The remedy to this catastrophe is unintuitive.
-Essentially, we adopt infinite values for the parameters of the Lagrangian ($m_0^2$ and $\lambda$ in $\phi^4$ theory).
-Since neither of these two quantities is ever measured directly, we do not have to assume that their values are finite.
-In practice, this technique is involved and consists of two steps: regularization and renormalization.
-
-\subsection{Regularization: fields on the lattice}
-Regularization is a process which introduces a new parameter into calculations. One example is a momentum cutoff. This technique transforms infinite momentum integrals as follows:  
-\begin{equation*}
-    \int_0^\infty dk \rightarrow \int_0^\Lambda dk,
-\end{equation*}
-introducing $\Lambda$ as a regularization parameter.
-This process makes results $\Lambda$-dependent, but finite.
-Another example is dimensional regularization, which calculates results in terms of the spacetime dimension $d$ and analytically continues this parameter into the real numbers.
-
-In this study, we employ lattice regularization.
-This process discretizes the field, modeling the field $\phi(x)$ as a lattice $\phi_i$ where $i$ indexes lattice sites.
-The inherent parameter in this case is the lattice spacing $a$ which measures the width of each lattice chunk.
-
-\subsection{Renormalization}
-After regularization, we redefine the Lagrangian parameters in terms of the regularization parameter, following a handful of boundary conditions.
-In this study, we assert that $L/\xi$ remains constant, where $L$ is the side length of the system and $\xi$ is the coherence length.
-In perturbation theory, this process is arduous and includes the introduction of counter-terms into the Lagrangian.
-In the case of the non-linear sigma model, it is impossible using counterterms\citeneeded but can be performed numerically.
-In this study, we use predetermined values from \cite{bietenholz2018}.
-
-At this point, we can calculate observables as functions of regularization parameters.
-To achieve physical values, we take the limit as the regularization parameters approach their physical values.
-With a momentum cutoff, we take $\Lambda \rightarrow \infty$ and with dimensional regularization we usually take $d\rightarrow 4$.
-With lattice regularization, we approach the continuum, taking the lattice spacing $a\rightarrow 0$.
-
-At this point, we have surely eliminated all divergences, right?
-Unfortunately, this is not always the case.
-The topological susceptibility $\chi_t$ is one such value that diverges in the continuum limit.
-As we decrease the width of each lattice site, high frequency modes become more significant, leading to an ultraviolet divergence in the operator.
-
-\subsection{The Gradient Flow}
-\label{sec:gradflow}
-To remove this ultraviolet divergence, we adopt a technique is ``smearing'', a local averaging of the field.\cite{solbrig2007} Specifically, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.
-
-Specifically, the gradient flow pushes field configuration toward classical minima of the action.
-Additionally, renormalized correlation functions remain renormalized at nonzero flow time.\cite{luscher2013} In 2D $\phi^4$ scalar field theory, the gradient flow is defined by the differential equation
-\begin{equation}
-    \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
-\end{equation}
-where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime and $\tau$ is the flow time.
-Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
-In the $\phi^4$ theory, we can solve this equation exactly to find \cite{monahan2016}
-\begin{equation}
-    \rho(\tau, x) = \frac{1}{4 \pi \tau} \int d^2 y e^{-(x-y)^2/4\tau} \phi(y).
-\end{equation}
-This function forms a Gaussian, smoothly dampening high-momentum modes and removing ultraviolet divergences from evolved correlation functions.\cite{makino2015a} We can visualize this by plotting the $\phi$ field, shown in Fig.~\ref{fig:flow}.
-These plots demonstrate the reduction of high momentum modes.
-
-\begin{figure}[h]
-  \centering
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=0.9\textwidth]{imgs/gf0.png}
-        \caption{$\tau=0$}
-      \end{subfigure}%
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=0.9\textwidth]{imgs/gf1.png}
-        \caption{$\tau=0.001$}
-      \end{subfigure}%
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=0.9\textwidth]{imgs/gf2.png}
-        \caption{$\tau=0.01$}
-      \end{subfigure}%
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=0.9\textwidth]{imgs/gf3.png}
-        \caption{$\tau=0.1$}
-      \end{subfigure}%
-      \caption{\label{fig:flow} Effect of flow time evolution on a random lattice in the symmetric phase. White represents positive values of $\phi$ while black represents negative.}
-\end{figure}
-
-Generally, we can choose any flow time equation that drives the field towards a classical minimum.
-Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields.
-In the non-linear sigma model, it has been shown that an appropriate manifestation of the flow time can resolve divergences \cite{makino2015a}.
-Therefore, it is possible to define a different flow equation for this model as well.
-Following \cite{bietenholz2018}, we can define the gradient flow in this model via the differential equation
-\begin{equation}
-    \label{eq:nsm_gradflow}
-    \partial_\tau \vec\rho (\tau,x) = \left( 1 - \vec\rho(\tau,x) \vec\rho(\tau,x)^T \right) \partial^2 \vec\rho(\tau,x),
-\end{equation}
-where $\partial^2$ is the Laplacian operator in Euclidean space\footnote{Explicitly, $\partial^2 = \frac{\partial^2}{\partial t^2} + \nabla^2$}.
-We solve this equation numerically using the boundary condition $\vec\rho(0,x) = \e(x)$.
-
-\chapter{Methods}
-\label{sec:methods}
-Our study of the gradient flow in the non-linear sigma model consists of a computational part and an analytical part.
-We begin by outlining a numerical Monte Carlo method to simulate the lattice in two and three dimensions.
-We verify our program with the well-studied $\phi^4$ scalar field theory.
-We then generalize our model to a vector field to simulate the non-linear sigma model.
-This simulation system provides data on vacuum states with which we study the gradient flow's effect on topology.
-
-We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories. 
-\section{Fields on the Lattice}
-To implement the lattice regularization technique, we must redefine the field action in terms of discrete positions, a process known as ``discretization''.
-We transition from $x$ as a continuous vector in $\mathbb{R}^2$ to $x_{i,j}$ where
-\begin{equation}
-    x_{i,j} = ia \hat{\mu}_t + j a \hat{\mu}_x.
-\end{equation}
-Here, $a$ is the lattice constant and $\hat{\mu}_0$ and $\hat{\mu}_1$ are unit vectors.
-This change effectively shifts the domain of the field from $\mathbb{R}^2$, which is uncountably infinite, to $\mathbb{Z}^2$, which is countably infinite.
-To achieve a finite domain, we impose periodic boundary conditions, such that
-\begin{equation}
-    \phi\left(x_{i+L,j}\right) = \phi\left(x_{i,j+L}\right) = \phi\left(x_{i,j}\right)
-\end{equation}
-where $L$ is the side length (in units of the lattice constant $a$) of the system. In this study we focus solely on square geometries and thus the side length $L$ is unambiguous.
-
-In the $\phi^4$ model, we can specify a discrete action using the Euclidean action from Eq. \ref{eq:phi4 euclidean action}. We begin by redefining the derivative operator as a difference:
-\begin{align}
-    \partial_\mu \phi = \frac{\phi\left(x + a \hat{\mu}\right) - \phi(x)}{a}
-\end{align}
-We can then define the kinetic term
-\begin{align}
-    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow& \frac{1}{2a^2}\left[ \left(\phi(x+a \hat{t}) -\phi(x)\right)^2 + \left(\phi(x+a \hat{x}) -\phi(x)\right)^2\right] \\
-    \begin{split} \rightarrow& \frac{1}{2a^2}\left[ \phi^2(x+a\hat{t}) + \phi^2(x+a\hat{x}) + 2\phi^2(x)\right.\\ &\qquad \left.- 2\phi(x+a\hat{t})\phi(x) - 2\phi(x+a\hat{x})\phi(x) \right]\hfilneg\end{split}
-\end{align}
-Since we will eventually sum over all sites $x$, the periodic boundary conditions imply that an overall shift in $x$ does not effect the final action.
-Therefore, we can combine the first two terms with the third term to produce
-\begin{equation}
-    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow \frac{1}{a^2}\left[ 2\phi^2(x) - \phi(x+a\hat{t})\phi(x) - \phi(x+a\hat{x})\phi(x) \right]
-\end{equation}
-Unlike the kinetic term, the mass and interaction terms remain unchanged under the discretization procedure. The only remaining change is a shift from an integral to a sum. This takes the form
-\begin{equation}
-    \int dtdx \rightarrow a^2 \sum_i
-\end{equation}
-such that the final discretized action becomes
-\begin{equation}
-    \label{eq:phi4 discretized action}
-    S_{\mathrm{lat}}[\phi] = \sum_i \left[-\phi(x_i + a\hat{t})\phi(x_i) - \phi(x_i + a\hat{x})\phi(x_i) + \left(2+\half m_0^2\right)\phi^2(x_i) + \frac{1}{4}\lambda \phi^4(x_i)\right]
-\end{equation}
-
-Likewise, we can discretize the non-linear sigma model. In this case, the derivative term becomes 
-\begin{equation}
-    \half \left(\partial_t\e\right)^2 + \half \left(\partial_x\e\right)^2 \rightarrow \frac{1}{a^2}\left[ 2 - \e(x+a\hat{t})\cdot\e(x) - \e(x+a\hat{x})\cdot\e(x) \right].
-\end{equation}
-Note that we have used the identity $\e\cdot\e = 1$. Inserting this into Eq.~\ref{eq:nlsm euclidean action} yields the discretized action
-\begin{equation}
-    \label{eq:nlsm discretized action}
-    S_\mathrm{lat}[\e] = \sum_i \left[ 2 - \e(x+a\hat{t})\cdot\e(x) - \e(x+a\hat{x})\cdot\e(x) \right].
-\end{equation}
-
-Finally, we redefine the gradient flow in on the lattice.
-Since the gradient flow is solved exactly in the $\phi^4$ model, we rely on a Fast Fourier Transform.
-\textit{TODO: explicit expression for this} In the NLSM, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes
-\begin{equation}
-    \label{eq:nsm_gradflow_disc}
-    \partial_\tau \e (\tau, x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T\right) \partial^2 \e(\tau,x),
-\end{equation}
- where the Laplacian operator $\partial^2$ is defined as
-\begin{equation*}
-    \partial^2 \e(\tau,x) = \e(\tau, x+a \hat{t}) + \e(\tau,x-a\hat t) + \e(\tau, x+a \hat{x}) + \e(\tau,x-a\hat x) - 2 d \e(t)_x.
-\end{equation*}
-%TODO: derive the laplacian operator
-
-\section{Monte Carlo Simulations}
-\label{sec:mc}
-We implement a Markov Chain Monte Carlo method following Schaich's thesis \cite{schaich2006}.
-This implementation utilizes a ``random walk,'' i.e.
-a set of random steps through phase space, to determine statistical values such as correlation functions across the lattice.
-By the definition of the Markov chain, the probability of adoption of each state, and therefore its inclusion in the Monte Carlo calculation, depends only on the current state and the proposed state.
-This probability is denoted as $P(\mu\rightarrow\nu)$ where $\mu$ and $\nu$ are the existing and proposed lattice configurations respectively.
-
-\subsection{Metropolis Algorithm}
-We primarily use the Metropolis algorithm for the calculation of new Markov chain configurations.
-We begin with a so-called ``hot start,'' where each field value at each lattice site is randomly selected.
-Then we propose a new value for a single lattice point, which is accepted with a probability
-\begin{equation}
-    P(\phi_a\rightarrow\phi_b) = \begin{cases} 
-        e^{-(S[\phi_b] - S_[\phi_a])} & S[\phi_b] < S_[\phi_a] \\
-        1 & \mathrm{otherwise} \\
-   \end{cases}
-\end{equation}
-where $\phi_a$ is the initial configuration and $\phi_b$ is the proposed configuration.
-This process is performed for each point on the lattice, making up a ``sweep.'' Repeating this sweep many times pushes the lattice toward the action minimum.
-
-\subsection{Wolff Cluster Algorithm}
-
-Though the Metropolis algorithm will slowly find the absolute minimum of the theory, the presence of local minima can greatly prolong the convergence.
-Both the $\phi^4$ model and the non-linear sigma model feature ``kinetic'' terms with gradients of $\phi$.
-Therefore, the presence of large similarly-valued regions in the lattice can lead to a local minimum.
-One method of removing these clusters involves identifying all clusters on the lattice and probabilistically flipping each, a technique known as the Swendsen-Wang algorithm\cite{swendsen1987}.
-
-A more efficient approach is the Wolff algorithm\cite{wolff1989}, which \textit{grows} one cluster probabilistically and flips it unconditionally.
-In the case of $\phi^4$ theory, this flipping takes the form of a simple sign change.
-In the non-linear sigma model we choose a random unit vector $\vec r$ and consider the projection of the field on this vector.
-When the cluster flips, each site is flipped along this direction.\citeneeded To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site, growing the cluster from a single, randomly selected ``seed''.
-Starting with the seed, the probability of adding each neighboring site is given by the source site $x$ and the proposed site $x'$.
-Wolff defines this probability for arbitrary sigma models as
-\begin{equation}
-    \label{eq:phi4 wolff padd}
-    P_{add}(\e(x),\e(x')) = \begin{cases} 
-        1 - e^{2\beta [\vec{r} \cdot \e(x)][\vec{r} \cdot \e(x')]} & \mathrm{sgn}[\vec{r}\cdot\e(x)]=\mathrm{sgn}[\vec{r}\cdot\e(x')]\\
-        0 & \mathrm{otherwise} \\
-   \end{cases}
-\end{equation}
-This expression is designed to preserve the detailed balance equations.
-We can demonstrate this quality, and motivate an equivalent expression for the $\phi^4$ model, by considering the probability $P(\phi\rightarrow f_C(\phi))$ of flipping some cluster $C$.
-Generally,
-\begin{equation}
-    P\big(\phi \rightarrow f_C\left(\phi\right)\big) \propto \prod_{\langle x,x'\rangle \in \partial C}  \Big[1 - P_{add}\big(\e(x),\e(x')\big)\Big]
-\end{equation}
-where $\partial C$ is the set of pairs of sites on the boundary of $C$.
-Since $P_{add}=0$ for unaligned sites, these pairs contribute nothing to the value.
-We can also find the probability $P(f_C(\phi) \rightarrow \phi)$ with the same expression:
-\begin{equation}
-P\big(f_C(\phi)\rightarrow \phi \big) \propto \prod_{\langle x,x'\rangle \in \partial C}  \Big[1 - P_{add}\big(\e(x),R\;\e(x')\big)\Big],
-\end{equation}
-where the matrix $R$ which is a reflection matrix along the vector $\vec{r}$.
-
-From the discretized action of the NLSM model (Eq.~\ref{eq:nlsm discretized action}) and the detailed balance equation (Eq.~\ref{eq:detailedbalance2}), we derive
-\begin{equation}
-    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\e(x),\e(x')\big)}{ 1 - P_{add}\big(\e(x),R\;\e(x')\big)}  = \mathrm{exp}\left\{\beta\sum_{\langle x,x'\rangle \in \partial C} \e(x)\cdot[R-1]\e(x')\right\}.
-\end{equation}
-Note that all the pairs within and outside the cluster cancel in the fraction on the left and the difference on the right. Using the definition of the reflection matrix
-\begin{equation}
-    R\;\e = \e - 2(\e\cdot\vec{r})\e,
-\end{equation}
-we can simplify the equation to be 
-\begin{equation}
-    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\e(x),\e(x')\big)}{ 1 - P_{add}\big(\e(x),R\;\e(x')\big)}  = \prod_{\langle x,x'\rangle \in \partial C}\mathrm{exp}\big\{2\beta[r\cdot\e(x)][r\cdot\e(x')]\big\}.
-\end{equation}
-By plugging in Eq.~\ref{eq:phi4 wolff padd}, it is clear to see this equation is satisfied.
-
-Using this same reasoning, we can deduce an expression for $P_{add}$ in the $\phi^4$ model.
-Since this model is one dimensional, the reflection matrix $R$ is equivalent to $-1$.
-Adapting the NLSM detailed balance equation for the $\phi$ field, we find
-\begin{equation}
-    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\phi(x),\phi(x')\big)}{ 1 - P_{add}\big(\phi(x),-\phi(x')\big)}  = \prod_{\langle x,x'\rangle \in \partial C}\mathrm{exp}\{-2 \phi(x)\phi(x')\}.
-\end{equation}
-This equation is satisfied by the ansatz
-\begin{equation}
-    \label{eq:nlsm wolff padd}
-    P_{add}(\phi(x),\phi(x')) = \begin{cases} 
-        1 - e^{-2\phi(x)\phi(x')]} & \mathrm{sgn}[\phi(x)]=\mathrm{sgn}[\phi(x')]\\
-        0 & \mathrm{otherwise} \\
-   \end{cases}.
-\end{equation}
-We will use this expression in the computational implementation of this algorithm.
-
-Fig.~\ref{fig:wolff} shows a real demonstration of this process.
-We can see a large cluster of negative field values becoming positive.
-Note that periodic boundary conditions apply, so the small island of black at the bottom is actually part of the larger cluster.
-Furthermore, this visualization demonstrates the probabilistic nature of the Wolff algorithm.
-Since states are added probabilistically, there are some small holes in the cluster.
-These will be removed by following Metropolis sweeps.
-\begin{figure}[h]
-  \centering
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/wolffa.png}
-        \caption{before cluster flip}
-      \end{subfigure}%
-      \hfill
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/wolffb.png}
-        \caption{after cluster flip}
-      \end{subfigure}
-      \hfill
-      \caption{\label{fig:wolff} An example of the Wolff cluster algorithm in the $\phi^4$ model.
-White represents positive values of $\phi$ while black represents negative.
-$\lambda=0.5$, $m_0^2=-0.9$}
-
-\end{figure}
-
-\subsection{Checkerboard algorithm}
-
-In order to parallelize the Metropolis algorithm, we use a checkerboard algorithm.
-Since the Lagrangian density at each site does not depend on any sites diagonal to it, the lattice can be split into ``white'' sites and ``black'' sites, like the tiles on a checkerboard.
-Each white site is independent of every other white site and likewise with black sites.
-Therefore, we can split the sites of each color into separate parallel processing nodes and independently run the Metropolis algorithm, ensuring that no site affects the Lagrangian density at any other site.
-We use this method to parallelize the code through the Message Passing Interface (MPI).
-
-\subsection{Thermalization}
-Since we begin with a random lattice, the first few configurations of our Monte Carlo simulation will be far from the lattice minimum.
-We plot the action as a function of metropolis sweeps in Fig.~\ref{fig:therm}.
-\begin{figure}[h]
-  \centering
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/therm24.png}
-        \caption{$L=24$}
-      \end{subfigure}%
-      \hfill
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/therm404.png}
-        \caption{$L=404$}
-      \end{subfigure}
-      \hfill
-      \caption{\label{fig:therm} Plots of the action as a function of Monte Carlo time, starting with a random NLSM lattice.}
-\end{figure}
-Based on this plot, we determine that 1000 sweeps will give sufficient time for the system to reach the classical action minimum. We use this value for the remainder of this study.
-\subsection{Autocorrelation times and thermalization}
-One important aspect is the correlation between different states of the Markov chain.
-Ideally, each sample of the lattice would be completely independent, but this is not the case.
-Since each configuration is based on previous configurations, there is a correlation between each pair of members in the Markov chain, decreasing exponentially based on the number of steps between.
-Explicitly, the autocorrelation scales as \[e^{t/\tau_{int}}\] where $t$ is the number of steps between configurations and $\tau_{int}$ is the autocorrelation time.\footnote{Though it is called a time, $\tau_{int}$ is in units of Markov Chain steps.} When performing simulations of the lattice, the number of sweeps between measurements should be much larger than $\tau_{int}$.
-
-    We use Wolff's automatic windowing procedure \cite{wolff2007} and the magnetic susceptibility $\chi_m$ to estimate the autocorrelation.
-Using Wolff's public MatLab code\footnote{\url{https://www.physik.hu-berlin.de/de/com/ALPHAsoft}}, we estimate the autocorrelation time for $L=24$ and $L=404$ lattices.
-This algorithm identifies the optimal window size with which to calculate the autocorrelation time.
-We perform this process on $L=24$ and $L=404$ lattices using a thermalization of $1000$ sweeps and $500$ total measurements.
-Note that this calculation included a Wolff cluster algorithm every $5$ sweeps.
-The result of this calculation is shown in Fig.~\ref{fig:tauint}.
-\begin{figure}[h]
-  \centering
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/tauint24.png}
-        \caption{$L=24$, $\tau_{int}=4.43$}
-      \end{subfigure}%
-      \hfill
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/tauint404.png}
-        \caption{$L=404$, $\tau_{int}=12.81$}
-      \end{subfigure}
-      \hfill
-      \caption{\label{fig:tauint} Plots of automatic windowing procedure used to calculate $\tau_{int}$ for the NLSM model.}
-\end{figure}
-
-Based on these two values for $\tau_{int}$, we decide to measure every $50$ sweeps for each simulation. This value will ensure that each measurement is effectively independent.
-
-\subsection{Runge-Kutta Algorithm}
-In order to calculate the gradient flow, we numerically solve the differential equation using a fourth-order Runga-Kutta approximation. This algorithm refines the Euler method 
-\begin{equation*}
-    \e(\tau+h)_x \approx \e(\tau)_x + h f(\e(\tau)_x).
-\end{equation*}
-where $f(\e)$ is defined for convenience as 
-\begin{equation}
-    f(\e)=\partial_\tau \e (\tau)_x  = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \partial^2 \e(\tau)_x,
-\end{equation}
-following from Eq.~\ref{eq:nsm_gradflow_disc}. To the fourth order, this approximation becomes 
-%
-\begin{align}
-    \label{eq:rungekutta}
-    k_1 &= h f\left(\e\left(\tau\right)_x\right) \\ 
-    k_2 &= h f\left(\e\left(\tau\right)_x + \frac{k_1}{2}\right) \\ 
-    k_3 &= h f\left(\e\left(\tau\right)_x + \frac{k_2}{2}\right) \\ 
-    k_4 &= h f\left(\e\left(\tau\right)_x + k_3\right) \\ 
-    \e(\tau+h)_x &= \e(\tau)_x + \frac{k_1}{6} + \frac{k_2}{3} + \frac{k_3}{3} + \frac{k_4}{6} + O(h^5).
-\end{align}
-This method is usually superior to Euler's method and the midpoint method \cite{vetterling1992}.
-
-To increase the efficiency of this algorithm, we implement the step-doubling algorithm to adaptively adjust $h$.
-If the error of a Runge-Kutta step is greater than the tolerance, the same step is repeated with half the step size.
-Alternatively, if the error is less than half of the tolerance, the step size is doubled for the next calculation.
-Finally, if the step size is greater than the distance to the next measurement, that distance is used as the step size, using the normal value afterwards.
-Otherwise, the algorithm proceeds with the consistent step size.
-
-\section{Topological charge with a $\theta$ term}
-In Sec.~\ref{sec:topological charge}, we discussed the introduction of a $\theta$ term into the action.
-This change made the theory ``topological'', pushing $\langle Q \rangle$ away from zero.
-In order to calculate $\langle Q \rangle$ as a function of $Q$, we consider the path integral
-\begin{align}
-    \langle Q \rangle_\theta &=\int \mathcal{D}\e\:Q[\e]e^{-S[\e]+i\theta Q[\e]} \\
-                             &=\int \mathcal{D}\e\:\left( Q[\e]e^{i\theta Q[\e]} \right) e^{-S[\e]} \\
-                             &=\langle Q e^{i \theta Q} \rangle_{\theta=0}.
-\end{align}
-Therefore, we can calculate $\langle Q \rangle_\theta$ for arbitrary $\theta$ using the same simulation framework as the $\theta=0$ case.
-
-We can also relate this function to the topological susceptibility. By expanding the exponent as a Taylor series around $\theta=0$, we find that  
-\begin{equation}
-    \langle Q \rangle_\theta = \langle Q \rangle_{\theta=0} + i \theta \langle Q^2 \rangle_{\theta=0} + O(\theta^2),
-\end{equation}
-such that 
-\begin{align}
-    -i \frac{\partial}{\partial \theta} \langle Q \rangle_\theta &= \langle Q^2 \rangle_{\theta=0}\\
-                                                              &= \chi_t L^2.
-\end{align}
-We expect to see the slope of this function at zero increase for large lattice sizes.
-
-\chapter{Results}
-\section{$\phi^4$ model results}
-
-We initially implemented the $\phi^4$ model using the Monte Carlo method described in Sec.~\ref{sec:methods} to verify the results of our system.
-According to previous studies (\cite{monahan2016}, \cite{schaich2006}), the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ for $\lambda = 0.5$.
-We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
-      \caption{\label{fig:phi4} Simulation of the phase transition in the $\phi^4$ model: Magnetization $\langle |\bar\phi|\rangle$, magnetic susceptibility $\chi_m$, Binder cumulant $U$ and bimodality $B$ for various values of the mass squared $m_0^2$.}
-\end{figure}
-
-\section{Non-linear sigma model results}
-After confirming the phase transition 
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=\textwidth]{imgs/internal_energy.png}
-      \caption{\label{fig:bergluscher}}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=\textwidth]{imgs/theta.png}
-      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$. Note the different scaling of the $y$-axis.}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=\textwidth]{imgs/bietenholz.png}
-      \caption{\label{fig:bietenholz} $\langle Q^2 \rangle=\chi_t L^2$ as a  }
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=0.6\textwidth]{imgs/hist.png}
-      \caption{\label{fig:hist} Histogram of topological charge values $Q$ for trivial NLSM. $L=404$, 10,000 measurements, measurements very 50 sweps, 1,000 sweep thermalization, $\tau=0$}
-\end{figure}
-
-%\chapter{Statistical Analysis}
-
-\chapter{Conclusion \& Outlook}
+Other future work may include explorations of higher order non-linear sigma models or other topological effective field theories.
+The latter of these models has applications in topological quantum computing and other technologies.
 
@@
 \appendix
-\chapter{C++ NLSM Monte Carlo Program}
-\lstset{language=C++,
-                basicstyle=\scriptsize\ttfamily,
-                keywordstyle=\color{blue}\ttfamily,
-                stringstyle=\color{red}\ttfamily,
-                commentstyle=\color{green}\ttfamily,
-                morecomment=[l][\color{magenta}]{\#},
-                breaklines=true,
-                postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}
-}
-
-\codeexample{sweep.h}
-\codeexample{sweep.cpp}
-\codeexample{lattice.h}
-\codeexample{lattice.cpp}
-\codeexample{phi.h}
-\codeexample{phi.cpp}
-\codeexample{observables.h}
-\codeexample{constants.h}
-
-\bibliographystyle{unsrt}
-\bibliography{library}
-
-\end{document}

### 2021-04-19 final-report.latex/intro.tex
split into chapter files

--- 2021-04-02 final-report.latex/main.tex
@@
-\documentclass[12pt]{report}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\setlength{\oddsidemargin}{0.5in}
-\setlength{\evensidemargin}{0.5in}
-\setlength{\textwidth}{6.0in}
-\setlength{\topmargin}{.75in}
-\setlength{\textheight}{8.6in}
-
-\newcommand{\e}{\vec e}
-\newcommand{\citeneeded}{}
-
-% Modified by S. Aubin for the class of 2019 Physics theses
-
-%
-% This is a LaTeX template for an undergraduate honors or
-%   senior thesis in physics at William & Mary. Created from
-%   several theses from previous students of D.S. Armstrong,
-%   pared down to bare bones.
-%
-%   you will need this present file (thesis_template.tex)
-%   as well as the file  figure1.eps  to make the whole thing
-%   (the latter is just an encapsulated postscript figure)
-%     For simplicity, we don't use bibtex here; experts can choose
-%   to use that more powerful way of dealing with citations.
-%
-%   To compile it, do:
-% > pdflatex thesis_template.tex
-% > pdflatex thesis_template.tex  (yes, you need to do it twice to
-%                                  fill in the list of tables, figures
-%                                  and references).
-
-\begin{document}
-\setlength{\baselineskip}{24pt}
-\begin{titlepage}
-\LARGE
-\begin{center}
-{\bf A Wonderful Piece of Scientific Research}\\[2.3cm]
-
-\normalsize
-A thesis submitted in partial fulfillment of the requirement \\
-for the degree of Bachelor of Science with Honors in \\
-Physics from the College of William and Mary in Virginia,\\[0.5cm]
-by\\[0.5cm]
-Stuart Thomas \\[2.5cm]
-\end{center}
-\normalsize
-\begin{flushright}
-%\hfill Accepted for Honors \\[.5cm]
-\hfill \hrulefill \\
-\hfill \hfill Advisor: Prof. Christopher J. Monahan\\[.6cm]
-\hfill \hrulefill \\Prof. Todd Averett \\[.6cm]
-\hfill \hrulefill \\Prof. Andreas Stathopoulos \\[.6cm]
-\end{flushright}
-\begin{center}
-Williamsburg, Virginia\\
-May 2021
-\end{center}
-\end{titlepage}
-
-\setlength{\topmargin}{0.0in}
-
-\pagenumbering{roman}
-\tableofcontents
-\chapter*{Acknowledgments}
-\addcontentsline{toc}{chapter}{Acknowledgments}
-\paragraph
-\indent I would like to thank lots of people, and this is where
-I will do it...
-
-\listoffigures
-\addcontentsline{toc}{chapter}{List of Figures}
-\listoftables
-\addcontentsline{toc}{chapter}{List of Tables}
-
-\begin{abstract}
-\setcounter{page}{5}
-\addcontentsline{toc}{chapter}{Abstract}
-\paragraph
-\indent \textit{Abstract}
-
-\end{abstract}
-
-\pagenumbering{arabic}
-\chapter{Introduction}
-
-Quantum field theory (QFT) is an extraordinarily successful framework which can be applied to a range of physical phenomena.
+Quantum field theory (QFT) is an extraordinarily successful framework which describes a range of physical phenomena.
 Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics.
@@
 
-In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time).
+In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions one dimension of space, one dimension of time).
 This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action.
@@
 %\section{Main Experimental Goal}
-
-\chapter{Theory}
-
-This thesis incorporates two main bodies of knowledge: quantum field theory and statistical simulation.
-Through the path integral formulation of quantum field theory, we are able to describe the physics of the former with the established mathematics of the latter.
-
-\section{Quantum Field Theory}
-
-In this section we outline a rough description of quantum field theory.
-A full introduction is beyond the scope of this paper, however we do assume knowledge of nonrelativistic quantum mechanics and classical field theory.
-
-The fundamental hypothesis of quantum field theory (QFT) describes particles as discrete packets of energy on a quantum field.
-But what is a quantum field?
-Like in classical mechanics, a field is a function of spacetime with some mathematical object assigned to each point in space and time.
-In the case of the electric field, this object is a three-dimensional vector, while the electric potential is a scalar field.
-Classical and quantum fields have Lagrangians which define how they evolve in time.
-What differentiates a quantum field from a classical field is superposition: where classical fields have a definite configuration, quantum fields exist in a superposition of all possible configurations.
-It is possible---though nontrivial and outside the scope of this description--- to motivate the appearance of discrete particles from this superposition.
-
-This general description allows QFT to easily incorporate special relativity.
-By ensuring that the Lagrangian of a theory is invariant under Lorentz transformations, we can ensure that the theory is physical.
-% is this enough?
-
-Though most introductions to QFT use the ``second quantization,'' we will use a alternative formulation known as the path integral formulation (for a similar introduction see Zee's textbook \cite{zee2010}).
-
-\subsection{Path Integral Formulation}
-\label{sec:pathintegral}
-
-As stated above, we can model a quantum field theory as a superposition of all possible classical fields.
-Like single-particle quantum mechanics, each configuration has a probability amplitude.
-To measure expectation values of observables, we simply take an average over all configurations weighted by this complex amplitude.
-We can formalize this notion using the fundamental formula\footnote{Since this study concerns vacua, we do not include a source term.}
-\begin{equation}
-    \label{eq:pathintegral}
-    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; e^{iS[\phi]}
-\end{equation}
-% not sure if the semicolons is good form
-where $\langle \hat O \rangle$ is the expectation value of an arbitrary operator $\hat O$; $Z$ is a normalization constant; $S$ is the action functional, defined from the field's Lagrangian; and $\int \mathcal{D}\phi$ represents the eponymous path integral.
-Though it is possible to define this integral rigorously, for our purposes we can equate it to a sum over all possible configurations.
-This form is the quantum analog of the classical principle of least action and reduces to such for large values of the action.
-For a more pedagogical explanation, see Richard Feynman's lectures on physics.\cite{feynman1963a}
-
-At first glance, Eq.~\ref{eq:pathintegral} is remarkably similar to the statistics of the canonical ensemble.
-Through this similarity, we will be able to use mathematical tools from statistical mechanics to study quantum field theories.
-However, the factor of $i$ in the exponent currently prohibits us from making this jump.
-To remedy this issue, we perform a ``Wick rotation'' which shifts spacetime into Euclidean coordinates.
-In normal spacetime, defined by the Minkowski metric, the Lorentz-invariant distance is given as
-\begin{equation}
-    s^2 = x^2_0 - x^2_1- x^2_3- x^2_3
-\end{equation}
-where $x_0=ct$ and $\vec{x} = (x_1, x_2, x_3)^T$. By redefining the time coordinate of a spacetime point $x$ to be $x_4=ix_0$, we find that the quantity
-\begin{equation}
-    s_E^2 = x^2_1+ x^2_3+ x^2_3 + x^2_4,
-\end{equation}
-is equivalently Lorentz invariant, which is representative of a four-dimensional Euclidean space. Furthermore, we find that
-\begin{align}
-    d^4x_E &= d^3\vec{x}dx_4 \\
-    &= i d^3\vec{x}dx_0 \\
-    &= i d^4x. \label{eq:wickdifferential}
-\end{align}
-
-We can use this transformation to redefine the Lagrangian $\mathcal{L}$ in Euclidean space as $\mathcal{L}_E$, replacing all $x_0$ with $ix_0$.
-Since a Lorentz-invariant Lagrangian must include only even powers and derivatives of $x$, the Euclidean Lagrangian remains real.
-Subsequently, we can define a Euclidean action based on the differential in Eq.~\ref{eq:wickdifferential}:
-\begin{align}
-    S_E &= \int d^4x_E \mathcal{L}_E \\
-    &= i \int d^4x \mathcal{L}_E,
-\end{align}
-allowing us to redefine the path integral as 
-\begin{equation}
-    \label{eq:pathintegraleuclidean}
-    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O (\phi)\; e^{-S_E[\phi]}.
-\end{equation}
-
-By replacing the Minkowski action $S$ with a Euclidean action $S_E$, we have transformed the amplitude $e^{iS}$ to a statistical Boltzmann factor $e^{-S_E}$.
-This new form will allow us to use statistical techniques to simulate quantum fields.
-
-\subsection{$\phi^4$ model}
-\label{sec:phi4}
-
-The simplest interacting field theory is known as the $\phi^4$ model. This theory describes a spin-0 boson and consists of a real scalar field given by the action
-\begin{equation}
-    \label{eq:phi4 action}
-    S[\phi] = \int d^D x \left[ \frac{1}{2}\partial^\mu \phi \partial_\mu\phi - \frac{1}{2} m_0^2 \phi^2 - \frac{\lambda}{4!}\phi^4\right].
-\end{equation}
-The first two terms describe a free relativistic particle of mass $m_0$ while the last term describes an interaction with strength $\lambda$.
-Note that we have generalized the action for $D$ dimensions.
-Per Einstein summation notation, there is an implicit sum over spacetime dimensions $\mu\in\{0,1,2,3\}$ indexing the derivative vectors\footnote{This canonical representation of the kinetic term $\frac{1}{2}\partial^\mu\phi\partial_\mu\phi$ is equivalent to $\frac{1}{2}\dot\phi^2-\frac{1}{2}\left(\nabla \phi\right)^2$.}
-\begin{align*}
-    \partial^\mu &= \left( \frac{\partial}{\partial t}, \frac{\partial}{\partial x},\frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) \\
-    \partial_\mu &= \left( \frac{\partial}{\partial t}, -\frac{\partial}{\partial x}, -\frac{\partial}{\partial y}, -\frac{\partial}{\partial z} \right).
-\end{align*}
-
-This field features spontaneous symmetry breaking at a critical value of $m_0^2$.
-In practice this property causes the field to spontaneously align, similarly to spins aligning in a ferromagnet.
-The name ``symmetry breaking'' refers to the transformation $\phi\rightarrow-\phi$, which changes the values of observables in the aligned regime but not the disordered regime.
-
-\subsection{Ultraviolet Divergences}
-
-In Section~\rec{sec:pathintegral},  we defined a fundamental equation of quantum fields using an ``path integral'' which encompasses an uncountably infinite configuration space.
-Likewise in perturbation theory, infinite integrals over momenta are commonplace.
-Unfortunately, these integrals often diverge, even when calculating physical values.
-
-The accepted remedy is unintuitive.
-Essentially, we adopt infinite values for the parameters of the Lagragian ($m_0^2$ and $\lambda$ in $\phi^4$ theory).
-In practice, this technique is involved and consists of two steps: regularization and renormalization.
-
-Regularization is a process which introduces a new parameter into calculations. One example is a momentum cutoff. This technique transforms infinite momentum integral integrals as 
-\begin{equation*}
-    \int_0^\infty dk \rightarrow \int_0^\Lambda dk,
-\end{equation*}
-introducing $\Lambda$ as a regularization parameter.
-
-\subsection{Fields on the lattice}
-
-\section{Markov Chain Monte Carlo}
-
-To accomplish a statistical analysis of quantum fields, we use a Monte Carlo simulation.
-This method amounts to producing a large number of configurations and calculating statistics on the sample.
-A brute-force calculation over all possible configurations---as Eq.~\ref{eq:pathintegraleuclidean} suggests--- is clearly infinite and computationally infeasible.
-However, the exponential nature of the Boltzmann factor dictates that only configurations near the action minimum contribute to observable statistics.
-Therefore, by selecting a sample of configurations near this minimum, we are able to extract meaningful results with a finite computation.
-
-\subsection{The Markov Chain}
-The Markov chain is an effective method to identity these low-action states.
-Essentially, we begin with a random configuration and then make small adjustments, gradually lowering the action.
-Starting with a field configuration $\phi_a$, we propose a new field $\phi_b$.
-The probability of accepting this change, thereby adding the configuration to the chain, is given by the function
-\[P(\phi_a \rightarrow \phi_b).\]
-
-There are four requirements that this function must obey to produce a Boltzmann distribution of samples:
-\begin{enumerate}
-    \item $P(\phi_a \rightarrow \phi_b)$ must depend only on the configurations $\phi_a$ and $\phi_b$.
-    \item The probability must be properly normalized, i.e. $\sum_{\phi} P(\phi_a \rightarrow \phi) = 1$.
-    \item Every configuration must be reachable in a finite number of steps. In other words, the chain must be ergodic.
-    \item In order to reach equilibrium, the chain must be reversible.
-In other words, the probability of a $\phi_a\rightarrow\phi_b$ transition must be equal to the probability of a $\phi_b\rightarrow\phi_a$ transition.
-Mathematically, this condition takes the form of \textit{detailed balance equations}:
-\begin{equation}
-    \label{eq:detailedbalance}
-    P(\phi_a)\,P(\phi_a\rightarrow\phi_b) = P(\phi_b)\,P(\phi_b\rightarrow\phi_a),
-\end{equation}
-where $P(\phi)$ is the probability of a system existing in state $\phi$.
-% this needs citation
-\end{enumerate}
-
-This final condition will allow us to explicitly define the transition probability using the action. From the Boltzmann distribution, we know
-\begin{equation}
-    P(\phi) = \frac{1}{Z} e^{-S_E[\phi]}.
-\end{equation}
-Therefore, be rearranging Eq.~\ref{eq:detailedbalance}, we find
-\begin{equation}
-    \frac{P(\phi_a\rightarrow\phi_b)}{P(\phi_b\rightarrow\phi_a)} = e^{-(S_E[\phi_b] - S_E[\phi_a])}.
-\end{equation}
-
-This formula will provide the explicit probabilities of the Metropolis and Wolff algorithms.
-
-\section{Gradient Flow}
-\label{sec:gradflow}
-To remove ultraviolet fluctuations, we introduce a momentum cutoff which reduces ultraviolet divergences.
-In a gauge theory, such as QCD, a hard cutoff violates gauge invariance, so more nuanced techniques are required.
-One such technique is ``smearing,'' a local averaging designed to remove diverging fluctuations \cite{solbrig2007}.
-In this study, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.
-
-Specifically, the gradient flow pushes field configuration toward classical minima of the action.
-Additionally, renormalized correlation functions remain renormalized at nonzero flow time.\cite{luscher2013} In 2D $\phi^4$ scalar field theory, the gradient flow is defined by the differential equation
-\begin{equation}
-    \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
-\end{equation}
-where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime\footnote{Explicitly, $\partial^2 = \frac{\partial^2}{\partial t^2} + \nabla^2$} and $\tau$ is the flow time.
-Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
-In the $\phi^4$ theory, we can solve this equation exactly to find \cite{monahan2016}
-\begin{equation}
-    \rho(\tau, x) = \frac{1}{4 \pi \tau} \int d^2 y e^{-(x-y)^2/4\tau} \phi(y).
-\end{equation}
-This function forms a Gaussian, smoothly dampening high-momentum modes and removing ultraviolet divergences from evolved correlation functions.\cite{makino2015a}
-
-Generally, we can choose any flow time equation that drives the field towards a classical minimum.
-Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields.
-In the non-linear sigma model, it has been shown that an appropriate manifestation of the flow time can resolve divergences \cite{makino2015a}.
-Therefore, it is possible to define a different flow equation for this model as well.
-
-\section{Non-linear sigma model}
-
-The non-linear sigma model is a prototypical theory for many physical phenomena, including applications in string theory.
-As a simple nonperturbative model, it provides an ideal starting point for lattice QCD studies.
-Specifically, the non-linear sigma model exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
-Furthermore, it has an exact application to Heisenberg ferromagnets in condensed matter.
-The theory is defined by the Euclidean action
-\begin{equation}
-    S_E = \frac{1}{2g^2} \int d^Dx \partial^\mu \vec\phi \cdot \partial_\mu \vec\phi
-\end{equation}
-subject to the constraint that $\vec\phi\cdot\vec\phi = 1$. In this report, we primarily consider $O(3)$, where $\vec\phi$ is a three dimensional vector.
-
-Following \cite{bietenholz2018}, we can define the gradient flow in this model via the differential equation
-\begin{equation}
-    \label{eq:nsm_gradflow}
-    \partial_\tau \vec\rho (\tau,x) = \left( 1 - \vec\rho(\tau,x) \vec\rho(\tau,x)^T \right) \nabla^2 \vec\rho(\tau,x),
-\end{equation}
-where $\nabla^2$ is the Laplacian operator in Euclidean space. We solve this equation numerically using the boundary condition $\vec\rho(0,x) = \phi(x)$.
-
-In order to use this theory on the lattice, we discretize the action.
-We replace the derivatives with a simple difference and impose periodic boundary conditions on the system.
-For a more in-depth discussion, see Section~\ref{sec:gradflow}.
-
-\section{Topological Charge}
-
-The $O(3)$ non-linear sigma model model features topological properties originating from two properties: 
-\begin{enumerate}
-    \item At $x\rightarrow\infty$, the field must become uniform since the Lagrangian must vanish.
-This allows us to model $x\rightarrow\infty$ as a single point on the field, forming a Riemann sphere in three dimensions.
-    \item The elements of the $O(3)$ non-linear sigma model are three-dimensional unit vectors, thereby existing on a three dimensional unit sphere. 
-\end{enumerate}
-
-With these two properties, we can view the field as a continuous mapping between two 3D spheres, denoted as $S^2$, and associate an integer number of wrappings to each mapping from $S^2$ to $S^2$.
-We can envision a tangible metaphor for this wrapping with a balloon and a baseball: by simply inserting the baseball into the balloon, we have established a mapping from every point on the balloon to every point on the baseball.
-We can create an equally valid map by twisting the balloon's mouth and wrapping the baseball again.
-In a purely mathematical world, we perform this process an infinite number of times, thereby associating every possible mapping with an integer.
-The group of integers is known as the \textit{homotopy group} of the non-linear sigma model.
-We associate every field configuration with an element of this group, known as the \textit{topological charge}, which we denote as $Q$.
-
-\chapter{Methods}
-\label{sec:methods}
-Our study of the gradient flow in the non-linear sigma model consists of a computational part and an analytical part.
-We begin by outlining a numerical Monte Carlo method to simulate the lattice in two and three dimensions.
-We verify our program with the well-studied $\phi^4$ scalar field theory.
-We then generalize our model to a vector field to simulate the non-linear sigma model.
-This simulation system provides data on vacuum states with which we study the gradient flow's effect on topology.
-
-We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories.
-
-\section{Monte Carlo Simulations}
-\label{sec:mc}
-
-We implement a Markov Chain Monte Carlo method following Schaich's thesis \cite{schaich2006}.
-This implementation utilizes a ``random walk,'' i.e.
-a set of random steps through phase space, to determine statistical values such as correlation functions across the lattice.
-By the definition of the Markov chain, the probability of adoption of each state, and therefore its inclusion in the Monte Carlo calculation, depends only on the current state and the proposed state.
-This probability is denoted as $P(\mu\rightarrow\nu)$ where $\mu$ and $\nu$ are the existing and proposed lattice configurations respectively.
-We use a combination of the Metropolis and Wolff algorithms to determine this value.
-
-\subsection{Metropolis Algorithm}
-We primarily use the Metropolis algorithm for the calculation of new Markov chain configurations.
-We begin with a so-called ``hot start,'' where each field value at each lattice site is randomly selected.
-Then we propose a new value for a single lattice point, which is accepted with a probability
-\begin{equation}
-    P(\phi_a\rightarrow\phi_b) = \begin{cases} 
-        e^{-(S_\nu - S_\mu)} & S_\nu > S_\mu \\
-        1 & \mathrm{otherwise} \\
-   \end{cases}
-\end{equation}
-where $S_\nu$ and $S_\mu$ are the lattice actions of configuration before and after the change.
-This process is performed for each point on the lattice, making up a ``sweep.'' Repeating this sweep many times pushes the lattice toward the action minimum.
-
-\subsection{Wolff Cluster Algorithm}
-
-Though the Metropolis algorithm will slowly find the absolute minimum of the theory, the presence of local minima can greatly prolong the convergence.
-Both the $\phi^4$ model and the non-linear sigma model feature ``kinetic'' terms with gradients of $\phi$.
-Therefore, the presence of large similarly-valued regions in the lattice can lead to a local minimum.
-The Wolff algorithm helps reduce the presence of these clusters through two steps: identifying a cluster and flipping it along some arbitrary vector.
-In the case of $\phi^4$ theory, this flipping takes the form of a simple sign change.
-In the non-linear sigma model we choose a random unit vector $\vec r$ and consider the projection of the field on this vector.
-When the cluster flips, each site is flipped along this direction.\citeneeded To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site.
-In the non-linear sigma model, this value is derived as
-\begin{equation}
-    P_{add} = 1-e^{- \beta \vec e_a\vec e_b}.
-\end{equation}
- % TODO: derive this
-
-\subsection{Checkerboard algorithm}
-
-In order to parallelize the Metropolis algorithm, we use a checkerboard algorithm.
-Since the Lagrangian density at each site does not depend on any sites diagonal to it, the lattice can be split into ``white'' sites and ``black'' sites, like the tiles on a checkerboard.
-Each white site is independent of every other white site and likewise with black sites.
-Therefore, we can split the sites of each color into separate parallel processing nodes and independently run the Metropolis algorithm, ensuring that no site affects the Lagrangian density at any other site.
-We use this method to parallelize the code through the Message Passing Interface (MPI).
-
-\subsection{Autocorrelation times and thermalization}
-One important aspect is the correlation between different states of the Markov chain.
-Ideally, each sample of the lattice would be completely independent, but this is not the case.
-Therefore, we must sweep the lattice sufficiently to produce an effective Monte Carlo result.
-We use Wolff's automatic windowing procedure \cite{wolff2007} and the magnetic susceptibility $\chi_m$ to estimate the autocorrelation and determine an appropriate number of sweep between measurements for each theory.
-Furthermore, the initial values of our Markov chain will be highly correlated with the initial random start.
-Therefore, we also wait a set number of sweeps before taking any measurements so that the lattice can approach a minimum of the action.
-
-\section{Gradient Flow}
-\label{sec:gradflow_disc}
-On the lattice, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes 
-\begin{equation}
-    \label{eq:nsm_gradflow_disc}
-    \partial_\tau \e (\tau)_x = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \nabla^2 \e(\tau)_x,
-\end{equation}
- where the Laplace operator $\nabla^2$ is defined as
-\begin{equation*}
-    \nabla^2 \e(t)_x = \sum^d_{\mu=1} \left( \e(t)_{x+\hat \mu} - \e(t)_{x-\hat \mu}\right) - 2 d \e(t)_x.
-\end{equation*}
-
-\subsection{Runge-Kutta Algorithm}
-In order to calculate the gradient flow, we numerically solve the differential equation using a fourth-order Runga-Kutta approximation. This algorithm refines the Euler method 
-\begin{equation*}
-    \e(\tau+h)_x \approx \e(\tau)_x + h f(\e(\tau)_x).
-\end{equation*}
-where $f(\e)$ is defined for convenience as 
-\begin{equation}
-    f(\e)=\partial_\tau \e (\tau)_x  = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \nabla^2 \e(\tau)_x,
-\end{equation}
-following from Eq.~\ref{eq:nsm_gradflow_disc}. To the fourth order, this approximation becomes 
-%
-\begin{align}
-    \label{eq:rungekutta}
-    k_1 &= h f\left(\e\left(\tau\right)_x\right) \\ 
-    k_2 &= h f\left(\e\left(\tau\right)_x + \frac{k_1}{2}\right) \\ 
-    k_3 &= h f\left(\e\left(\tau\right)_x + \frac{k_2}{2}\right) \\ 
-    k_4 &= h f\left(\e\left(\tau\right)_x + k_3\right) \\ 
-    \e(\tau+h)_x &= \e(\tau)_x + \frac{k_1}{6} + \frac{k_2}{3} + \frac{k_3}{3} + \frac{k_4}{6} + O(h^5).
-\end{align}
-This method is usually superior to Euler's method and the midpoint method \cite{vetterling1992}.
-
-To increase the efficiency of this algorithm, we implement the step-doubling algorithm to adaptively adjust $h$.
-If the error of a Runge-Kutta step is greater than the tolerance, the same step is repeated with half the step size.
-Alternatively, if the error is less than half of the tolerance, the step size is doubled for the next calculation.
-Finally, if the step size is greater than the distance to the next measurement, that distance is used as the step size, using the normal value afterwards.
-Otherwise, the algorithm proceeds with the consistent step size.
-
-\chapter{Results}
-\section{$\phi^4$ model}
-
-We initially implemented the $\phi^4$ model using the Monte Carlo method described in Sec.~\ref{sec:methods} to verify the results of our system.
-According to previous studies (\cite{monahan2016}, \cite{schaich2006}), the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ for $\lambda = 0.5$.
-We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
-
-\begin{figure}[h]
-  \centering
-      \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=\textwidth]{imgs/0.png}
-        \caption{$\tau=0$}
-      \end{subfigure}
-      \hfill
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=\textwidth]{imgs/0_5.png}
-        \caption{$\tau=0.5$}
-      \end{subfigure}
-      \hfill
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=\textwidth]{imgs/2.png}
-        \caption{$\tau=2$}
-      \end{subfigure}
-      \hfill
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-            \includegraphics[width=\textwidth]{imgs/10.png}
-            \caption{$\tau=10$}
-      \end{subfigure}
-      \hfill
-      \caption{\label{fig:flow} Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
-
-\end{figure}
-
-\chapter{Statistical Analysis}
-
-\chapter{Conclusion \& Outlook}
-
-% MPI is unnecessary in this case
-
-\appendix
-\chapter{My Great Computer Program}
-
-\section{Code sample}
-\bibliographystyle{unsrt}
-\bibliography{library}
-
-\end{document}

### 2021-04-19 final-report.latex/main.tex
split into chapter files

--- 2021-04-09 final-report.latex/main.tex
@@
 \usepackage{amsmath}
+\usepackage{tabularx}
 \usepackage{amsfonts}
@@
 \chapter{Introduction}
-
-Quantum field theory (QFT) is an extraordinarily successful framework which can be applied to a range of physical phenomena.
-Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics.
-In condensed matter physics, effective field theories model emergent effects such as phonons and quasiparticles.
-Compared to experiment, QFT is remarkably accurate, famously matching the experimental value for the electron $g$-factor to eight significant figures.\cite{weisskopf1981}
-
-However, this power comes at a cost: the study of quantum fields is rife with infinities.
-A na\"ive treatment of quantum field theory produces divergent values for physical quantities, a clearly impossible result.
-Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the so-called \textit{renormalization group}.
-However, the technique fails with perturbatively nonrenormalizable theories.
-
-One such example is the \textit{non-linear sigma model}, a prototypical theory in both condensed matter and particle physics.
-In solid-state systems, this model describes Heisenberg ferromagnets and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), exhibiting characteristic features such as a mass gap and asymptotic freedom.
-\citeneeded
-
-In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time).
-This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action.
-
-Since the non-linear sigma model cannot be renormalized perturbatively, we cannot study these topological effects with normal perturbative techniques.
-An alternative solution is placing the field on a discretized lattice, a technique originally used for quantum chromodynamics.
-In this scenario, field configurations become computationally calculable.
-This process introduces an nonphysical length scale $a$, the lattice spacing.
-Therefore, we expect any physical result to converge in the continuum limit, i.e.
-when $a\rightarrow 0$.
-However, this is not always the case as observables mix on the lattice, leading to divergences.
-As an example, states of definite angular momentum mix when discretized, a clear violation of the angular momentum commutation relations.
-
-The gradient flow is a technique designed to remove these divergences.
-By dampening high-momentum fluctuations, the gradient flow reduces power-divergent mixing and make observables finite on the lattice.\cite{monahan2016} In quantum chromodynamics, previous studies have verified the ability of the gradient flow to make observables finite\citeneeded.
-Due to its success in QCD, there has been interest in using the gradient flow to finitize the topological susceptibility in the 1+1 $O(3)$ non-linear sigma model.\cite{bietenholz2018}.
-
-%goals here
-
-\section{Method Overview}
-
-To numerically study the topological qualities of the non-linear sigma model, we first implement a Markov Chain Monte Carlo simulation.
-We initially construct a proof-of-concept Python program that models the simpler $\phi^4$ model (see Sec.~\ref{sec:phi4}).
-After comparing with existing literature, we transition to a C++ simulation for efficiency, implementing the non-linear sigma model in larger lattices.
-Since the gradient flow has no exact solution in the non-linear sigma model, we implement a numerical solution using a fourth-order Runge-Kutta approximation.
-By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.
-
-\section{Conventions}
-\begin{itemize}
-    \item Throughout this paper, we use natural units, i.e. $\hbar = 1$ and $c=1$.
-    \item We use Einstein summation notation, an implicit sum over repeated spacetime indices. For example, if $x^\mu$ is a spacetime four-vector and $x_\mu$ is its covariant form, the term
-        \begin{align*}
-            x^\mu x_\mu &= \sum^4_{\mu=0} x^\mu x_\mu \\
-            &= x_0^2-x_1^2-x_2^2-x_3^2.
-        \end{align*}
-\end{itemize}
-
-%\section{Main Experimental Goal}
-
+\input{intro.tex}
 \chapter{Theory}
-
-This thesis incorporates two main bodies of knowledge: quantum field theory and statistical simulation.
-Through the path integral formulation of quantum field theory, we are able to describe the physics of the former with the established mathematics of the latter.
-
-\section{Quantum Field Theory}
-
-In this section we outline a rough description of quantum field theory.
-A full introduction is beyond the scope of this paper, however we do assume knowledge of nonrelativistic quantum mechanics and classical field theory.
-
-The fundamental hypothesis of quantum field theory (QFT) describes particles as discrete packets of energy on a quantum field.
-But what is a quantum field?
-Like in classical mechanics, a field is a function of spacetime with some mathematical object assigned to each point in space and time.
-In the case of the electric field, this object is a three-dimensional vector, while the electric potential is a scalar field.
-Classical and quantum fields have Lagrangians which define how they evolve in time.
-What differentiates a quantum field from a classical field is superposition: where classical fields have a definite configuration, quantum fields exist in a superposition of all possible configurations.
-It is possible---though nontrivial and outside the scope of this description--- to motivate the appearance of discrete particles from this superposition.
-
-This general description allows QFT to easily incorporate special relativity.
-By ensuring that the Lagrangian of a theory is invariant under Lorentz transformations, we can ensure that the theory is physical.
-% is this enough?
-
-Though most introductions to QFT use the ``second quantization,'' we will use a alternative formulation known as the path integral formulation (for a similar introduction see Zee's textbook \cite{zee2010}).
-
-\subsection{Path Integral Formulation}
-\label{sec:pathintegral}
-
-As stated above, we can model a quantum field theory as a superposition of all possible classical fields.
-Like single-particle quantum mechanics, each configuration has a probability amplitude.
-To measure expectation values of observables, we simply take an average over all configurations weighted by this complex amplitude.
-We can formalize this notion using the fundamental formula\footnote{Since this study concerns vacua, we do not include a source term.}
-\begin{equation}
-    \label{eq:pathintegral}
-    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; e^{iS[\phi]}
-\end{equation}
-% not sure if the semicolons is good form
-where $\langle \hat O \rangle$ is the expectation value of an arbitrary operator $\hat O$; $Z$ is a normalization constant; $S$ is the action functional, defined from the field's Lagrangian; and $\int \mathcal{D}\phi$ represents the eponymous path integral.
-Though it is possible to define this integral rigorously, for our purposes we can equate it to a sum over all possible configurations.
-This form is the quantum analog of the classical principle of least action and reduces to such for large values of the action.
-For a more pedagogical explanation, see Richard Feynman's lectures on physics.\cite{feynman1963a}
-
-At first glance, Eq.~\ref{eq:pathintegral} is remarkably similar to the statistics of the canonical ensemble.
-Through this similarity, we will be able to use mathematical tools from statistical mechanics to study quantum field theories.
-However, the factor of $i$ in the exponent currently prohibits us from making this jump.
-To remedy this issue, we perform a ``Wick rotation'' which shifts spacetime into Euclidean coordinates.
-In normal spacetime, defined by the Minkowski metric, the Lorentz-invariant distance is given as
-\begin{equation}
-    s^2 = x^2_0 - x^2_1- x^2_3- x^2_3
-\end{equation}
-where $x_0=ct$ and $\vec{x} = (x_1, x_2, x_3)^T$. By redefining the time coordinate of a spacetime point $x$ to be $x_4=ix_0$, we find that the quantity
-\begin{equation}
-    s_E^2 = x^2_1+ x^2_3+ x^2_3 + x^2_4,
-\end{equation}
-is equivalently Lorentz invariant, which is representative of a four-dimensional Euclidean space. Furthermore, we find that
-\begin{align}
-    d^4x_E &= d^3\vec{x}dx_4 \\
-    &= i d^3\vec{x}dx_0 \\
-    &= i d^4x. \label{eq:wickdifferential}
-\end{align}
-
-We can use this transformation to redefine the Lagrangian $\mathcal{L}$ in Euclidean space as $\mathcal{L}_E$, replacing all $x_0$ with $ix_0$.
-Since a Lorentz-invariant Lagrangian must include only even powers and derivatives of $x$, the Euclidean Lagrangian remains real.
-Subsequently, we can define a Euclidean action based on the differential in Eq.~\ref{eq:wickdifferential}:
-\begin{align}
-    S_E &= \int d^4x_E \mathcal{L}_E \\
-    &= i \int d^4x \mathcal{L}_E,
-\end{align}
-allowing us to redefine the path integral as 
-\begin{equation}
-    \label{eq:pathintegraleuclidean}
-    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O (\phi)\; e^{-S_E[\phi]}.
-\end{equation}
-
-By replacing the Minkowski action $S$ with a Euclidean action $S_E$, we have transformed the amplitude $e^{iS}$ to a statistical Boltzmann factor $e^{-S_E}$.
-This new form will allow us to use statistical techniques to simulate quantum fields.
-
-\subsection{$\phi^4$ model}
-\label{sec:phi4}
-
-The simplest interacting field theory is known as the $\phi^4$ model. This theory describes a spin-0 boson and consists of a real scalar field given by the action
-\begin{equation}
-    \label{eq:phi4 action}
-    S[\phi] = \int d^D x \left[ \frac{1}{2}\partial^\mu \phi \partial_\mu\phi - \frac{1}{2} m_0^2 \phi^2 - \frac{\lambda}{4}\phi^4\right].
-\end{equation}
-The first two terms describe a free relativistic particle of mass $m_0$ while the last term describes an interaction with strength $\lambda$.
-Note that we have generalized the action for $D$ dimensions.
-Per Einstein summation notation, there is an implicit sum over spacetime dimensions $\mu\in\{0,1,2,3\}$ indexing the derivative vectors\footnote{This canonical representation of the kinetic term $\frac{1}{2}\partial^\mu\phi\partial_\mu\phi$ is equivalent to $\frac{1}{2}\dot\phi^2-\frac{1}{2}\left(\nabla \phi\right)^2$.}
-\begin{align*}
-    \partial^\mu &= \left( \frac{\partial}{\partial t}, \frac{\partial}{\partial x},\frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) \\
-    \partial_\mu &= \left( \frac{\partial}{\partial t}, -\frac{\partial}{\partial x}, -\frac{\partial}{\partial y}, -\frac{\partial}{\partial z} \right).
-\end{align*}
-Following Sec.~\ref{sec:pathintegral}, we calculate the Euclidean action in 1+1 dimensions as 
-\begin{equation}
-    \label{eq:phi4 euclidean action}
-    S_E[\phi] = \int d^2 x_E \left[\frac{1}{2}\left(\partial_t \phi\right) + \frac{1}{2} \left(\partial_x \phi \right) + \frac{1}{2} m_0^2 \phi^2 + \frac{\lambda}{4}\phi^4\right].
-\end{equation}
-where $\partial_t$ and $\partial_x$ are the two derivatives in 1+1 Euclidean spacetime.
-
-This field features spontaneous symmetry breaking at a critical value of $m_0^2$.
-In practice this property causes the field to spontaneously align, similarly to spins aligning in a ferromagnet.
-The name ``symmetry breaking'' refers to the transformation $\phi\rightarrow-\phi$, which changes the values of observables in the aligned regime but not the disordered regime.
-These two phases are known as the ``broken'' and ``symmetric'' phases and their transition is well understood.
-
-\subsection{Non-linear sigma model}
-
-The non-linear sigma model (NLSM) is a prototypical theory for many physical phenomena, including applications in string theory.
-As a simple nonperturbative model, it provides an ideal starting point for lattice QCD studies.
-Specifically, the NLSM exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
-Furthermore, it has an exact application to Heisenberg ferromagnets in condensed matter.
-
-Unlike the $\phi^4$ model, which consists of a real value at each point in spacetime, the $O(3)$ NLSM consists of a 3D unit vector at each point.
-For this reason, every transformation of the field must be ``norm preserving''.
-Per its name, the non-linear sigma model features a global symmetry under the 3D orthogonal group $O(3)$.
-In other words, the theory is invariant under rotating all vectors.
-To differentiate it from the $\phi^4$ model, we denote the NLSM field as $\e(x)$.
-
-The theory is defined by the Euclidean action 
-\begin{equation}
-    \label{eq:nlsm euclidean action}
-    S_E = \frac{\beta}{2} \int d^Dx \partial^\mu \e \cdot \partial_\mu \e
-\end{equation}
-subject to the constraint that $\e\cdot\e = 1$. Here, $\beta$ is the inverse coupling.
-
-\section{Markov Chain Monte Carlo}
-
-To accomplish a statistical analysis of quantum fields, we use a Monte Carlo simulation.
-This method amounts to producing a large number of configurations and calculating statistics on the sample.
-A brute-force calculation over all possible configurations---as Eq.~\ref{eq:pathintegraleuclidean} suggests--- is clearly infinite and computationally infeasible.
-However, the exponential nature of the Boltzmann factor dictates that only configurations near the action minimum contribute to observable statistics.
-Therefore, by selecting a sample of configurations near this minimum, we are able to extract meaningful results with a finite computation.
-
-\subsection{The Markov Chain}
-The Markov chain is an effective method to identity these low-action states.
-Essentially, we begin with a random configuration and then make small adjustments, gradually lowering the action.
-Starting with a field configuration $\phi_a$, we propose a new field $\phi_b$.
-The probability of accepting this change, thereby adding the configuration to the chain, is given by the function
-\[P(\phi_a \rightarrow \phi_b).\]
-
-There are four requirements that this function must obey to produce a Boltzmann distribution of samples:
-\begin{enumerate}
-    \item $P(\phi_a \rightarrow \phi_b)$ must depend only on the configurations $\phi_a$ and $\phi_b$.
-    \item The probability must be properly normalized, i.e. $\sum_{\phi} P(\phi_a \rightarrow \phi) = 1$.
-    \item Every configuration must be reachable in a finite number of steps. In other words, the chain must be ergodic.
-    \item In order to reach equilibrium, the chain must be reversible.
-In other words, the probability of a $\phi_a\rightarrow\phi_b$ transition must be equal to the probability of a $\phi_b\rightarrow\phi_a$ transition.
-Mathematically, this condition takes the form of \textit{detailed balance equations}:
-\begin{equation}
-    \label{eq:detailedbalance}
-    P(\phi_a)\,P(\phi_a\rightarrow\phi_b) = P(\phi_b)\,P(\phi_b\rightarrow\phi_a),
-\end{equation}
-where $P(\phi)$ is the probability of a system existing in state $\phi$.
-% this needs citation
-\end{enumerate}
-
-This final condition will allow us to explicitly define the transition probability using the action. From the Boltzmann distribution, we know
-\begin{equation}
-    P(\phi) = \frac{1}{Z} e^{-S_E[\phi]}.
-\end{equation}
-Therefore, by rearranging Eq.~\ref{eq:detailedbalance}, we find
-\begin{equation}
-    \label{eq:detailedbalance2}
-    \frac{P(\phi_a\rightarrow\phi_b)}{P(\phi_b\rightarrow\phi_a)} = e^{S_E[\phi_a] - S_E[\phi_b]}.
-\end{equation}
-
-This formula will provide the explicit probabilities of the Metropolis and Wolff algorithms.
-
-\section{Observables}
-
-To extract physics from Monte Carlo simulations, we define a set of ``observables''.
-These quantities manifest as expectation values of operators, calculated using the Euclidean path integral formula (Eq.~\ref{eq:pathintegraleuclidean}).
-We can classify these observables into two categories: primary and secondary observables.
-Primary observables are calculated as expectation values of global operators while secondary observables are derived from these quantities.
-
-\subsection{Primary Observables}
-Each primary observable is defined on each configuration independently, meaning they do not encode ensemble statistics of the Markov Chain.\footnote{One slight exception is the magnetic susceptibility, which can be defined both ways.} In the $\phi^4$ model, we can develop an intuition around these quantities by visualizing the symmetric and broken phases.
-Fig~\ref{fig:primary observables} and Tab.~\ref{tab:primary observables} show examples of these quantities in three different configurations: one in the broken phase, one in the symmetric phase, and one at the transition.
-
-There are two potential points of confusion here.
-The first lies in the definition of ``broken`` phase.
-Though the symmetric phase more closely resembles a pane of broken glass, it leaves $\phi\rightarrow-\phi$ symmetry \textit{un}-broken, thereby giving the title ``broken'' to the more visually uniform configuration.
-An additional potential pitfall is the distinction between the \textit{lattice average} and the \textit{ensemble average}.
-The first is a mean over all lattice sites while the second is a mean over all configurations in the Markov Chain.
-
-\begin{figure}[h]
-    \begin{center}
-      \begin{subfigure}[b]{0.3\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/broken.png}
-        \caption{broken phase \\($m_0^2=-1.0$)}
-      \end{subfigure}%
-      \hfill
-      \begin{subfigure}[b]{0.3\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/transition.png}
-        \caption{transition \\($m_0^2=-0.7$)}
-      \end{subfigure}%
-      \hfill
-      \begin{subfigure}[b]{0.3\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/symmetric.png}
-        \caption{symmetric phase \\($m_0^2=-0.4$)}
-      \end{subfigure}
-      \hfill
-      \caption{\label{fig:primary observables} Visualization of broken phase, symmetric phase and transition. Simulation run on $64\times64$ lattice, plotted after 1000 sweep thermalization.}
-  \end{center}
-\end{figure}
-
-\begin{table}[h]
-    \begin{center}
-    {\renewcommand{\arraystretch}{1.2} %
-    \begin{tabular}{ c|| c | c | c}
-        & broken & transition & symmetric \\ \hline
-        $|\bar\phi|$ & 0.56 & 0.07 & 0.02 \\ 
-        $S_E/L^2$ & 0.29 & 0.40 & 0.44
-    \end{tabular}}
-
-    \end{center}
-    \caption{\label{tab:primary observables} Average magnetization with average action per site corresponding to the particular configurations in Fig.~\ref{fig:primary observables}.}
-\end{table}
-
-\subsubsection{Average Magnetization}
-\label{sec:avg mag}
-The average magnetization quantifies the total alignment of the field.
-In both theories, a value of zero indicates the symmetric phase while a nonzero value indicates broken symmetry.
-In the NLSM, a magnitude of one represents total alignment.
-
-Due to the $\phi\rightarrow-\phi$ symmetry of the $\phi^4$ model, the ensemble mean of the average magnetization $\langle \phi \rangle$ is $0$.
-Likewise, the $O(3)$ symmetry in the NLSM enforces $\langle \e \rangle$.
-To measure the alignment, we therefore use the magnitude of this quantity, defined in the $\phi^4$ model as
-\begin{equation}
-|\bar\phi| \equiv \frac{1}{L^2}\left| \sum_i^{L^2} \phi(x_i)\right|.
-\end{equation}
-and in the NLSM as
-\begin{equation}
-    |\bar\e| \equiv \frac{1}{L^2}\left|\sum_i^{L^2} \e(x_i)\right|.
-\end{equation}
-
-In the symmetric phase, both $\langle \bar\e \rangle$ and $\langle \bar\phi \rangle = 0$ while in the broken phase they are both nonzero.
-
-\subsubsection{Magnetic Susceptibility}
-Despite being the prototypical indicator of a phase transition, the magnitude of the magnetization is not always clear on a discretized lattice.
-For this reason, we consider other observables such as the magnetic susceptibility.
-Generally, this value is defined as
-\begin{equation}
-    \chi_m \equiv L^2 \big(\langle {\bar e}^2 \rangle - \langle \bar e \rangle^2\big)
-\end{equation}
-in the NLSM (the $\phi^2$ expression is nearly identical).
-Manifestly, this expression appears as a secondary observable since it is not defined for each lattice configuration.
-However, due to the rotation invariance of both fields, the second term disappears (see Sec.~\ref{sec:avg mag}), yielding
-\begin{align}
-    \chi_m &= L^2 \left\langle \left( \sum_x e(x) \right)^2\right\rangle  \\
-           &= L^2 \left\langle \sum_{x,y}\e(x)\cdot \e(y) \right\rangle. 
-\end{align}
-Due to translational invariance from the periodic boundary conditions, this expression becomes
-\begin{equation}
-    \chi_m = \left\langle \sum_x \e(0)\cdot\e(x) \right\rangle
-\end{equation}
-for the NLSM. Following the same logic, we derive 
-\begin{equation}
-    \chi_m = \left\langle \sum_x \phi(0)\phi(x) \right\rangle
-\end{equation}
-for the $\phi^4$ model.
-
-\subsubsection{Internal Energy}
-The internal energy is defined as \cite{berg1981}
-\begin{equation}
-    E = \frac{2}{\beta L^2} \langle S \rangle
-\end{equation}
-in the NLSM.
-
-\subsection{Secondary Observables}
-Unlike primary observables, secondary observables are defined for each \textit{ensemble}, not each configuration.
-To further ensure an accurate perspective of the phase transition in the $\phi^4$ model, we incorporate two indicators: the Binder cumulant and the bimodality.
-Since we do not measure these observables on the NLSM, we do not include their analogous expressions here.
-
-\subsubsection{Binder Cumulant}
-We define the Binder cumulant $U$ as \cite{landau2000}
-\begin{equation}
-    U \equiv 1-\frac{\langle \bar{\phi^4} \rangle}{3\langle \bar{\phi^2}\rangle^2}.
-\end{equation}
-Similar to the magnitude of the average magnetization, this formula yields $0$ in the symmetric phase and a nonzero value in the broken phase.
-This nonzero value i $2/3$ for the Binder cumulant.
-The advantage of this metric is a sharper transition at the critical $m_0^2$.
-
-\subsection{Bimodality}
-The final phase transition indicator we use is the bimodality.
-In the symmetric phase, the average magnetization $\bar\phi$ centers around $0$ while in the broken phase, these values cluster around two peaks.
-Qualitatively, this metric measures the separation of these peaks.
-
-We begin by measuring $\bar\phi$ for each configuration.
-We separate each value into an odd number number of bins, ensuring that there is a bin centered at $\bar\phi=0$.
-We then calculate the number of configurations $n_0$ in the center bin and the number of configurations $n_{max}$ in the fullest bin.
-The bimodality is then calculated as
-\begin{equation}
-    B = 1 - \frac{n_0}{n_{max}}.
-\end{equation}
-When the configurations are centered around $\bar\phi=0$, i.e.
-the symmetric phase, this value is $0$.
-When the configurations are aligned such that $\bar\phi\neq0$, i.e.
-the broken phase, this value becomes $1$.
-\subsection{Jackknife Method}
-Though it simple to measure the uncertainties associated with primary observables, secondary observables make this process more complicated.
-While we could propagate the uncertainty of the Binder cumulant, such a process is not clear for the bimodality.
-Therefore, we utilize a method known Jackknife resampling.
-
-We begin by calculating some observable $O$ on an ensemble of $N$ configurations.
-Then, for each configuration $i$, we calculate the same observable but exclude said configuration.
-This leaves us with a set of $N$ observables $O_i$.
-Assuming independent measurements, we can calculate the variance of $O$ as
-\begin{equation}
-    \mathrm{Var}(O) = \sum_{i} \left(C_i - C\right)^2.
-\end{equation}
-We use this formula to calculate all uncertainties in this study.
-\section{Topological Observables}
-\label{sec:topological charge}
-The $O(3)$ non-linear sigma model model features topological properties originating from two properties: 
-\begin{enumerate}
-    \item At $x\rightarrow\infty$, the field must become uniform since the Lagrangian must vanish.
-This allows us to model $x\rightarrow\infty$ as a single point on the field, forming a Riemann sphere in three dimensions.
-    \item The elements of the $O(3)$ non-linear sigma model are three-dimensional unit vectors, thereby existing on a three dimensional unit sphere. 
-\end{enumerate}
-
-With these two properties, we can view the field as a continuous mapping between two 3D spheres, denoted as $S^2$, and associate an integer number of wrappings to each mapping from $S^2$ to $S^2$.
-We can envision a tangible metaphor for this wrapping with a balloon and a baseball: by simply inserting the baseball into the balloon, we have established a mapping from every point on the balloon to every point on the baseball.
-We can create an equally valid map by twisting the balloon's mouth and wrapping the baseball again.
-In a purely mathematical world, we perform this process an infinite number of times, thereby associating every possible mapping with an integer.
-The group of integers is known as the \textit{homotopy group} of the non-linear sigma model.
-We associate every field configuration with an element of this group, known as the \textit{topological charge}, which we denote as $Q$.
-
-In practice, this quantity is nontrivial to calculate.
-To begin, we define a local topological charge density $q$, defined for each square of adjacent lattice points.
-This square, known as a \textit{plaquette}, is denoted $x^*$.
-The global charge $Q$ is the sum of all local charges:
-\begin{equation}
-    Q \equiv \sum_{x^*} q(x^*).
-\end{equation}
-As a function of $x^*$, the charge density is a function of the field on the plaquette vertices, an idea visualized in Fig.~\ref{fig:plaquette}.
-\begin{figure}
-    \centering
-    \begin{tikzpicture}
-        \draw[step=4cm,gray,thin] (-2,-2) grid (6,6);
-        \draw[dotted] (0,0) -- (4,4);
-        \draw[thick](0,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_1$} -- 
-                    (0,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_2$} -- 
-                    (4,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_3$} -- 
-                    (4,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_4$} -- cycle ;
-                    \draw[] (2,2) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north east]{$x^*$};
-
-        \draw[-stealth] (0.2,1) -- (0.2,3);
-        \draw[-stealth] (1,3.8) -- (3,3.8);
-        \draw[-stealth] (3,3.5) -- (1,1.5);
-
-        \draw[-stealth] (1,0.5) -- (3,2.5);
-        \draw[-stealth] (3.8,3) -- (3.8,1);
-        \draw[-stealth] (3,0.2) -- (1,0.2);
-    \end{tikzpicture}
-    \caption{\label{fig:plaquette} Visualization of plaquette $x^*$.
-Dotted line separated plaquette into two signed areas which are used to define the topological charge density $q(x^*)$.
-Arrows represent order of signed area.}
-\end{figure}
-In the NLSM, the field at each of these vertices is a point on the sphere $S^2$.
-Therefore, there is a signed area $A$ on the sphere associated with each triplet of points, as shown in Fig.~\ref{fig:sphere} (the sign changes with odd permutations of the ordering).
-We follow the derivation in \cite{berg1981} and split the plaquette into two triangles, as shown in Fig.~\ref{fig:plaquette}, with the ordering determining the sign.
-The topological charge density is defined using this signed area as
-\begin{equation}
-    q(x^*) = \frac{1}{4\pi} \bigg[A\Big(\e(x_1), \e(x_3), \e(x_3)\Big) + A\Big(\e(x_1), \e(x_3), \e(x_4)\Big) \bigg].
-\end{equation}
-This sign is defined if $A\neq 0, 2\pi$, or in other words, as long as the three points on the sphere are distinct and do not form a hemisphere.
-In numerical calculations, these points can be ignored.
-Therefore, we impose that the signed area is defined on the smallest spherical triangle, or mathematically
-\begin{equation}
-    -2\pi < A < 2\pi.
-\end{equation}
-
-Following \cite{berg1981}, this yields an expression for the signed area
-\begin{equation}
-    A(\e_1, \e_2, \e_1) = 2 \:\mathrm{arg}\Big(1+\e_1\cdot\e_2 + \e_2\cdot\e_3 + \e_3\cdot\e_1 + i\e_1 \cdot (\e_2\times\e_3)\Big).
-\end{equation}
-Under periodic boundary conditions, these triangles on the sphere necessarily wrap $S^2$ an integer number of times, ensuring $Q$ is an integer.
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}
-    \pgfdeclarelayer{nodelayer}
-    \pgfdeclarelayer{edgelayer}
-    \pgfdeclareradialshading{sphere4}{\pgfpoint{-0.2cm}{0.5cm}}% 
-        {rgb(0cm)=(1,1,1);
-        rgb(1cm)=(0.5,0.5,0.5); rgb(1.05cm)=(1,1,1)}
-        %rgb(0.7cm)=(0.1,0.1,0.1); rgb(1cm)=(0.5,0.05,0); rgb(1.05cm)=(1,1,1)}
-    \pgfsetlayers{nodelayer,edgelayer}
-    \tikzstyle{label}=[fill=none, draw=none, shape=circle]
-    \tikzstyle{point}=[inner sep=0pt, minimum size=0.2cm,fill=black, draw, shape=circle]
-    %\tikzstyle{interior line}=[{Stealth[scale=1.5]}-,dotted,thick]
-    \tikzstyle{interior line}=[dotted,thick]
-
-    \tikzstyle{triangle}=[thick]
-    \tikzstyle{arrow}=[->, thick]
-	\begin{pgfonlayer}{nodelayer}
-		\node [style=point] (4) at (0.5, 2.5) {};
-		\node [style=point] (5) at (0.55, 0.625) {};
-		\node [style=point] (6) at (2.25, 0.75) {};
-		\node (7) at (0, 0) {};
-		\node (8) at (1.25, 1.375) {};
-		\node (9) at (2.425, 2.525) {};
-	\end{pgfonlayer}
-	\begin{pgfonlayer}{edgelayer}
-        \draw (0,0) circle (3cm);
-        %\shade[inner color=white,outer color=lightgray] (0,0) circle (3cm);
-        \shade[shading=sphere4] (0,0) circle (3cm);
-		\draw [bend left=15,style=triangle] (4.center) to (6.center);
-		\draw [bend left=355,style=triangle] (6.center) to (5.center);
-		\draw [bend right=5,style=triangle] (5.center) to (4.center);
-		\draw [style=interior line] (4.center) to (7.center);
-		\draw [style=interior line] (5.center) to (7.center);
-        \draw [style=interior line] (6.center) to (7.center);
-        \draw [fill=gray!80] (4.center) to [bend left=15] (6.center) to [bend left=355] (5.center) to [bend right=5] cycle;
-
-		%\draw [style=arrow] (8.center) to (9.center);
-
-        \node [style=label, anchor=north] at (2.25, 0.75) {\small $\e(x_1)$};
-        \node [style=label, anchor=north] at (0.55, 0.625) {\small $\e(x_2)$};
-        \node [style=label, anchor=east] at (0.5, 2.5) {\small $\e(x_3)$};
-		\node [style=point] at (4){};
-		\node [style=point] at (5){};
-		\node [style=point] at (6){};
-
-        \node [style=label] at (8) {\small $A$};
-	\end{pgfonlayer}
-\end{tikzpicture}
-\caption{\label{fig:sphere} Visualization of signed area $A$ on the sphere $S^2$ traced out by field at points $x_1$, $x_2$ and $x_3$.}
-\end{figure}
-
-Following this quantity, we can define a topological susceptibility $\chi_t$
-\begin{equation}
-\chi_t \equiv \frac{1}{L^2} \Big( \langle Q^2 \rangle - \langle Q \rangle^2 \Big).
-\end{equation}
-In the trivial case, $\langle Q \rangle$ disappears and   
-\begin{equation}
-    \chi_t = \frac{1}{L^2} \sum_{x,y} \langle q(x)q(y)\rangle.
-\end{equation}
-Assuming periodic boundary conditions and therefore translational symmetry, this expression simplifies to 
-\begin{equation}
-    \chi_t = \frac{1}{L^2} \sum_{x} \langle q(x)q(0)\rangle.
-\end{equation}
-
-On the lattice, this quantity is known to diverge in the continuum limit.\cite{bietenholz2018}
-
-\subsection{NLSM $\theta$ term}
-
-In the nonlinear sigma model, we expect $\langle Q \rangle=0$. By introducing a $\theta$ term into the action,
-\begin{equation}
-    S[\e] \rightarrow S[\e] - i \theta Q[\phi],
-\end{equation}
-we can construct a topologically nontrivial model (that is, $\langle Q \rangle \neq 0$).
-
-\section{Ultraviolet Divergences}
-
-In Section~\ref{sec:pathintegral},  we defined a fundamental equation of quantum fields using a ``path integral'' which encompasses an uncountably infinite configuration space.
-However, we said nothing of the integral's convergence.
-In fact, many fundamental processes in QFT have divergent amplitudes, yielding nonsensical results.
-The most type of divergence stems from high-momentum states, giving them the name ``ultraviolet divergences''.
-The remedy to this catastrophe is unintuitive.
-Essentially, we adopt infinite values for the parameters of the Lagrangian ($m_0^2$ and $\lambda$ in $\phi^4$ theory).
-Since neither of these two quantities is ever measured directly, we do not have to assume that their values are finite.
-In practice, this technique is involved and consists of two steps: regularization and renormalization.
-
-\subsection{Regularization: fields on the lattice}
-Regularization is a process which introduces a new parameter into calculations. One example is a momentum cutoff. This technique transforms infinite momentum integrals as follows:  
-\begin{equation*}
-    \int_0^\infty dk \rightarrow \int_0^\Lambda dk,
-\end{equation*}
-introducing $\Lambda$ as a regularization parameter.
-This process makes results $\Lambda$-dependent, but finite.
-Another example is dimensional regularization, which calculates results in terms of the spacetime dimension $d$ and analytically continues this parameter into the real numbers.
-
-In this study, we employ lattice regularization.
-This process discretizes the field, modeling the field $\phi(x)$ as a lattice $\phi_i$ where $i$ indexes lattice sites.
-The inherent parameter in this case is the lattice spacing $a$ which measures the width of each lattice chunk.
-
-\subsection{Renormalization}
-After regularization, we redefine the Lagrangian parameters in terms of the regularization parameter, following a handful of boundary conditions.
-In this study, we assert that $L/\xi$ remains constant, where $L$ is the side length of the system and $\xi$ is the coherence length.
-In perturbation theory, this process is arduous and includes the introduction of counter-terms into the Lagrangian.
-In the case of the non-linear sigma model, it is impossible using counterterms\citeneeded but can be performed numerically.
-In this study, we use predetermined values from \cite{bietenholz2018}.
-
-At this point, we can calculate observables as functions of regularization parameters.
-To achieve physical values, we take the limit as the regularization parameters approach their physical values.
-With a momentum cutoff, we take $\Lambda \rightarrow \infty$ and with dimensional regularization we usually take $d\rightarrow 4$.
-With lattice regularization, we approach the continuum, taking the lattice spacing $a\rightarrow 0$.
-
-At this point, we have surely eliminated all divergences, right?
-Unfortunately, this is not always the case.
-The topological susceptibility $\chi_t$ is one such value that diverges in the continuum limit.
-As we decrease the width of each lattice site, high frequency modes become more significant, leading to an ultraviolet divergence in the operator.
-
-\subsection{The Gradient Flow}
-\label{sec:gradflow}
-To remove this ultraviolet divergence, we adopt a technique is ``smearing'', a local averaging of the field.\cite{solbrig2007} Specifically, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.
-
-Specifically, the gradient flow pushes field configuration toward classical minima of the action.
-Additionally, renormalized correlation functions remain renormalized at nonzero flow time.\cite{luscher2013} In 2D $\phi^4$ scalar field theory, the gradient flow is defined by the differential equation
-\begin{equation}
-    \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
-\end{equation}
-where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime and $\tau$ is the flow time.
-Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
-In the $\phi^4$ theory, we can solve this equation exactly to find \cite{monahan2016}
-\begin{equation}
-    \rho(\tau, x) = \frac{1}{4 \pi \tau} \int d^2 y e^{-(x-y)^2/4\tau} \phi(y).
-\end{equation}
-This function forms a Gaussian, smoothly dampening high-momentum modes and removing ultraviolet divergences from evolved correlation functions.\cite{makino2015a} We can visualize this by plotting the $\phi$ field, shown in Fig.~\ref{fig:flow}.
-These plots demonstrate the reduction of high momentum modes.
-
-\begin{figure}[h]
-  \centering
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=0.9\textwidth]{imgs/gf0.png}
-        \caption{$\tau=0$}
-      \end{subfigure}%
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=0.9\textwidth]{imgs/gf1.png}
-        \caption{$\tau=0.001$}
-      \end{subfigure}%
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=0.9\textwidth]{imgs/gf2.png}
-        \caption{$\tau=0.01$}
-      \end{subfigure}%
-      \begin{subfigure}[b]{0.2\textwidth}\centering
-        \includegraphics[width=0.9\textwidth]{imgs/gf3.png}
-        \caption{$\tau=0.1$}
-      \end{subfigure}%
-      \caption{\label{fig:flow} Effect of flow time evolution on a random lattice in the symmetric phase. White represents positive values of $\phi$ while black represents negative.}
-\end{figure}
-
-Generally, we can choose any flow time equation that drives the field towards a classical minimum.
-Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields.
-In the non-linear sigma model, it has been shown that an appropriate manifestation of the flow time can resolve divergences \cite{makino2015a}.
-Therefore, it is possible to define a different flow equation for this model as well.
-Following \cite{bietenholz2018}, we can define the gradient flow in this model via the differential equation
-\begin{equation}
-    \label{eq:nsm_gradflow}
-    \partial_\tau \vec\rho (\tau,x) = \left( 1 - \vec\rho(\tau,x) \vec\rho(\tau,x)^T \right) \partial^2 \vec\rho(\tau,x),
-\end{equation}
-where $\partial^2$ is the Laplacian operator in Euclidean space\footnote{Explicitly, $\partial^2 = \frac{\partial^2}{\partial t^2} + \nabla^2$}.
-We solve this equation numerically using the boundary condition $\vec\rho(0,x) = \e(x)$.
-
+\input{theory.tex}
 \chapter{Methods}
 \label{sec:methods}
-Our study of the gradient flow in the non-linear sigma model consists of a computational part and an analytical part.
-We begin by outlining a numerical Monte Carlo method to simulate the lattice in two and three dimensions.
-We verify our program with the well-studied $\phi^4$ scalar field theory.
-We then generalize our model to a vector field to simulate the non-linear sigma model.
-This simulation system provides data on vacuum states with which we study the gradient flow's effect on topology.
-
-We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories. 
-\section{Fields on the Lattice}
-To implement the lattice regularization technique, we must redefine the field action in terms of discrete positions, a process known as ``discretization''.
-We transition from $x$ as a continuous vector in $\mathbb{R}^2$ to $x_{i,j}$ where
-\begin{equation}
-    x_{i,j} = ia \hat{\mu}_t + j a \hat{\mu}_x.
-\end{equation}
-Here, $a$ is the lattice constant and $\hat{\mu}_0$ and $\hat{\mu}_1$ are unit vectors.
-This change effectively shifts the domain of the field from $\mathbb{R}^2$, which is uncountably infinite, to $\mathbb{Z}^2$, which is countably infinite.
-To achieve a finite domain, we impose periodic boundary conditions, such that
-\begin{equation}
-    \phi\left(x_{i+L,j}\right) = \phi\left(x_{i,j+L}\right) = \phi\left(x_{i,j}\right)
-\end{equation}
-where $L$ is the side length (in units of the lattice constant $a$) of the system. In this study we focus solely on square geometries and thus the side length $L$ is unambiguous.
-
-In the $\phi^4$ model, we can specify a discrete action using the Euclidean action from Eq. \ref{eq:phi4 euclidean action}. We begin by redefining the derivative operator as a difference:
-\begin{align}
-    \partial_\mu \phi = \frac{\phi\left(x + a \hat{\mu}\right) - \phi(x)}{a}
-\end{align}
-We can then define the kinetic term
-\begin{align}
-    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow& \frac{1}{2a^2}\left[ \left(\phi(x+a \hat{t}) -\phi(x)\right)^2 + \left(\phi(x+a \hat{x}) -\phi(x)\right)^2\right] \\
-    \begin{split} \rightarrow& \frac{1}{2a^2}\left[ \phi^2(x+a\hat{t}) + \phi^2(x+a\hat{x}) + 2\phi^2(x)\right.\\ &\qquad \left.- 2\phi(x+a\hat{t})\phi(x) - 2\phi(x+a\hat{x})\phi(x) \right]\hfilneg\end{split}
-\end{align}
-Since we will eventually sum over all sites $x$, the periodic boundary conditions imply that an overall shift in $x$ does not effect the final action.
-Therefore, we can combine the first two terms with the third term to produce
-\begin{equation}
-    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow \frac{1}{a^2}\left[ 2\phi^2(x) - \phi(x+a\hat{t})\phi(x) - \phi(x+a\hat{x})\phi(x) \right]
-\end{equation}
-Unlike the kinetic term, the mass and interaction terms remain unchanged under the discretization procedure. The only remaining change is a shift from an integral to a sum. This takes the form
-\begin{equation}
-    \int dtdx \rightarrow a^2 \sum_i
-\end{equation}
-such that the final discretized action becomes
-\begin{equation}
-    \label{eq:phi4 discretized action}
-    S_{\mathrm{lat}}[\phi] = \sum_i \left[-\phi(x_i + a\hat{t})\phi(x_i) - \phi(x_i + a\hat{x})\phi(x_i) + \left(2+\half m_0^2\right)\phi^2(x_i) + \frac{1}{4}\lambda \phi^4(x_i)\right]
-\end{equation}
-
-Likewise, we can discretize the non-linear sigma model. In this case, the derivative term becomes 
-\begin{equation}
-    \half \left(\partial_t\e\right)^2 + \half \left(\partial_x\e\right)^2 \rightarrow \frac{1}{a^2}\left[ 2 - \e(x+a\hat{t})\cdot\e(x) - \e(x+a\hat{x})\cdot\e(x) \right].
-\end{equation}
-Note that we have used the identity $\e\cdot\e = 1$. Inserting this into Eq.~\ref{eq:nlsm euclidean action} yields the discretized action
-\begin{equation}
-    \label{eq:nlsm discretized action}
-    S_\mathrm{lat}[\e] = \sum_i \left[ 2 - \e(x+a\hat{t})\cdot\e(x) - \e(x+a\hat{x})\cdot\e(x) \right].
-\end{equation}
-
-Finally, we redefine the gradient flow in on the lattice.
-Since the gradient flow is solved exactly in the $\phi^4$ model, we rely on a Fast Fourier Transform.
-\textit{TODO: explicit expression for this} In the NLSM, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes
-\begin{equation}
-    \label{eq:nsm_gradflow_disc}
-    \partial_\tau \e (\tau, x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T\right) \partial^2 \e(\tau,x),
-\end{equation}
- where the Laplacian operator $\partial^2$ is defined as
-\begin{equation*}
-    \partial^2 \e(\tau,x) = \e(\tau, x+a \hat{t}) + \e(\tau,x-a\hat t) + \e(\tau, x+a \hat{x}) + \e(\tau,x-a\hat x) - 2 d \e(t)_x.
-\end{equation*}
-%TODO: derive the laplacian operator
-
-\section{Monte Carlo Simulations}
-\label{sec:mc}
-We implement a Markov Chain Monte Carlo method following Schaich's thesis \cite{schaich2006}.
-This implementation utilizes a ``random walk,'' i.e.
-a set of random steps through phase space, to determine statistical values such as correlation functions across the lattice.
-By the definition of the Markov chain, the probability of adoption of each state, and therefore its inclusion in the Monte Carlo calculation, depends only on the current state and the proposed state.
-This probability is denoted as $P(\mu\rightarrow\nu)$ where $\mu$ and $\nu$ are the existing and proposed lattice configurations respectively.
-
-\subsection{Metropolis Algorithm}
-We primarily use the Metropolis algorithm for the calculation of new Markov chain configurations.
-We begin with a so-called ``hot start,'' where each field value at each lattice site is randomly selected.
-Then we propose a new value for a single lattice point, which is accepted with a probability
-\begin{equation}
-    P(\phi_a\rightarrow\phi_b) = \begin{cases} 
-        e^{-(S[\phi_b] - S_[\phi_a])} & S[\phi_b] < S_[\phi_a] \\
-        1 & \mathrm{otherwise} \\
-   \end{cases}
-\end{equation}
-where $\phi_a$ is the initial configuration and $\phi_b$ is the proposed configuration.
-This process is performed for each point on the lattice, making up a ``sweep.'' Repeating this sweep many times pushes the lattice toward the action minimum.
-
-\subsection{Wolff Cluster Algorithm}
-
-Though the Metropolis algorithm will slowly find the absolute minimum of the theory, the presence of local minima can greatly prolong the convergence.
-Both the $\phi^4$ model and the non-linear sigma model feature ``kinetic'' terms with gradients of $\phi$.
-Therefore, the presence of large similarly-valued regions in the lattice can lead to a local minimum.
-One method of removing these clusters involves identifying all clusters on the lattice and probabilistically flipping each, a technique known as the Swendsen-Wang algorithm\cite{swendsen1987}.
-
-A more efficient approach is the Wolff algorithm\cite{wolff1989}, which \textit{grows} one cluster probabilistically and flips it unconditionally.
-In the case of $\phi^4$ theory, this flipping takes the form of a simple sign change.
-In the non-linear sigma model we choose a random unit vector $\vec r$ and consider the projection of the field on this vector.
-When the cluster flips, each site is flipped along this direction.\citeneeded To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site, growing the cluster from a single, randomly selected ``seed''.
-Starting with the seed, the probability of adding each neighboring site is given by the source site $x$ and the proposed site $x'$.
-Wolff defines this probability for arbitrary sigma models as
-\begin{equation}
-    \label{eq:phi4 wolff padd}
-    P_{add}(\e(x),\e(x')) = \begin{cases} 
-        1 - e^{2\beta [\vec{r} \cdot \e(x)][\vec{r} \cdot \e(x')]} & \mathrm{sgn}[\vec{r}\cdot\e(x)]=\mathrm{sgn}[\vec{r}\cdot\e(x')]\\
-        0 & \mathrm{otherwise} \\
-   \end{cases}
-\end{equation}
-This expression is designed to preserve the detailed balance equations.
-We can demonstrate this quality, and motivate an equivalent expression for the $\phi^4$ model, by considering the probability $P(\phi\rightarrow f_C(\phi))$ of flipping some cluster $C$.
-Generally,
-\begin{equation}
-    P\big(\phi \rightarrow f_C\left(\phi\right)\big) \propto \prod_{\langle x,x'\rangle \in \partial C}  \Big[1 - P_{add}\big(\e(x),\e(x')\big)\Big]
-\end{equation}
-where $\partial C$ is the set of pairs of sites on the boundary of $C$.
-Since $P_{add}=0$ for unaligned sites, these pairs contribute nothing to the value.
-We can also find the probability $P(f_C(\phi) \rightarrow \phi)$ with the same expression:
-\begin{equation}
-P\big(f_C(\phi)\rightarrow \phi \big) \propto \prod_{\langle x,x'\rangle \in \partial C}  \Big[1 - P_{add}\big(\e(x),R\;\e(x')\big)\Big],
-\end{equation}
-where the matrix $R$ which is a reflection matrix along the vector $\vec{r}$.
-
-From the discretized action of the NLSM model (Eq.~\ref{eq:nlsm discretized action}) and the detailed balance equation (Eq.~\ref{eq:detailedbalance2}), we derive
-\begin{equation}
-    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\e(x),\e(x')\big)}{ 1 - P_{add}\big(\e(x),R\;\e(x')\big)}  = \mathrm{exp}\left\{\beta\sum_{\langle x,x'\rangle \in \partial C} \e(x)\cdot[R-1]\e(x')\right\}.
-\end{equation}
-Note that all the pairs within and outside the cluster cancel in the fraction on the left and the difference on the right. Using the definition of the reflection matrix
-\begin{equation}
-    R\;\e = \e - 2(\e\cdot\vec{r})\e,
-\end{equation}
-we can simplify the equation to be 
-\begin{equation}
-    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\e(x),\e(x')\big)}{ 1 - P_{add}\big(\e(x),R\;\e(x')\big)}  = \prod_{\langle x,x'\rangle \in \partial C}\mathrm{exp}\big\{2\beta[r\cdot\e(x)][r\cdot\e(x')]\big\}.
-\end{equation}
-By plugging in Eq.~\ref{eq:phi4 wolff padd}, it is clear to see this equation is satisfied.
-
-Using this same reasoning, we can deduce an expression for $P_{add}$ in the $\phi^4$ model.
-Since this model is one dimensional, the reflection matrix $R$ is equivalent to $-1$.
-Adapting the NLSM detailed balance equation for the $\phi$ field, we find
-\begin{equation}
-    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\phi(x),\phi(x')\big)}{ 1 - P_{add}\big(\phi(x),-\phi(x')\big)}  = \prod_{\langle x,x'\rangle \in \partial C}\mathrm{exp}\{-2 \phi(x)\phi(x')\}.
-\end{equation}
-This equation is satisfied by the ansatz
-\begin{equation}
-    \label{eq:nlsm wolff padd}
-    P_{add}(\phi(x),\phi(x')) = \begin{cases} 
-        1 - e^{-2\phi(x)\phi(x')]} & \mathrm{sgn}[\phi(x)]=\mathrm{sgn}[\phi(x')]\\
-        0 & \mathrm{otherwise} \\
-   \end{cases}.
-\end{equation}
-We will use this expression in the computational implementation of this algorithm.
-
-Fig.~\ref{fig:wolff} shows a real demonstration of this process.
-We can see a large cluster of negative field values becoming positive.
-Note that periodic boundary conditions apply, so the small island of black at the bottom is actually part of the larger cluster.
-Furthermore, this visualization demonstrates the probabilistic nature of the Wolff algorithm.
-Since states are added probabilistically, there are some small holes in the cluster.
-These will be removed by following Metropolis sweeps.
-\begin{figure}[h]
-  \centering
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/wolffa.png}
-        \caption{before cluster flip}
-      \end{subfigure}%
-      \hfill
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/wolffb.png}
-        \caption{after cluster flip}
-      \end{subfigure}
-      \hfill
-      \caption{\label{fig:wolff} An example of the Wolff cluster algorithm in the $\phi^4$ model.
-White represents positive values of $\phi$ while black represents negative.
-$\lambda=0.5$, $m_0^2=-0.9$}
-
-\end{figure}
-
-\subsection{Checkerboard algorithm}
-
-In order to parallelize the Metropolis algorithm, we use a checkerboard algorithm.
-Since the Lagrangian density at each site does not depend on any sites diagonal to it, the lattice can be split into ``white'' sites and ``black'' sites, like the tiles on a checkerboard.
-Each white site is independent of every other white site and likewise with black sites.
-Therefore, we can split the sites of each color into separate parallel processing nodes and independently run the Metropolis algorithm, ensuring that no site affects the Lagrangian density at any other site.
-We use this method to parallelize the code through the Message Passing Interface (MPI).
-
-\subsection{Thermalization}
-Since we begin with a random lattice, the first few configurations of our Monte Carlo simulation will be far from the lattice minimum.
-We plot the action as a function of metropolis sweeps in Fig.~\ref{fig:therm}.
-\begin{figure}[h]
-  \centering
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/therm24.png}
-        \caption{$L=24$}
-      \end{subfigure}%
-      \hfill
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/therm404.png}
-        \caption{$L=404$}
-      \end{subfigure}
-      \hfill
-      \caption{\label{fig:therm} Plots of the action as a function of Monte Carlo time, starting with a random NLSM lattice.}
-\end{figure}
-Based on this plot, we determine that 1000 sweeps will give sufficient time for the system to reach the classical action minimum. We use this value for the remainder of this study.
-\subsection{Autocorrelation times and thermalization}
-One important aspect is the correlation between different states of the Markov chain.
-Ideally, each sample of the lattice would be completely independent, but this is not the case.
-Since each configuration is based on previous configurations, there is a correlation between each pair of members in the Markov chain, decreasing exponentially based on the number of steps between.
-Explicitly, the autocorrelation scales as \[e^{t/\tau_{int}}\] where $t$ is the number of steps between configurations and $\tau_{int}$ is the autocorrelation time.\footnote{Though it is called a time, $\tau_{int}$ is in units of Markov Chain steps.} When performing simulations of the lattice, the number of sweeps between measurements should be much larger than $\tau_{int}$.
-
-    We use Wolff's automatic windowing procedure \cite{wolff2007} and the magnetic susceptibility $\chi_m$ to estimate the autocorrelation.
-Using Wolff's public MatLab code\footnote{\url{https://www.physik.hu-berlin.de/de/com/ALPHAsoft}}, we estimate the autocorrelation time for $L=24$ and $L=404$ lattices.
-This algorithm identifies the optimal window size with which to calculate the autocorrelation time.
-We perform this process on $L=24$ and $L=404$ lattices using a thermalization of $1000$ sweeps and $500$ total measurements.
-Note that this calculation included a Wolff cluster algorithm every $5$ sweeps.
-The result of this calculation is shown in Fig.~\ref{fig:tauint}.
-\begin{figure}[h]
-  \centering
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/tauint24.png}
-        \caption{$L=24$, $\tau_{int}=4.43$}
-      \end{subfigure}%
-      \hfill
-      \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.6\textwidth]{imgs/tauint404.png}
-        \caption{$L=404$, $\tau_{int}=12.81$}
-      \end{subfigure}
-      \hfill
-      \caption{\label{fig:tauint} Plots of automatic windowing procedure used to calculate $\tau_{int}$ for the NLSM model.}
-\end{figure}
-
-Based on these two values for $\tau_{int}$, we decide to measure every $50$ sweeps for each simulation. This value will ensure that each measurement is effectively independent.
-
-\subsection{Runge-Kutta Algorithm}
-In order to calculate the gradient flow, we numerically solve the differential equation using a fourth-order Runga-Kutta approximation. This algorithm refines the Euler method 
-\begin{equation*}
-    \e(\tau+h)_x \approx \e(\tau)_x + h f(\e(\tau)_x).
-\end{equation*}
-where $f(\e)$ is defined for convenience as 
-\begin{equation}
-    f(\e)=\partial_\tau \e (\tau)_x  = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \partial^2 \e(\tau)_x,
-\end{equation}
-following from Eq.~\ref{eq:nsm_gradflow_disc}. To the fourth order, this approximation becomes 
-%
-\begin{align}
-    \label{eq:rungekutta}
-    k_1 &= h f\left(\e\left(\tau\right)_x\right) \\ 
-    k_2 &= h f\left(\e\left(\tau\right)_x + \frac{k_1}{2}\right) \\ 
-    k_3 &= h f\left(\e\left(\tau\right)_x + \frac{k_2}{2}\right) \\ 
-    k_4 &= h f\left(\e\left(\tau\right)_x + k_3\right) \\ 
-    \e(\tau+h)_x &= \e(\tau)_x + \frac{k_1}{6} + \frac{k_2}{3} + \frac{k_3}{3} + \frac{k_4}{6} + O(h^5).
-\end{align}
-This method is usually superior to Euler's method and the midpoint method \cite{vetterling1992}.
-
-To increase the efficiency of this algorithm, we implement the step-doubling algorithm to adaptively adjust $h$.
-If the error of a Runge-Kutta step is greater than the tolerance, the same step is repeated with half the step size.
-Alternatively, if the error is less than half of the tolerance, the step size is doubled for the next calculation.
-Finally, if the step size is greater than the distance to the next measurement, that distance is used as the step size, using the normal value afterwards.
-Otherwise, the algorithm proceeds with the consistent step size.
-
-\section{Topological charge with a $\theta$ term}
-In Sec.~\ref{sec:topological charge}, we discussed the introduction of a $\theta$ term into the action.
-This change made the theory ``topological'', pushing $\langle Q \rangle$ away from zero.
-In order to calculate $\langle Q \rangle$ as a function of $Q$, we consider the path integral
-\begin{align}
-    \langle Q \rangle_\theta &=\int \mathcal{D}\e\:Q[\e]e^{-S[\e]+i\theta Q[\e]} \\
-                             &=\int \mathcal{D}\e\:\left( Q[\e]e^{i\theta Q[\e]} \right) e^{-S[\e]} \\
-                             &=\langle Q e^{i \theta Q} \rangle_{\theta=0}.
-\end{align}
-Therefore, we can calculate $\langle Q \rangle_\theta$ for arbitrary $\theta$ using the same simulation framework as the $\theta=0$ case.
-
-We can also relate this function to the topological susceptibility. By expanding the exponent as a Taylor series around $\theta=0$, we find that  
-\begin{equation}
-    \langle Q \rangle_\theta = \langle Q \rangle_{\theta=0} + i \theta \langle Q^2 \rangle_{\theta=0} + O(\theta^2),
-\end{equation}
-such that 
-\begin{align}
-    -i \frac{\partial}{\partial \theta} \langle Q \rangle_\theta &= \langle Q^2 \rangle_{\theta=0}\\
-                                                              &= \chi_t L^2.
-\end{align}
-We expect to see the slope of this function at zero increase for large lattice sizes.
-
+\input{methods.tex}
 \chapter{Results}
-\section{$\phi^4$ model results}
-
-We initially implemented the $\phi^4$ model using the Monte Carlo method described in Sec.~\ref{sec:methods} to verify the results of our system.
-According to previous studies (\cite{monahan2016}, \cite{schaich2006}), the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ for $\lambda = 0.5$.
-We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
-      \caption{\label{fig:phi4} Simulation of the phase transition in the $\phi^4$ model: Magnetization $\langle |\bar\phi|\rangle$, magnetic susceptibility $\chi_m$, Binder cumulant $U$ and bimodality $B$ for various values of the mass squared $m_0^2$.}
-\end{figure}
-
-\section{Non-linear sigma model results}
-After confirming the phase transition 
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=\textwidth]{imgs/internal_energy.png}
-      \caption{\label{fig:bergluscher}}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=\textwidth]{imgs/theta.png}
-      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$. Note the different scaling of the $y$-axis.}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=\textwidth]{imgs/bietenholz.png}
-      \caption{\label{fig:bietenholz} $\langle Q^2 \rangle=\chi_t L^2$ as a  }
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=0.6\textwidth]{imgs/hist.png}
-      \caption{\label{fig:hist} Histogram of topological charge values $Q$ for trivial NLSM. $L=404$, 10,000 measurements, measurements very 50 sweps, 1,000 sweep thermalization, $\tau=0$}
-\end{figure}
-
-%\chapter{Statistical Analysis}
-
+\input{results.tex}
 \chapter{Conclusion \& Outlook}
-
-% MPI is unnecessary in this case
-
+\input{conclusion.tex}
 \appendix

### 2021-04-19 final-report.latex/methods.tex
split into chapter files

Our study of the gradient flow in the non-linear sigma model consists of a computational part and an analytical part. We begin by outlining a numerical Monte Carlo method to simulate the lattice in two and three dimensions. We verify our program with the well-studied $\phi^4$ scalar field theory. We then generalize our model to a vector field to simulate the non-linear sigma model. This simulation system provides data on vacuum states with which we study the gradient flow's effect on topology.

We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories. 
\section{Fields on the Lattice}
To implement the lattice regularization technique, we must redefine the field action in terms of discrete positions, a process known as ``discretization''. We transition from $x$ as a continuous vector in $\mathbb{R}^2$ to $x_{i,j}$ where
\begin{equation}
    x_{i,j} = ia \hat{\mu}_t + j a \hat{\mu}_x.
\end{equation}
Here, $a$ is the lattice constant and $\hat{\mu}_0$ and $\hat{\mu}_1$ are unit vectors. This change effectively shifts the domain of the field from $\mathbb{R}^2$, which is uncountably infinite, to $\mathbb{Z}^2$, which is countably infinite. To achieve a finite domain, we impose periodic boundary conditions, such that 
\begin{equation}
    \phi\left(x_{i+L,j}\right) = \phi\left(x_{i,j+L}\right) = \phi\left(x_{i,j}\right)
\end{equation}
where $L$ is the side length (in units of the lattice constant $a$) of the system. In this study we focus solely on square geometries and thus the side length $L$ is unambiguous.

In the $\phi^4$ model, we can specify a discrete action using the Euclidean action from Eq. \ref{eq:phi4 euclidean action}. We begin by redefining the derivative operator as a difference:
\begin{align}
    \partial_\mu \phi = \frac{\phi\left(x + a \hat{\mu}\right) - \phi(x)}{a}
\end{align}
We can then define the kinetic term
\begin{align}
    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow& \frac{1}{2a^2}\left[ \left(\phi(x+a \hat{t}) -\phi(x)\right)^2 + \left(\phi(x+a \hat{x}) -\phi(x)\right)^2\right] \\
    \begin{split} \rightarrow& \frac{1}{2a^2}\left[ \phi^2(x+a\hat{t}) + \phi^2(x+a\hat{x}) + 2\phi^2(x)\right.\\ &\qquad \left.- 2\phi(x+a\hat{t})\phi(x) - 2\phi(x+a\hat{x})\phi(x) \right]\hfilneg\end{split}
\end{align}
Since we will eventually sum over all sites $x$, the periodic boundary conditions imply that an overall shift in $x$ does not effect the final action. Therefore, we can combine the first two terms with the third term to produce 
\begin{equation}
    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow \frac{1}{a^2}\left[ 2\phi^2(x) - \phi(x+a\hat{t})\phi(x) - \phi(x+a\hat{x})\phi(x) \right]
\end{equation}
Unlike the kinetic term, the mass and interaction terms remain unchanged under the discretization procedure. The only remaining change is a shift from an integral to a sum. This takes the form
\begin{equation}
    \int dtdx \rightarrow a^2 \sum_i
\end{equation}
such that the final discretized action becomes
\begin{equation}
    \label{eq:phi4 discretized action}
    S_{\mathrm{lat}}[\phi] = \sum_i \left[-\phi(x_i + a\hat{t})\phi(x_i) - \phi(x_i + a\hat{x})\phi(x_i) + \left(2+\half m_0^2\right)\phi^2(x_i) + \frac{1}{4}\lambda \phi^4(x_i)\right]
\end{equation}

Likewise, we can discretize the non-linear sigma model. In this case, the derivative term becomes 
\begin{equation}
    \half \left(\partial_t\e\right)^2 + \half \left(\partial_x\e\right)^2 \rightarrow \frac{1}{a^2}\left[ 2 - \e(x+a\hat{t})\cdot\e(x) - \e(x+a\hat{x})\cdot\e(x) \right].
\end{equation}
Note that we have used the identity $\e\cdot\e = 1$. Inserting this into Eq.~\ref{eq:nlsm euclidean action} yields the discretized action
\begin{equation}
    \label{eq:nlsm discretized action}
    S_\mathrm{lat}[\e] = \sum_i \left[ 2 - \e(x+a\hat{t})\cdot\e(x) - \e(x+a\hat{x})\cdot\e(x) \right].
\end{equation}

Finally, we redefine the gradient flow in on the lattice. Since the gradient flow is solved exactly in the $\phi^4$ model, we rely on a Fast Fourier Transform. \textit{TODO: explicit expression for this} In the NLSM, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes 
\begin{equation}
    \label{eq:nsm_gradflow_disc}
    \partial_\tau \e (\tau, x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T\right) \partial^2 \e(\tau,x),
\end{equation}
 where the Laplacian operator $\partial^2$ is defined as
\begin{equation*}
    \partial^2 \e(\tau,x) = \e(\tau, x+a \hat{t}) + \e(\tau,x-a\hat t) + \e(\tau, x+a \hat{x}) + \e(\tau,x-a\hat x) - 2 d \e(t)_x.
\end{equation*}
%TODO: derive the laplacian operator

\subsection{Discretized Observables}

\section{Defining the topological charge}
On the lattice, the topological charge is nontrivial to calculate. Primarily, there are multiple possible mappings between field configurations and the integers. In this work, we use the definition found in \cite{berg1981}. To begin, we define a local topological charge density $q$, defined for each square of adjacent lattice points. This square, known as a \textit{plaquette}, is denoted $x^*$. The global charge $Q$ is the sum of all local charges:
\begin{equation}
    Q \equiv \sum_{x^*} q(x^*).
\end{equation}
As a function of $x^*$, the charge density is a function of the field on the plaquette vertices, an idea visualized in Fig.~\ref{fig:plaquette}.
\begin{figure}
    \centering
    \begin{tikzpicture}
        \draw[step=4cm,gray,thin] (-2,-2) grid (6,6);
        \draw[dotted] (0,0) -- (4,4);
        \draw[thick](0,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_1$} -- 
                    (0,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_2$} -- 
                    (4,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_3$} -- 
                    (4,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_4$} -- cycle ;
                    \draw[] (2,2) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north east]{$x^*$};

        \draw[-stealth] (0.2,1) -- (0.2,3);
        \draw[-stealth] (1,3.8) -- (3,3.8);
        \draw[-stealth] (3,3.5) -- (1,1.5);

        \draw[-stealth] (1,0.5) -- (3,2.5);
        \draw[-stealth] (3.8,3) -- (3.8,1);
        \draw[-stealth] (3,0.2) -- (1,0.2);
    \end{tikzpicture}
    \caption{\label{fig:plaquette} Visualization of plaquette $x^*$. Dotted line separated plaquette into two signed areas which are used to define the topological charge density $q(x^*)$. Arrows represent order of signed area.}
\end{figure}
In the NLSM, the field at each of these vertices is a point on the sphere $S^2$. Therefore, there is a signed area $A$ on the sphere associated with each triplet of points, as shown in Fig.~\ref{fig:sphere} (the sign changes with odd permutations of the ordering). We follow the derivation in \cite{berg1981} and split the plaquette into two triangles, as shown in Fig.~\ref{fig:plaquette}, with the ordering determining the sign. The topological charge density is defined using this signed area as
\begin{equation}
    q(x^*) = \frac{1}{4\pi} \bigg[A\Big(\e(x_1), \e(x_3), \e(x_3)\Big) + A\Big(\e(x_1), \e(x_3), \e(x_4)\Big) \bigg].
\end{equation}
This sign is defined if $A\neq 0, 2\pi$, or in other words, as long as the three points on the sphere are distinct and do not form a hemisphere. In numerical calculations, these points can be ignored. Therefore, we impose that the signed area is defined on the smallest spherical triangle, or mathematically
\begin{equation}
    -2\pi < A < 2\pi.
\end{equation}

Following \cite{berg1981}, this yields an expression for the signed area
\begin{equation}
    A(\e_1, \e_2, \e_1) = 2 \:\mathrm{arg}\Big(1+\e_1\cdot\e_2 + \e_2\cdot\e_3 + \e_3\cdot\e_1 + i\e_1 \cdot (\e_2\times\e_3)\Big).
\end{equation}
Under periodic boundary conditions, these triangles on the sphere necessarily wrap $S^2$ an integer number of times, ensuring $Q$ is an integer. In the continuum limit, the eff\ref{sec:topological charge}
\begin{figure}[h]
\centering
\begin{tikzpicture}
    \pgfdeclarelayer{nodelayer}
    \pgfdeclarelayer{edgelayer}
    \pgfdeclareradialshading{sphere4}{\pgfpoint{-0.2cm}{0.5cm}}% 
        {rgb(0cm)=(1,1,1);
        rgb(1cm)=(0.5,0.5,0.5); rgb(1.05cm)=(1,1,1)}
        %rgb(0.7cm)=(0.1,0.1,0.1); rgb(1cm)=(0.5,0.05,0); rgb(1.05cm)=(1,1,1)}
    \pgfsetlayers{nodelayer,edgelayer}
    \tikzstyle{label}=[fill=none, draw=none, shape=circle]
    \tikzstyle{point}=[inner sep=0pt, minimum size=0.2cm,fill=black, draw, shape=circle]
    %\tikzstyle{interior line}=[{Stealth[scale=1.5]}-,dotted,thick]
    \tikzstyle{interior line}=[dotted,thick]

    \tikzstyle{triangle}=[thick]
    \tikzstyle{arrow}=[->, thick]
	\begin{pgfonlayer}{nodelayer}
		\node [style=point] (4) at (0.5, 2.5) {};
		\node [style=point] (5) at (0.55, 0.625) {};
		\node [style=point] (6) at (2.25, 0.75) {};
		\node (7) at (0, 0) {};
		\node (8) at (1.25, 1.375) {};
		\node (9) at (2.425, 2.525) {};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
        \draw (0,0) circle (3cm);
        %\shade[inner color=white,outer color=lightgray] (0,0) circle (3cm);
        \shade[shading=sphere4] (0,0) circle (3cm);
		\draw [bend left=15,style=triangle] (4.center) to (6.center);
		\draw [bend left=355,style=triangle] (6.center) to (5.center);
		\draw [bend right=5,style=triangle] (5.center) to (4.center);
		\draw [style=interior line] (4.center) to (7.center);
		\draw [style=interior line] (5.center) to (7.center);
        \draw [style=interior line] (6.center) to (7.center);
        \draw [fill=gray!80] (4.center) to [bend left=15] (6.center) to [bend left=355] (5.center) to [bend right=5] cycle;

		%\draw [style=arrow] (8.center) to (9.center);

        \node [style=label, anchor=north] at (2.25, 0.75) {\small $\e(x_1)$};
        \node [style=label, anchor=north] at (0.55, 0.625) {\small $\e(x_2)$};
        \node [style=label, anchor=east] at (0.5, 2.5) {\small $\e(x_3)$};
		\node [style=point] at (4){};
		\node [style=point] at (5){};
		\node [style=point] at (6){};

        \node [style=label] at (8) {\small $A$};
	\end{pgfonlayer}
\end{tikzpicture}
\caption{\label{fig:sphere} Visualization of signed area $A$ on the sphere $S^2$ traced out by field at points $x_1$, $x_2$ and $x_3$.}
\end{figure}
In the NLSM described by Eq.~\ref{eq:nlsm discrete action}, the values of $Q$ are roughly Gaussian around $Q=0$, as shown in Fig.~\ref{fig:hist}.
\begin{figure}[h]
    \centering
      \includegraphics[width=0.6\textwidth]{imgs/hist.png}
      \caption{\label{fig:hist} Histogram of topological charge values $Q$ for trivial NLSM. $L=404$, 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization, $\tau=0$}
\end{figure}

Following this quantity, we can define a topological susceptibility $\chi_t$
\begin{equation}
\chi_t \equiv \frac{1}{L^2} \Big( \langle Q^2 \rangle - \langle Q \rangle^2 \Big).
\end{equation}
In the trivial case, $\langle Q \rangle$ is equal to $0$ and   
\begin{equation}
    \chi_t = \frac{1}{L^2} \sum_{x,y} \langle q(x)q(y)\rangle.
\end{equation}
Assuming periodic boundary conditions and therefore translational symmetry, this expression simplifies to 
\begin{equation}
    \chi_t = \frac{1}{L^2} \sum_{x} \langle q(x)q(0)\rangle.
\end{equation}

On the lattice, this quantity is known to diverge in the continuum limit \cite{bietenholz2018}. % TODO: highlight that this is true for QCD and the NLSM.

\section{Monte Carlo Simulations}
\label{sec:mc}
We implement a Markov Chain Monte Carlo method following Schaich's thesis \cite{schaich2006}. This implementation utilizes a ``random walk,'' i.e. a set of random steps through phase space, to determine statistical values such as correlation functions across the lattice. By the definition of the Markov chain, the probability of adoption of each state, and therefore its inclusion in the Monte Carlo calculation, depends only on the current state and the proposed state. This probability is denoted as $P(\mu\rightarrow\nu)$ where $\mu$ and $\nu$ are the existing and proposed lattice configurations respectively.

\subsection{Metropolis Algorithm}
We primarily use the Metropolis algorithm for the calculation of new Markov chain configurations. We begin with a so-called ``hot start,'' where each field value at each lattice site is randomly selected. Then we propose a new value for a single lattice point, which is accepted with a probability
\begin{equation}
    P(\phi_a\rightarrow\phi_b) = \begin{cases} 
        e^{-(S[\phi_b] - S_[\phi_a])} & S[\phi_b] < S_[\phi_a] \\
        1 & \mathrm{otherwise} \\
   \end{cases}
\end{equation}
where $\phi_a$ is the initial configuration and $\phi_b$ is the proposed configuration. This process is performed for each point on the lattice, making up a ``sweep.'' Repeating this sweep many times pushes the lattice toward the action minimum.

\subsection{Wolff Cluster Algorithm}

Though the Metropolis algorithm will slowly find the absolute minimum of the theory, the presence of local minima can greatly prolong the convergence. Both the $\phi^4$ model and the non-linear sigma model feature ``kinetic'' terms with gradients of $\phi$. Therefore, the presence of large similarly-valued regions in the lattice can lead to a local minimum. One method of removing these clusters involves identifying all clusters on the lattice and probabilistically flipping each, a technique known as the Swendsen-Wang algorithm\cite{swendsen1987}.

A more efficient approach is the Wolff algorithm\cite{wolff1989}, which \textit{grows} one cluster probabilistically and flips it unconditionally. In the case of $\phi^4$ theory, this flipping takes the form of a simple sign change. In the non-linear sigma model we choose a random unit vector $\vec r$ and consider the projection of the field on this vector. When the cluster flips, each site is flipped along this direction.\citeneeded To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site, growing the cluster from a single, randomly selected ``seed''. Starting with the seed, the probability of adding each neighboring site is given by the source site $x$ and the proposed site $x'$. Wolff defines this probability for arbitrary sigma models as 
\begin{equation}
    \label{eq:phi4 wolff padd}
    P_{add}(\e(x),\e(x')) = \begin{cases} 
        1 - e^{2\beta [\vec{r} \cdot \e(x)][\vec{r} \cdot \e(x')]} & \mathrm{sgn}[\vec{r}\cdot\e(x)]=\mathrm{sgn}[\vec{r}\cdot\e(x')]\\
        0 & \mathrm{otherwise} \\
   \end{cases}
\end{equation}
This expression is designed to preserve the detailed balance equations. We can demonstrate this quality, and motivate an equivalent expression for the $\phi^4$ model, by considering the probability $P(\phi\rightarrow f_C(\phi))$ of flipping some cluster $C$. Generally,
\begin{equation}
    P\big(\phi \rightarrow f_C\left(\phi\right)\big) \propto \prod_{\langle x,x'\rangle \in \partial C}  \Big[1 - P_{add}\big(\e(x),\e(x')\big)\Big]
\end{equation}
where $\partial C$ is the set of pairs of sites on the boundary of $C$. Since $P_{add}=0$ for unaligned sites, these pairs contribute nothing to the value. We can also find the probability $P(f_C(\phi) \rightarrow \phi)$ with the same expression:
\begin{equation}
P\big(f_C(\phi)\rightarrow \phi \big) \propto \prod_{\langle x,x'\rangle \in \partial C}  \Big[1 - P_{add}\big(\e(x),R\;\e(x')\big)\Big],
\end{equation}
where the matrix $R$ which is a reflection matrix along the vector $\vec{r}$.

From the discretized action of the NLSM model (Eq.~\ref{eq:nlsm discretized action}) and the detailed balance equation (Eq.~\ref{eq:detailedbalance2}), we derive
\begin{equation}
    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\e(x),\e(x')\big)}{ 1 - P_{add}\big(\e(x),R\;\e(x')\big)}  = \mathrm{exp}\left\{\beta\sum_{\langle x,x'\rangle \in \partial C} \e(x)\cdot[R-1]\e(x')\right\}.
\end{equation}
Note that all the pairs within and outside the cluster cancel in the fraction on the left and the difference on the right. Using the definition of the reflection matrix
\begin{equation}
    R\;\e = \e - 2(\e\cdot\vec{r})\e,
\end{equation}
we can simplify the equation to be 
\begin{equation}
    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\e(x),\e(x')\big)}{ 1 - P_{add}\big(\e(x),R\;\e(x')\big)}  = \prod_{\langle x,x'\rangle \in \partial C}\mathrm{exp}\big\{2\beta[r\cdot\e(x)][r\cdot\e(x')]\big\}.
\end{equation}
By plugging in Eq.~\ref{eq:phi4 wolff padd}, it is clear to see this equation is satisfied.

Using this same reasoning, we can deduce an expression for $P_{add}$ in the $\phi^4$ model. Since this model is one dimensional, the reflection matrix $R$ is equivalent to $-1$. Adapting the NLSM detailed balance equation for the $\phi$ field, we find
\begin{equation}
    \prod_{\langle x,x'\rangle \in \partial C}  \frac{ 1 - P_{add}\big(\phi(x),\phi(x')\big)}{ 1 - P_{add}\big(\phi(x),-\phi(x')\big)}  = \prod_{\langle x,x'\rangle \in \partial C}\mathrm{exp}\{-2 \phi(x)\phi(x')\}.
\end{equation}
This equation is satisfied by the ansatz
\begin{equation}
    \label{eq:nlsm wolff padd}
    P_{add}(\phi(x),\phi(x')) = \begin{cases} 
        1 - e^{-2\phi(x)\phi(x')]} & \mathrm{sgn}[\phi(x)]=\mathrm{sgn}[\phi(x')]\\
        0 & \mathrm{otherwise} \\
   \end{cases}.
\end{equation}
We will use this expression in the computational implementation of this algorithm.

Fig.~\ref{fig:wolff} shows a real demonstration of this process. We can see a large cluster of negative field values becoming positive. Note that periodic boundary conditions apply, so the small island of black at the bottom is actually part of the larger cluster. Furthermore, this visualization demonstrates the probabilistic nature of the Wolff algorithm. Since states are added probabilistically, there are some small holes in the cluster. These will be removed by following Metropolis sweeps.
\begin{figure}[h]
  \centering
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=0.6\textwidth]{imgs/wolffa.png}
        \caption{before cluster flip}
      \end{subfigure}%
      \hfill
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=0.6\textwidth]{imgs/wolffb.png}
        \caption{after cluster flip}
      \end{subfigure}
      \hfill
      \caption{\label{fig:wolff} An example of the Wolff cluster algorithm in the $\phi^4$ model. White represents positive values of $\phi$ while black represents negative. $\lambda=0.5$, $m_0^2=-0.9$}

\end{figure}

\subsection{Checkerboard algorithm}

In order to parallelize the Metropolis algorithm, we use a checkerboard algorithm. Since the Lagrangian density at each site does not depend on any sites diagonal to it, the lattice can be split into ``white'' sites and ``black'' sites, like the tiles on a checkerboard. Each white site is independent of every other white site and likewise with black sites. Therefore, we can split the sites of each color into separate parallel processing nodes and independently run the Metropolis algorithm, ensuring that no site affects the Lagrangian density at any other site. We use this method to parallelize the code through the Message Passing Interface (MPI).

\subsection{Thermalization}
\label{sec:thermalization}
Since we begin with a random lattice, the first few configurations of our Monte Carlo simulation will be far from the lattice minimum. We plot the action as a function of metropolis sweeps in Fig.~\ref{fig:therm}.
\begin{figure}[h]
  \centering
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=0.9\textwidth]{imgs/therm24.png}
        \caption{$L=24$}
      \end{subfigure}%
      \hfill
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=0.9\textwidth]{imgs/therm404.png}
        \caption{$L=404$}
      \end{subfigure}
      \hfill
      \caption{\label{fig:therm} Plots of the action as a function of Monte Carlo time, starting with a random NLSM lattice.}
\end{figure}
Based on this plot, we determine that 1000 sweeps will give sufficient time for the system to reach the classical action minimum. We use this value for the remainder of this study.
\subsection{Autocorrelation}
One important aspect is the correlation between different states of the Markov chain. Ideally, each sample of the lattice would be completely independent, but this is not the case. Since each configuration is based on previous configurations, there is a correlation between each pair of members in the Markov chain, decreasing exponentially based on the number of steps between. Explicitly, the autocorrelation scales as $e^{t/\tau_{int}}$ where $t$ is the number of steps between configurations and $\tau_{int}$ is the autocorrelation time.\footnote{Though it is called a time, $\tau_{int}$ is in units of Markov Chain steps.} When performing simulations of the lattice, the number of sweeps between measurements should be much larger than $\tau_{int}$.

    We use Wolff's automatic windowing procedure \cite{wolff2007} and the magnetic susceptibility $\chi_m$ to estimate the autocorrelation. Using Wolff's public MatLab code\footnote{\url{https://www.physik.hu-berlin.de/de/com/ALPHAsoft}}, we estimate the autocorrelation time for $L=24$ and $L=404$ lattices. This algorithm identifies the optimal window size with which to calculate the autocorrelation time. We perform this process on $L=24$ and $L=404$ lattices using a thermalization of $1000$ sweeps and $500$ total measurements. Note that this calculation included a Wolff cluster algorithm every $5$ sweeps. The result of this calculation is shown in Fig.~\ref{fig:tauint}.
\begin{figure}[h]
  \centering
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=0.9\textwidth]{imgs/tauint24.png}
        \caption{$L=24$, $\tau_{int}=4.43$}
      \end{subfigure}%
      \hfill
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=0.9\textwidth]{imgs/tauint404.png}
        \caption{$L=404$, $\tau_{int}=12.81$}
      \end{subfigure}
      \hfill
      \caption{\label{fig:tauint} Plots of automatic windowing procedure used to calculate $\tau_{int}$ for the NLSM model.}
\end{figure}

Based on these two values for $\tau_{int}$, we decide to measure every $50$ sweeps for each simulation. This value will ensure that each measurement is effectively independent.

\subsection{Runge-Kutta Algorithm}
In order to calculate the gradient flow, we numerically solve the differential equation using a fourth-order Runga-Kutta approximation. This algorithm refines the Euler method 
\begin{equation*}
    \e(\tau+h)_x \approx \e(\tau)_x + h f(\e(\tau)_x).
\end{equation*}
where $f(\e)$ is defined for convenience as 
\begin{equation}
    f(\e)=\partial_\tau \e (\tau)_x  = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \partial^2 \e(\tau)_x,
\end{equation}
following from Eq.~\ref{eq:nsm_gradflow_disc}. To the fourth order, this approximation becomes 
%
\begin{align}
    \label{eq:rungekutta}
    k_1 &= h f\left(\e\left(\tau\right)_x\right) \\ 
    k_2 &= h f\left(\e\left(\tau\right)_x + \frac{k_1}{2}\right) \\ 
    k_3 &= h f\left(\e\left(\tau\right)_x + \frac{k_2}{2}\right) \\ 
    k_4 &= h f\left(\e\left(\tau\right)_x + k_3\right) \\ 
    \e(\tau+h)_x &= \e(\tau)_x + \frac{k_1}{6} + \frac{k_2}{3} + \frac{k_3}{3} + \frac{k_4}{6} + O(h^5).
\end{align}
This method is usually superior to Euler's method and the midpoint method \cite{vetterling1992}.

To increase the efficiency of this algorithm, we implement the step-doubling algorithm to adaptively adjust $h$. If the error of a Runge-Kutta step is greater than the tolerance, the same step is repeated with half the step size. Alternatively, if the error is less than half of the tolerance, the step size is doubled for the next calculation. Finally, if the step size is greater than the distance to the next measurement, that distance is used as the step size, using the normal value afterwards. Otherwise, the algorithm proceeds with the consistent step size.

\section{Topological charge with a $\theta$ term}
\label{sec:topotheta}
In Sec.~\ref{sec:topological charge}, we discussed the introduction of a $\theta$ term into the action. This change made the theory ``topological'', pushing $\langle Q \rangle$ away from zero. In order to calculate $\langle Q \rangle$ as a function of $Q$, we consider the path integral
\begin{align}
    \langle Q \rangle_\theta &=\int \mathcal{D}\e\:Q[\e]e^{-S[\e]+i\theta Q[\e]} \\
                             &=\int \mathcal{D}\e\:\left( Q[\e]e^{i\theta Q[\e]} \right) e^{-S[\e]} \\
                             &=\langle Q e^{i \theta Q} \rangle_{\theta=0}.
\end{align}
Therefore, we can calculate $\langle Q \rangle_\theta$ for arbitrary $\theta$ using the same simulation framework as the $\theta=0$ case.

We can also relate this function to the topological susceptibility. By expanding the exponent as a Taylor series around $\theta=0$, we find that  
\begin{equation}
    \langle Q \rangle_\theta = \langle Q \rangle_{\theta=0} + i \theta \langle Q^2 \rangle_{\theta=0} + O(\theta^2),
\end{equation}
such that 
\begin{align}
    -i \frac{\partial}{\partial \theta}\Big|_{\theta=0} \langle Q \rangle_\theta &= \langle Q^2 \rangle_{\theta=0}\\
                                                              &= \chi_t L^2.
\end{align}
Since $\chi_t$ diverges at $\theta=0$, we expect the plot of $\langle Q\rangle_\theta$ to approach a vertical line in the in continuum limit.

### 2021-04-19 final-report.latex/results.tex
split into chapter files

We initially implement the $\phi^4$ model to verify the results of our system. According to previous studies \cite{monahan2016, schaich2006}, the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ when $\lambda = 0.5$. We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the magnetic susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
\begin{figure}[h]
    \centering
      \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
      \caption{\label{fig:phi4} The lattice average $|\langle\bar\phi\rangle|$, the lattice susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$. $L=32$, $\lambda=0.5$. The lattice was thermalized from a hot start for 1000 sweeps. Afterwards, 100 measurements were taken with 100 sweeps between each. The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
\end{figure}
This figure confirms a phase transition near $m_0^2=-0.72$ and supports the accuracy of our model.

\section{Non-linear sigma model results}
After confirming the phase transition in the $\phi^4$ model, we turn to the NLSM. In order to confirm the accuracy of the model, we compare results with existing literature. We first compare our results to Berg \& L\"uscher \cite{berg1981}, specifically measuring the internal energy and magnetic susceptibility.

\subsection{Comparison with Existing Literature}
Following \cite{berg1981}, we approximate the internal energy in the strong ($\beta<1$) and weak ($\beta>2$) regimes as 
\begin{equation}
    \label{eq:analytic_energy}
E \approx \begin{cases} 
    4-4y-8y^3-\frac{48}{5}y^5& \beta<1 \\
    \frac{2}{\beta} + \frac{4\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2. 
\end{cases}
\end{equation}
We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{fig:bergluscher}.
\begin{figure}[h]
    \centering
      \includegraphics[width=0.6\textwidth]{imgs/internal_energy.png}
      \caption{\label{fig:bergluscher}Comparison with \cite{berg1981}. First panel: internal energy compared with analytic  energy (Eq.~\ref{eq:analytic_energy}). Second panel: magnetic susceptibility compared with literature values.}
\end{figure}
These two charts show a high degree of agreement with the current literature.

We also seek to confirm the results from Bietenholz et al.\cite{bietenholz2018}. Specifically, that the topological susceptibility $\chi_t$ diverges in the continuum limit even at finite flow time. Since $\chi_t$ is in units of inverse distance squared, we multiply by $\xi_2^2$, the square of the second moment correlation length, to achieve a scale-invariant value $\chi_t\xi_2^2$. Additionally, we use a parameter $t_0$ to scale the flow time such that $t_0\sim L^2$. In our Monte Carlo simulation, we adapt the same values as \cite{bietenholz2018} for $\xi_2$.

To begin the comparison, we plot $\chi_t\xi_2^2$ as a function of flow time $\tau$, shown in Fig.~\ref{fig:bietenholz}.
\begin{figure}[h]
    \centering
      \includegraphics[width=\textwidth]{imgs/bietenholz.png}
      \caption{\label{fig:bietenholz} $\chi_t L^2$ as a  function of flow time $\tau$. Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization.}
\end{figure}
We find that the flow time effectively decreases the topological susceptibility by dampening high-momentum modes. To analyze the divergence of $\chi_t$ in the continuum limit, we plot $\chi_t \xi_2^2$ as a function of lattice size $L$. We perform this simulation at flow time $\tau=5t_0$.
\begin{figure}[h]
    \begin{center}
      \begin{subfigure}[b]{\textwidth}
          \includegraphics[width=\textwidth]{imgs/divergence.png}
          \caption{$\tau = 0$}
      \end{subfigure}

      \begin{subfigure}[b]{\textwidth}
          \includegraphics[width=\textwidth]{imgs/divergence_flowed.png}
          \caption{$\tau = 5t_0$}
      \end{subfigure}
      \caption{\label{fig:bietenholz} $\chi_t\xi_2^2$ as a function of $L$. The data is fit with both a log and power fit. Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization. In the $\tau=0$ case, we have compared our result with the curve fit found in \cite{bietenholz2018}.}
    \end{center}
\end{figure}
The data is fit with two options: a log fit
\begin{equation}
    \chi_t \xi_2^2 = A \mathrm{log}(B L) + C
\end{equation}
and a power fit
\begin{equation}
    \chi_t \xi_2^2 = A B^L + C.
\end{equation}
The data fits these functions with $\chi^2/DOF$ of $31.3$ and $761.6$ respectively. Both of these function diverge as $L\rightarrow \infty$, indicating that the topological susceptibility does as well. This is still a slower divergence than the linear divergence featured in the $\tau=0$ case.

\subsection{Topological charge when $\theta\neq 0$}
Following the method explained in Sec.~\ref{sec:topotheta}, we calculate the imaginary part of $\langle Q \rangle$ for arbitrary $\theta$. We perform this calculation for four values of the flow time $\tau$, shown in Fig~\ref{fig:theta}.
\begin{figure}[h]
    \centering
      \includegraphics[width=\textwidth]{imgs/theta.png}
      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$. Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization. Note the different scaling of the $y$-axis.}
\end{figure}
These plots demonstrate the divergence of the continuum limit in the $\tau=0$ and the flowed regimes. In the $\tau=0$ case, the slope increases sharply, reflecting the linear divergence of $\chi_t$. However in the flowed regime, this divergence is much slower, reflecting the logarithmic or power-law relationship shown in Fig.~\ref{fig:bietenholz}.

%\chapter{Statistical Analysis}

### 2021-04-19 final-report.latex/theory.tex
split into chapter files

This thesis incorporates two main bodies of knowledge: quantum field theory and statistical simulation. Through the path integral formulation of quantum field theory, we are able to describe the physics of the former with the established mathematics of the latter.

\section{Quantum Field Theory}

In this section we outline a rough description of quantum field theory. A full introduction is beyond the scope of this paper, however we do assume knowledge of nonrelativistic quantum mechanics and classical field theory.

The fundamental hypothesis of quantum field theory (QFT) describes particles as discrete packets of energy on a quantum field. But what is a quantum field? Like in classical mechanics, a field is a function of spacetime with some mathematical object assigned to each point in space and time. In the case of the electric field, this object is a three-dimensional vector, while the electric potential is a scalar field. Classical and quantum fields have Lagrangians which define how they evolve in space and time. What differentiates a quantum field from a classical field is superposition: where classical fields have a definite configuration, quantum fields exist in a superposition of all possible configurations. It is possible---though nontrivial and outside the scope of this description--- to motivate the appearance of discrete particles from this superposition \cite{zee2010}. This formulation of QFT is known as the ``path integral formulation'', which differs from the ``second quantization'' used by most textbooks.

This general description allows QFT to easily incorporate special relativity. By ensuring that the Lagrangian of a theory is invariant under Lorentz transformations, we can ensure that the theory itself is Lorentz invariant. This is a necessary condition of physical theories.

\subsection{Path Integral Formulation}
\label{sec:pathintegral}

We can model a quantum field theory as a superposition of all possible classical fields. Like single-particle quantum mechanics, each configuration has a probability amplitude. To measure expectation values of observables, we simply take an average over all configurations weighted by this complex amplitude. We can formalize this notion using the fundamental formula\footnote{Since this study concerns vacua, we do not include a source term.}
\begin{equation}
    \label{eq:pathintegral}
    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; e^{iS[\phi]}
\end{equation}
% not sure if the semicolons is good form
where $\langle \hat O \rangle$ is the expectation value of an arbitrary operator $\hat O$; $Z$ is a normalization constant; $S$ is the action functional, defined from the theory's Lagrangian; and $\int \mathcal{D}\phi$ represents the eponymous path integral. Though it is possible to define this integral rigorously, for our purposes we can equate it to a sum over all possible configurations. This form is the quantum analog of the classical principle of least action and reduces to such for large values of the action. For a more pedagogical explanation, see Richard Feynman's lectures on physics \cite{feynman1963a}.

At first glance, Eq.~\ref{eq:pathintegral} is remarkably similar to the statistics of the canonical ensemble. Through this similarity, we will be able to use mathematical tools from statistical mechanics to study quantum field theories. However, the factor of $i$ in the exponent currently prohibits us from making this jump. To remedy this issue, we perform a ``Wick rotation'' which shifts spacetime into Euclidean coordinates. In normal spacetime, defined by the Minkowski metric, the Lorentz-invariant distance is given as
\begin{equation}
    s^2 = x^2_0 - x^2_1- x^2_3- x^2_3
\end{equation}
where $x_0=ct$ and $\vec{x} = (x_1, x_2, x_3)^T$. By redefining the time coordinate of a spacetime point $x$ to be $x_4=ix_0$, we find that the quantity
\begin{equation}
    s_E^2 = x^2_1+ x^2_3+ x^2_3 + x^2_4,
\end{equation}
is therefore invariant under $SO(4)$ transformations, which is representative of a four-dimensional Euclidean space. Furthermore, we find that
\begin{align}
    d^4x_E &= d^3\vec{x}dx_4\nonumber \\
    &= i d^3\vec{x}dx_0 \nonumber \\
    &= i d^4x. \label{eq:wickdifferential} % TODO: is this better?
\end{align}

We can use this transformation to redefine the Lagrangian $\mathcal{L}$ in Euclidean space as $\mathcal{L}_E$, replacing all $x_0$ with $ix_0$. Since a Lorentz-invariant Lagrangian must only include even powers and derivatives of $x$, the Euclidean Lagrangian remains real. Subsequently, we can define a Euclidean action based on the differential in Eq.~\ref{eq:wickdifferential}:
\begin{align}
    S_E &= \int d^4x_E \mathcal{L}_E \\
    &= i \int d^4x \mathcal{L}_E,
\end{align}
allowing us to redefine the path integral as 
\begin{equation}
    \label{eq:pathintegraleuclidean}
    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O (\phi)\; e^{-S_E[\phi]}.
\end{equation}

By replacing the Minkowski action $S$ with a Euclidean action $S_E$, we have transformed the amplitude $e^{iS}$ to a statistical Boltzmann factor $e^{-S_E}$. This new form will allow us to use statistical techniques to simulate quantum fields.

\subsection{$\phi^4$ model}
\label{sec:phi4}

One of the simplest interacting field theories is known as the $\phi^4$ model. This theory describes a spin-0 boson and consists of a real scalar field given by the four-dimensional Minkowski action
\begin{equation}
    \label{eq:phi4 action}
    S[\phi] = \int d^4 x \left[ \frac{1}{2}\partial^\mu \phi \partial_\mu\phi - \frac{1}{2} m_0^2 \phi^2 - \frac{\lambda}{4}\phi^4\right].
\end{equation}
The first two terms describe a free relativistic particle of mass $m_0$ while the last term describes an interaction with strength $\lambda$. Per Einstein summation notation, there is an implicit sum over spacetime dimensions $\mu\in\{0,1,2,3\}$ indexing the derivative vectors\footnote{This canonical representation of the kinetic term $\frac{1}{2}\partial^\mu\phi\partial_\mu\phi$ is equivalent to $\frac{1}{2}\dot\phi^2-\frac{1}{2}\left(\nabla \phi\right)^2$.}
\begin{align}
    \partial^\mu &= \left( \frac{\partial}{\partial t}, \frac{\partial}{\partial x},\frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) \\
    \partial_\mu &= \left( \frac{\partial}{\partial t}, -\frac{\partial}{\partial x}, -\frac{\partial}{\partial y}, -\frac{\partial}{\partial z} \right).
\end{align}

In this study, we specifically consider fields in 1+1 spacetime dimensions. Following Sec.~\ref{sec:pathintegral}, we convert Eq.~\ref{eq:phi4 action} from 3+1 Minkowski spacetime to 1+1 Euclidean spacetime, yielding
\begin{equation}
    \label{eq:phi4 euclidean action}
    S_E[\phi] = \int d^2 x_E \left[\frac{1}{2}\left(\partial_t \phi\right)^2 + \frac{1}{2} \left(\partial_x \phi \right)^2 + \frac{1}{2} m_0^2 \phi^2 + \frac{\lambda}{4}\phi^4\right].
\end{equation}
where $\partial_t$ and $\partial_x$ are the two partial derivatives in 1+1 Euclidean spacetime.

This field features spontaneous symmetry breaking at a critical value of $m_0^2$. In practice this property causes the field to spontaneously align, similarly to spins aligning in a ferromagnet. The name ``symmetry breaking'' refers to the transformation $\phi\rightarrow-\phi$, which changes the values of observables in the aligned regime but not the disordered regime. These two phases are known as the ``broken'' and ``symmetric'' phases and their transition is well understood.

\subsection{Non-linear sigma model}

The non-linear sigma model (NLSM) is a prototypical theory for a variety of unrelated physical phenomena, including applications in string theory \cite{callan1985} and ferromagnetism \cite{polyakov1975}. As a simple nonperturbative model, it also provides an ideal starting point for lattice QCD studies. Specifically, the NLSM exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.

Unlike the $\phi^4$ model, which consists of a real value at each point in spacetime, the $O(3)$ NLSM consists of a 3D unit vector at each point. For this reason, every transformation of the field must be ``norm preserving''. Per its name, the $O(3)$ NLSM features a global symmetry under the 3D orthogonal group $O(3)$. In other words, the theory remains the same if all vectors are rotated equivalently. To differentiate it from the $\phi^4$ model, we denote the NLSM field as $\e(x)$.

The theory is defined by the Euclidean action 
\begin{equation}
    \label{eq:nlsm euclidean action}
    S_E = \frac{\beta}{2} \int d^Dx \; \partial^\mu \e \cdot \partial_\mu \e
\end{equation}
subject to the constraint that $\e\cdot\e = 1$. Here, $\beta$ is the inverse coupling and $D$ is the number of spacetime dimensions.

\section{Markov Chain Monte Carlo}

To accomplish a statistical analysis of quantum fields, we use a Monte Carlo simulation. This method amounts to producing a large number of configurations and calculating statistics on the sample. A brute-force calculation over all possible configurations---as Eq.~\ref{eq:pathintegraleuclidean} suggests--- is clearly infinite and computationally infeasible. However, the exponential nature of the Boltzmann factor dictates that only configurations near the action minima contribute to observable statistics. Therefore, by selecting a sample of configurations near this minimum, we are able to extract meaningful results with a finite computation.

\subsection{The Markov Chain}
The Markov chain is an effective method to identity field configurations that minimize the action. Essentially, we begin with a predetermined configuration and then make small adjustments, gradually lowering the action. By measuring states after a certain amount of time has passed, called the ``thermalization'', we can form a set of configurations near the action minima and approximate the observables of the system.

Each step consists of two parts: proposing a change and accepting the new configuration. The proposal creates a new configuration $\phi_b$ based on the current configuration $\phi_a$. In order to direct the Markov chain towards the action minima, we do not immediately accept this change. We calculate a probability
\begin{equation}
P(\phi_a \rightarrow \phi_b)
\end{equation}
of adding $\phi_b$ to the Markov chain, depending only on the actions of each configuration. This stochastic process ensures the chain will seek action minima.

There are four requirements that this function must obey to produce a Boltzmann distribution of samples:
\begin{enumerate}
    \item $P(\phi_a \rightarrow \phi_b)$ must depend only on the configurations $\phi_a$ and $\phi_b$.
    \item The probability must be properly normalized, i.e. $\sum_{\phi} P(\phi_a \rightarrow \phi) = 1$.
    \item Every configuration must be reachable in a finite number of steps. In other words, the chain must be ergodic.
    \item In order to reach equilibrium, the chain must be reversible. In other words, the probability of a $\phi_a\rightarrow\phi_b$ transition must be equal to the probability of a $\phi_b\rightarrow\phi_a$ transition. Mathematically, this condition takes the form of \textit{detailed balance equations}:
\begin{equation}
    \label{eq:detailedbalance}
    P(\phi_a)\,P(\phi_a\rightarrow\phi_b) = P(\phi_b)\,P(\phi_b\rightarrow\phi_a),
\end{equation}
where $P(\phi)$ is the probability of a system existing in state $\phi$.
% this needs citation
\end{enumerate}

This final condition will allow us to explicitly define the transition probability using the action. From the Boltzmann distribution, we know
\begin{equation}
    P(\phi) = \frac{1}{Z} e^{-S_E[\phi]}.
\end{equation}
Therefore, by rearranging Eq.~\ref{eq:detailedbalance}, we find
\begin{equation}
    \label{eq:detailedbalance2}
    \frac{P(\phi_a\rightarrow\phi_b)}{P(\phi_b\rightarrow\phi_a)} = e^{S_E[\phi_a] - S_E[\phi_b]}.
\end{equation}

This formula will provide the explicit transition probabilities for the Metropolis and Wolff algorithms.

\section{Observables}

To extract physics from Monte Carlo simulations, we define a set of ``observables''. These quantities manifest as expectation values of operators, calculated using the Euclidean path integral formula (Eq.~\ref{eq:pathintegraleuclidean}). We can classify these observables into two categories: primary and secondary observables. Primary observables are calculated as expectation values of global operators while secondary observables are derived from these quantities.

\subsection{Primary Observables}
Each primary observable is defined on each configuration independently, meaning they do not encode ensemble statistics of the Markov Chain.\footnote{One slight exception is the magnetic susceptibility, which can be defined both ways.} In the $\phi^4$ model, we can develop an intuition around these quantities by visualizing the symmetric and broken phases. Fig~\ref{fig:primary observables} and Tab.~\ref{tab:primary observables} show examples of these quantities in three different configurations: one in the broken phase, one in the symmetric phase, and one at the transition.

There are two potential points of confusion here. The first lies in the definition of ``broken'' phase. Though the symmetric phase more closely resembles a pane of broken glass, it leaves $\phi\rightarrow-\phi$ symmetry \textit{un}-broken, thereby giving the title ``broken'' to the more visually uniform configuration. An additional potential pitfall is the distinction between the \textit{lattice average} and the \textit{ensemble average}. The first is a mean over all lattice sites while the second is a mean over all configurations in the Markov Chain.

\begin{figure}[h]
    \begin{center}
      \begin{subfigure}[b]{0.3\textwidth}\centering
        \includegraphics[width=0.6\textwidth]{imgs/broken.png}
        \caption{broken phase ($m_0^2=-1.0$)}
      \end{subfigure}%
      \hfill
      \begin{subfigure}[b]{0.3\textwidth}\centering
        \includegraphics[width=0.6\textwidth]{imgs/transition.png}
        \caption{transition ($m_0^2=-0.7$)}
      \end{subfigure}%
      \hfill
      \begin{subfigure}[b]{0.3\textwidth}\centering
        \includegraphics[width=0.6\textwidth]{imgs/symmetric.png}
        \caption{symmetric phase ($m_0^2=-0.4$)}
      \end{subfigure}
      \hfill
      \caption{\label{fig:primary observables} Visualization of broken phase, symmetric phase and transition. Simulation run on $64\times64$ lattice, plotted after 1000 sweep thermalization (see Sec~\ref{sec:thermalization})}
  \end{center}
\end{figure}

\begin{table}[h]
    \begin{center}
    {\renewcommand{\arraystretch}{1.2} %
    \begin{tabular}{c c c c}
        \hline\hline & broken & transition & symmetric \\ \hline
        $|\bar\phi|$ & 0.56 & 0.07 & 0.02 \\ 
        $S_E/L^2$ & 0.29 & 0.40 & 0.44  \\ \hline\hline
    \end{tabular}}

    \end{center}
    \caption{\label{tab:primary observables} Average magnetization with average action per site corresponding to the particular configurations in Fig.~\ref{fig:primary observables}.}
\subsubsection{Average Magnetization}
\label{sec:avg mag}
The average magnetization quantifies the total alignment of the field. In both the $\phi^4$ model and the NLSM, a value of zero indicates the symmetric phase while a nonzero value indicates broken symmetry. In the NLSM, a magnitude of one represents total alignment.

Due to the $\phi\rightarrow-\phi$ symmetry of the $\phi^4$ model, the ensemble mean of the average magnetization $\langle \phi \rangle$ is $0$. Likewise, the $O(3)$ symmetry in the NLSM enforces $\langle \e \rangle$ = 0. To measure the alignment, we therefore use the magnitude of this quantity, defined in the $\phi^4$ model as 
\begin{equation}
|\bar\phi| \equiv \frac{1}{L^2}\left| \sum_i^{L^2} \phi(x_i)\right|.
\end{equation}
and in the NLSM as
\begin{equation}
    |\bar\e| \equiv \frac{1}{L^2}\left|\sum_i^{L^2} \e(x_i)\right|.
\end{equation}

In the symmetric phase, both $\langle \bar\e\, \rangle$ and $\langle \bar\phi \rangle = 0$ while in the broken phase they are both nonzero.

\subsubsection{Internal Energy}
The internal energy is defined as \cite{berg1981}
\begin{equation}
    E = \frac{2}{\beta L^2} \langle S \rangle
\end{equation}
in the NLSM.

\subsection{Secondary Observables}
Unlike primary observables, secondary observables are defined for each \textit{ensemble}, not each configuration. We define three secondary observables: the magnetic susceptibility, the Binder cumulant and the bimodality. Since the Binder cumulant and the bimodality provide different perspectives on phase transitions, we will restrict our usage of these observables to the $\phi^4$ model.

\subsubsection{Magnetic Susceptibility}
Though the magnitude of the average magnetization is the main phase transition indicator, its behaviour on the lattice makes the critical point difficult to identify. With a finite lattice spacing $a$, the transition becomes smoother around the critical point. An alternative metric is the magnetic susceptibility. The magnetic susceptibility is proportional to the variance of the magnetization and features a peak at the critical point. This peak is more identifiable than the smooth transition of the magnetization.

Mathematically, this value is defined as 
\begin{equation}
    \chi_m \equiv L^2 \big(\langle {\bar\e\,}^2 \rangle - \langle \bar\e\, \rangle^2\big)
\end{equation}
in the NLSM and
\begin{equation}
    \chi_m \equiv L^2 \big(\langle {\bar\phi}^2 \rangle - \langle \bar\phi \rangle^2\big)
\end{equation}
in the $\phi^4$ model. Manifestly, this expression appears as a secondary observable since it is not defined for each lattice configuration. However, due to the rotation invariance of both fields, the second term disappears (see Sec.~\ref{sec:avg mag}), yielding
\begin{align}
    \chi_m &= L^2 \left\langle \left( \sum_x e(x) \right)^2\right\rangle  \\
           &= L^2 \left\langle \sum_{x,y}\e(x)\cdot \e(y) \right\rangle. 
\end{align}
Due to translational invariance from the periodic boundary conditions, this expression becomes
\begin{equation}
    \chi_m = \left\langle \sum_x \e(0)\cdot\e(x) \right\rangle
\end{equation}
for the NLSM. Following the same logic, we derive 
\begin{equation}
    \chi_m = \left\langle \sum_x \phi(0)\phi(x) \right\rangle
\end{equation}
for the $\phi^4$ model.

\subsubsection{Binder Cumulant}
We define the Binder cumulant $U$ as \cite{binder1981}
\begin{equation}
    U \equiv 1-\frac{\langle \bar{\phi^4} \rangle}{3\langle \bar{\phi^2}\rangle^2}.
\end{equation}
Similar to the magnitude of the average magnetization, this formula yields $0$ in the symmetric phase and a nonzero value in the broken phase. This nonzero value is $U = 2/3$ for the Binder cumulant. The advantage of this metric is a fixed point with regards to a scaling transformation that corresponds with the critical point of the phase transition \cite{landau2000}. In our study, we use the Binder cumulant as further evidence of a phase transition.

\subsection{Bimodality}
The final phase transition indicator we use is the bimodality. In the symmetric phase, the average magnetization $\bar\phi$ centers around $0$ while in the broken phase, these values cluster around two peaks. Qualitatively, this metric measures the separation of these peaks.

We begin by measuring $\bar\phi$ for each configuration. We separate each value into an odd number number of bins, ensuring that there is a bin centered at $\bar\phi=0$. We then calculate the number of configurations $n_0$ in the center bin and the number of configurations $n_{max}$ in the fullest bin. The bimodality is then calculated as
\begin{equation}
    B = 1 - \frac{n_0}{n_{max}}.
\end{equation}
When the configurations are centered around $\bar\phi=0$, i.e. in the symmetric phase, this value is $B=0$. When the configurations are aligned such that $\bar\phi\neq0$, i.e. in the broken phase, this value becomes $B=1$.
\subsection{Jackknife Method}
Though the uncertainties of primary observables are simple to calculate, this process is more complicated for secondary observables. While we could propagate the uncertainty of the Binder cumulant, such a process is not clear for the bimodality. Therefore, we utilize a method known as Jackknife resampling.

We begin by calculating some observable $O$ on an ensemble of $N$ configurations. Then, for each configuration $i$, we calculate the same observable but exclude said configuration. This leaves us with a set of $N$ observables $O_i$. Assuming independent measurements, we can calculate the variance of $O$ as
\begin{equation}
    \mathrm{Var}(O) = \sum_{i} \left(C_i - C\right)^2.
\end{equation}
We use this formula to calculate all uncertainties in this study.
\section{Topological Observables}
\label{sec:topological charge}
The $O(3)$ NLSM features topological features originating from two properties: 
\begin{enumerate}
    \item At $x\rightarrow\infty$, the field must become uniform since the Lagrangian must vanish. This allows us to model $x\rightarrow\infty$ as a single point on the field, forming a Riemann sphere in three dimensions.
    \item The elements of the $O(3)$ non-linear sigma model are three-dimensional unit vectors, thereby existing on a three dimensional unit sphere. 
\end{enumerate}

With these two properties, we can view the field as a continuous mapping between two 3D spheres, denoted as $S^2$, and associate an integer number of wrappings to each mapping from $S^2$ to $S^2$. We can envision a tangible metaphor for this wrapping with a balloon and a baseball: by simply inserting the baseball into the balloon, we have established a mapping from every point on the balloon to every point on the baseball. We can create an equally valid map by twisting the balloon's mouth and wrapping the baseball again. In a purely mathematical world, we perform this process an infinite number of times, thereby associating every possible mapping with an integer. The group of integers is known as the \textit{homotopy group} of the non-linear sigma model. We associate every field configuration with an element of this group, known as the \textit{topological charge}, which we denote as $Q$. Configurations with $|Q|=1$ are known as instantons.

Following this quantity, we can define a topological susceptibility $\chi_t$
\begin{equation}
\chi_t \equiv \frac{1}{L^2} \Big( \langle Q^2 \rangle - \langle Q \rangle^2 \Big).
\end{equation}
In the trivial case, $\langle Q \rangle$ disappears and   
\begin{equation}
    \chi_t = \frac{\langle Q^2 \rangle - \langle}{L^2}.
\end{equation}
% TODO Highlight this is true for QCD and NLSM. Do you know why?

\subsection{NLSM $\theta$ term}
In the NLSM, we expect $\langle Q \rangle=0$. By introducing a $\theta$ term into the action,
\begin{equation}
    S[\e] \rightarrow S[\e] - i \theta Q[\phi],
\end{equation}
we can construct a topologically nontrivial model (that is, $\langle Q \rangle \neq 0$).

\section{Ultraviolet Divergences}

In Section~\ref{sec:pathintegral},  we defined a fundamental equation of quantum fields using a ``path integral'' which encompasses an uncountably infinite configuration space. However, we said nothing of the integral's convergence. In fact, many fundamental processes in QFT have divergent amplitudes, yielding nonsensical results. The most common type of divergence stems from high-momentum states, giving them the name ``ultraviolet divergences''. The remedy to this catastrophe is unintuitive. Essentially, we adopt infinite values for the parameters of the Lagrangian ($m_0^2$ and $\lambda$ in $\phi^4$ theory). Since neither of these two quantities is ever measured directly, we do not have to assume that their values are finite. In practice, this technique is involved and consists of two steps: regularization and renormalization.

\subsection{Regularization: fields on the lattice}
Regularization is a process which introduces a new parameter into calculations. One example is a momentum cutoff. This technique transforms infinite momentum integrals as follows:  
\begin{equation*}
    \int_0^\infty dk \rightarrow \int_0^\Lambda dk,
\end{equation*}
introducing $\Lambda$ as a regularization parameter. This process makes results $\Lambda$-dependent, but finite. Another example is dimensional regularization, which calculates results in terms of the spacetime dimension $D$ and analytically continues this parameter from just the integers into the real numbers.

In this study, we employ lattice regularization. This process discretizes the field, modeling the field $\phi(x)$ as a lattice $\phi_i$ where $i$ indexes lattice sites. The inherent parameter in this case is the lattice spacing $a$ which measures the width of each lattice chunk.

% TODO: Mention what was summer work

\subsection{Renormalization}
After regularization, we redefine the Lagrangian parameters in terms of the regularization parameter, following a handful of boundary conditions. In this study, we assert that $L/\xi$ remains constant, where $L$ is the side length of the system and $\xi$ is the coherence length. In perturbation theory, this process is arduous and includes the introduction of counter-terms into the Lagrangian. In the case of the non-linear sigma model, it is impossible using counterterms\citeneeded but can be performed numerically. In this study, we use predetermined values from \cite{bietenholz2018}.

At this point, we can calculate observables as functions of regularization parameters. To achieve physical values, we take the limit as the regularization parameters approach their physical values. With a momentum cutoff, we take $\Lambda \rightarrow \infty$ and with dimensional regularization we usually take $d\rightarrow 4$. With lattice regularization, we approach the continuum, taking the lattice spacing $a\rightarrow 0$.

At this point, we have surely eliminated all divergences, right? Unfortunately, this is not always the case. The topological susceptibility $\chi_t$ is one such value that diverges in the continuum limit. As we decrease the width of each lattice site, high frequency modes become more significant, leading to an ultraviolet divergence in the operator.

\subsection{The Gradient Flow}
\label{sec:gradflow}
To remove this ultraviolet divergence, we adopt a technique is ``smearing'', a local averaging of the field.\cite{solbrig2007} Specifically, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.

Specifically, the gradient flow pushes field configuration toward classical minima of the action. Additionally, renormalized correlation functions remain renormalized at nonzero flow time.\cite{luscher2013} In 2D $\phi^4$ scalar field theory, the gradient flow is defined by the differential equation 
\begin{equation}
    \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
\end{equation}
where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime and $\tau$ is the flow time. Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$. In the $\phi^4$ theory, we can solve this equation exactly to find \cite{monahan2016}
\begin{equation}
    \rho(\tau, x) = \frac{1}{4 \pi \tau} \int d^2 y e^{-(x-y)^2/4\tau} \phi(y).
\end{equation}
This function forms a Gaussian, smoothly dampening high-momentum modes and removing ultraviolet divergences from evolved correlation functions.\cite{makino2015a} We can visualize this by plotting the $\phi$ field, shown in Fig.~\ref{fig:flow}. These plots demonstrate the reduction of high momentum modes.

\begin{figure}[h]
  \centering
      \begin{subfigure}[b]{0.2\textwidth}\centering
        \includegraphics[width=0.9\textwidth]{imgs/gf0.png}
        \caption{$\tau=0$}
      \end{subfigure}%
      \begin{subfigure}[b]{0.2\textwidth}\centering
        \includegraphics[width=0.9\textwidth]{imgs/gf1.png}
        \caption{$\tau=0.001$}
      \end{subfigure}%
      \begin{subfigure}[b]{0.2\textwidth}\centering
        \includegraphics[width=0.9\textwidth]{imgs/gf2.png}
        \caption{$\tau=0.01$}
      \end{subfigure}%
      \begin{subfigure}[b]{0.2\textwidth}\centering
        \includegraphics[width=0.9\textwidth]{imgs/gf3.png}
        \caption{$\tau=0.1$}
      \end{subfigure}%
      \caption{\label{fig:flow} Effect of flow time evolution on a random lattice in the symmetric phase. White represents positive values of $\phi$ while black represents negative.}
\end{figure}

Generally, we can choose any flow time equation that drives the field towards a classical minimum. Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields. In the non-linear sigma model, it has been shown that an appropriate manifestation of the flow time can resolve divergences \cite{makino2015a}. Therefore, it is possible to define a different flow equation for this model as well. Following \cite{bietenholz2018}, we can define the gradient flow in this model via the differential equation
\begin{equation}
    \label{eq:nsm_gradflow}
    \partial_\tau \vec\rho (\tau,x) = \left( 1 - \vec\rho(\tau,x) \vec\rho(\tau,x)^T \right) \partial^2 \vec\rho(\tau,x),
\end{equation}
where $\partial^2$ is the Laplacian operator in Euclidean space\footnote{Explicitly, $\partial^2 = \frac{\partial^2}{\partial t^2} + \nabla^2$}. We solve this equation numerically using the boundary condition $\vec\rho(0,x) = \e(x)$.

### 2021-04-19 final-report.latex/methods.tex
fix Chris' notes in Methods

@@
-Our study of the gradient flow in the non-linear sigma model consists of a computational part and an analytical part.
+Our study of the gradient flow in the non-linear sigma model consists of computational system to simulate quantum fields numerically.
 We begin by outlining a numerical Monte Carlo method to simulate the lattice in two and three dimensions.
@@
 We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories.
+
 \section{Fields on the Lattice}
@@
 \end{equation}
-where $L$ is the side length (in units of the lattice constant $a$) of the system. In this study we focus solely on square geometries and thus the side length $L$ is unambiguous.
+where $L$ is the side length (in units of the lattice spacing $a$) of the system. In this study we focus solely on square geometries and thus the side length $L$ is unambiguous.
 
@@
 \begin{equation}
+    \label{eq:disc def}
     \int dtdx \rightarrow a^2 \sum_i
@@
 Finally, we redefine the gradient flow in on the lattice.
-Since the gradient flow is solved exactly in the $\phi^4$ model, we rely on a Fast Fourier Transform.
-\textit{TODO: explicit expression for this} In the NLSM, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes
+Since the gradient flow is solved exactly in the $\phi^4$ model, we rely on a discrete Fourier Transform.
+This method isolates the momentum modes of the field and dampens them by a factor of $e^-{\tau p^2}$ where $\tau$ is the flow time and $p$ is the momentum.
+In the NLSM, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes
 \begin{equation}
@@
 \begin{equation*}
-    \partial^2 \e(\tau,x) = \e(\tau, x+a \hat{t}) + \e(\tau,x-a\hat t) + \e(\tau, x+a \hat{x}) + \e(\tau,x-a\hat x) - 2 d \e(t)_x.
+    \partial^2 \e(\tau,x) = \e(\tau, x+a \hat{t}) + \e(\tau,x-a\hat t) + \e(\tau, x+a \hat{x}) + \e(\tau,x-a\hat x) - 4 \e(t)_x.
 \end{equation*}
-%TODO: derive the laplacian operator
+Unlike the $\phi^4$ gradient flow, this differential equation has no analytic solution.
+Therefore, we numerically solve for the gradient flow using a fourth-order Runge-Kutta approximation (see Sec.~\label{sec:runga-kutta}).
 
 \subsection{Discretized Observables}
+Following the definitions in Sec.~\ref{sec:primary observables}, we redefine the primary and secondary observables on the discretized lattice.
+Using the discretization definition in Eq.~\ref{eq:disc def}, we express the average magnetization as
+\begin{equation}
+|\bar\phi| \equiv \frac{1}{L^2}\left| \sum_i^{L^2} \phi(x_i)\right|.
+\end{equation}
+in the $\phi^4$ model and 
+\begin{equation}
+    |\bar\e| \equiv \frac{1}{L^2}\left|\sum_i^{L^2} \e(x_i)\right|.
+\end{equation}
+in the NLSM, where $L^2 = V/a^2$.
+
+We can also simplify the expression for the magnetic susceptibility.
+Due to the rotation invariance of both fields, the second terms in Eq.~\ref{eq:chim nlsm} and Eq.~\ref{eq:chim phi4} disappear (see Sec.~\ref{sec:avg mag}).
+In the NLSM, the susceptibility therefore becomes
+\begin{align}
+    \chi_m &= L^2 \left\langle \left( \sum_x e(x) \right)^2\right\rangle  \\
+           &= L^2 \left\langle \sum_{x,y}\e(x)\cdot \e(y) \right\rangle. 
+\end{align}
+Due to translational invariance from the periodic boundary conditions, we simplify this expression to be
+\begin{equation}
+    \chi_m = \left\langle \sum_x \e(0)\cdot\e(x) \right\rangle.
+\end{equation}
+Following the same logic, we derive 
+\begin{equation}
+    \chi_m = \left\langle \sum_x \phi(0)\phi(x) \right\rangle
+\end{equation}
+for the $\phi^4$ model.
+
+The definitions for the internal energy, Binder cumulant and bimodality remain unchanged on the lattice.
 
@@
 
+We begin this random walk with a so-called ``hot start'' where each field value at each lattice site is randomly selected.
+As an alternative to the hot start, we could use a ``cold start'' where the field begins completely aligned.
+With an appropriate thermalization, these initial configurations have no effect on the Monte Carlo statistics, a fact we verify in Sec.~\ref{thermalization}.
+
 \subsection{Metropolis Algorithm}
 We primarily use the Metropolis algorithm for the calculation of new Markov chain configurations.
-We begin with a so-called ``hot start,'' where each field value at each lattice site is randomly selected.
-Then we propose a new value for a single lattice point, which is accepted with a probability
+The method begins with proposing a new value for a single lattice point, which is accepted with a probability
 \begin{equation}
     P(\phi_a\rightarrow\phi_b) = \begin{cases} 
-        e^{-(S[\phi_b] - S_[\phi_a])} & S[\phi_b] < S_[\phi_a] \\
+        e^{-(S[\phi_b] - S[\phi_a])} & S[\phi_b] < S_[\phi_a] \\
         1 & \mathrm{otherwise} \\
@@
 where $\phi_a$ is the initial configuration and $\phi_b$ is the proposed configuration.
-This process is performed for each point on the lattice, making up a ``sweep.'' Repeating this sweep many times pushes the lattice toward the action minimum.
+This process is performed for each point on the lattice, making up a ``sweep''.
+Repeating this sweep many times pushes the lattice toward the action minimum.
 
 \subsection{Wolff Cluster Algorithm}
-
 Though the Metropolis algorithm will slowly find the absolute minimum of the theory, the presence of local minima can greatly prolong the convergence.
@@
 Therefore, the presence of large similarly-valued regions in the lattice can lead to a local minimum.
-One method of removing these clusters involves identifying all clusters on the lattice and probabilistically flipping each, a technique known as the Swendsen-Wang algorithm\cite{swendsen1987}.
+One method of removing these clusters involves identifying all clusters on the lattice and probabilistically flipping each, a technique known as the Swendsen-Wang algorithm \cite{swendsen1987}.
 
-A more efficient approach is the Wolff algorithm\cite{wolff1989}, which \textit{grows} one cluster probabilistically and flips it unconditionally.
+A more efficient approach is the Wolff algorithm \cite{wolff1989}, which \textit{grows} one cluster probabilistically and flips it unconditionally.
 In the case of $\phi^4$ theory, this flipping takes the form of a simple sign change.
 In the non-linear sigma model we choose a random unit vector $\vec r$ and consider the projection of the field on this vector.
-When the cluster flips, each site is flipped along this direction.\citeneeded To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site, growing the cluster from a single, randomly selected ``seed''.
+When the cluster flips, each site is flipped along this direction.
+To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site, growing the cluster from a single, randomly selected ``seed''.
 Starting with the seed, the probability of adding each neighboring site is given by the source site $x$ and the proposed site $x'$.
@@
 \begin{equation}
-    R\;\e = \e - 2(\e\cdot\vec{r})\e,
+    R\,\e = \e - 2(\e\cdot\vec{r})\e,
 \end{equation}
@@
     P_{add}(\phi(x),\phi(x')) = \begin{cases} 
-        1 - e^{-2\phi(x)\phi(x')]} & \mathrm{sgn}[\phi(x)]=\mathrm{sgn}[\phi(x')]\\
+        1 - e^{-2\phi(x)\phi(x')} & \mathrm{sgn}[\phi(x)]=\mathrm{sgn}[\phi(x')]\\
         0 & \mathrm{otherwise} \\
@@
 In order to parallelize the Metropolis algorithm, we use a checkerboard algorithm.
-Since the Lagrangian density at each site does not depend on any sites diagonal to it, the lattice can be split into ``white'' sites and ``black'' sites, like the tiles on a checkerboard.
-Each white site is independent of every other white site and likewise with black sites.
+We begin by splitting the lattice into ``white'' sites and ``black'' sites, like the tiles on a checkerboard.
+Since the Lagrangian density at each site does not depend on diagonal neighbors, each white site is independent of every other white site and likewise for black sites.
 Therefore, we can split the sites of each color into separate parallel processing nodes and independently run the Metropolis algorithm, ensuring that no site affects the Lagrangian density at any other site.
@@
 \label{sec:thermalization}
-Since we begin with a random lattice, the first few configurations of our Monte Carlo simulation will be far from the lattice minimum.
-We plot the action as a function of metropolis sweeps in Fig.~\ref{fig:therm}.
+In order to determine the necessary thermalization, we plot the action as a function of Metropolis sweeps in Fig.~\ref{fig:therm}.
 \begin{figure}[h]
@@
 Based on this plot, we determine that 1000 sweeps will give sufficient time for the system to reach the classical action minimum. We use this value for the remainder of this study.
+
+We also compare the hot and cold starts by plotting a histogram of the actions in Fig.~\ref{fig:coldstart}. This plot qualitatively demonstrates the irrelevance of the initial configuration.
+\begin{figure}[h]
+  \centering
+    \includegraphics[width=0.6\textwidth]{imgs/coldstart.png}
+  \caption{\label{fig:coldstart} Comparison of cold and hot starts after 1000 sweep thermalization in $L=404$ lattice. 1000 measurements taken every 50 sweeps.}
+\end{figure}
+
 \subsection{Autocorrelation}
-One important aspect is the correlation between different states of the Markov chain.
-Ideally, each sample of the lattice would be completely independent, but this is not the case.
-Since each configuration is based on previous configurations, there is a correlation between each pair of members in the Markov chain, decreasing exponentially based on the number of steps between.
-Explicitly, the autocorrelation scales as $e^{t/\tau_{int}}$ where $t$ is the number of steps between configurations and $\tau_{int}$ is the autocorrelation time.\footnote{Though it is called a time, $\tau_{int}$ is in units of Markov Chain steps.} When performing simulations of the lattice, the number of sweeps between measurements should be much larger than $\tau_{int}$.
+Due to the nature of the Markov chain, each member of the ensemble is correlated to every other.
+Since each configuration is based on previous configurations, each pair of members in the Markov chain has a correlation which decreases exponentially based on the number of steps between.
+This value is known as the ``autocorrelation'' and scales as $e^{t/\tau_{int}}$ where $t$ is the number of steps between configurations and $\tau_{int}$ is the autocorrelation time.\footnote{Though it is called a time, $\tau_{int}$ is in units of Markov Chain steps.} When performing simulations of the lattice, the number of sweeps between measurements should be much larger than $\tau_{int}$ since Monte Carlo methods generally assume independent observations.
 
@@
 \subsection{Runge-Kutta Algorithm}
+\label{sec:runga-kutta}
 In order to calculate the gradient flow, we numerically solve the differential equation using a fourth-order Runga-Kutta approximation. This algorithm refines the Euler method 
 \begin{equation*}
-    \e(\tau+h)_x \approx \e(\tau)_x + h f(\e(\tau)_x).
+    \e(\tau+h,x) \approx \e(\tau)_x + h f(\e(\tau,x))
 \end{equation*}
@@
 \begin{equation}
-    f(\e)=\partial_\tau \e (\tau)_x  = \left( 1 - \e(\tau)_x \e(\tau)_x^T\right) \partial^2 \e(\tau)_x,
+    f(\e)=\partial_\tau \e (\tau,x)  = \left( 1 - \e(\tau,x) \e(\tau,x)^T\right) \partial^2 \e(\tau,x),
 \end{equation}
@@
     \label{eq:rungekutta}
-    k_1 &= h f\left(\e\left(\tau\right)_x\right) \\ 
-    k_2 &= h f\left(\e\left(\tau\right)_x + \frac{k_1}{2}\right) \\ 
-    k_3 &= h f\left(\e\left(\tau\right)_x + \frac{k_2}{2}\right) \\ 
-    k_4 &= h f\left(\e\left(\tau\right)_x + k_3\right) \\ 
-    \e(\tau+h)_x &= \e(\tau)_x + \frac{k_1}{6} + \frac{k_2}{3} + \frac{k_3}{3} + \frac{k_4}{6} + O(h^5).
+    k_1 &= h f\left(\e\left(\tau,x\right)\right) \\ 
+    k_2 &= h f\left(\e\left(\tau,x\right) + \frac{k_1}{2}\right) \\ 
+    k_3 &= h f\left(\e\left(\tau,x\right) + \frac{k_2}{2}\right) \\ 
+    k_4 &= h f\left(\e\left(\tau,x\right) + k_3\right) \\ 
+    \e(\tau+h)_x &= \e(\tau,x) + \frac{k_1}{6} + \frac{k_2}{3} + \frac{k_3}{3} + \frac{k_4}{6} + O(h^5).
 \end{align}
@@
 Since $\chi_t$ diverges at $\theta=0$, we expect the plot of $\langle Q\rangle_\theta$ to approach a vertical line in the in continuum limit.
+The numerical demonstration of this hypothesis is a goal of this work.

### 2021-04-19 final-report.latex/theory.tex
fix Chris' notes in Methods

@@
 
-The theory is defined by the Euclidean action 
+The theory is defined by the 1+1 dimensional Euclidean action 
 \begin{equation}
     \label{eq:nlsm euclidean action}
-    S_E = \frac{\beta}{2} \int d^Dx \; \partial^\mu \e \cdot \partial_\mu \e
+    S_E = \frac{\beta}{2} \int d^2x \; \partial^\mu \e \cdot \partial_\mu \e
 \end{equation}
-subject to the constraint that $\e\cdot\e = 1$. Here, $\beta$ is the inverse coupling and $D$ is the number of spacetime dimensions.
+subject to the constraint that $\e\cdot\e = 1$. Here, $\beta$ is the inverse coupling.
 
@@
 \subsection{Primary Observables}
+\label{sec:primary observables}
 Each primary observable is defined on each configuration independently, meaning they do not encode ensemble statistics of the Markov Chain.\footnote{One slight exception is the magnetic susceptibility, which can be defined both ways.} In the $\phi^4$ model, we can develop an intuition around these quantities by visualizing the symmetric and broken phases.
@@
     \caption{\label{tab:primary observables} Average magnetization with average action per site corresponding to the particular configurations in Fig.~\ref{fig:primary observables}.}
+\end{table}
+
 \subsubsection{Average Magnetization}
@@
 \begin{equation}
-|\bar\phi| \equiv \frac{1}{L^2}\left| \sum_i^{L^2} \phi(x_i)\right|.
+|\bar\phi| \equiv \frac{1}{V}\left| \int d^2x \phi(x)\right|.
 \end{equation}
@@
 \begin{equation}
-    |\bar\e| \equiv \frac{1}{L^2}\left|\sum_i^{L^2} \e(x_i)\right|.
+    \label{eq:chim nlsm}
+    |\bar\e| \equiv \frac{1}{V}\left|\int d^2x \e(x)\right|.
 \end{equation}
@@
 \begin{equation}
-    E = \frac{2}{\beta L^2} \langle S \rangle
+    E = \frac{2}{\beta V} \langle S \rangle
 \end{equation}
@@
 \begin{equation}
-    \chi_m \equiv L^2 \big(\langle {\bar\e\,}^2 \rangle - \langle \bar\e\, \rangle^2\big)
+    \chi_m \equiv V \big(\langle {\bar\e\,}^2 \rangle - \langle \bar\e\, \rangle^2\big)
 \end{equation}
@@
 \begin{equation}
-    \chi_m \equiv L^2 \big(\langle {\bar\phi}^2 \rangle - \langle \bar\phi \rangle^2\big)
+    \chi_m \equiv V \big(\langle {\bar\phi}^2 \rangle - \langle \bar\phi \rangle^2\big)
 \end{equation}
-in the $\phi^4$ model.
-Manifestly, this expression appears as a secondary observable since it is not defined for each lattice configuration.
-However, due to the rotation invariance of both fields, the second term disappears (see Sec.~\ref{sec:avg mag}), yielding
-\begin{align}
-    \chi_m &= L^2 \left\langle \left( \sum_x e(x) \right)^2\right\rangle  \\
-           &= L^2 \left\langle \sum_{x,y}\e(x)\cdot \e(y) \right\rangle. 
-\end{align}
-Due to translational invariance from the periodic boundary conditions, this expression becomes
-\begin{equation}
-    \chi_m = \left\langle \sum_x \e(0)\cdot\e(x) \right\rangle
-\end{equation}
-for the NLSM. Following the same logic, we derive 
-\begin{equation}
-    \chi_m = \left\langle \sum_x \phi(0)\phi(x) \right\rangle
-\end{equation}
-for the $\phi^4$ model.
+in the $\phi^4$ model. Manifestly, this expression appears as a secondary observable since it is not defined for each lattice configuration.
 
@@
 
-\subsection{Regularization: fields on the lattice}
+\subsection{Regularization}
 Regularization is a process which introduces a new parameter into calculations. One example is a momentum cutoff. This technique transforms infinite momentum integrals as follows:  
@@
 The inherent parameter in this case is the lattice spacing $a$ which measures the width of each lattice chunk.
+This discretization effectively imposes a hard momentum cutoff of $k=\pi/a$.
+According to Bloch's theorem, any mode above this cutoff is equivalent to a lower-momentum mode since the high frequency information disappears on a discrete lattice.
 
-% TODO: Mention what was summer work
+We can view this cutoff in momentum space by considering a square with side length $2\pi/a$ centered at the origin.
+On the lattice, any mode outside this zone contains no more information than a corresponding mode inside.
+This area is known as a ``Brillouin zone'' and contains all possible momentum modes on the lattice, effectively imposing a hard cutoff.
+One of the main strengths of lattice regularization is preservation of gauge invariance, a property that makes lattice methods useful for QCD.
 
 \subsection{Renormalization}
-After regularization, we redefine the Lagrangian parameters in terms of the regularization parameter, following a handful of boundary conditions.
-In this study, we assert that $L/\xi$ remains constant, where $L$ is the side length of the system and $\xi$ is the coherence length.
-In perturbation theory, this process is arduous and includes the introduction of counter-terms into the Lagrangian.
-In the case of the non-linear sigma model, it is impossible using counterterms\citeneeded but can be performed numerically.
+After regularization, we redefine the Lagrangian parameters in terms of the regularization parameter using a handful of boundary conditions.
+Following \cite{bietenholz2018}, we require that $L/\xi$ remains constant, where $L$ is the side length of the system and $\xi$ is the coherence length.
+In perturbation theory, the renormalization process is arduous and includes the introduction of counter-terms into the Lagrangian.
+In the case of the NLSM it is impossible using counter-terms but can be performed numerically.
 In this study, we use predetermined values from \cite{bietenholz2018}.
+% TODO: cite this
 
-At this point, we can calculate observables as functions of regularization parameters.
-To achieve physical values, we take the limit as the regularization parameters approach their physical values.
+At this point, we can calculate renormalized parameters of the Lagrangian.
+To achieve a physical theory, we take the limit as the regularization parameters approach their physical values.
 With a momentum cutoff, we take $\Lambda \rightarrow \infty$ and with dimensional regularization we usually take $d\rightarrow 4$.
@@
 Unfortunately, this is not always the case.
+External operators may also diverge due to regularization procedures.
+A prototypical example is the angular momentum operator on the lattice.
+Since a square lattice breaks continuous rotational symmetry, orthogonal angular momentum operators mix on the lattice and can cause divergences \cite{monahan2016}.
+
 The topological susceptibility $\chi_t$ is one such value that diverges in the continuum limit.
@@
 \label{sec:gradflow}
-To remove this ultraviolet divergence, we adopt a technique is ``smearing'', a local averaging of the field.\cite{solbrig2007} Specifically, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.
+To remove this ultraviolet divergence, we adopt a technique is ``smearing'', a local averaging of the field \cite{solbrig2007}.
+Specifically, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.
 
-Specifically, the gradient flow pushes field configuration toward classical minima of the action.
-Additionally, renormalized correlation functions remain renormalized at nonzero flow time.\cite{luscher2013} In 2D $\phi^4$ scalar field theory, the gradient flow is defined by the differential equation
+Specifically, the gradient flow pushes field configurations toward classical minima of the action.
+Additionally, renormalized correlation functions remain renormalized at nonzero flow time for gauge theories such as QCD \cite{luscher2013}.
+In 2D $\phi^4$ scalar field theory, the gradient flow is defined by the differential equation
 \begin{equation}
@@
 where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime and $\tau$ is the flow time.
-Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
+Here, $\rho$ is the field flowed into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
 In the $\phi^4$ theory, we can solve this equation exactly to find \cite{monahan2016}
@@
 \end{equation}
-This function forms a Gaussian, smoothly dampening high-momentum modes and removing ultraviolet divergences from evolved correlation functions.\cite{makino2015a} We can visualize this by plotting the $\phi$ field, shown in Fig.~\ref{fig:flow}.
+This function forms a Gaussian, smoothly dampening high-momentum modes and removing ultraviolet divergences from evolved correlation functions \cite{makino2015a}.
+We can visualize this by plotting the $\phi$ field, shown in Fig.~\ref{fig:flow}.
 These plots demonstrate the reduction of high momentum modes.
-
 \begin{figure}[h]
@@
 Generally, we can choose any flow time equation that drives the field towards a classical minimum.
-Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields.
-In the non-linear sigma model, it has been shown that an appropriate manifestation of the flow time can resolve divergences \cite{makino2015a}.
-Therefore, it is possible to define a different flow equation for this model as well.
-Following \cite{bietenholz2018}, we can define the gradient flow in this model via the differential equation
+Following \cite{bietenholz2018}, we can define the gradient flow for the NLSM via the differential equation
 \begin{equation}

### 2021-04-19 final-report.latex/wordcount.sh

@@ -1,0 +1,4 @@
+#!/bin/bash
+
+
+texcount conclusion.tex intro.tex main.tex methods.tex results.tex theory.tex

### 2021-04-21 final-report.latex/intro.tex
fix Chris' notes in Methods
beef up intro

@@
 Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics.
-In condensed matter physics, effective field theories model emergent effects such as phonons and quasiparticles.
-Compared to experiment, QFT is remarkably accurate, famously matching the experimental value for the electron $g$-factor to eight significant figures.\cite{weisskopf1981}
+QFT is also plays an critical role in condensed matter physics through effective field theories which model emergent effects such as phonons and quasiparticles.
+Compared to experiment, QFT is remarkably accurate, famously matching the experimental value for the electron $g$-factor to eleven significant figures \cite{odom2006}, arguably the most accurate prediction in all of science.
 
-However, this power comes at a cost: the study of quantum fields is rife with infinities.
+However this power comes at a cost: the study of quantum fields is rife with infinities.
 A na\"ive treatment of quantum field theory produces divergent values for physical quantities, a clearly impossible result.
-Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the so-called \textit{renormalization group}.
-However, the technique fails with perturbatively nonrenormalizable theories.
+Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the renormalization group which exploits the freedom of parameters such as mass and electric charge.
+Since these values cannot be directly measured, renormalization allows them to become infinite, thereby cancelling the infinities in physical results.
+However, there exists a set of theories which as not perturbatively renormalizable.
 
-One such example is the \textit{non-linear sigma model}, a prototypical theory in both condensed matter and particle physics.
-In solid-state systems, this model describes Heisenberg ferromagnets and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), exhibiting characteristic features such as a mass gap and asymptotic freedom.
-\citeneeded
+One such example is the \textit{non-linear sigma model} (NLSM), a prototypical theory in both condensed matter and particle physics.
+In solid-state systems, this model describes Heisenberg ferromagnets \cite{callan1985} and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), the gauge theory that describes the strong nuclear force.
+In general, the NLSM shares key features with non-Abelian gauge theories (such as QCD) like a mass gap and asymptotic freedom \cite{polyakov1975} and therefore is a useful model for exploring these phenomena in simpler systems.
 
-In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions one dimension of space, one dimension of time).
-This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action.
+In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time).
+This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action in Euclidean space.
+Solutions such as these are ``topologically protected'', meaning they cannot evolve into the vacuum state via small fluctuations.
+Therefore, it is critically important to take topology into account when considering quantum field theories in a cosmology and high energy physics \cite{goddard1986}.
+Additionally, topological stability may become a key tool for fault-tolerant quantum computers \cite{kitaev1997}.
+In these devices, topology could protect the delicate quantum states necessary for information processing.
 
-Since the non-linear sigma model cannot be renormalized perturbatively, we cannot study these topological effects with normal perturbative techniques.
-An alternative solution is placing the field on a discretized lattice, a technique originally used for quantum chromodynamics.
-In this scenario, field configurations become computationally calculable.
-This process introduces an nonphysical length scale $a$, the lattice spacing.
+Since the NLSM in not perturbatively renormalizable, we require nonperturbative techniques to study topological effects.
+An alternative solution is to place the field on a discretized Euclidean lattice, a technique originally used for quantum chromodynamics \cite{wilson1974}.
+After this transformation, fields become computationally calculable.
+This process introduces the lattice spacing as a length scale $a$.
+% something here
 Therefore, we expect any physical result to converge in the continuum limit, i.e.
@@
 
-The gradient flow is a technique designed to remove these divergences.
-By dampening high-momentum fluctuations, the gradient flow reduces power-divergent mixing and make observables finite on the lattice.\cite{monahan2016} In quantum chromodynamics, previous studies have verified the ability of the gradient flow to make observables finite\citeneeded.
-Due to its success in QCD, there has been interest in using the gradient flow to finitize the topological susceptibility in the 1+1 $O(3)$ non-linear sigma model.\cite{bietenholz2018}.
+In this study, we focus on one observable that diverges in the continuum: the topological susceptibility.
+The convergence of the topological susceptibility in the 1+1 $O(3)$ NLSM has been the subject of debate for the past four decades \cite{berg1981,bietenholz2018}.
+In QCD, mathematical techniques have proven that this quantity vanishes in the continuum limit \cite{giusti2004}, however a similar line of reasoning for the NLSM has not been successful \cite{bietenholz2018}.
 
-%goals here
+To remedy this issue, we consider the gradient flow, a technique designed to remove divergences in operators.
+By dampening high-frequency fluctuations, the gradient flow reduces terms which scale with the inverse lattice spacing, making some observables finite on the lattice \cite{monahan2016}.
+In QCD, the gradient flow successfully makes the topological susceptibility finite in numerical calculations \cite{bruno2014}, corroborating the analytical result in \cite{giusti2004}.
+This success has motivated the usage of the gradient flow to calculate the topological susceptibility in the 1+1 $O(3)$ NLSM, though the observable still diverges in the continuum limit \cite{bietenholz2018}.
+
+A second perspective on the topological susceptibility arises from the introduction of a $\theta$ term into the field Lagrangian.
+This term drives the vacuum state into a topological phase \cite{alles2007}.
+Differentiating the field's partition function with respect to $\theta$ yields a value proportional to the topological susceptibility.
+The effect of nonzero $\theta$ on the theory therefore should reflect the divergence in the continuum limit.
+
+In this paper we verify the result the divergence of the topological susceptibility and develop a clearer picture of how the $\theta$-term affects the topology of the NLSM.
 
@@
 By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.
+
+\section{Summer Research}
+
+A portion of this work began during the summer of 2021 using funding from the 1693 Scholars Program.
+This preliminary research consisted of literature review and numerical tests with an Ising model simulation as well as an initial implementation of the $\phi^4$ model.
+The $\phi^4$ calculation performed in this study and the entirety of the NLSM portion occurred during the academic year as part of the PHYS 495/496 Honor course.
 
@@
 \end{itemize}
-
-%\section{Main Experimental Goal}

### 2021-04-22 final-report.latex/conclusion.tex
second round of Chris' notes

@@
-Using a Monte Carlo simulation, we have analyzed two main quantum field theories: the $\phi^4$ model and the $O(3)$ non-linear sigma model.
-To perform the calculation, we run a series of Metropolis sweeps interspersed with Wolff cluster algorithms, creating a Markov chain of samples.
-We use these field configurations to calculate operators using a Wick-rotated path integral.
+Using a Monte Carlo simulation, we have analyzed two main quantum field theories in 1+1 spacetime dimensions: the $\phi^4$ model and the $O(3)$ non-linear sigma model.
+We used the path integral formulation of QFT to simulate these quantum fields as statistical systems on a Euclidean lattice and extract information about the phase transitions and topology.
+To perform these calculations, we run a series of Metropolis sweeps interspersed with Wolff cluster steps, creating a Markov chain of samples.
+We ensure that each measurement is effectively independent by measuring the autocorrelation and thermalize the lattice to keep all measurements near the action minima.
 
@@
 This process verified the effectiveness of the Monte Carlo computational system.
-We then generalized our system to the $O(3)$ non-linear sigma model, confirming our calculations with known results.
+We then generalized our system to the $O(3)$ non-linear sigma model and transitioned to a \texttt{C++} code base.
+We confirmed our calculations with known results by measuring the internal energy and magnetic susceptibility.
 We calculated the topological susceptibility $\chi_t$ to analyze its divergence in the continuum limit, confirming that this is indeed the case even at finite flow time.
-Specifically, the gradient flow reduces the linear divergence of the susceptibility to either a power-law relationship or a logarithmic relationship with $\chi^2/DOF$ of $31.3$ and $761.6$ respectively.
+Specifically, the topological susceptibility under the gradient flow follows either a power-law relationship or a logarithmic relationship as a function of lattice size, both of which diverge as $L\rightarrow\infty$.
+At flow time $\tau=5t_0$, the $\chi^2/DOF$ goodness of fit values are of \powchi and \logchi respectively.
 
-Finally, we produced a plot demonstrating the relationship between the topological charge and the $\theta$ in the topological action.
-This plot demonstrates qualitatively the different divergences of the topological susceptibility, which is equivalent to the slope at $\theta=0$.
-In the $\tau=0$ case, this slope became vertical rapidly for higher lattice sizes while this process occurred more slowly at nonzero flow time.
+Finally, we demonstrated the relationship between the topological charge and the $\theta$ in the topological action.
+This plot demonstrates quantitatively the different divergences of the topological susceptibility, which is equivalent to the slope at $\theta=0$.
+In the $\tau=0$ case, we see the slope rapidly approach positive infinity at $\theta=0$.
+While this transition occurs more slowly with the application of the gradient flow, the slope continues to increase.
 
 Though the logarithmic and power-law functions visually fit the susceptibility data well, the $\chi_2$ values are high.
-This imprecision can be attributed to errors that are too small.
-Future work could include more independent simulations which may reduce the variance between points or increase the statistical errors.
-Furthermore, more and larger lattices may help reduce the $\chi^2$ value.
+This imprecision can be attributed to underestimated errors.
+Since the $\tau=0$ case features a more acceptable fit ($\chi^2/DOF=1.8$ for the power-law), the gradient flow seemingly contributes to this error.
+While the Jackknife method accurately estimates statistical errors of the sample, we did not incorporate any systematic errors arising from the gradient flow calculation.
+Future work could reduce the tolerance of the adaptive step size algorithm to make this calculation more accurate, though this increases computational requirements substantially.
+Furthermore, a larger number of lattice sizes may provide a more complete picture of the continuum limit.
 
-Procedurally, we found that the MPI parallelization and checkerboard algorithm was unnecessary in this calculation.
+Procedurally, we found that the MPI parallelization and checkerboard algorithm were unnecessary in this calculation.
 The computation time of the Runge-Kutta algorithm for computing the gradient flow far outweighed that of the Monte Carlo simulation.
 A possible improvement could be a parallelization of the Runge-Kutta algorithm or a more accurate approximation to the gradient flow.
+Additionally, the performance of the Python simulation was slower by up to two orders of magnitude.
+Future studies should therefore rely solely on an efficient programming language for Monte Carlo techniques.
+
+In the context of the long-standing question of topological suscpetibility in the NLSM, this study has diminished the plausibility that ultraviolet fluctuations cause the divergence.
+Instead, we are left to consider the other two options outlined in \cite{berg1981}: that the definition of the topological charge density is problematic or that the NLSM does not have a decent continuum limit.
+Future work could include similar numerical calculations using a different definition of the topological charge.
 
@@
 Though the calculation of the $\chi_t$ divergence confirms existing literature, the relationship between the topological charge and $\theta$ in the flow time was previously unexplored.
-
-Other future work may include explorations of higher order non-linear sigma models or other topological effective field theories.
-The latter of these models has applications in topological quantum computing and other technologies.
-
-% MPI is unnecessary in this case
-
-\appendix
+Beyond the convergence of $\chi_t$, this relationship has applications in condensed matter where different values of $\theta$ can describe spin-chains of fermions and bosons \cite{bogli2012}.
+Furthermore, the mass gap has a strong relationship with the $\theta$-terms in topological NLSM, featuring massive and massless regimes \cite{allessalom2008}.

### 2021-04-22 final-report.latex/intro.tex
second round of Chris' notes

@@
-Quantum field theory (QFT) is an extraordinarily successful framework which describes a range of physical phenomena.
+Quantum field theory (QFT) is an framework used to describe a range of physical phenomena to a remarkable degree of accuracy.
 Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics.
-QFT is also plays an critical role in condensed matter physics through effective field theories which model emergent effects such as phonons and quasiparticles.
-Compared to experiment, QFT is remarkably accurate, famously matching the experimental value for the electron $g$-factor to eleven significant figures \cite{odom2006}, arguably the most accurate prediction in all of science.
+QFT also plays a critical role in condensed matter physics through effective field theories which model emergent phenomena such as phonons and quasiparticles.
+Compared to experiment, QFT is remarkably accurate, famously predicting the electron $g$-factor to eleven significant figures \cite{odom2006}, arguably the most accurate prediction in all of science.
 
@@
 A na\"ive treatment of quantum field theory produces divergent values for physical quantities, a clearly impossible result.
-Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the renormalization group which exploits the freedom of parameters such as mass and electric charge.
-Since these values cannot be directly measured, renormalization allows them to become infinite, thereby cancelling the infinities in physical results.
-However, there exists a set of theories which as not perturbatively renormalizable.
+Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the renormalization group.
+This counter-intuitive technique exploits the freedom of parameters such as mass and electric charge.
+Since these constants cannot be directly measured, renormalization allows them to become infinite, thereby cancelling the infinities in physical results.
+Unfortunately, not all theories are perturbatively renormalizable.
 
@@
 In solid-state systems, this model describes Heisenberg ferromagnets \cite{callan1985} and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), the gauge theory that describes the strong nuclear force.
-In general, the NLSM shares key features with non-Abelian gauge theories (such as QCD) like a mass gap and asymptotic freedom \cite{polyakov1975} and therefore is a useful model for exploring these phenomena in simpler systems.
+In general, the NLSM shares key features with non-Abelian gauge theories such as QCD, including a mass gap and asymptotic freedom \cite{polyakov1975}.
+Therefore, the NLSM is a useful model for exploring the effect of these properties using a simpler system.
 
@@
 This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action in Euclidean space.
-Solutions such as these are ``topologically protected'', meaning they cannot evolve into the vacuum state via small fluctuations.
+Solutions such as these are ``topologically protected,'' meaning they cannot evolve into the vacuum state via small fluctuations.
 Therefore, it is critically important to take topology into account when considering quantum field theories in a cosmology and high energy physics \cite{goddard1986}.
@@
 After this transformation, fields become computationally calculable.
-This process introduces the lattice spacing as a length scale $a$.
-% something here
-Therefore, we expect any physical result to converge in the continuum limit, i.e.
-when $a\rightarrow 0$.
-However, this is not always the case as observables mix on the lattice, leading to divergences.
-As an example, states of definite angular momentum mix when discretized, a clear violation of the angular momentum commutation relations.
+This process introduces the lattice spacing as a length scale $a$ where the \textit{continuum limit} is defined as taking $a$ to zero.
+We expect physical observables to converge in the continuum limit, however this is not always the case.
+As an example, states of definite angular momentum mix when discretized on a rectangular lattice due to a breaking of continuous rotational symmetry.
+This causes some operators that depend on angular momentum to suffer divergences.
 
 In this study, we focus on one observable that diverges in the continuum: the topological susceptibility.
-The convergence of the topological susceptibility in the 1+1 $O(3)$ NLSM has been the subject of debate for the past four decades \cite{berg1981,bietenholz2018}.
-In QCD, mathematical techniques have proven that this quantity vanishes in the continuum limit \cite{giusti2004}, however a similar line of reasoning for the NLSM has not been successful \cite{bietenholz2018}.
+The topological susceptibility in the 1+1 $O(3)$ NLSM has been the subject of debate for the past four decades \cite{bietenholz2018} since it is unclear if a convergent solution exists.
+While the some analytical arguments argue the topological susceptibility should approach zero in the continuum limit, numerical results predict infinities.
+In QCD, mathematical techniques proved that the susceptibility vanishes in the continuum limit \cite{giusti2004}, a fact supported by numerical calculation \cite{giusti2004}.
 
@@
 
-A second perspective on the topological susceptibility arises from the introduction of a $\theta$ term into the field Lagrangian.
-This term drives the vacuum state into a topological phase \cite{alles2007}.
+A second perspective on the topological susceptibility arises from the introduction of a $\theta$-term into the field Lagrangian.
+This term drives the vacuum state into a topological phase \cite{allessalom2008}.
 Differentiating the field's partition function with respect to $\theta$ yields a value proportional to the topological susceptibility.
@@
 
-In this paper we verify the result the divergence of the topological susceptibility and develop a clearer picture of how the $\theta$-term affects the topology of the NLSM.
+In this work we verify the divergence of the topological susceptibility and develop a clearer picture of how the $\theta$-term affects the topology of the 1+1 $O(3)$ NLSM.
 
@@
 
-To numerically study the topological qualities of the non-linear sigma model, we first implement a Markov Chain Monte Carlo simulation.
+To numerically study the topological qualities of the non-linear sigma model, we first implement a Markov chain Monte Carlo simulation.
 We initially construct a proof-of-concept Python program that models the simpler $\phi^4$ model (see Sec.~\ref{sec:phi4}).
 After comparing with existing literature, we transition to a C++ simulation for efficiency, implementing the non-linear sigma model in larger lattices.
-Since the gradient flow has no exact solution in the non-linear sigma model, we implement a numerical solution using a fourth-order Runge-Kutta approximation.
+Since the gradient flow has no exact solution in the non-linear sigma model, we implement a numerical solution using a fourth-order Runge-Kutta approximation with automatic step sized.
 By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.
@@
 This preliminary research consisted of literature review and numerical tests with an Ising model simulation as well as an initial implementation of the $\phi^4$ model.
-The $\phi^4$ calculation performed in this study and the entirety of the NLSM portion occurred during the academic year as part of the PHYS 495/496 Honor course.
+The $\phi^4$ calculation performed in this study and the entirety of the NLSM portion occurred during the academic year as part of the PHYS 495/496 Honors course.
 

### 2021-04-22 final-report.latex/main.tex
second round of Chris' notes

@@
 \usepackage{listings}
+%\usepackage[%
+    %style=numeric, 
+    %sorting=none,
+    %doi=false,
+    %url=false,
+    %giveninits=true,
+    %eprint=false,
+    %isbn=false,
+    %clearlang=true,
+    %date=year,
+    %hyperref]{biblatex}
+\usepackage[style=phys,biblabel=brackets,maxbibnames=3]{biblatex}
+%\renewbibmacro{in:}{%
+  %\ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
+
 \setlength{\oddsidemargin}{0.5in}
@@
 
+\newcommand{\logchi}{7.4 }
+\newcommand{\powchi}{8.8 }
+
+\addbibresource{library.bib}
+
 \newcommand{\e}{\vec e}
-\newcommand{\citeneeded}{\footnote{\textit{citation needed}}}
 \newcommand{\half}{\frac{1}{2}}
@@
 I also acknowledge the help the High Performance Computing group at William \& Mary, whose cluster performed many of these simulations.
-And finally, I would like to acknowledge the financial help of the 1693 Scholars Program which funded the summer portion of this research.
+And finally, I would like to acknowledge the financial help of the 1693 Scholars Program which funded preliminary research over the summer.
 
@@
 \paragraph
-\indent The $O(3)$ non-linear model features topological effects such as instantons but cannot be perturbatively renormalized.
-Other nonperturbative theories such as QCD have motivated the development of lattice quantum field theory, where fields are simulated as discrete arrays and extrapolated to the continuum limit.
-Unfortunately, operators such as the topological susceptibility diverge in this limit.
-The gradient flow is a strong contender to remove these infinites, having prior success in QCD.
-We confirm a previous result\cite{bietenholz2018} that the gradient flow does not make the topological susceptibility infinite.
-We also study the nontrivial field theory by introducing a $\theta$ term into the action and plot the topological charge as a function of $\theta$.
+\indent Quantum field theory is an extraordinarily successful framework that describes phenomena in particle physics and condensed matter.
+The $O(3)$ non-linear sigma model (NLSM) is a specific theory used in both of these fields, describing ferromagnets and acting as a prototype for the strong nuclear force.
+It features topologically stable configurations known as instantons which cannot continuously evolve to the ground state.
+The topological susceptibility is a parameter that describes this stability and is predicted to vanish in physical theories, however numerical simulations find that the topological susceptibility diverges in the continuum limit \cite{berg1981}.
+This issue has motivated the application of the ``gradient flow'', a smoothing of high-frequency modes.
+We study the effect of the gradient flow on this divergence, finding that it persists as a logarithmic divergence.
+This result supports a previous study \cite{bietenholz2018} and indicates that either the definition of topological charge is problematic or the NLSM has no well-defined continuum limit.
+We also study the nontrivial field theory by introducing a $\theta$-term into the action and analyze the topological charge as a function of $\theta$ under the gradient flow.
 
@@
 \pagenumbering{arabic}
+
 \chapter{Introduction}
@@
 
-\bibliographystyle{unsrt}
-\bibliography{library}
+%\bibliographystyle{unsrt}
+%\bibliography{library}
+
+\printbibliography
 

### 2021-04-22 final-report.latex/methods.tex
second round of Chris' notes

@@
 
-We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories.
+We implement these two algorithms first in Python for the $\phi^4$ model.
+Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories.
+For the C++ simulation, we use the \texttt{gcc} compiler with the highest level of optimization.
 
@@
 
-In the $\phi^4$ model, we can specify a discrete action using the Euclidean action from Eq. \ref{eq:phi4 euclidean action}. We begin by redefining the derivative operator as a difference:
+In the $\phi^4$ model, we can specify a discrete action using the Euclidean action from Eq. \ref{eq:phi4 euclidean action}. We begin by redefining the derivative operator as a finite difference:
 \begin{align}
@@
 \begin{align}
-    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow& \frac{1}{2a^2}\left[ \left(\phi(x+a \hat{t}) -\phi(x)\right)^2 + \left(\phi(x+a \hat{x}) -\phi(x)\right)^2\right] \\
+    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow& \frac{1}{2a^2}\left[ \left(\phi(x+a \hat{t}) -\phi(x)\right)^2 + \left(\phi(x+a \hat{x}) -\phi(x)\right)^2\right]\nonumber \\
     \begin{split} \rightarrow& \frac{1}{2a^2}\left[ \phi^2(x+a\hat{t}) + \phi^2(x+a\hat{x}) + 2\phi^2(x)\right.\\ &\qquad \left.- 2\phi(x+a\hat{t})\phi(x) - 2\phi(x+a\hat{x})\phi(x) \right]\hfilneg\end{split}
@@
 Since the gradient flow is solved exactly in the $\phi^4$ model, we rely on a discrete Fourier Transform.
-This method isolates the momentum modes of the field and dampens them by a factor of $e^-{\tau p^2}$ where $\tau$ is the flow time and $p$ is the momentum.
+This method isolates the momentum modes of the field and dampens them by a factor of $e^{-\tau p^2}$ where $\tau$ is the flow time and $p$ is the momentum.
 In the NLSM, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes
@@
 Unlike the $\phi^4$ gradient flow, this differential equation has no analytic solution.
-Therefore, we numerically solve for the gradient flow using a fourth-order Runge-Kutta approximation (see Sec.~\label{sec:runga-kutta}).
+Therefore, we numerically solve for the gradient flow using a fourth-order Runge-Kutta approximation (see Sec.~\ref{sec:runga-kutta}).
 
@@
 We can also simplify the expression for the magnetic susceptibility.
-Due to the rotation invariance of both fields, the second terms in Eq.~\ref{eq:chim nlsm} and Eq.~\ref{eq:chim phi4} disappear (see Sec.~\ref{sec:avg mag}).
+Due to the spin-flip invariance of both fields, the second terms in Eq.~\ref{eq:chim nlsm} and Eq.~\ref{eq:chim phi4} disappear (see Sec.~\ref{sec:avg mag}).
 In the NLSM, the susceptibility therefore becomes
@@
 Under periodic boundary conditions, these triangles on the sphere necessarily wrap $S^2$ an integer number of times, ensuring $Q$ is an integer.
-In the continuum limit, the eff\ref{sec:topological charge}
+In the continuum limit, field must be uniform at $x\rightarrow\infty$ and therefore effectively forms a sphere (see Sec.~\ref{sec:topological charge}).
 \begin{figure}[h]
@@
 \end{figure}
-In the NLSM described by Eq.~\ref{eq:nlsm discrete action}, the values of $Q$ are roughly Gaussian around $Q=0$, as shown in Fig.~\ref{fig:hist}.
+In the NLSM described by Eq.~\ref{eq:nlsm discretized action}, the values of $Q$ are roughly Gaussian around $Q=0$, as shown in Fig.~\ref{fig:hist}.
 \begin{figure}[h]
@@
 
-On the lattice, this quantity is known to diverge in the continuum limit \cite{bietenholz2018}. % TODO: highlight that this is true for QCD and the NLSM.
+On the lattice, this quantity is known to diverge in the continuum limit owing solely to the $x=0$ term \cite{bietenholz2018}. This divergence exists in QCD as well \cite{bruno2014}.
 
@@
 As an alternative to the hot start, we could use a ``cold start'' where the field begins completely aligned.
-With an appropriate thermalization, these initial configurations have no effect on the Monte Carlo statistics, a fact we verify in Sec.~\ref{thermalization}.
+With an appropriate thermalization, these initial configurations have no effect on the Monte Carlo statistics, a fact we verify in Sec.~\ref{sec:thermalization}.
 
@@
       \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.9\textwidth]{imgs/therm24.png}
+        \includegraphics[width=\textwidth]{imgs/therm24.png}
         \caption{$L=24$}
@@
       \begin{subfigure}[b]{0.5\textwidth}\centering
-        \includegraphics[width=0.9\textwidth]{imgs/therm404.png}
+        \includegraphics[width=\textwidth]{imgs/therm404.png}
         \caption{$L=404$}
@@
     \includegraphics[width=0.6\textwidth]{imgs/coldstart.png}
-  \caption{\label{fig:coldstart} Comparison of cold and hot starts after 1000 sweep thermalization in $L=404$ lattice. 1000 measurements taken every 50 sweeps.}
+  \caption{\label{fig:coldstart} Histogram of lattice-averaged actions $S/L^2$ with hot and cold starts. 1,000 sweep thermalization in $L=404$ lattice, 1,000 measurements taken every 50 sweeps.}
 \end{figure}

### 2021-04-22 final-report.latex/results.tex
fix Chris' notes in Methods
second round of Chris' notes

@@
+Using the Monte Carlo procedure outlined in Sec.~\ref{sec:methods}, we run statistical simulations on the $\phi^4$ model and NLSM.
+\section{$\phi^4$ results}
 We initially implement the $\phi^4$ model to verify the results of our system.
-According to previous studies \cite{monahan2016, schaich2006}, the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ when $\lambda = 0.5$.
+According to previous studies \cite{chang1976, monahan2016, schaich2006}, the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ when $\lambda = 0.5$.
 We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the magnetic susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
-\begin{figure}[h]
+\begin{figure}[h!]
     \centering
-      \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
+      \includegraphics[width=0.9\textwidth]{imgs/phi4.png}
       \caption{\label{fig:phi4} The lattice average $|\langle\bar\phi\rangle|$, the lattice susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$.
-$L=32$, $\lambda=0.5$.
+$L=64$, $\lambda=0.5$.
 The lattice was thermalized from a hot start for 1000 sweeps.
-Afterwards, 100 measurements were taken with 100 sweeps between each.
+Afterwards, 1000 measurements were taken with 50 sweeps between each.
 The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
@@
     4-4y-8y^3-\frac{48}{5}y^5& \beta<1 \\
-    \frac{2}{\beta} + \frac{4\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2. 
+\frac{2}{\beta} + \frac{4}{\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2
 \end{cases}
 \end{equation}
+where 
+\begin{equation}
+    y=\mathrm{coth}\beta-\frac{1}{\beta}.
+\end{equation}
 We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{fig:bergluscher}.
-\begin{figure}[h]
+\begin{figure}[h!]
     \centering
@@
 \end{figure}
-These two charts show a high degree of agreement with the current literature.
+These two charts show a high degree of agreement with the current literature however there is a slight discrepancy, prehaps arising from different Monte Carlo methods.
 
-We also seek to confirm the results from Bietenholz et al.\cite{bietenholz2018}.
-Specifically, that the topological susceptibility $\chi_t$ diverges in the continuum limit even at finite flow time.
+We also seek to confirm the results from Bietenholz et al.
+\cite{bietenholz2018}.
+Specifically, we show the topological susceptibility $\chi_t$ diverges in the continuum limit even at finite flow time.
 Since $\chi_t$ is in units of inverse distance squared, we multiply by $\xi_2^2$, the square of the second moment correlation length, to achieve a scale-invariant value $\chi_t\xi_2^2$.
@@
 To begin the comparison, we plot $\chi_t\xi_2^2$ as a function of flow time $\tau$, shown in Fig.~\ref{fig:bietenholz}.
-\begin{figure}[h]
+\begin{figure}[h!]
     \centering
       \includegraphics[width=\textwidth]{imgs/bietenholz.png}
-      \caption{\label{fig:bietenholz} $\chi_t L^2$ as a  function of flow time $\tau$. Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization.}
+      \caption{\label{fig:bietenholz} $\chi_t L^2$ as a  function of flow time $\tau$. Simulation run with 10,000 measurements every 50 sweeps, 1,000 sweep thermalization.}
 \end{figure}
@@
 To analyze the divergence of $\chi_t$ in the continuum limit, we plot $\chi_t \xi_2^2$ as a function of lattice size $L$.
-We perform this simulation at flow time $\tau=5t_0$.
-\begin{figure}[h]
+We perform this simulation at flow times $\tau=0$ and $\tau=5t_0$ (Fig.~\ref{fig:divergence}).
+\begin{figure}[h!]
     \begin{center}
       \begin{subfigure}[b]{\textwidth}
-          \includegraphics[width=\textwidth]{imgs/divergence.png}
+          \centering
+          \includegraphics[width=0.8\textwidth]{imgs/divergence.png}
           \caption{$\tau = 0$}
@@
       \begin{subfigure}[b]{\textwidth}
-          \includegraphics[width=\textwidth]{imgs/divergence_flowed.png}
+          \centering
+          \includegraphics[width=0.8\textwidth]{imgs/divergence_flowed.png}
           \caption{$\tau = 5t_0$}
       \end{subfigure}
-      \caption{\label{fig:bietenholz} $\chi_t\xi_2^2$ as a function of $L$.
-The data is fit with both a log and power fit.
-Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization.
+      \caption{\label{fig:divergence} $\chi_t\xi_2^2$ as a function of $L$.
+The data is fit with both a logarithmic and power fit.
+Simulation run with 10,000 measurements, measurements every 50 sweeps, 1,000 sweep thermalization.
 In the $\tau=0$ case, we have compared our result with the curve fit found in \cite{bietenholz2018}.}
@@
 \begin{equation}
-    \chi_t \xi_2^2 = A \mathrm{log}(B L) + C
+    \chi_t \xi_2^2 = a \mathrm{log}(b L + c)
 \end{equation}
-and a power fit
+and a power law fit
 \begin{equation}
-    \chi_t \xi_2^2 = A B^L + C.
+    \chi_t \xi_2^2 = A L^B + C.
 \end{equation}
-The data fits these functions with $\chi^2/DOF$ of $31.3$ and $761.6$ respectively.
-Both of these function diverge as $L\rightarrow \infty$, indicating that the topological susceptibility does as well.
-This is still a slower divergence than the linear divergence featured in the $\tau=0$ case.
+We calculate the parameters to these functions using the \texttt{curve\_fit} tool in the \texttt{scipy} Python package \cite{virtanen2020}.
+When $\tau=5t_0$, the data fits these functions with $\chi^2/DOF$ of \logchi and \powchi respectively, indicating errors were underestimated.
+Both of these function diverge as $L\rightarrow \infty$, indicating that the topological susceptibility does as well in the continuum limit.
+Though there is a clear difference between the quantitative fit from \cite{bietenholz2018} and the fit calculated in this work, both demonstrate divergent behavior.
+This result supports the inherent divergence of $\chi_t$ in the continuum limit.
 
@@
 We perform this calculation for four values of the flow time $\tau$, shown in Fig~\ref{fig:theta}.
-\begin{figure}[h]
+\begin{figure}[h!]
     \centering
@@
 
-%\chapter{Statistical Analysis}
+\section{Implications}
+
+Berg \& L\"uscher \cite{berg1981} originally illuminated this discrepancy between the renormalization group hypothesis (that $\chi_t \rightarrow 0$) and the numerical results, providing three possible causes:
+\begin{enumerate}
+    \item The definition of the topological charge does not scale to the continuum.
+    \item There are ultraviolet divergences.
+    \item There is no reasonable continuum limit.
+\end{enumerate}
+Since the gradient flow suppresses ultraviolet fluctuations, the persistence of a divergent topological susceptibility under the gradient flow undermines the second option, thereby supporting the other two.

### 2021-04-22 final-report.latex/theory.tex
second round of Chris' notes

@@
 \subsection{Non-linear sigma model}
-
 The non-linear sigma model (NLSM) is a prototypical theory for a variety of unrelated physical phenomena, including applications in string theory \cite{callan1985} and ferromagnetism \cite{polyakov1975}.

### 2021-04-22 code/cpp_code/lattice.cpp
clean comments, fix random seed

@@ -1,5 +1,3 @@
-// lattice.cpp
-
 #include <iostream>
 #include "lattice.h"
 #include <algorithm>
@@ -77,7 +75,6 @@ double Lattice2D::full_action() {
 
 
 double Lattice2D::lagrangian(const Phi phi, const Phi nphi_sum) {
-    // S[phi] = beta/2 int (dphi . dphi)
     return beta * (D - phi * nphi_sum); // note that the sum over dimension has already been made
 }
 

### 2021-04-22 code/cpp_code/lattice.h

@@ -34,9 +34,8 @@ class Lattice2D {
         Lattice2D& operator*=(const double & factor);
         Lattice2D& operator/=(const double & factor);
         Lattice2D operator+ (const Lattice2D & other) const;
-        //Lattice2D operator+ (const Lattice2D & other) const;
 
-        // Iterator stuff
+        // Iterators 
         typedef datatype::iterator iterator;
         typedef datatype::const_iterator const_iterator;
 

### 2021-04-22 code/cpp_code/observables.h

@@ -44,20 +44,6 @@ namespace observables {
             const string name() const { return "chi_m"; };
     };
 
-    //class C : public BaseObservable {
-        //public:
-            //double operator()(const Lattice2D& lat) const {
-                //double val = 0;
-                //const Phi phi0 = lat[0];
-                ////for (const Phi& phi : as_const(lat)) {
-                //for (auto it = lat.cbegin(); it!=lat.cend(); ++it) {
-                    //val += (*it)*phi0;
-                //}
-                //return val;
-            //};
-            //const string name() const { return "chi_m"; };
-    //};
-
     class F : public BaseObservable {
         public:
             double operator()(const Lattice2D& lat) const {
@@ -65,11 +51,12 @@ namespace observables {
 
                 int x=0;
                 int y=0;
+                const double two_pi_L = 2*M_PI*lat.L;
 
                 //for (const Phi& phi : as_const(lat)) {
                 for (auto itx = lat.cbegin(); itx!=lat.cend(); ++itx) {
                     for (auto ity = lat.cbegin(); ity!=lat.cend(); ++ity) {
-                        val += (*itx)*(*ity) * cos( 2pi_L * (x - y));
+                        val += (*itx)*(*ity) * cos( two_pi_L * (x - y));
                         x++;
                         if (x==lat.L) x=0; // This ensures that x is the Euclidean space dimension
                     }

### 2021-04-22 code/cpp_code/phi.cpp

@@ -10,9 +10,6 @@ Phi::Phi(array<double, N> phi){
 
 void Phi::init_as_zero(){
     phi = {0,0,0};
-    //for (int i=0; i<N; i++) {
-        //phi[i] = 0;
-    //}
 }
 
 double Phi::norm_sq() const {
@@ -47,9 +44,6 @@ Phi& Phi::operator/=(const double & a) {
 
 Phi Phi::operator+ (const Phi & aPhi) const {
     Phi new_phi(phi);
-    //for (int i=0; i<N; i++) {
-        //new_phi[i] = phi[i] + aPhi[i];
-    //}
     new_phi += aPhi;
     return new_phi;
 }

### 2021-04-22 code/cpp_code/sweep.cpp

@@ -166,34 +166,11 @@ double Sweeper::rand_dist(double r) {
     return 3 * r - 1.5;
 }
 
-/*    Phi Sweeper::new_value(Phi old_phi) {*/
-
-
-    //Phi dphi;
-    //double sum_of_squares = 0;
-    //for (int i = 0; i<N-1; i++) {
-        //dphi[i] = rand_dist(randf());
-
-        //sum_of_squares += pow(old_phi[i] + dphi[i], 2);
-    //}
-    //dphi[N-1] = (2*randint(2) - 1) * sqrt(1-sum_of_squares) - old_phi[N-1]; //the first term randomizes the sign
-
-    //// test?
-    //cout << "Dot: " << (old_phi + dphi) * (old_phi + dphi) << endl;
-    //return dphi;
-/*}*/
-
 Phi Sweeper::new_value(Phi old_phi) {
-    //Phi dphi;
-    //dphi[0] = rand_dist(randf());
-    //return dphi;
     return random_phi();
 }
 
 Phi Sweeper::proj_vec () {
-    //for (int i = 0; i<N-1; i++) {
-        //r[i] = 
-    //}
     return Phi();
 }
 
@@ -222,7 +199,6 @@ void write_gif_frame(Lattice2D& lat, GifWriter* gif_writer, int delay, double ra
             vec[4*i + j] = static_cast<uint8_t>((aPhi[j]+1)*128);
         }
         vec[4*i + 3] = 255;
-        //cout << vec[0] <<" "<< vec[1] <<" "<< vec[2] <<" "<< vec[3] <<" "<< endl;
     }
     GifWriteFrame(gif_writer, vec.data(), DIM, DIM, delay);
 }
@@ -241,7 +217,6 @@ void Sweeper::assert_action(double tol) {
 
 void Sweeper::full_sweep(Recorder* recorder, const sweep_args& args = sweep_args()) {
    
-    //if (args.cluster_algorithm != WOLFF) throw invalid_argument("Currently, Wolff is the only allowed algorithm");
 
     shared_ptr<progress_bar> progress;
     progress = args.progress ? make_shared<progress_bar>(cout, 70u, "Working") :  nullptr; 
@@ -270,25 +245,18 @@ void Sweeper::full_sweep(Recorder* recorder, const sweep_args& args = sweep_args
             }
             broadcast_lattice();
 
-            //cout << "Met: " << action << " vs " << full_action() << endl;
             dS = sweep(color);
             collect_changes(dS, color);
[…]

### 2021-04-22 code/cpp_code/sweep.h

@@ -93,7 +93,6 @@ class Sweeper {
         const int DIM;
         Lattice2D lat;
 
-
         Sweeper();
         ~Sweeper();
         Sweeper(int DIM, MPI_Comm c);
@@ -121,13 +120,9 @@ class Sweeper {
 
 };
 
-
 double randf();
 int randint(int n);
 Phi sign(Phi x);
 double sign(double x);
-
-
-
 
 #endif

### 2021-04-22 code/lattice.py

@@ -1,0 +1,173 @@
+from functools import lru_cache
+from itertools import product
+from colorsys import hls_to_rgb
+from random import randrange
+import abc
+from .croutines import lagrangian, rand_dist
+from copy import copy
+import sys
+
+if sys.platform == 'darwin':
+    import matplotlib as mpl
+    mpl.rcParams['axes.formatter.useoffset'] = False
+    from matplotlib import pyplot as plt
+
+# from matplotlib import pyplot as plt
+import numpy as np
+from mpi4py import MPI
+
+from .keys import *
+
+COMM = MPI.COMM_WORLD
+RANK = COMM.Get_rank()
+SIZE = COMM.Get_size()
+
+REL_COORDS = ((1,0),(0,1),(-1,0),(0,-1))
+
+def show(data, figsize=(6,6), show=True):
+    plt.figure(figsize=figsize)
+    im = plt.imshow(c)
+    if show:
+        plt.show()
+    return im
+
+class Lattice(object, metaclass=abc.ABCMeta):
+
+    def __init__(self, data=None, dim=None):
+        self._construct(data, dim)
+
+
+    def _construct(self, data, dim):
+        if data is not None:
+            self.data = data
+            self.dim = data.shape[0]
+            self.size = data.size
+
+
+        elif dim is not None:
+            self.dim = dim
+            self.size = dim**2
+            if RANK==0:
+                self.data = np.vectorize(rand_dist) (np.random.rand(self.dim, self.dim)) # Hot start
+            else:
+                self.data = np.empty((self.dim, self.dim))
+
+        else:
+            raise Exception("dim and data cannot both be None")
+
+        self.sites = list(product(range(self.dim), repeat=2))
+        self.full_neighbors = {s:self.neighbors(s) for s in self.sites}
+        self.half_neighbors = {s:self.neighbors(s)[:2] for s in self.sites}
+
+        if RANK==0:
+            self.action = self.calculate_action()
+        else:
+            self.action = 0
+
+    def calculate_action(self):
+        action = 0
+        for c in self.sites:
+            action += self.lat_lagrangian(c)
+        return action
+
+    def __repr__(self):
+        return repr(self.data)
+
+    def __iter__(self):
+        return iter(self.data)
+
+    def __getitem__(self, k):
+        return self.data[k]
+
[…]

### 2021-04-22 code/leak.sh

@@ -1,0 +1,3 @@
+#!/bin/zsh
+
+./cpp_code/bin/sweep -i inputs/params_0.5.yml -o TEST.out

### 2021-04-22 code/mcmc/phi4.py

@@ -1,0 +1,88 @@
+from time import time, sleep
+import sys
+import os
+from getopt import getopt
+from mpi4py import MPI
+
+from lattice import Lattice, show
+from random_walk import RandomWalk
+from keys import *
+from recorder import Recorder
+import numpy as np
+from parallel_utils import COMM, RANK, SIZE
+
+
+if 'line_profiler' not in globals():
+    def profile(f):
+        return f
+
+MPI_RES_CHUNKSIZE = 0
+
+def main():
+    opts, args = getopt(sys.argv[1:],'pL:s:S')
+    dopts = dict(opts)
+
+    if '-s' in dopts:
+        seed = int(dopts['-s'])
+        seed(a=seed)
+        np.random.seed(seed)
+
+    L = int(dopts.get('-L', 64))
+    m02 = -0.68
+    lam = 0.5
+    sweeps = 100
+    cluster_method = None
+
+
+
+    l = Lattice(dim=L, m02=m02, l=lam)
+
+
+
+# Test gradient flow
+#     recorder = Recorder()
+
+    # def laplacian(a):
+        # return np.gradient(np.gradient(a, axis=0), axis=0) + np.gradient(np.gradient(a, axis=1), axis=1)
+
+    # last_rho = None
+    # for tau in np.linspace(0.,0.005,100):
+        # rho = l.rho(tau)
+        # recorder.save(rho)
+        # if last_rho is not None:
+            # dt_data = rho.data - last_rho.data
+            # grad_data = laplacian(rho.data)
+        # last_rho = rho
+    # recorder.plot(title=f"Effect of gradient flow on observables: $L={L}$, $\\lambda={lam}$, $\\mu_0^2={m02}$", fname="gradient_flow", xlabel=r'$\tau$ (flow time)', show=True)
+
+    # recorder.save_gif("plots/gradient_flow.gif", fps=30)
+
+    # quit() # just testing the gradient flow, so stop here
+
+    #set the site assignments (factor of 2 for checkerboard)
+
+
+    cluster_rate = 1
+    record_rate = 1
+    thermalization = 1
+
+    recorder = Recorder()
+    recorder.save(l)
+
+    rw = RandomWalk(l)
+
+    start = time()
+
+    rw.run(sweeps, cluster_method, cluster_rate, record_rate, thermalization, recorder)
+
+    if RANK==0:
+        exec_time = time() - start
+        print(f"{exec_time}")
+
[…]

### 2021-04-22 code/phi4_serial2.py

@@ -14,15 +14,6 @@ import multiprocessing as mp
 import pickle
 import inspect
 
-from matplotlib import rc, rcParams
-rc('text',usetex=True) #related to latex for the labels
-rcParams['font.size'] = 20.0
-rcParams['axes.titlesize'] = 22.0
-rcParams['xtick.labelsize'] = 20.0
-rcParams['ytick.labelsize'] = 20.0
-rcParams['legend.fontsize'] = 22.0
-letters = 24.0;
-
 if 'line_profiler' not in globals():
     def profile(f):
         return f
@@ -39,7 +30,7 @@ class Lattice(object):
         self.m = m
         self.l = l
 
-        self.__redefined_mass = 2 + 1/2 * self.m
+        self.__redefined_mass = 4 + 1/2 * self.m
         self.__quarter_lambda = 1/4 * self.l
 
         if data:
@@ -95,7 +86,6 @@ class Lattice(object):
         dphi = new_phi - phi
         self.data[i] = new_phi
         self._update_action(i, dphi)
-
 
     @profile
     def revert(self):
@@ -188,38 +178,14 @@ class Lattice(object):
         return sum(self) / self.size
 
     def susceptibility(self):
-        # m = self.magnetization()
-        # m2 = 1
-        # return (m2 - m**2)
-        cumsum = 0
-        for phia in self:
-            for phib in self:
-                cumsum += phia*phib
-
-        return cumsum / self.size
+        m = self.magnetization()
+        m2 = 1
+        return (m2 - m**2)
 
     def binder_cumulant(self):
         phi_sq = sum(phi**2 for phi in self)
         phi_qu = sum(phi**4 for phi in self)
         return 1 - phi_qu / (3 * phi_sq**2)
-
-    def flow_evolution(self, tau, action=False):
-        lat = np.resize(np.array(self.data), (self.dim, self.dim))
-        fft = np.fft.fft2(lat)
-        ps = np.concatenate([np.arange(self.dim//2), -np.arange(self.dim//2,0,-1)])
-        px, py = np.meshgrid(ps, ps)
-        new_lat = copy(self) # make shallow copy so as not to copy the entire field
-
-
-        p2 = px**2 + py**2
-
-        new_lat_np = np.real(np.fft.ifft2(np.exp(-tau*p2) * fft))
-        new_lat.data = np.resize(new_lat_np, (self.size,)).tolist()
-        if action:
[…]

### 2021-04-22 code/sweep.py

@@ -20,11 +20,11 @@ delay = 0.4
 cores = 1
 actions = []
 phibars = []
-# betas = [0.6, 0.8, 1.0,1.2, 1.4, 1.6, 1.8, 2.0, 2.2, 2.4, 2.6, 2.8, 3.0]
+betas = [0.5, 1.0, 1.1, 1.2, 1.3, 1.4, 1.5, 1.6, 1.7, 1.8, 1.9, 2.0, 2.5]
 orig_betas = [1.1, 1.2, 1.3, 1.4, 1.5, 1.6, 1.7, 1.8, 1.9, 2.0]
-betas = [0.5, 1.0, 1.5, 2.0]
+# betas = [0.5, 1.0, 1.5, 2.0]
 # betas = [1.4, 1.5, 1.6, 2.0]
-def o_filename(beta): return f"outputs/data_{beta}.csv"
+def o_filename(beta): return f"outputs/bergluscher/params{beta:.1f}..csv"
 def i_filename(beta): return f"inputs/params_{beta}.yml"
 o_filenames = [o_filename(beta) for beta in betas]
 processes = []
@@ -43,6 +43,7 @@ def parse_file(beta):
     data = []
     with open(o_filename(beta), 'r') as datafile:
         csv_data = csv.reader(datafile)
+        next(csv_data)
         for lines in csv_data:
             data.append([float(l) for l in lines])
     return np.array(data).T
@@ -69,12 +70,12 @@ class Observable(object):
 
 global_observables = dict(
     # magnetization =     Observable(r"\langle|\bar\phi|\rangle", lambda data: np.mean(np.abs(data[0])) ),
-    action          =   Observable(r"S", lambda data,beta: np.mean(data[0])/(beta*L**2) ),
-    internal_energy =   Observable(r"E", lambda data,beta: np.mean(data[0])/(0.5*beta*L**2) ),
-    chi_m           =   Observable(r"\chi_m", lambda data,beta: np.mean(data[1])/L**2, scale='log'),
+    action          =   Observable(r"S", lambda data,beta: np.mean(data[3])/(beta*L**2) ),
+    internal_energy =   Observable(r"E", lambda data,beta: np.mean(data[3])/(0.5*beta*L**2) ),
+    chi_m           =   Observable(r"\chi_m", lambda data,beta: np.mean(data[5])/L**2, scale='log'),
     delta_m         =   Observable(r"\delta_m",
                             lambda data,beta:
[…]

### 2021-04-23 final-report.latex/conclusion.tex
Final draft

@@
 
-In the $\phi^4$ model, we verified a phase transition at $m_0^2=-0.72$ using the magnitude of the average magnetization, the magnetic susceptibility, the Binder cumulant and the bimodality.
+In the $\phi^4$ model, we verified a phase transition at $m_0^2=-0.7$ using the magnitude of the average magnetization, the magnetic susceptibility, the Binder cumulant and the bimodality.
 This process verified the effectiveness of the Monte Carlo computational system.
-We then generalized our system to the $O(3)$ non-linear sigma model and transitioned to a \texttt{C++} code base.
+We then generalized our system to the $O(3)$ non-linear sigma model and transitioned to a C++ code base.
 We confirmed our calculations with known results by measuring the internal energy and magnetic susceptibility.
 We calculated the topological susceptibility $\chi_t$ to analyze its divergence in the continuum limit, confirming that this is indeed the case even at finite flow time.
-Specifically, the topological susceptibility under the gradient flow follows either a power-law relationship or a logarithmic relationship as a function of lattice size, both of which diverge as $L\rightarrow\infty$.
+Specifically, the topological susceptibility under the gradient flow follows either a power law relationship or a logarithmic relationship as a function of lattice size, both of which diverge as $L\rightarrow\infty$.
 At flow time $\tau=5t_0$, the $\chi^2/DOF$ goodness of fit values are of \powchi and \logchi respectively.
@@
 Finally, we demonstrated the relationship between the topological charge and the $\theta$ in the topological action.
-This plot demonstrates quantitatively the different divergences of the topological susceptibility, which is equivalent to the slope at $\theta=0$.
+This plot demonstrates quantitatively the different rates of divergence in the topological susceptibility.
 In the $\tau=0$ case, we see the slope rapidly approach positive infinity at $\theta=0$.
@@
 
-Though the logarithmic and power-law functions visually fit the susceptibility data well, the $\chi_2$ values are high.
+Though the logarithmic and power-law functions visually fit the susceptibility data well, the $\chi^2$ values are high.
 This imprecision can be attributed to underestimated errors.
@@
 While the Jackknife method accurately estimates statistical errors of the sample, we did not incorporate any systematic errors arising from the gradient flow calculation.
-Future work could reduce the tolerance of the adaptive step size algorithm to make this calculation more accurate, though this increases computational requirements substantially.
+Future work could reduce the tolerance of the adaptive step size algorithm to make this calculation more accurate, though this change increases computational requirements substantially.
 Furthermore, a larger number of lattice sizes may provide a more complete picture of the continuum limit.
@@
 
-Generally, this study has implications in both condensed matter and nuclear physics as the NLSM is used in both.
+Generally, this study has implications in both condensed matter and nuclear physics.
 Though the calculation of the $\chi_t$ divergence confirms existing literature, the relationship between the topological charge and $\theta$ in the flow time was previously unexplored.
-Beyond the convergence of $\chi_t$, this relationship has applications in condensed matter where different values of $\theta$ can describe spin-chains of fermions and bosons \cite{bogli2012}.
-Furthermore, the mass gap has a strong relationship with the $\theta$-terms in topological NLSM, featuring massive and massless regimes \cite{allessalom2008}.
+Beyond the convergence of $\chi_t$, this relationship has applications in condensed matter where different values of $\theta$ can describe spin-chains of either fermions or bosons \cite{bogli2012}.
+Furthermore, the mass gap has a strong relationship with the $\theta$-term in the topological NLSM, featuring massive and massless regimes \cite{allessalom2008}.

### 2021-04-23 final-report.latex/intro.tex
Final draft

@@
-Quantum field theory (QFT) is an framework used to describe a range of physical phenomena to a remarkable degree of accuracy.
+Quantum field theory (QFT) is a framework used to describe a range of physical phenomena to a remarkable degree of accuracy.
 Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics.
@@
 In general, the NLSM shares key features with non-Abelian gauge theories such as QCD, including a mass gap and asymptotic freedom \cite{polyakov1975}.
-Therefore, the NLSM is a useful model for exploring the effect of these properties using a simpler system.
+Therefore, the NLSM is a useful model for exploring the effect of these properties in a simpler system.
 
-In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time).
+In this study, we specifically consider the $O(3)$ NLSM in 1+1 dimensions (one dimension of space, one dimension of time).
 This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action in Euclidean space.
-Solutions such as these are ``topologically protected,'' meaning they cannot evolve into the vacuum state via small fluctuations.
-Therefore, it is critically important to take topology into account when considering quantum field theories in a cosmology and high energy physics \cite{goddard1986}.
+Solutions such as these are ``topologically protected'', meaning they cannot evolve into the vacuum state via small fluctuations.
+Due to this property, topology is critically important to quantum field theories in cosmology and high energy physics \cite{goddard1986}.
 Additionally, topological stability may become a key tool for fault-tolerant quantum computers \cite{kitaev1997}.
-In these devices, topology could protect the delicate quantum states necessary for information processing.
+In these devices, topology protects the delicate quantum states necessary for information processing.
 
 Since the NLSM in not perturbatively renormalizable, we require nonperturbative techniques to study topological effects.
-An alternative solution is to place the field on a discretized Euclidean lattice, a technique originally used for quantum chromodynamics \cite{wilson1974}.
-After this transformation, fields become computationally calculable.
+A solution is to place the field on a discretized Euclidean lattice, a technique originally used for quantum chromodynamics \cite{wilson1974}.
+After this transformation, fields become numerically calculable using modern computers.
 This process introduces the lattice spacing as a length scale $a$ where the \textit{continuum limit} is defined as taking $a$ to zero.
@@
 The topological susceptibility in the 1+1 $O(3)$ NLSM has been the subject of debate for the past four decades \cite{bietenholz2018} since it is unclear if a convergent solution exists.
-While the some analytical arguments argue the topological susceptibility should approach zero in the continuum limit, numerical results predict infinities.
-In QCD, mathematical techniques proved that the susceptibility vanishes in the continuum limit \cite{giusti2004}, a fact supported by numerical calculation \cite{giusti2004}.
+While the some analytical arguments argue the topological susceptibility should approach zero in the continuum limit, numerical results predict infinities \cite{berg1981}.
+In QCD, mathematical techniques proved that the susceptibility vanishes in the continuum limit \cite{giusti2004}, a fact supported by numerical calculation \cite{bruno2014}.
 
 To remedy this issue, we consider the gradient flow, a technique designed to remove divergences in operators.
-By dampening high-frequency fluctuations, the gradient flow reduces terms which scale with the inverse lattice spacing, making some observables finite on the lattice \cite{monahan2016}.
+By dampening high-frequency fluctuations, the gradient flow reduces terms that scale with the inverse lattice spacing, making some observables finite on the lattice \cite{monahan2016}.
 In QCD, the gradient flow successfully makes the topological susceptibility finite in numerical calculations \cite{bruno2014}, corroborating the analytical result in \cite{giusti2004}.
-This success has motivated the usage of the gradient flow to calculate the topological susceptibility in the 1+1 $O(3)$ NLSM, though the observable still diverges in the continuum limit \cite{bietenholz2018}.
+This success has motivated the usage of the gradient flow to calculate the topological susceptibility in the 1+1 $O(3)$ NLSM, though recent studies demonstrate that the observable still diverges in the continuum limit \cite{bietenholz2018}.
 
@@
 
-To numerically study the topological qualities of the non-linear sigma model, we first implement a Markov chain Monte Carlo simulation.
+To numerically study the topological qualities of the NLSM, we first implement a Markov chain Monte Carlo simulation.
 We initially construct a proof-of-concept Python program that models the simpler $\phi^4$ model (see Sec.~\ref{sec:phi4}).
-After comparing with existing literature, we transition to a C++ simulation for efficiency, implementing the non-linear sigma model in larger lattices.
-Since the gradient flow has no exact solution in the non-linear sigma model, we implement a numerical solution using a fourth-order Runge-Kutta approximation with automatic step sized.
+After comparing with existing literature, we transition to a C++ simulation for efficiency, implementing the NLSM on larger lattices.
+Since the gradient flow has no exact solution in the NLSM we implement a numerical solution using a fourth-order Runge-Kutta approximation with automatic step sizing.
 By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.
@@
 
-A portion of this work began during the summer of 2021 using funding from the 1693 Scholars Program.
+A portion of this work began during the summer of 2020 using funding from the 1693 Scholars Program.
 This preliminary research consisted of literature review and numerical tests with an Ising model simulation as well as an initial implementation of the $\phi^4$ model.
@@
     \item Throughout this paper, we use natural units, i.e. $\hbar = 1$ and $c=1$.
-    \item We use Einstein summation notation, an implicit sum over repeated spacetime indices. For example, if $x^\mu$ is a spacetime four-vector and $x_\mu$ is its covariant form, the term
+    \item We use Einstein summation notation, an implicit sum over repeated spacetime indices.
+For example, if $x^\mu$ is a four-vector in Minkowski spacetime and $x_\mu$ is its covariant form, the term
         \begin{align*}
@@
         \end{align*}
+
 \end{itemize}

### 2021-04-23 final-report.latex/main.tex
Final draft

@@
 \usepackage{amsmath}
+\usepackage{bbm}
 \usepackage{tabularx}
@@
   %\ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
+\DefineBibliographyStrings{english}{%
+  bibliography = {References},
+}
 
@@
 \paragraph
-\indent I would like to thank my advisor, Prof.
-Christopher Monahan.
+\indent I would like to thank my advisor, Professor Christopher Monahan.
 His constant helpfulness has made this research possible and enjoyable.
 I would also like to thank the entire faculty of the William \& Mary Physics Department.
-The amount of knowledge and wisdom they have bestowed upon me has been beyond my imagination.
+The amount of knowledge and wisdom they have given me has been beyond my imagination.
 
-I also acknowledge the help the High Performance Computing group at William \& Mary, whose cluster performed many of these simulations.
-And finally, I would like to acknowledge the financial help of the 1693 Scholars Program which funded preliminary research over the summer.
+I would like to acknowledge the help of the High Performance Computing group at William \& Mary, whose cluster performed many of these simulations.
+I also want to acknowledge the financial help of the 1693 Scholars Program which funded preliminary research over the summer.
+
+Finally, I would like to thank my friends and family who helped me reach this point in my academic and personal journey.
 
@@
 
-\printbibliography
+\printbibliography[heading=bibintoc]
 

### 2021-04-23 final-report.latex/results.tex
Final draft

@@
 Using the Monte Carlo procedure outlined in Sec.~\ref{sec:methods}, we run statistical simulations on the $\phi^4$ model and NLSM.
-\section{$\phi^4$ results}
+\section{$\phi^4$ Results}
 We initially implement the $\phi^4$ model to verify the results of our system.
-According to previous studies \cite{chang1976, monahan2016, schaich2006}, the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ when $\lambda = 0.5$.
-We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the magnetic susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
+According to previous studies \cite{chang1976, monahan2016, schaich2006}, the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, transitioning at $m_0^2 = -0.72$ when $\lambda = 0.5$.
+We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the magnetic susceptibility $\chi_m$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
 \begin{figure}[h!]
@@
       \includegraphics[width=0.9\textwidth]{imgs/phi4.png}
-      \caption{\label{fig:phi4} The lattice average $|\langle\bar\phi\rangle|$, the lattice susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$.
+      \caption{\label{fig:phi4} The lattice average $|\langle\bar\phi\rangle|$, the magnetic susceptibility $\chi_m$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$.
 $L=64$, $\lambda=0.5$.
@@
 Afterwards, 1000 measurements were taken with 50 sweeps between each.
-The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
+The red horizontal line indicates $U=2/3$, the broken phase limit of the Binder cumulant.}
 \end{figure}
-This figure confirms a phase transition near $m_0^2=-0.72$ and supports the accuracy of our model.
+This figure confirms a phase transition near $m_0^2=-0.7$ and supports the accuracy of our model.
 
-\section{Non-linear sigma model results}
+\section{Non-linear Sigma Model Results}
 After confirming the phase transition in the $\phi^4$ model, we turn to the NLSM.
 In order to confirm the accuracy of the model, we compare results with existing literature.
-We first compare our results to Berg \& L\"uscher \cite{berg1981}, specifically measuring the internal energy and magnetic susceptibility.
+We first consider the results of Berg \& L\"uscher \cite{berg1981}, specifically the internal energy and magnetic susceptibility.
 
@@
 Additionally, we use a parameter $t_0$ to scale the flow time such that $t_0\sim L^2$.
-In our Monte Carlo simulation, we adapt the same values as \cite{bietenholz2018} for $\xi_2$.
+In our Monte Carlo simulation, we utilize the same values as \cite{bietenholz2018} for $\xi_2$, $\beta$ and $t_0$.
 
@@
       \caption{\label{fig:divergence} $\chi_t\xi_2^2$ as a function of $L$.
-The data is fit with both a logarithmic and power fit.
-Simulation run with 10,000 measurements, measurements every 50 sweeps, 1,000 sweep thermalization.
+We fit the data with both a logarithmic and power fit.
+Simulation run with 10,000 measurements, once every 50 sweeps, 1,000 sweep thermalization.
 In the $\tau=0$ case, we have compared our result with the curve fit found in \cite{bietenholz2018}.}
@@
 \end{figure}
-The data is fit with two options: a log fit
+We fit the data with two options: a log fit
 \begin{equation}
@@
 \begin{equation}
-    \chi_t \xi_2^2 = A L^B + C.
+    \chi_t \xi_2^2 = a L^b + c.
 \end{equation}
@@
 When $\tau=5t_0$, the data fits these functions with $\chi^2/DOF$ of \logchi and \powchi respectively, indicating errors were underestimated.
-Both of these function diverge as $L\rightarrow \infty$, indicating that the topological susceptibility does as well in the continuum limit.
+Both of these functions diverge as $L\rightarrow \infty$, indicating that the topological susceptibility also diverges in the continuum limit.
 Though there is a clear difference between the quantitative fit from \cite{bietenholz2018} and the fit calculated in this work, both demonstrate divergent behavior.
@@
 
-\subsection{Topological charge when $\theta\neq 0$}
+\subsection{Topological Charge when $\theta\neq 0$}
+
 Following the method explained in Sec.~\ref{sec:topotheta}, we calculate the imaginary part of $\langle Q \rangle$ for arbitrary $\theta$.
-We perform this calculation for four values of the flow time $\tau$, shown in Fig~\ref{fig:theta}.
+We perform this calculation for three values of the flow time $\tau$, shown in Fig~\ref{fig:theta}.
 \begin{figure}[h!]
@@
 These plots demonstrate the divergence of the continuum limit in the $\tau=0$ and the flowed regimes.
-In the $\tau=0$ case, the slope increases sharply, reflecting the linear divergence of $\chi_t$.
-However in the flowed regime, this divergence is much slower, reflecting the logarithmic or power-law relationship shown in Fig.~\ref{fig:bietenholz}.
+In the $\tau=0$ case, the slope increases sharply, reflecting the rapid divergence of $\chi_t$.
+However in the flowed regime, this divergence is much slower, reflecting the decreased values of $\chi_t$ shown in Fig.~\ref{fig:bietenholz}.
 

### 2021-04-23 final-report.latex/theory.tex
Final draft

--- 2021-04-19 final-report.latex/theory.tex
@@
 What differentiates a quantum field from a classical field is superposition: where classical fields have a definite configuration, quantum fields exist in a superposition of all possible configurations.
-It is possible---though nontrivial and outside the scope of this description--- to motivate the appearance of discrete particles from this superposition \cite{zee2010}.
-This formulation of QFT is known as the ``path integral formulation'', which differs from the ``second quantization'' used by most textbooks.
+It is possible---though nontrivial and outside the scope of this description--- to motivate the appearance of discrete particles from this superposition (see \cite{zee2010}).
+This formulation of QFT is known as the ``path integral formulation'', which differs from the ``second quantization'' used by many textbooks.
 
@@
 
-We can model a quantum field theory as a superposition of all possible classical fields.
-Like single-particle quantum mechanics, each configuration has a probability amplitude.
+We can model a quantum field as a superposition of all possible classical fields.
+Like single-particle quantum mechanics, each configuration has a complex probability amplitude.
 To measure expectation values of observables, we simply take an average over all configurations weighted by this complex amplitude.
@@
 To remedy this issue, we perform a ``Wick rotation'' which shifts spacetime into Euclidean coordinates.
-In normal spacetime, defined by the Minkowski metric, the Lorentz-invariant distance is given as
+In physical spacetime, defined by the Minkowski metric, the Lorentz-invariant distance is given as
 \begin{equation}
@@
 \end{equation}
-is therefore invariant under $SO(4)$ transformations, which is representative of a four-dimensional Euclidean space. Furthermore, we find that
+is invariant under $SO(4)$ transformations and is therefore representative of a four-dimensional Euclidean space. Furthermore, we find that
 \begin{align}
@@
     &= i d^3\vec{x}dx_0 \nonumber \\
-    &= i d^4x. \label{eq:wickdifferential} % TODO: is this better?
+    &= i d^4x. \label{eq:wickdifferential} 
 \end{align}
 
-We can use this transformation to redefine the Lagrangian $\mathcal{L}$ in Euclidean space as $\mathcal{L}_E$, replacing all $x_0$ with $ix_0$.
+We can use this transformation to redefine the Lagrangian $\mathcal{L}$ in Euclidean space as $\mathcal{L}_E$, replacing all $x_0$ with $-ix_4$ and flipping the overall sign.
 Since a Lorentz-invariant Lagrangian must only include even powers and derivatives of $x$, the Euclidean Lagrangian remains real.
@@
 \begin{align}
-    S_E &= \int d^4x_E \mathcal{L}_E \\
-    &= i \int d^4x \mathcal{L}_E,
+    S_E &= \int d^4x_E \mathcal{L}_E \nonumber\\
+        &= i \int d^4x \left(-\mathcal{L}\right) \nonumber \\
+    &= -i S
 \end{align}
@@
     \label{eq:pathintegraleuclidean}
-    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O (\phi)\; e^{-S_E[\phi]}.
+    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; e^{-S_E[\phi]}.
 \end{equation}
@@
 
-\subsection{$\phi^4$ model}
+\subsection{$\phi^4$ Model}
 \label{sec:phi4}
@@
 
-In this study, we specifically consider fields in 1+1 spacetime dimensions.
-Following Sec.~\ref{sec:pathintegral}, we convert Eq.~\ref{eq:phi4 action} from 3+1 Minkowski spacetime to 1+1 Euclidean spacetime, yielding
+In this study, we specifically consider fields in $1+1$ spacetime dimensions.
+Following Sec.~\ref{sec:pathintegral}, we convert Eq.~\ref{eq:phi4 action} from 3+1 Minkowski spacetime to $1+1$ Euclidean spacetime, yielding
 \begin{equation}
@@
 \end{equation}
-where $\partial_t$ and $\partial_x$ are the two partial derivatives in 1+1 Euclidean spacetime.
+where $\partial_t$ and $\partial_x$ are the two partial derivatives in $1+1$ Euclidean spacetime.
 
-This field features spontaneous symmetry breaking at a critical value of $m_0^2$.
-In practice this property causes the field to spontaneously align, similarly to spins aligning in a ferromagnet.
+This field features spontaneous symmetry breaking at a critical value of $m_0^2$ \cite{chang1976}.
+This property causes the field to spontaneously align, similarly to spins aligning in a ferromagnet.
 The name ``symmetry breaking'' refers to the transformation $\phi\rightarrow-\phi$, which changes the values of observables in the aligned regime but not the disordered regime.
@@
 
-\subsection{Non-linear sigma model}
-
-The non-linear sigma model (NLSM) is a prototypical theory for a variety of unrelated physical phenomena, including applications in string theory \cite{callan1985} and ferromagnetism \cite{polyakov1975}.
+\subsection{Non-linear Sigma Model}
+The non-linear sigma model (NLSM) is a prototypical theory for a variety of physical phenomena including applications in string theory \cite{callan1985} and ferromagnetism \cite{polyakov1975}.
 As a simple nonperturbative model, it also provides an ideal starting point for lattice QCD studies.
-Specifically, the NLSM exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
+Specifically, the NLSM exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability \cite{polyakov1975}.
 
 Unlike the $\phi^4$ model, which consists of a real value at each point in spacetime, the $O(3)$ NLSM consists of a 3D unit vector at each point.
-For this reason, every transformation of the field must be ``norm preserving''.
-Per its name, the $O(3)$ NLSM features a global symmetry under the 3D orthogonal group $O(3)$.
-In other words, the theory remains the same if all vectors are rotated equivalently.
+For this reason, every transformation of the field must be norm preserving.
+Per its name, the $O(3)$ NLSM features a global symmetry under the 3D orthogonal group $O(3)$; in other words, the theory remains the same if all vectors are rotated equivalently.
 To differentiate it from the $\phi^4$ model, we denote the NLSM field as $\e(x)$.
 
-The theory is defined by the 1+1 dimensional Euclidean action 
+The theory is defined by the $1+1$ dimensional Euclidean action 
 \begin{equation}
     \label{eq:nlsm euclidean action}
-    S_E = \frac{\beta}{2} \int d^2x \; \partial^\mu \e \cdot \partial_\mu \e
+    % S_E = \frac{\beta}{2} \int d^2x \; \partial^\mu \e \cdot \partial_\mu \e
+    S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right]
 \end{equation}
@@
 
-To accomplish a statistical analysis of quantum fields, we use a Monte Carlo simulation.
-This method amounts to producing a large number of configurations and calculating statistics on the sample.
-A brute-force calculation over all possible configurations---as Eq.~\ref{eq:pathintegraleuclidean} suggests--- is clearly infinite and computationally infeasible.
+To accomplish a statistical analysis of quantum fields, we use a Monte Carlo simulation which produces a large number of configurations and calculates statistics on the sample.
+A brute-force calculation over all possible configurations, as Eq.~\ref{eq:pathintegraleuclidean} suggests, is clearly infinite and computationally infeasible.
 However, the exponential nature of the Boltzmann factor dictates that only configurations near the action minima contribute to observable statistics.
@@
 \subsection{The Markov Chain}
-The Markov chain is an effective method to identity field configurations that minimize the action.
+In order to determine this subset of configurations, we use a Markov chain.
+This method identities field configurations that minimize the action using a random walk through phase space.
 Essentially, we begin with a predetermined configuration and then make small adjustments, gradually lowering the action.
@@
 Each step consists of two parts: proposing a change and accepting the new configuration.
-The proposal creates a new configuration $\phi_b$ based on the current configuration $\phi_a$.
-In order to direct the Markov chain towards the action minima, we do not immediately accept this change.
-We calculate a probability
+The proposal creates a new configuration $\phi_b$ based on the current configuration $\phi_a$, a change that we accept with probability
 \begin{equation}
-P(\phi_a \rightarrow \phi_b)
+P(\phi_a \rightarrow \phi_b).
 \end{equation}
-of adding $\phi_b$ to the Markov chain, depending only on the actions of each configuration. This stochastic process ensures the chain will seek action minima.
+This probability determines if $\phi_b$ should be added to the Markov chain and depends on the change in action, stochastically ensuring that the chain will seek the action minima.
 
@@
 In other words, the probability of a $\phi_a\rightarrow\phi_b$ transition must be equal to the probability of a $\phi_b\rightarrow\phi_a$ transition.
-Mathematically, this condition takes the form of \textit{detailed balance equations}:
+Mathematically, this condition takes the form of a ``detailed balance equation'':
 \begin{equation}
@@
 \label{sec:primary observables}
-Each primary observable is defined on each configuration independently, meaning they do not encode ensemble statistics of the Markov Chain.\footnote{One slight exception is the magnetic susceptibility, which can be defined both ways.} In the $\phi^4$ model, we can develop an intuition around these quantities by visualizing the symmetric and broken phases.
+Each primary observable is defined on each configuration independently, meaning they do not encode ensemble statistics of the Markov chain.
+In the $\phi^4$ model, we can develop an intuition around these quantities by visualizing the symmetric and broken phases.
 Fig~\ref{fig:primary observables} and Tab.~\ref{tab:primary observables} show examples of these quantities in three different configurations: one in the broken phase, one in the symmetric phase, and one at the transition.
@@
         \includegraphics[width=0.6\textwidth]{imgs/broken.png}
-        \caption{broken phase ($m_0^2=-1.0$)}
+        \caption{broken\\($m_0^2=-1.0$)}
       \end{subfigure}%
@@
         \includegraphics[width=0.6\textwidth]{imgs/transition.png}
-        \caption{transition ($m_0^2=-0.7$)}
+        \caption{transition\\($m_0^2=-0.7$)}
       \end{subfigure}%
@@
         \includegraphics[width=0.6\textwidth]{imgs/symmetric.png}
-        \caption{symmetric phase ($m_0^2=-0.4$)}
+        \caption{symmetric\\($m_0^2=-0.4$)}
       \end{subfigure}
@@
       \caption{\label{fig:primary observables} Visualization of broken phase, symmetric phase and transition.
-Simulation run on $64\times64$ lattice, plotted after 1000 sweep thermalization (see Sec~\ref{sec:thermalization})}
+Simulation run on $64\times64$ lattice, plotted after 1000 sweep thermalization (see Sec~\ref{sec:thermalization}), $\lambda=0.5$.}
   \end{center}
@@
 
-Due to the $\phi\rightarrow-\phi$ symmetry of the $\phi^4$ model, the ensemble mean of the average magnetization $\langle \phi \rangle$ is $0$.
-Likewise, the $O(3)$ symmetry in the NLSM enforces $\langle \e \rangle$ = 0.
-To measure the alignment, we therefore use the magnitude of this quantity, defined in the $\phi^4$ model as
+Due to the $\phi\rightarrow-\phi$ symmetry of the $\phi^4$ model, the ensemble mean of the average magnetization $\langle \bar\phi \rangle$ is $0$.
+Likewise, the $O(3)$ symmetry in the NLSM enforces $\langle \bar \e\, \rangle$ = 0.
+To measure the alignment, we therefore use the magnitude of this quantity, defined in the 1+1 $\phi^4$ model as
 \begin{equation}
-|\bar\phi| \equiv \frac{1}{V}\left| \int d^2x \phi(x)\right|.
+|\bar\phi| \equiv \frac{1}{V}\left| \int d^2x \;\phi(x)\right|.
 \end{equation}
@@
 \begin{equation}
-    \label{eq:chim nlsm}
-    |\bar\e| \equiv \frac{1}{V}\left|\int d^2x \e(x)\right|.
+    |\bar\e\,| \equiv \frac{1}{V}\left|\int d^2x \;\e(x)\right|.
 \end{equation}
 
-In the symmetric phase, both $\langle \bar\e\, \rangle$ and $\langle \bar\phi \rangle = 0$ while in the broken phase they are both nonzero.
+In the broken phase, both $\langle |\bar\e\, |\rangle$ and $\langle |\bar\phi| \rangle$ are nonzero but vanish in the symmetric phase.
 
@@
 \end{equation}
-in the NLSM.
+in the NLSM. We use this metric to compare our simulation with existing literature.\footnote{The internal energy is not part of the $\phi^4$ portion of this work.}
 
@@
 We define three secondary observables: the magnetic susceptibility, the Binder cumulant and the bimodality.
-Since the Binder cumulant and the bimodality provide different perspectives on phase transitions, we will restrict our usage of these observables to the $\phi^4$ model.
+Since the Binder cumulant and the bimodality primarily provide different perspectives on phase transitions, we will restrict our usage of these observables to the $\phi^4$ model.
 
 \subsubsection{Magnetic Susceptibility}
-Though the magnitude of the average magnetization is the main phase transition indicator, its behaviour on the lattice makes the critical point difficult to identify.
-With a finite lattice spacing $a$, the transition becomes smoother around the critical point.
+Though the magnitude of the average magnetization is the main phase transition indicator, the values become smooth around the critical point as the lattice spacing increases.
+This behaviour on the lattice makes the critical point difficult to identify.
 An alternative metric is the magnetic susceptibility.
@@
 \begin{equation}
-    \chi_m \equiv V \big(\langle {\bar\e\,}^2 \rangle - \langle \bar\e\, \rangle^2\big)
+\chi_m \equiv V \big(\langle {\bar\e}\,^2 \rangle - \langle \bar\e\, \rangle^2\big)
 \end{equation}
@@
 \end{equation}
-in the $\phi^4$ model. Manifestly, this expression appears as a secondary observable since it is not defined for each lattice configuration.
+in the $\phi^4$ model. Following the symmetry argument in Sec.~\ref{sec:avg mag}, the second terms vanish and the susceptibility becomes 
+\begin{equation}
+    \label{eq:chim phi4}
+\chi_m = V \langle {\bar\e}\,^2 \rangle 
+\end{equation}
+in the NLSM and
+\begin{equation}
+    \label{eq:chim nlsm}
+    \chi_m = V \langle {\bar\phi^2} \rangle
+\end{equation}
+in the $\phi^4$ model.
 
@@
 \begin{equation}
-    U \equiv 1-\frac{\langle \bar{\phi^4} \rangle}{3\langle \bar{\phi^2}\rangle^2}.
+U \equiv 1-\frac{\langle {\bar\phi}^4 \rangle}{3\langle {\bar\phi}^2\rangle^2}.
 \end{equation}
 Similar to the magnitude of the average magnetization, this formula yields $0$ in the symmetric phase and a nonzero value in the broken phase.
-This nonzero value is $U = 2/3$ for the Binder cumulant.
-The advantage of this metric is a fixed point with regards to a scaling transformation that corresponds with the critical point of the phase transition \cite{landau2000}.
-In our study, we use the Binder cumulant as further evidence of a phase transition.
+The Binder cumulant of the broken phase exhibits a universal value of $U = 2/3$.
+The advantage of this metric is a fixed point of the scaling transformation which corresponds with the critical point of the phase transition \cite{landau2000}.
+In other words, the Binder cumulants for different length scales should all intersect at the critical point.
+In our study, we use the Binder cumulant as further evidence of a phase transition in the $\phi^4$ model.
 
 \subsection{Bimodality}
-The final phase transition indicator we use is the bimodality.
+The final phase transition indicator that we use is the bimodality.
 In the symmetric phase, the average magnetization $\bar\phi$ centers around $0$ while in the broken phase, these values cluster around two peaks.
-Qualitatively, this metric measures the separation of these peaks.
+This metric quantitatively measures the separation of these peaks.
 
-We begin by measuring $\bar\phi$ for each configuration.
+To calculate the bimodality, we begin by measuring $\bar\phi$ for each configuration.
 We separate each value into an odd number number of bins, ensuring that there is a bin centered at $\bar\phi=0$.
@@
 in the broken phase, this value becomes $B=1$.
+In this study, we separate $\phi$ into bins of width $\Delta \phi = 0.1$.
 \subsection{Jackknife Method}
@@
 While we could propagate the uncertainty of the Binder cumulant, such a process is not clear for the bimodality.
-Therefore, we utilize a method known as Jackknife resampling.
+Therefore, we utilize a method known as Jackknife resampling to calculate statistical errors of secondary observables.
 
-We begin by calculating some observable $O$ on an ensemble of $N$ configurations.
-Then, for each configuration $i$, we calculate the same observable but exclude said configuration.
-This leaves us with a set of $N$ observables $O_i$.
-Assuming independent measurements, we can calculate the variance of $O$ as
+We begin by calculating the expectation value $O$ of some observable $\hat O$ on an ensemble of $N$ configurations.
+Then, for each configuration $i$, we calculate the expectation value of $\hat O$ while excluding said configuration, calling this quantity $O_i$.
+This leaves us with a set of $N$ expectation values $O_i$.
+Assuming independent measurements, we can calculate the variance of $\hat O$ as
 \begin{equation}
-    \mathrm{Var}(O) = \sum_{i} \left(C_i - C\right)^2.
+    \mathrm{Var}(\hat O) = \sum_{i} \left(O_i - O\right)^2.
 \end{equation}
-We use this formula to calculate all uncertainties in this study.
+We use this formula to calculate all statistical errors in this study.
 \section{Topological Observables}
 \label{sec:topological charge}
-The $O(3)$ NLSM features topological features originating from two properties: 
+The 1+1 $O(3)$ NLSM features topological features originating from two properties: 
 \begin{enumerate}
     \item At $x\rightarrow\infty$, the field must become uniform since the Lagrangian must vanish.
-This allows us to model $x\rightarrow\infty$ as a single point on the field, forming a Riemann sphere in three dimensions.
-    \item The elements of the $O(3)$ non-linear sigma model are three-dimensional unit vectors, thereby existing on a three dimensional unit sphere. 
+This allows us to model $x\rightarrow\infty$ as a single point on the field, forming a Riemann sphere with a two-dimensional surface.
+    \item The elements of the $O(3)$ NLSM are three-dimensional unit vectors, and thus also exist on a sphere. 
 \end{enumerate}
 
-With these two properties, we can view the field as a continuous mapping between two 3D spheres, denoted as $S^2$, and associate an integer number of wrappings to each mapping from $S^2$ to $S^2$.
+With these two properties, we can view the field as a continuous mapping between two three-dimensional spheres, each denoted as $S^2$, and associate an integer number of wrappings to each mapping from $S^2$ to $S^2$.
 We can envision a tangible metaphor for this wrapping with a balloon and a baseball: by simply inserting the baseball into the balloon, we have established a mapping from every point on the balloon to every point on the baseball.
@@
 In a purely mathematical world, we perform this process an infinite number of times, thereby associating every possible mapping with an integer.
-The group of integers is known as the \textit{homotopy group} of the non-linear sigma model.
+The group of integers is known as the \textit{homotopy group} of the $1+1$ $O(3)$ NLSM.
 We associate every field configuration with an element of this group, known as the \textit{topological charge}, which we denote as $Q$.
@@
 \end{equation}
-In the trivial case, $\langle Q \rangle$ disappears and   
+This quantity relates to the stability of topological phases and we expect it to approach zero in the continuum limit \cite{berg1981}.
+\subsection{NLSM $\theta$ term}
+The NLSM is invariant under the transformation $\e \rightarrow -\e$, a change that flips the sign of the topological charge.
+This implies that $\langle Q \rangle$ disappears and therefore, that the NLSM vacuum is topologically trivial.
+We can construct a nontrivial vacuum by introducing a $\theta$ term into the action:
 \begin{equation}
-    \chi_t = \frac{\langle Q^2 \rangle - \langle}{L^2}.
+    S[\e\,] \rightarrow S[\e\,] - i \theta Q[\e\,].
 \end{equation}
-% TODO Highlight this is true for QCD and NLSM. Do you know why?
+With this topological action, $\langle Q \rangle = 0$ if and only if $\theta=0$.
 
-\subsection{NLSM $\theta$ term}
-In the NLSM, we expect $\langle Q \rangle=0$. By introducing a $\theta$ term into the action,
+We can also redefine the topological susceptibility for the $\theta=0$ case as
 \begin{equation}
-    S[\e] \rightarrow S[\e] - i \theta Q[\phi],
+    \label{eq:chit simple}
+    \chi_t = \frac{\langle Q^2 \rangle}{L^2}.
 \end{equation}
-we can construct a topologically nontrivial model (that is, $\langle Q \rangle \neq 0$).
 
@@
 
-In Section~\ref{sec:pathintegral},  we defined a fundamental equation of quantum fields using a ``path integral'' which encompasses an uncountably infinite configuration space.
+In Section~\ref{sec:pathintegral},  we defined a fundamental equation of quantum fields using a path integral which encompasses an uncountably infinite configuration space.
 However, we said nothing of the integral's convergence.
@@
 The remedy to this catastrophe is unintuitive.
-Essentially, we adopt infinite values for the parameters of the Lagrangian ($m_0^2$ and $\lambda$ in $\phi^4$ theory).
+Essentially, we adopt infinite values for the parameters of the Lagrangian ($m_0^2$ and $\lambda$ in $\phi^4$ theory and $\beta$ in the NLSM).
 Since neither of these two quantities is ever measured directly, we do not have to assume that their values are finite.
-In practice, this technique is involved and consists of two steps: regularization and renormalization.
+In practice, this technique consists of two steps: regularization and renormalization.
 
 \subsection{Regularization}
+
 Regularization is a process which introduces a new parameter into calculations. One example is a momentum cutoff. This technique transforms infinite momentum integrals as follows:  
@@
 This process makes results $\Lambda$-dependent, but finite.
-Another example is dimensional regularization, which calculates results in terms of the spacetime dimension $D$ and analytically continues this parameter from just the integers into the real numbers.
+Another example is dimensional regularization, which calculates results in terms of the spacetime dimension $D$ and analytically continues this parameter from the integers into the real numbers.
 
-In this study, we employ lattice regularization.
-This process discretizes the field, modeling the field $\phi(x)$ as a lattice $\phi_i$ where $i$ indexes lattice sites.
+In this study, we employ a third technique: lattice regularization.
+This process discretizes the field, modeling $\phi(x)$ as a lattice $\phi(x_i)$ where $i$ indexes lattice sites.
 The inherent parameter in this case is the lattice spacing $a$ which measures the width of each lattice chunk.
@@
 According to Bloch's theorem, any mode above this cutoff is equivalent to a lower-momentum mode since the high frequency information disappears on a discrete lattice.
-
 We can view this cutoff in momentum space by considering a square with side length $2\pi/a$ centered at the origin.
@@
 This area is known as a ``Brillouin zone'' and contains all possible momentum modes on the lattice, effectively imposing a hard cutoff.
+
 One of the main strengths of lattice regularization is preservation of gauge invariance, a property that makes lattice methods useful for QCD.
@@
 \subsection{Renormalization}
-After regularization, we redefine the Lagrangian parameters in terms of the regularization parameter using a handful of boundary conditions.
+After regularization, we redefine the Lagrangian parameters in terms of the new parameter using a handful of boundary conditions.
 Following \cite{bietenholz2018}, we require that $L/\xi$ remains constant, where $L$ is the side length of the system and $\xi$ is the coherence length.
@@
 In the case of the NLSM it is impossible using counter-terms but can be performed numerically.
-In this study, we use predetermined values from \cite{bietenholz2018}.
-% TODO: cite this
+In this study, we use the second moment of the correlation length, notated $\xi_2$, which more precisely estimates the $\xi$ on the lattice \cite{bietenholz2018}.
+The specific values of $\beta$ and $\xi_2$ used in this work are taken from \cite{bietenholz2018}.
 
-At this point, we can calculate renormalized parameters of the Lagrangian.
 To achieve a physical theory, we take the limit as the regularization parameters approach their physical values.
-With a momentum cutoff, we take $\Lambda \rightarrow \infty$ and with dimensional regularization we usually take $d\rightarrow 4$.
+With a momentum cutoff, we take $\Lambda \rightarrow \infty$ and with dimensional regularization we usually take $D\rightarrow 4$.
 With lattice regularization, we approach the continuum, taking the lattice spacing $a\rightarrow 0$.
@@
 External operators may also diverge due to regularization procedures.
+As we decrease the width of each lattice site, high frequency modes become more significant, leading to ultraviolet divergences.
 A prototypical example is the angular momentum operator on the lattice.
@@
 
-The topological susceptibility $\chi_t$ is one such value that diverges in the continuum limit.
-As we decrease the width of each lattice site, high frequency modes become more significant, leading to an ultraviolet divergence in the operator.
+The topological susceptibility $\chi_t$ is one such value that diverges in the continuum limit, though the reasons for this divergence are not fully understood \cite{berg1981}.
+This question is central to this work.
 
@@
 \label{sec:gradflow}
-To remove this ultraviolet divergence, we adopt a technique is ``smearing'', a local averaging of the field \cite{solbrig2007}.
-Specifically, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.
+To remove this ultraviolet divergence, we adopt a technique known as ``smearing'', a local averaging of the field \cite{solbrig2008}.
+Specifically, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension\footnote{The term ``half-dimension'' indicates that the flow time is exclusively positive.} called ``flow time''.
+The flow time $\tau$ parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.
 
@@
 \end{equation}
-where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime and $\tau$ is the flow time.
+where $\partial^2$ is the Laplacian in 2D Euclidean spacetime\footnote{Explicitly, $\partial^2 = \frac{\partial^2}{\partial t^2} + \nabla^2$} and $\tau$ is the flow time.
 Here, $\rho$ is the field flowed into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$.
@@
 \begin{equation}
-    \rho(\tau, x) = \frac{1}{4 \pi \tau} \int d^2 y e^{-(x-y)^2/4\tau} \phi(y).
+    \rho(\tau, x) = \frac{1}{4 \pi \tau} \int d^2 y \; e^{-(x-y)^2/4\tau} \phi(y).
 \end{equation}
@@
     \label{eq:nsm_gradflow}
-    \partial_\tau \vec\rho (\tau,x) = \left( 1 - \vec\rho(\tau,x) \vec\rho(\tau,x)^T \right) \partial^2 \vec\rho(\tau,x),
+  \partial_\tau \e (\tau,x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T \right) \partial^2 \e(\tau,x).
 \end{equation}
-where $\partial^2$ is the Laplacian operator in Euclidean space\footnote{Explicitly, $\partial^2 = \frac{\partial^2}{\partial t^2} + \nabla^2$}.
-We solve this equation numerically using the boundary condition $\vec\rho(0,x) = \e(x)$.
+We solve this equation numerically using the boundary condition $\e(\tau=0,x) = \e(x)$, a process described in Sec.~\ref{sec:runga-kutta}.

### 2021-04-23 final-report.latex/theory.tex.bk.tex
Final draft

--- 2021-04-19 final-report.latex/theory.tex
@@
 But what is a quantum field?
-Like in classical mechanics, a field is a function of spacetime with some mathematical object assigned to each point in space and time.
+Like in classical mechanics, a field is a function of spacetimewith some mathematical object assigned to each point in space and time.
 In the case of the electric field, this object is a three-dimensional vector, while the electric potential is a scalar field.
@@
 \begin{align}
-    S_E &= \int d^4x_E \mathcal{L}_E \\
+    S_E &= \int d^4x_E \mathcal{L}_E \nonumber \\
     &= i \int d^4x \mathcal{L}_E,
@@
 \subsection{Non-linear sigma model}
-
-The non-linear sigma model (NLSM) is a prototypical theory for a variety of unrelated physical phenomena, including applications in string theory \cite{callan1985} and ferromagnetism \cite{polyakov1975}.
+The non-linear sigma model (NLSM) is a prototypical theory for a variety of physical phenomena with applications in string theory \cite{callan1985} and ferromagnetism \cite{polyakov1975}.
 As a simple nonperturbative model, it also provides an ideal starting point for lattice QCD studies.
-Specifically, the NLSM exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
+Specifically, the NLSM exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability \cite{polyakov1975}.
 
 Unlike the $\phi^4$ model, which consists of a real value at each point in spacetime, the $O(3)$ NLSM consists of a 3D unit vector at each point.
-For this reason, every transformation of the field must be ``norm preserving''.
-Per its name, the $O(3)$ NLSM features a global symmetry under the 3D orthogonal group $O(3)$.
-In other words, the theory remains the same if all vectors are rotated equivalently.
+For this reason, every transformation of the field must be norm preserving.
+Per its name, the $O(3)$ NLSM features a global symmetry under the 3D orthogonal group $O(3)$; in other words, the theory remains the same if all vectors are rotated equivalently.
 To differentiate it from the $\phi^4$ model, we denote the NLSM field as $\e(x)$.
@@
     \label{eq:nlsm euclidean action}
-    S_E = \frac{\beta}{2} \int d^2x \; \partial^\mu \e \cdot \partial_\mu \e
+    %S_E = \frac{\beta}{2} \int d^2x \; \partial^\mu \e \cdot \partial_\mu \e
+    S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right]
 \end{equation}
@@
 
-To accomplish a statistical analysis of quantum fields, we use a Monte Carlo simulation.
-This method amounts to producing a large number of configurations and calculating statistics on the sample.
-A brute-force calculation over all possible configurations---as Eq.~\ref{eq:pathintegraleuclidean} suggests--- is clearly infinite and computationally infeasible.
+To accomplish a statistical analysis of quantum fields, we use a Monte Carlo simulation which produces a large number of configurations and calculates statistics on the sample.
+A brute-force calculation over all possible configurations, as Eq.~\ref{eq:pathintegraleuclidean} suggests, is clearly infinite and computationally infeasible.
 However, the exponential nature of the Boltzmann factor dictates that only configurations near the action minima contribute to observable statistics.
-Therefore, by selecting a sample of configurations near this minimum, we are able to extract meaningful results with a finite computation.
+Therefore, by selecting a sample of configurations near these minima, we are able to extract meaningful results with a finite computation.
 
 \subsection{The Markov Chain}
-The Markov chain is an effective method to identity field configurations that minimize the action.
+In order to determine this subset of configurations, we use a Markov chain.
+This method identities field configurations that minimize the action using a random walk through phase space.
 Essentially, we begin with a predetermined configuration and then make small adjustments, gradually lowering the action.
@@
 Each step consists of two parts: proposing a change and accepting the new configuration.
-The proposal creates a new configuration $\phi_b$ based on the current configuration $\phi_a$.
-In order to direct the Markov chain towards the action minima, we do not immediately accept this change.
-We calculate a probability
+The proposal creates a new configuration $\phi_b$ based on the current configuration $\phi_a$, a change that we accept with probability
+
 \begin{equation}
-P(\phi_a \rightarrow \phi_b)
+P(\phi_a \rightarrow \phi_b).
 \end{equation}
-of adding $\phi_b$ to the Markov chain, depending only on the actions of each configuration. This stochastic process ensures the chain will seek action minima.
+This probability determines if $\phi_b$ should be added to the Markov chain and depends on the change in action, stochastically ensuring that the chain will seek the action minima.
 
@@
 In other words, the probability of a $\phi_a\rightarrow\phi_b$ transition must be equal to the probability of a $\phi_b\rightarrow\phi_a$ transition.
-Mathematically, this condition takes the form of \textit{detailed balance equations}:
+Mathematically, this condition takes the form of ``detailed balance equations'':
 \begin{equation}
@@
 \end{enumerate}
-
 This final condition will allow us to explicitly define the transition probability using the action. From the Boltzmann distribution, we know
@@
 
-Due to the $\phi\rightarrow-\phi$ symmetry of the $\phi^4$ model, the ensemble mean of the average magnetization $\langle \phi \rangle$ is $0$.
-Likewise, the $O(3)$ symmetry in the NLSM enforces $\langle \e \rangle$ = 0.
+Due to the $\phi\rightarrow-\phi$ symmetry of the $\phi^4$ model, the ensemble mean of the average magnetization $\langle \bar\phi \rangle$ is $0$.
+Likewise, the $O(3)$ symmetry in the NLSM enforces $\langle \bar \e\, \rangle$ = 0.
 To measure the alignment, we therefore use the magnitude of this quantity, defined in the $\phi^4$ model as
 \begin{equation}
-|\bar\phi| \equiv \frac{1}{V}\left| \int d^2x \phi(x)\right|.
+|\bar\phi| \equiv \frac{1}{V}\left| \int d^2x \;\phi(x)\right|.
 \end{equation}
@@
     \label{eq:chim nlsm}
-    |\bar\e| \equiv \frac{1}{V}\left|\int d^2x \e(x)\right|.
+    |\bar\e| \equiv \frac{1}{V}\left|\int d^2x \;\e(x)\right|.
 \end{equation}
 
-In the symmetric phase, both $\langle \bar\e\, \rangle$ and $\langle \bar\phi \rangle = 0$ while in the broken phase they are both nonzero.
+In the symmetric phase, both $\langle |\bar\e\,| \rangle$ and $\langle |\bar\phi| \rangle = 0$ while in the broken phase they are both nonzero.
 
@@
 \end{equation}
-in the NLSM.
+in the NLSM.\footnote{The internal energy is not part of the $\phi^4$ portion of this work.}
 
@@
 We define three secondary observables: the magnetic susceptibility, the Binder cumulant and the bimodality.
-Since the Binder cumulant and the bimodality provide different perspectives on phase transitions, we will restrict our usage of these observables to the $\phi^4$ model.
+Since the Binder cumulant and the bimodality primarily provide different perspectives on phase transitions, we will restrict our usage of these observables to the $\phi^4$ model.
 
@@
 We use this formula to calculate all uncertainties in this study.
+
 \section{Topological Observables}
@@
 \label{sec:gradflow}
-To remove this ultraviolet divergence, we adopt a technique is ``smearing'', a local averaging of the field \cite{solbrig2007}.
+To remove this ultraviolet divergence, we adopt a technique is ``smearing'', a local averaging of the field \cite{solbrig2008}.
 Specifically, we use a technique known as the ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time'', or $\tau$.\footnote{The term ``half-dimension'' indicates that $\tau>0$.}  The flow time parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences.

### 2021-04-26 final-presentation.latex/presentation_1.tex
add final presentation

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Beamer Presentation
% LaTeX Template
% Version 1.0 (10/11/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND THEMES
%----------------------------------------------------------------------------------------

\documentclass{beamer}

\mode<presentation> {

% The Beamer class comes with a number of default slide themes
% which change the colors and layouts of slides. Below this is a list
% of all the themes, uncomment each in turn to see what they look like.

\usetheme{default}
%\usetheme{AnnArbor}
%\usetheme{Antibes}
%\usetheme{Bergen}
%\usetheme{Berkeley}
%\usetheme{Berlin}
%\usetheme{Boadilla}
%\usetheme{CambridgeUS}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{Dresden}
%\usetheme{Frankfurt}
%\usetheme{Goettingen}
%\usetheme{Hannover}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Luebeck}
%\usetheme{Madrid}
%\usetheme{Malmoe}
%\usetheme{Marburg}
%\usetheme{Montpellier}
%\usetheme{PaloAlto}
%\usetheme{Pittsburgh}
%\usetheme{Rochester}
%\usetheme{Singapore}
%\usetheme{Szeged}
%\usetheme{Warsaw}

\setbeamerfont{caption}{size=\scriptsize}

% As well as themes, the Beamer class has a number of color themes
% for any slide theme. Uncomment each of these in turn to see how it
% changes the colors of your current slide theme.

%\usecolortheme{albatross}
%\usecolortheme{beaver}
%\usecolortheme{beetle}
%\usecolortheme{crane}
%\usecolortheme{dolphin}
%\usecolortheme{dove}
%\usecolortheme{fly}
%\usecolortheme{lily}
%\usecolortheme{orchid}
%\usecolortheme{rose}
%\usecolortheme{seagull}
%\usecolortheme{seahorse}
%\usecolortheme{whale}
%\usecolortheme{wolverine}

%\setbeamertemplate{footline} % To remove the footer line in all slides uncomment this line
%\setbeamertemplate{footline}[page number] % To replace the footer line in all slides with a simple slide count uncomment this line

%\setbeamertemplate{navigation symbols}{} % To remove the navigation symbols from the bottom of all slides uncomment this line
\usefonttheme{serif}
}

\usepackage{graphicx} % Allows including images

\usepackage{subfigure}
\usepackage{booktabs} % Allows the use of \toprule, \midrule and \bottomrule in tables

%----------------------------------------------------------------------------------------
%	TITLE PAGE
%----------------------------------------------------------------------------------------

\title[Short title]{Defining condensates in low-dimensional spin models through the gradient flow} % The short title appears at the bottom of every slide, the full title is only on the title page

\author{Stuart Thomas} % Your name
\institute[W\&M] % Your institution as it will appear on the bottom of every slide, may be shorthand to save space
{
Christopher Monahan \\ % Your institution for the title page
\medskip
}
\date{November 16, 2020} % Date, can be changed to a custom date

\begin{document}

\begin{frame}
\titlepage % Print the title page as the first slide
\end{frame}

\begin{frame}
\frametitle{Lattice Quantum Chromodynamics}
\begin{itemize}
    \item Method to solve non-perturbative theories
    \item Discretizes spacetime to numerically calculate fields.
    \item Uses path integral formulation of Quantum Field Theory to solve for correlation functions.
\end{itemize}

\begin{block}{Problem}
    As lattice spacing approaches zero, some operators do not remain finite. 
\end{block}
\end{frame}

\begin{frame}
\frametitle{Power-divergent mixing in lattice QCD}
\begin{itemize}
    \item On the lattice, some operators mix and lead to divergences in the continuum limit.
    \item Example: breaking of rotational symmetry
    \item Operators must have different mass dimension
    \item In QCD:
        \begin{itemize}
            \item occurs with twist-2 operators
            \item used to calculate parton distribution functions
        \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Possible Solution: Smeared Operator Product Expansion (sOPE) }
    \begin{itemize}
        \item Expand non-local operators in a basis of smeared operators.
        \item Takes the form of a Laurent series
    \end{itemize}
    \textbf{Smeared Operator}: locally averaged operator
\end{frame}

\begin{frame}
\frametitle{The Gradient Flow}
\begin{itemize}
    \item A form of smearing
    \item Introduces a new dimension, $\tau$, called the ``flow time,'' through which the field evolves as follows:
    \begin{equation}
        \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
    \end{equation}
    with the boundary condition $\rho(0, x)=\phi(x)$.

    \item Suppresses infinities while keeping observables constant.

        \begin{figure}[h]
  \centering
      \subfigure[$\tau=0$]{\includegraphics[width=0.20\textwidth]{imgs/0.png}}
      \subfigure[$\tau=0.5$]{\includegraphics[width=0.20\textwidth]{imgs/0_5.png}}
      \subfigure[$\tau=2$]{\includegraphics[width=0.20\textwidth]{imgs/2.png}} 
      \subfigure[$\tau=10$]{\includegraphics[width=0.20\textwidth]{imgs/10.png}}
          \caption{Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
  \label{fig:flow}
\end{figure}

\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Non-Linear $\sigma$ Model}
\begin{itemize}
    \item a nonperturbative model in Quantum Field Theory which provides a good toy model for Quantum Chromodynamics
    \item characterized by the action
        \begin{equation*}
            S = \frac{1}{2g} \int d^2x \: \partial_\mu \vec{\phi}(x) \cdot \partial^\mu \vec{\phi}(x)
        \end{equation*}
    where the $\vec{\phi}(x)$ is a 3-component vector with norm 1 and $g$ is a coupling constant. 
\item Merits 
    \begin{itemize}
        \item mass gap
        \item asymptotic freedom
        \item $O(2)$ renormalizability
        \item in condensed matter, models Heisenberg ferromagnets
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Goals}
    \begin{itemize}
        \item Computationally implement the simpler $\phi^4$ theory to verify code.
        \item Transition to $O(3)$ nonlinear $\sigma$ model.
        \item Calculate twist-2 operators and vacuum expectation value with gradient flow applied.
        \item Other future paths:
            \begin{itemize}
                \item Consider topological charge
                \item Explore Wilson coefficients 
            \end{itemize}

    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Broad Motivation}
    To understand Deep Inelastic Scattering (DIS)
    \begin{itemize}
        \item Twist-2 operators $\rightarrow$ Mellin moments $\rightarrow$ Parton Distribution Functions $\rightarrow$ DIS Cross Sections
    \end{itemize}
\end{frame}
%------------------------------------------------

\begin{frame}
    \frametitle{Computational Methods}
\begin{itemize}
    \item Use a Markov Chain Monte Carlo method to simulate lattice in 2D and 3D
        \begin{itemize}
            \item Metropolis algorithm to generate random changes
            \item Wolff cluster algorithm to remove metastable states
            \item Implement checkerboard algorithm to parallelize Metropolis algorithm
        \end{itemize}

    \item \textbf{By sampling the lattice at set intervals, we can calculate statical estimates of arbitrary operators.}
\end{itemize}

\end{frame}

\begin{frame}
    \frametitle{Markov Chain Monte Carlo}
\begin{enumerate}
    \item Start with random lattice (known as ``hot start'')
    \item Repeatedly transition from old configuration $\mu$ to new configuration $\nu$ with probability $P(\mu\rightarrow\nu)$.
    \item After a thermalization period, record observables at certain intervals. 
    \item Take the statistical mean
\end{enumerate}
\end{frame}

\begin{frame}
    \frametitle{Metropolis Algorithm}
    For each site, propose a new value and accept change with probability
        \begin{equation*}
    P(\mu\rightarrow\nu) = \begin{cases} 
        e^{-(S_\nu - S_\mu)} & S_\nu > S_\mu \\
        1 & \mathrm{S_\nu \leq S_\mu} \\
       \end{cases}
        \end{equation*} 
        A \textbf{sweep} consists of perfoming this on every site.

\end{frame}

\begin{frame}
    \frametitle{Wolff Cluster Algorithm}
        \textbf{Problem}: Markov Chain gets ``stuck'' in metastable states, i.e. local minima of the action.

        \textbf{Solution}: Identify cluster of similar-values points and flip the sign.
    \begin{enumerate}
        \item Recursively grow a cluster with the probability of adding a new site
        \begin{equation*}
            P_{add} = 1-e^{-2\phi_a\cdot\phi_b}.
        \end{equation*}
        \item Flip sign of sites

    \end{enumerate}
    \begin{figure}[h]

      \centering
          \subfigure[before]{\includegraphics[width=0.20\textwidth]{imgs/wolff1}}
          \subfigure[after]{\includegraphics[width=0.20\textwidth]{imgs/wolff2}}
          \caption{Wolff cluster algorithm applied on $\phi^4$ lattice in $1+1$ dimensions, red corresponding to positive values and green to negative}
      \label{fig:flow}
    \end{figure}
\end{frame}

\begin{frame}
    \frametitle{Checkerboard Parallelization Algorithm}
    \begin{itemize}
        \item Split lattice into ``white'' and ``black'' sites. 
        \item Then, perform metropolis sweep on white sites in parallel, then black sites.
        \item Ensures independence of sweeps.
\end{itemize}

    \begin{figure}[h]

      \centering
          \includegraphics[width=0.5\textwidth]{imgs/checkerboard}
          \caption{Illustration of checkerboard algorithm\footnote{Yang, Kun, et al. "High performance Monte Carlo simulation of Ising model on TPU clusters." Proceedings of the International Conference for High Performance Computing, Networking, Storage and Analysis. 2019.}}
    \end{figure}

\end{frame}

\begin{frame}
    \frametitle{$\phi^4$ model}
    \begin{itemize}
        \item Given by the Euclidean action
            \begin{equation}
                S_E = \int d^D x \left[ \partial^\mu \phi \partial_\mu\phi + \frac{1}{2} m_0^2 \phi^2 + \frac{\lambda}{4}\phi^4\right]
            \end{equation}

        \item Simplest interacting field
        \item In 1+1 dimensions, exhibits spontaneous symmetry breaking at critical $m_0^2$.

    \end{itemize}

    We measure\ldots
\begin{itemize}
    \item Magnetization: $\langle |\bar\phi|\rangle$
    \item Susceptibility: $\chi = \langle {\bar\phi}^2\rangle - \langle\bar\phi\rangle^2$
    \item Binder cumulant: $U = 1 - \frac{\langle {\bar\phi}^4\rangle}{3 \langle {\bar\phi}^2\rangle^2}$
\end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Computational Implementation}
    \begin{enumerate}
        \item Begin with simple Python $\phi^4$ model with Cython for efficiency
        \item Transition to C++ code, generalized to nonlinear $\sigma$ model
        \item Compare results of Python and C++ code.
    \end{enumerate}
    \begin{block}{Parallelization}
        Message Passing Interface: shared RAM between processes.
    \end{block}
\end{frame}

\begin{frame}
\frametitle{Spontaneous Symmetry Breaking in $\phi^4$ model}
\begin{figure}[h]
  \centering
      \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
      \caption{Average $\bar\phi$, Binder cumulant and susceptibility for various $m_0^2$ values. $\lambda=0.5$, $L=64$, $1000$ measurements, $1000$ sweep thermalization with measurements taken every $100$ sweeps.}
  \label{fig:flow}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Comparison between Python and C++ code}
\begin{figure}[h]
  \centering
      \includegraphics[width=0.9\textwidth]{imgs/compare.png}
      \caption{Absolute difference between Python and C++ results of average $\bar\phi$, Binder cumulant and susceptibility for various $m_0^2$ values. $\lambda=0.5$, $L=64$, $1000$ measurements, $1000$ sweep thermalization with measurements taken every $100$ sweeps. }

  \label{fig:flow}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Future Work}
\begin{itemize}
    \item Fully implement nonlinear $\sigma$ model.
    \item Calculate twist-2 operators and vacuum expectation value
    \item Expand to 3 dimensions
    \item Profile and optimize C++ code
    \item Mathematical analysis of results
\end{itemize}

\end{frame}

%%------------------------------------------------

%\begin{frame}
%\frametitle{Multiple Columns}
%\begin{columns}[c] % The "c" option specifies centered vertical alignment while the "t" option is used for top vertical alignment

%\column{.45\textwidth} % Left column and width
%\textbf{Heading}
%\begin{enumerate}
%\item Statement
%\item Explanation
%\item Example
%\end{enumerate}

%\column{.5\textwidth} % Right column and width
%Lorem ipsum dolor sit amet, consectetur adipiscing elit. Integer lectus nisl, ultricies in feugiat rutrum, porttitor sit amet augue. Aliquam ut tortor mauris. Sed volutpat ante purus, quis accumsan dolor.

%\end{columns}
%\end{frame}
%\begin{frame}
%\frametitle{Blocks of Highlighted Text}

%\begin{block}{Block 2}
%Pellentesque sed tellus purus. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos. Vestibulum quis magna at risus dictum tempor eu vitae velit.
%\end{block}

%\begin{block}{Block 3}
%Suspendisse tincidunt sagittis gravida. Curabitur condimentum, enim sed venenatis rutrum, ipsum neque consectetur orci, sed blandit justo nisi ac lacus.
%\end{block}
%\end{frame}

%%------------------------------------------------

%\begin{frame}
%\frametitle{Multiple Columns}
%\begin{columns}[c] % The "c" option specifies centered vertical alignment while the "t" option is used for top vertical alignment

%\column{.45\textwidth} % Left column and width
%\textbf{Heading}
%\begin{enumerate}
%\item Statement
%\item Explanation
%\item Example
%\end{enumerate}

%\column{.5\textwidth} % Right column and width
%Lorem ipsum dolor sit amet, consectetur adipiscing elit. Integer lectus nisl, ultricies in feugiat rutrum, porttitor sit amet augue. Aliquam ut tortor mauris. Sed volutpat ante purus, quis accumsan dolor.

%\end{columns}
%\end{frame}

%%------------------------------------------------
%\section{Second Section}
%%------------------------------------------------

%\begin{frame}
%\frametitle{Table}
%\begin{table}
%\begin{tabular}{l l l}
%\toprule
%\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2}\\
%\midrule
%Treatment 1 & 0.0003262 & 0.562 \\
%Treatment 2 & 0.0015681 & 0.910 \\
%Treatment 3 & 0.0009271 & 0.296 \\
%\bottomrule
%\end{tabular}
%\caption{Table caption}
%\end{table}
%\end{frame}

%%------------------------------------------------

%\begin{frame}
%\frametitle{Theorem}
%\begin{theorem}[Mass--energy equivalence]
%$E = mc^2$
%\end{theorem}
%\end{frame}

%%------------------------------------------------

%\begin{frame}[fragile] % Need to use the fragile option when verbatim is used in the slide
%\frametitle{Verbatim}
%\begin{example}[Theorem Slide Code]
%\begin{verbatim}
%\begin{frame}
%\frametitle{Theorem}
%\begin{theorem}[Mass--energy equivalence]
%$E = mc^2$
%\end{theorem}
%\end{frame}\end{verbatim}
%\end{example}
%\end{frame}

%%------------------------------------------------

%\begin{frame}
%\frametitle{Figure}
%Uncomment the code on this slide to include your own image from the same directory as the template .TeX file.
%%\begin{figure}
%%\includegraphics[width=0.8\linewidth]{test}
%%\end{figure}
%\end{frame}

%%------------------------------------------------

%\begin{frame}[fragile] % Need to use the fragile option when verbatim is used in the slide
%\frametitle{Citation}
%An example of the \verb|\cite| command to cite within the presentation:\\~

%This statement requires citation \cite{p1}.
%\end{frame}

%%------------------------------------------------

%\begin{frame}
%\frametitle{References}
%\footnotesize{
%\begin{thebibliography}{99} % Beamer does not support BibTeX so references must be inserted manually as below
%\bibitem[Smith, 2012]{p1} John Smith (2012)
%\newblock Title of the publication
%\newblock \emph{Journal Name} 12(3), 45 -- 678.
%\end{thebibliography}
%}
%\end{frame}

%%------------------------------------------------

%\begin{frame}
%\Huge{\centerline{The End}}
%\end{frame}

%%----------------------------------------------------------------------------------------

\end{document}

### 2021-04-29 final-presentation.latex/main.tex
add final presentation

--- 2021-04-26 final-presentation.latex/presentation_1.tex
@@
 
-\setbeamerfont{caption}{size=\scriptsize}
+%\setbeamerfont{caption}{size=\scriptsize}
 
@@
 \usepackage{graphicx} % Allows including images
-
-\usepackage{subfigure}
+\usepackage{tikz}
+\usepackage[duration=30,disablemarkdown]{pdfpc}
+\usepackage[font=tiny, labelfont=bf]{caption}
+\usepackage{subcaption}
 \usepackage{booktabs} % Allows the use of \toprule, \midrule and \bottomrule in tables
 
+\renewcommand{\note}{\pdfpcnote}
 %----------------------------------------------------------------------------------------
@@
 
-\title[Short title]{Defining condensates in low-dimensional spin models through the gradient flow} % The short title appears at the bottom of every slide, the full title is only on the title page
+\title[Short title]{Topology of the $O(3)$ non-linear sigma model under the gradient flow} % The short title appears at the bottom of every slide, the full title is only on the title page
 
@@
 }
-\date{November 16, 2020} % Date, can be changed to a custom date
+\date{May 3, 2021} % Date, can be changed to a custom date
+
+\newcommand{\e}{\vec e}
+\renewcommand{\theequation}{\relax}
 
@@
 \titlepage % Print the title page as the first slide
+\note{ - Test note one \\
+- Test note two%
+}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Quantum Field Theory}
+\begin{equation}
+    \label{eq:pathintegral}
+    \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; e^{iS[\phi]}
+\end{equation}
+\end{frame}
+
+\begin{frame}
+\frametitle{$\phi^4$ model}
+\begin{equation}
+    \label{eq:phi4 euclidean action}
+    S_E[\phi] = \int d^2 x_E \left[\frac{1}{2}\left(\partial_t \phi\right)^2 + \frac{1}{2} \left(\partial_x \phi \right)^2 + \frac{1}{2} m_0^2 \phi^2 + \frac{\lambda}{4}\phi^4\right].
+\end{equation}
+\end{frame}
+
+\begin{frame}
+\frametitle{Non-Linear $\sigma$ Model}
+\begin{equation}
+    \label{eq:nlsm euclidean action}
+    % S_E = \frac{\beta}{2} \int d^2x \; \partial^\mu \e \cdot \partial_\mu \e
+    S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right]
+\end{equation}
+\begin{itemize}
+    \item a nonperturbative model in Quantum Field Theory which provides a good toy model for Quantum Chromodynamics
+    \item characterized by the action
+        \begin{equation*}
+            S = \frac{1}{2g} \int d^2x \: \partial_\mu \vec{\phi}(x) \cdot \partial^\mu \vec{\phi}(x)
+        \end{equation*}
+    where the $\vec{\phi}(x)$ is a 3-component vector with norm 1 and $g$ is a coupling constant. 
+\item Merits 
+    \begin{itemize}
+        \item mass gap
+        \item asymptotic freedom
+        \item $O(2)$ renormalizability
+        \item in condensed matter, models Heisenberg ferromagnets
+    \end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Monte Carlo and the Markov Chain}
+\end{frame}
+
+\begin{frame}
+\frametitle{Topological Charge}
+\begin{figure}
+    \includegraphics[width=0.3\paperwidth]{imgs/homotopy.png}
+    \caption{\label{fig:homotopy} Homotopy group}
+\end{figure}
+\note{- Test note one \\ - Test note two}
+\end{frame}
+
+\begin{frame}
+\frametitle{Topological Action}
+\begin{equation}
+    S[\e\,] \rightarrow S[\e\,] - i \theta Q[\e\,].
+\end{equation}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Topology Visualization}
+\begin{figure}[h]
+    \begin{subfigure}[b]{0.4\paperwidth}
+    \centering
+    \begin{tikzpicture}[scale=0.5, every node/.style={scale=0.7}]
+        \draw[step=4cm,gray,thin] (-2,-2) grid (6,6);
+        \draw[dotted] (0,0) -- (4,4);
+        \draw[thick](0,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_1$} -- 
+                    (0,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_2$} -- 
+                    (4,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_3$} -- 
+                    (4,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_4$} -- cycle ;
+                    \draw[] (2,2) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north east]{$x^*$};
+
+        \draw[-stealth] (0.2,1) -- (0.2,3);
+        \draw[-stealth] (1,3.8) -- (3,3.8);
+        \draw[-stealth] (3,3.5) -- (1,1.5);
+
+        \draw[-stealth] (1,0.5) -- (3,2.5);
+        \draw[-stealth] (3.8,3) -- (3.8,1);
+        \draw[-stealth] (3,0.2) -- (1,0.2);
+    \end{tikzpicture}
+    %\caption{\label{fig:plaquette} Visualization of plaquette $x^*$.
+The dotted line separates the plaquette into two signed areas which are used to define the topological charge density $q(x^*)$.
+Arrows represent order of signed area.}
+    \end{subfigure} %
+    \begin{subfigure}[b]{0.4\paperwidth}
+    \centering
+    \begin{tikzpicture}[scale=0.7, every node/.style={scale=0.7}]
+        \pgfdeclarelayer{nodelayer}
+        \pgfdeclarelayer{edgelayer}
+        \pgfdeclareradialshading{sphere4}{\pgfpoint{-0.2cm}{0.5cm}}% 
+            {rgb(0cm)=(1,1,1);
+            rgb(1cm)=(0.5,0.5,0.5); rgb(1.05cm)=(1,1,1)}
+            %rgb(0.7cm)=(0.1,0.1,0.1); rgb(1cm)=(0.5,0.05,0); rgb(1.05cm)=(1,1,1)}
+        \pgfsetlayers{nodelayer,edgelayer}
+        \tikzstyle{label}=[fill=none, draw=none, shape=circle]
+        \tikzstyle{point}=[inner sep=0pt, minimum size=0.2cm,fill=black, draw, shape=circle]
+        %\tikzstyle{interior line}=[{Stealth[scale=1.5]}-,dotted,thick]
+        \tikzstyle{interior line}=[dotted,thick]
+
+        \tikzstyle{triangle}=[thick]
+        \tikzstyle{arrow}=[->, thick]
+        \begin{pgfonlayer}{nodelayer}
+            \node [style=point] (4) at (0.5, 2.5) {};
+            \node [style=point] (5) at (0.55, 0.625) {};
+            \node [style=point] (6) at (2.25, 0.75) {};
+            \node (7) at (0, 0) {};
+            \node (8) at (1.25, 1.375) {};
+            \node (9) at (2.425, 2.525) {};
+        \end{pgfonlayer}
+        \begin{pgfonlayer}{edgelayer}
+            \draw (0,0) circle (3cm);
+            %\shade[inner color=white,outer color=lightgray] (0,0) circle (3cm);
+            \shade[shading=sphere4] (0,0) circle (3cm);
+            \draw [bend left=15,style=triangle] (4.center) to (6.center);
+            \draw [bend left=355,style=triangle] (6.center) to (5.center);
+            \draw [bend right=5,style=triangle] (5.center) to (4.center);
+            \draw [style=interior line] (4.center) to (7.center);
+            \draw [style=interior line] (5.center) to (7.center);
+            \draw [style=interior line] (6.center) to (7.center);
+            \draw [fill=gray!80] (4.center) to [bend left=15] (6.center) to [bend left=355] (5.center) to [bend right=5] cycle;
+
+            %\draw [style=arrow] (8.center) to (9.center);
+
+            \node [style=label, anchor=north] at (2.25, 0.75) {\small $\e(x_1)$};
+            \node [style=label, anchor=north] at (0.55, 0.625) {\small $\e(x_2)$};
+            \node [style=label, anchor=east] at (0.5, 2.5) {\small $\e(x_3)$};
+            \node [style=point] at (4){};
+            \node [style=point] at (5){};
+            \node [style=point] at (6){};
+
+            \node [style=label] at (8) {\small $A$};
+        \end{pgfonlayer}
+    \end{tikzpicture}
+    \end{subfigure}
+    \caption{some plots}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}
+    \frametitle{Wolff Cluster Algorithm}
+
+    \begin{figure}[h]
+      \centering
+          \begin{subfigure}[b]{0.5\textwidth}\centering
+            \includegraphics[width=0.6\textwidth]{imgs/wolffa.png}
+            \caption{before cluster flip}
+          \end{subfigure}%
+          \hfill
+          \begin{subfigure}[b]{0.5\textwidth}\centering
+            \includegraphics[width=0.6\textwidth]{imgs/wolffb.png}
+            \caption{after cluster flip}
+          \end{subfigure}
+          \hfill
+          \caption{\label{fig:wolff} An example of the Wolff cluster algorithm in the $\phi^4$ model.
+White represents positive values of $\phi$ while black represents negative.
+$\lambda=0.5$, $m_0^2=-0.9$.}
+
+    \end{figure}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Power-divergent mixing in lattice QCD}
-\begin{itemize}
-    \item On the lattice, some operators mix and lead to divergences in the continuum limit.
-    \item Example: breaking of rotational symmetry
-    \item Operators must have different mass dimension
-    \item In QCD:
-        \begin{itemize}
-            \item occurs with twist-2 operators
-            \item used to calculate parton distribution functions
-        \end{itemize}
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-    \frametitle{Possible Solution: Smeared Operator Product Expansion (sOPE) }
-    \begin{itemize}
-        \item Expand non-local operators in a basis of smeared operators.
-        \item Takes the form of a Laurent series
-    \end{itemize}
-    \textbf{Smeared Operator}: locally averaged operator
-\end{frame}
-
-\begin{frame}
 \frametitle{The Gradient Flow}
@@
 
-        \begin{figure}[h]
-  \centering
-      \subfigure[$\tau=0$]{\includegraphics[width=0.20\textwidth]{imgs/0.png}}
-      \subfigure[$\tau=0.5$]{\includegraphics[width=0.20\textwidth]{imgs/0_5.png}}
-      \subfigure[$\tau=2$]{\includegraphics[width=0.20\textwidth]{imgs/2.png}} 
-      \subfigure[$\tau=10$]{\includegraphics[width=0.20\textwidth]{imgs/10.png}}
-          \caption{Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
-  \label{fig:flow}
-\end{figure}
+        %\begin{figure}[h]
+  %\centering
+      %\subfigure[$\tau=0$]{\includegraphics[width=0.20\textwidth]{imgs/0.png}}
+      %\subfigure[$\tau=0.5$]{\includegraphics[width=0.20\textwidth]{imgs/0_5.png}}
+      %\subfigure[$\tau=2$]{\includegraphics[width=0.20\textwidth]{imgs/2.png}} 
+      %\subfigure[$\tau=10$]{\includegraphics[width=0.20\textwidth]{imgs/10.png}}
+          %\caption{Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
+  %\label{fig:flow}
+%\end{figure}
 
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Non-Linear $\sigma$ Model}
-\begin{itemize}
-    \item a nonperturbative model in Quantum Field Theory which provides a good toy model for Quantum Chromodynamics
-    \item characterized by the action
-        \begin{equation*}
-            S = \frac{1}{2g} \int d^2x \: \partial_\mu \vec{\phi}(x) \cdot \partial^\mu \vec{\phi}(x)
-        \end{equation*}
-    where the $\vec{\phi}(x)$ is a 3-component vector with norm 1 and $g$ is a coupling constant. 
-\item Merits 
-    \begin{itemize}
-        \item mass gap
-        \item asymptotic freedom
-        \item $O(2)$ renormalizability
-        \item in condensed matter, models Heisenberg ferromagnets
-    \end{itemize}
 \end{itemize}
@@
     \end{enumerate}
-    \begin{figure}[h]
+    %\begin{figure}[h]
 
-      \centering
-          \subfigure[before]{\includegraphics[width=0.20\textwidth]{imgs/wolff1}}
-          \subfigure[after]{\includegraphics[width=0.20\textwidth]{imgs/wolff2}}
-          \caption{Wolff cluster algorithm applied on $\phi^4$ lattice in $1+1$ dimensions, red corresponding to positive values and green to negative}
-      \label{fig:flow}
-    \end{figure}
-\end{frame}
-
-\begin{frame}
-    \frametitle{Checkerboard Parallelization Algorithm}
-    \begin{itemize}
-        \item Split lattice into ``white'' and ``black'' sites. 
-        \item Then, perform metropolis sweep on white sites in parallel, then black sites.
-        \item Ensures independence of sweeps.
-\end{itemize}
-
-    \begin{figure}[h]
-
-      \centering
-          \includegraphics[width=0.5\textwidth]{imgs/checkerboard}
-          \caption{Illustration of checkerboard algorithm\footnote{Yang, Kun, et al.
-"High performance Monte Carlo simulation of Ising model on TPU clusters." Proceedings of the International Conference for High Performance Computing, Networking, Storage and Analysis.
-2019.}}
-    \end{figure}
-
+      %\centering
+          %\subfigure[before]{\includegraphics[width=0.20\textwidth]{imgs/wolff1}}
+          %\subfigure[after]{\includegraphics[width=0.20\textwidth]{imgs/wolff2}}
+          %\caption{Wolff cluster algorithm applied on $\phi^4$ lattice in $1+1$ dimensions, red corresponding to positive values and green to negative}
+      %\label{fig:flow}
+    %\end{figure}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Comparison between Python and C++ code}
+    \frametitle{Effect of $\theta$-term on $\langle Q \rangle$}
 \begin{figure}[h]
   \centering
-      \includegraphics[width=0.9\textwidth]{imgs/compare.png}
-      \caption{Absolute difference between Python and C++ results of average $\bar\phi$, Binder cumulant and susceptibility for various $m_0^2$ values.
+      \includegraphics[width=0.8\textwidth]{imgs/theta.png}
+      %\caption{Absolute difference between Python and C++ results of average $\bar\phi$, Binder cumulant and susceptibility for various $m_0^2$ values.
 $\lambda=0.5$, $L=64$, $1000$ measurements, $1000$ sweep thermalization with measurements taken every $100$ sweeps.

### 2021-10-01 proceedings/skeleton.tex
add draft  of proceedings

% Please make sure you insert your
% data according to the instructions in PoSauthmanual.pdf
\documentclass[a4paper,11pt]{article}
\usepackage{pos}

\title{Topology of the $O(3)$ non-linear sigma model under the gradient flow}
%% \ShortTitle{Short Title for header}

\author*[a]{Stuart Thomas}
\author[a]{Christopher Monahan}

\affiliation[a]{Department of Physics, William \& Mary,  Williamsburg, Virginia 23187, USA}

%\affiliation[b]{Department, University,\\
%Street number, City, Country}

\emailAdd{snthomas01@email.wm.edu}
\emailAdd{cjmonahan@wm.edu}

\abstract{
The $O(3)$ non-linear sigma model (NLSM) is a prototypical field theory for QCD and ferromagnetism, featuring topological qualities. Though the topological susceptibility $\chi_t$ should vanish in physical theories, lattice simulations of the NLSM find that $\chi_t$ diverges in the continuum limit. We study the effect of the gradient flow on this quantity using a Markov Chain Monte Carlo method, finding that a logarithmic divergence persists. This result supports a previous study and indicates that either the definition of topological charge is problematic or the NLSM has no well-defined continuum limit. We also introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow.}

\FullConference{%
 The 38th International Symposium on Lattice Field Theory, LATTICE2021
  26th-30th July, 2021
  Zoom/Gather@Massachusetts Institute of Technology
}

\begin{document}
\maketitle

\section{...}

\begin{thebibliography}{99}
\bibitem{...}
....

\end{thebibliography}

\end{document}

### 2021-10-07 proceedings/draft.tex
add proceeding template

--- 2021-10-01 proceedings/skeleton.tex
@@
 
-\section{...}
+\section{The Non-Linear Sigma Model}
+            We study the $O(3)$ non-linear sigma model (NLSM) in 1+1 dimensions, defined by the Euclidean action
+            \begin{equation*}
+                \label{eq:nlsm euclidean action}
+                S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right]
+            \end{equation*}
+        where, $\e$ is 3-component real vector constrained by $|\e\,|=1$ and $\beta$ is the inverse coupling constant.
+In solid-state systems, this model describes Heisenberg ferromagnets \cite{callan1985} and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), the gauge theory that describes the strong nuclear force.
+In general, the NLSM shares key features with non-Abelian gauge theories such as QCD, including a mass gap and asymptotic freedom \cite{polyakov1975}.
+Therefore, the NLSM is a useful model for exploring the effect of these properties in a simpler system.
+
+            %\item Prototypical model for strong nuclear force
+                    %\item Models Heisenberg ferromagnets
+                    %\item Applications to string theory
+                %\end{itemize}
+            %\item<4-> Merits
+                %\begin{itemize}
+                    %\item mass gap
+                    %\item asymptotic freedom
+                %\end{itemize}
+        \end{itemize}
 

### 2021-10-07 final-presentation.latex/main.tex
update presentation

@@
 
-\usetheme{default}
+%\usetheme{default}
 %\usetheme{AnnArbor}
@@
 %\usetheme{Luebeck}
-%\usetheme{Madrid}
+\usetheme{Madrid}
 %\usetheme{Malmoe}
@@
 %\usetheme{Warsaw}
-
+\setbeamertemplate{navigation symbols}{}%remove navigation symbols
 %\setbeamerfont{caption}{size=\scriptsize}
@@
 }
-
+\usepackage{transparent}
 \usepackage{graphicx} % Allows including images
@@
 \usepackage{booktabs} % Allows the use of \toprule, \midrule and \bottomrule in tables
+\usepackage[%
+    style=phys,%
+    articletitle=false,biblabel=brackets,%
+    chaptertitle=false,pageranges=false,%
+    doi=false,maxnames=2,minnames=1,eprint=false%
+  ]
+{biblatex}
+\addbibresource{library.bib}
+\makeatletter
+\renewcommand{\@makefnmark}{\makebox{\normalfont[\@thefnmark]}}
+\renewcommand\@makefntext[1]{%
+    {{\normalfont[\@thefnmark]}\enspace #1}
+  }
+\makeatother
+%\renewcommand{\cite}[1]{[\footfullcite{#1}]}
+\renewenvironment{equation}
+    {
+    \begin{equation*}
+    }
+    { 
+    \end{equation*}
+    }
 
+%\renewenvironment{align}
+    %{
+    %\begin{align*}
+    %}
+    %{ 
+    %\end{align*}
+    %}
 \renewcommand{\note}{\pdfpcnote}
@@
 
-\title[Short title]{Topology of the $O(3)$ non-linear sigma model under the gradient flow} % The short title appears at the bottom of every slide, the full title is only on the title page
+\title[Honors Presentation]{Topology of the $O(3)$ non-linear sigma model under the gradient flow} % The short title appears at the bottom of every slide, the full title is only on the title page
 
@@
 \newcommand{\e}{\vec e}
-\renewcommand{\theequation}{\relax}
 
 \begin{document}
+\setbeamerfont{footnote}{size=\tiny}
+\begin{frame}
+\titlepage % Print the title page as the first slide
+\end{frame}
 
@@
 \titlepage % Print the title page as the first slide
-\note{ - Test note one \\
-- Test note two%
-}
 \end{frame}
@@
 \frametitle{Quantum Field Theory}
-\begin{equation}
+   \centering
+   \Huge particles $\Rightarrow$ fields
+   \vspace{1cm}
+\Large
+   \begin{itemize}
+       \item<2-> particle physics and condensed matter
+       \item<3-> incorporates many-particle quantum mechanics and special relativity
+       \item<4-> remarkably accurate \footfullcite{odom2006}
+
+   \end{itemize}
+   \note{What is a field? a mathematical object at every point of space and time.}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Path Integral Formulation}
+    \Large\centering
+``quantum principle of least action'' 
+\begin{equation*}
     \label{eq:pathintegral}
     \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; e^{iS[\phi]}
-\end{equation}
+\end{equation*}
+\note{Note vacuum}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Wick Rotation}
+    \Large
+    \begin{align*}
+        t &\Rightarrow it \\
+        \textrm{Minkowski Spacetime} &\Rightarrow \textrm{Euclidean Spacetime}
+    \end{align*}
+\vspace{0.1cm}
+\hrule
+\vspace{0.1cm}
+    \onslide<2->{\begin{equation*}
+        \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; e^{iS[\phi]}
+    \end{equation*}
+    \[\Downarrow\]}
+    \only<2>{\begin{equation*}
+        \label{eq:pathintegraleuclidean}
+        \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; e^{-S_E[\phi]}
+    \end{equation*}}
+    \onslide<3>{\begin{equation*}
+        \langle \hat O \rangle = \frac{1}{Z} \int \mathcal{D}\phi \: \hat O [\phi]\; \textcolor{red}{e^{-S_E[\phi]}}
+    \end{equation*}}
+\centering
+\note{Downside: only possible for time-independent observables}
 \end{frame}
@@
 \frametitle{$\phi^4$ model}
-\begin{equation}
+\begin{block}{Euclidean Action}
+\begin{equation*}
     \label{eq:phi4 euclidean action}
-    S_E[\phi] = \int d^2 x_E \left[\frac{1}{2}\left(\partial_t \phi\right)^2 + \frac{1}{2} \left(\partial_x \phi \right)^2 + \frac{1}{2} m_0^2 \phi^2 + \frac{\lambda}{4}\phi^4\right].
-\end{equation}
+    S_E[\phi] = \int d^2 x_E \left[\frac{1}{2}\left(\partial_t \phi\right)^2 + \frac{1}{2} \left(\partial_x \phi \right)^2 + \frac{1}{2} m_0^2 \phi^2 + \frac{\lambda}{4}\phi^4\right]
+\end{equation*}
+\end{block}
+\begin{itemize}
+    \item real scalar field
+    \item describes a boson
+    \item spontaneous symmetry breaking at $\lambda=0.5$, $m_0^2=-0.72$
+\end{itemize}
+\note{Check our simulation}
 \end{frame}
@@
 \frametitle{Non-Linear $\sigma$ Model}
-\begin{equation}
+\begin{block}{Euclidean Action}
+\begin{equation*}
     \label{eq:nlsm euclidean action}
@@
     S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right]
-\end{equation}
+\end{equation*}
+\end{block}
 \begin{itemize}
-    \item a nonperturbative model in Quantum Field Theory which provides a good toy model for Quantum Chromodynamics
-    \item characterized by the action
-        \begin{equation*}
-            S = \frac{1}{2g} \int d^2x \: \partial_\mu \vec{\phi}(x) \cdot \partial^\mu \vec{\phi}(x)
-        \end{equation*}
-    where the $\vec{\phi}(x)$ is a 3-component vector with norm 1 and $g$ is a coupling constant. 
-\item Merits 
+    \item<1-> $O(3)$ non-linear sigma model (NLSM) in 1+1 dimensions
+    \item<2-> $\e$ is 3-component vector constrained by $|\e\,|=1$
+    \item<3-> Applications
+        \begin{itemize}
+            \item Prototypical model for strong nuclear force
+            \item Models Heisenberg ferromagnets
+            \item Applications to string theory
+        \end{itemize}
+    \item<4-> Merits 
         \begin{itemize}
@@
             \item asymptotic freedom
-        \item $O(2)$ renormalizability
-        \item in condensed matter, models Heisenberg ferromagnets
         \end{itemize}
@@
 \begin{frame}
-\frametitle{Monte Carlo and the Markov Chain}
+    \frametitle{Topology of 1+1 $O(3)$ NLSM}
+    \begin{columns}
+
+    \begin{column}{0.5\textwidth}
+    \begin{enumerate}
+        \item<1-> Envision spacetime as a sphere
+        \item<2-> The NLSM field $\e$ becomes mapping between Riemann spheres ($S^2 \rightarrow S^2$).
+        \item<3-> Associates every configuration with topological charge $Q$
+        \item<4-> Topology is important to cosmology\footnotemark[2]\ and fault-tolerant quantum computing\footnotemark[3]. 
+    \end{enumerate}
+    \end{column}
+    \begin{column}{0.5\textwidth}
+        \begin{figure}
+            \includegraphics[width=0.3\paperwidth]{imgs/homotopy.png}
+            \caption{\label{fig:homotopy} Homotopy group of $S^1\rightarrow S^1$}
+        \end{figure}
+    \end{column}
+    \end{columns}
+    \footnotetext[2]{\fullcite{goddard1986}}
+    \footnotetext[3]{\fullcite{kitaev1997}}
+\note{Many people are not familiar with topology \\%
+    1. Since the Lagrangian must disappear at infinity, the field must be uniform for all points as x -> infty. \\%
+    2. We can classify every mapping with an integer. Use ball and balloon metaphor}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Topological Charge}
-\begin{figure}
-    \includegraphics[width=0.3\paperwidth]{imgs/homotopy.png}
-    \caption{\label{fig:homotopy} Homotopy group}
-\end{figure}
-\note{- Test note one \\ - Test note two}
+    \frametitle{Non-trivial NLSM}
+    \begin{equation*}
+        S[\e\,] \rightarrow S[\e\,] - i \theta Q[\e\,].
+    \end{equation*}
+    Nonzero $\theta$ implies nonzero $\langle Q \rangle$.
 \end{frame}
@@
 \begin{frame}
-\frametitle{Topological Action}
+    \frametitle{Topological Susceptibility}
+    \centering
+    Proportional to variance of topological charge
     \begin{equation}
-    S[\e\,] \rightarrow S[\e\,] - i \theta Q[\e\,].
+        \chi_t \equiv \frac{\langle Q^2 \rangle - \langle Q \rangle^2}{L^2}
+    \end{equation}
+    \onslide<2->{
+        \vfill
+        \hrule
+        \vfill
+        \begin{columns}
+            \begin{column}{0.5\textwidth}
+                \centering
+            In trivial NLSM:
+            \begin{equation}
+                \chi_t = \frac{\langle Q^2 \rangle}{L^2}
+            \end{equation}
+            \end{column}
+
+        %\only<3->{
+            \begin{column}{0.5\textwidth}
+                \centering
+            In nontrivial NLSM:
+            \begin{equation}
+                \chi_t \propto \frac{\partial\langle Q \rangle}{\partial \theta} 
+            \end{equation}
+            \end{column}
+        %}
+        \end{columns}
+    }
+    \onslide<3->\begin{block}{Main Problem}
+        Topological stability predicts $\chi_t=0$, but $\chi_t$ diverges in numerical results \footfullcite{berg1981}.
+    \end{block}
+\end{frame}
+
+\begin{frame}
+\frametitle{The Gradient Flow}
+\begin{itemize}
+    \item<1-> Successful in removing $\chi_t$ divergence in QCD (model of the strong nuclear force).
+    \item<2-> Reduces high-momentum modes
+    \item<3-> Introduces a new dimension, $\tau$, called the ``flow time,'' which pushes fields towards action minima 
+    \item<4-> In the $\phi^4$ model
+    \begin{equation*}
+        \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
+    \end{equation*}
+    with the boundary condition $\rho(0, x)=\phi(x)$.
+
+\end{itemize}
+\onslide<4->\begin{figure}[h]
+  \centering
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+        \includegraphics[width=0.9\textwidth]{imgs/gf0.png}
+        \caption{$\tau=0$}
+      \end{subfigure}%
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+        \includegraphics[width=0.9\textwidth]{imgs/gf1.png}
+        \caption{$\tau=0.001$}
+      \end{subfigure}%
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+        \includegraphics[width=0.9\textwidth]{imgs/gf2.png}
+        \caption{$\tau=0.01$}
+      \end{subfigure}%
+      \begin{subfigure}[b]{0.2\textwidth}\centering
+        \includegraphics[width=0.9\textwidth]{imgs/gf3.png}
+        \caption{$\tau=0.1$}
+      \end{subfigure}%
+      \caption{\label{fig:flow} Effect of flow time evolution on a random lattice in the symmetric phase. White represents positive values of $\phi$ while black represents negative.}
+\end{figure}
+\note{Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
+\end{frame}
+
+\begin{frame}
+   \centering
+   \begin{block}{Research Question}
+       Can the gradient flow remove the topological susceptibility divergence in the NLSM?
+   \end{block}
+\end{frame}
+
+\begin{frame}
+\frametitle{Methods}
+\begin{enumerate}
+    \item Construct Monte Carlo simulation following Euclidean path integral formalism
+    \item Apply gradient flow to configurations
+    \item Measure topological susceptibility
+\end{enumerate}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Fields on the Lattice}
+    \begin{align*}
+        x & \rightarrow ia \hat{t} + j a \hat{x} \\
+        \int dtdx &\rightarrow a^2 \sum_i
+    \end{align*}
+
+    where $a$ is the lattice spacing.
+
+    \vspace{0.3cm}\hrule\vspace{0.3cm}
+
+    Periodic boundary conditions:
+    \begin{equation}
+        \phi\left(x_{i+L,j}\right) = \phi\left(x_{i,j+L}\right) = \phi\left(x_{i,j}\right)
     \end{equation}
@@
 \begin{frame}
-    \frametitle{Topology Visualization}
+    \frametitle{Markov chain Monte Carlo}
+    \onslide<1-> Due to Boltzmann factor $e^{-S}$, we need only consider configurations near action minima.
+    \begin{itemize}
+        \item<2-> Starting at configuration $\phi_a$, we transition to configuration $\phi_b$ with probability 
+        \[P(\phi_a \rightarrow \phi_b).\]
+    \end{itemize}
+    \note{Mention infinite path integral}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Metropolis Algorithm}
+    Changes a single site randomly
+    \begin{equation}
+        P(\phi_a\rightarrow\phi_b) = \begin{cases} 
+            e^{S[\phi_a] - S[\phi_b]} & S[\phi_b] > S[\phi_a] \\
+            1 & \mathrm{otherwise} \\
+       \end{cases}
+    \end{equation}
+    Performing this process on every site forms a ``sweep''
+\end{frame}
+
+\begin{frame}
+    \frametitle{Wolff Cluster Algorithm}
+    Grows one cluster probabilistically, flips sign~\footfullcite{wolff1989}
+    \begin{figure}[h]
+      \centering
+          \begin{subfigure}[b]{0.5\textwidth}\centering
+            \includegraphics[width=0.6\textwidth]{imgs/wolffa.png}
+            \caption{before cluster flip}
+          \end{subfigure}%
+          \hfill
+          \begin{subfigure}[b]{0.5\textwidth}\centering
+            \includegraphics[width=0.6\textwidth]{imgs/wolffb.png}
+            \caption{after cluster flip}
+          \end{subfigure}
+          \hfill
+          %\caption{\label{fig:wolff} An example of the Wolff cluster algorithm in the $\phi^4$ model.
+White represents positive values of $\phi$ while black represents negative.
+$\lambda=0.5$, $m_0^2=-0.9$.}
+    \end{figure}
+    \note{What about NLSM?}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Thermalization}
+
+    How do we start the Markov chain?
+    \begin{columns}
+
+        \begin{column}{0.5\textwidth}
+            \begin{figure}[h]
+              \centering
+                  \begin{subfigure}[b]{0.8\textwidth}\centering
+                    \includegraphics[width=\textwidth]{imgs/therm24.png}
+                    \caption{$L=24$}
+                  \end{subfigure}
+                  \begin{subfigure}[b]{0.8\textwidth}\centering
+                    \includegraphics[width=\textwidth]{imgs/therm404.png}
+                    \caption{$L=404$}
+                  \end{subfigure}
+                  \hfill
+            \end{figure}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \onslide<2->{
+            \begin{figure}[h]
+              \centering
+                \includegraphics[width=0.9\textwidth]{imgs/coldstart.png}
+                \caption{Action histogram comparing hot vs. cold start after 1000 sweep thermalization}
+            \end{figure}
+            }
+        \end{column}
+    \end{columns}
+      \note{Plots of the action as a function of Monte Carlo time, starting with a random NLSM lattice.
+\\ Histogram of lattice-averaged actions $S/L^2$ with hot and cold starts.
+1,000 sweep thermalization in $L=404$ lattice, 1,000 measurements taken every 50 sweeps.}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Autocorrelation}
+    How many sweeps per measurement?
+    \begin{figure}[h]
+      \centering
+          \begin{subfigure}[b]{0.5\textwidth}\centering
+            \includegraphics[width=0.9\textwidth]{imgs/tauint24.png}
+            \caption{$L=24$, $\tau_{int}=4.43$}
+          \end{subfigure}%
+          \hfill
+          \begin{subfigure}[b]{0.5\textwidth}\centering
+            \includegraphics[width=0.9\textwidth]{imgs/tauint404.png}
+            \caption{$L=404$, $\tau_{int}=12.81$}
+          \end{subfigure}
+          \hfill
+          \caption{\label{fig:tauint} Plots of automatic windowing\footfullcite{wolff2007} procedure used to calculate $\tau_{int}$ for the NLSM model. $W$ is summation window size.}
+    \end{figure}
+    Therefore, we measure every 50 sweeps with a Wolff cluster step every 5 sweeps.
+\end{frame}
+
+\begin{frame}
+    \frametitle{Topological Charge on the Lattice}
+    Following Berg \& L\"usher~\footfullcite{berg1981}, define topological charge density $q(x^*)$ for each plaquette $x^*$ such that
+    \begin{equation}
+        Q = \sum_{x^*} q(x^*)
+    \end{equation}
     \begin{figure}[h]
@@
     \centering
-    \begin{tikzpicture}[scale=0.5, every node/.style={scale=0.7}]
-        \draw[step=4cm,gray,thin] (-2,-2) grid (6,6);
+    \begin{tikzpicture}[scale=0.4, every node/.style={scale=0.7}]
+        \draw[step=4cm,gray,thin] (-1,-1) grid (5,5);
         \draw[dotted] (0,0) -- (4,4);
@@
     \end{tikzpicture}
+    \caption{a plaquette}
     %\caption{\label{fig:plaquette} Visualization of plaquette $x^*$.
@@
     \centering
-    \begin{tikzpicture}[scale=0.7, every node/.style={scale=0.7}]
+    \begin{tikzpicture}[scale=0.5, every node/.style={scale=0.7}]
         \pgfdeclarelayer{nodelayer}
@@
     \end{tikzpicture}
+    \caption{signed area of triangle}
     \end{subfigure}
-    \caption{some plots}
     \end{figure}
-
+    \begin{equation}
+        q(x^*) = \frac{1}{4\pi} \bigg[A\Big(\e(x_1), \e(x_2), \e(x_3)\Big) + A\Big(\e(x_1), \e(x_3), \e(x_4)\Big) \bigg].
+    \end{equation}
 \end{frame}
@@
 \begin{frame}
-    \frametitle{Wolff Cluster Algorithm}
-
+    \frametitle{Topological Charge Measurement}
     \begin{figure}[h]
     \centering
-          \begin{subfigure}[b]{0.5\textwidth}\centering
-            \includegraphics[width=0.6\textwidth]{imgs/wolffa.png}
-            \caption{before cluster flip}
-          \end{subfigure}%
-          \hfill
-          \begin{subfigure}[b]{0.5\textwidth}\centering
-            \includegraphics[width=0.6\textwidth]{imgs/wolffb.png}
-            \caption{after cluster flip}
-          \end{subfigure}
-          \hfill
-          \caption{\label{fig:wolff} An example of the Wolff cluster algorithm in the $\phi^4$ model.
-White represents positive values of $\phi$ while black represents negative.
-$\lambda=0.5$, $m_0^2=-0.9$.}
+      \includegraphics[width=0.6\textwidth]{imgs/hist.png}
+      \caption{Histogram of topological charge values $Q$ for trivial NLSM. $L=404$, 10,000 measurements.}
+\end{figure}
+      \note{Histogram of topological charge values $Q$ for trivial NLSM. $L=404$, 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization, $\tau=0$}
+\end{frame}
 
+\begin{frame}
+    \frametitle{Gradient flow on NLSM}
+    We solve the Gradient Flow numerically using
+    \begin{itemize}
+        \item fourth-order Runge Kutta
+        \item adaptive step size
+    \end{itemize}
+\end{frame}
+
+\begin{frame}
+    \frametitle{$\phi^4$ results}
+    \begin{figure}[h!]
+        \centering
+          \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
+          \caption{\label{fig:phi4} The lattice average $|\langle\bar\phi\rangle|$, the magnetic susceptibility $\chi_m$, the Binder cumulant $U\:$ ~\footfullcite{binder1981} and the bimodality $B$ plotted as functions of $m_0^2$.
+$L=64$, $\lambda=0.5$.
+1000 measurements}
+    \end{figure}
+    \note{First mention broken and symmetric phases}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Comparison with Berg \& L\"uscher}
+    \begin{figure}[h!]
+        \centering
+          \includegraphics[width=0.9\textwidth]{imgs/internal_energy.png}
+          \caption{\label{fig:bergluscher}Comparison with~\footfullcite{berg1981}, $L=100$, 1000 measurements.}
     \end{figure}
@@
 \begin{frame}
-\frametitle{Lattice Quantum Chromodynamics}
-\begin{itemize}
-    \item Method to solve non-perturbative theories
-    \item Discretizes spacetime to numerically calculate fields.
-    \item Uses path integral formulation of Quantum Field Theory to solve for correlation functions.
-\end{itemize}
-
-\begin{block}{Problem}
-    As lattice spacing approaches zero, some operators do not remain finite. 
-\end{block}
+    \frametitle{Topological Susceptibility in Flow Time}
+    \begin{figure}[h!]
+        \centering
+          \includegraphics[width=\textwidth]{imgs/bietenholz.png}
+          \caption{\label{fig:bietenholz} $\chi_t \xi_2^2$ as a  function of flow time $\tau$. Simulation run with 10,000 measurements }
+    \end{figure}
 \end{frame}
@@
 \begin{frame}
-\frametitle{The Gradient Flow}
-\begin{itemize}
-    \item A form of smearing
-    \item Introduces a new dimension, $\tau$, called the ``flow time,'' through which the field evolves as follows:
-    \begin{equation}
-        \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
-    \end{equation}
-    with the boundary condition $\rho(0, x)=\phi(x)$.
+    \frametitle{$\chi_t$ Divergence}
+    \begin{figure}[h!]
+        \begin{center}
+          \begin{subfigure}[b]{0.45\paperwidth}
 
-    \item Suppresses infinities while keeping observables constant.
-
-        %\begin{figure}[h]
-  %\centering
-      %\subfigure[$\tau=0$]{\includegraphics[width=0.20\textwidth]{imgs/0.png}}
-      %\subfigure[$\tau=0.5$]{\includegraphics[width=0.20\textwidth]{imgs/0_5.png}}
-      %\subfigure[$\tau=2$]{\includegraphics[width=0.20\textwidth]{imgs/2.png}} 
-      %\subfigure[$\tau=10$]{\includegraphics[width=0.20\textwidth]{imgs/10.png}}
-          %\caption{Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
-  %\label{fig:flow}
-%\end{figure}
-
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-    \frametitle{Goals}
-    \begin{itemize}
-        \item Computationally implement the simpler $\phi^4$ theory to verify code.
-        \item Transition to $O(3)$ nonlinear $\sigma$ model.
-        \item Calculate twist-2 operators and vacuum expectation value with gradient flow applied.
-        \item Other future paths:
-            \begin{itemize}
-                \item Consider topological charge
-                \item Explore Wilson coefficients 
-            \end{itemize}
-
-    \end{itemize}
-\end{frame}
-
-\begin{frame}
-    \frametitle{Broad Motivation}
-    To understand Deep Inelastic Scattering (DIS)
-    \begin{itemize}
-        \item Twist-2 operators $\rightarrow$ Mellin moments $\rightarrow$ Parton Distribution Functions $\rightarrow$ DIS Cross Sections
-    \end{itemize}
-\end{frame}
-%------------------------------------------------
-
-\begin{frame}
-    \frametitle{Computational Methods}
-\begin{itemize}
-    \item Use a Markov Chain Monte Carlo method to simulate lattice in 2D and 3D
-        \begin{itemize}
-            \item Metropolis algorithm to generate random changes
-            \item Wolff cluster algorithm to remove metastable states
-            \item Implement checkerboard algorithm to parallelize Metropolis algorithm
-        \end{itemize}
-
-    \item \textbf{By sampling the lattice at set intervals, we can calculate statical estimates of arbitrary operators.}
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}
-    \frametitle{Markov Chain Monte Carlo}
-\begin{enumerate}
-    \item Start with random lattice (known as ``hot start'')
-    \item Repeatedly transition from old configuration $\mu$ to new configuration $\nu$ with probability $P(\mu\rightarrow\nu)$.
-    \item After a thermalization period, record observables at certain intervals. 
-    \item Take the statistical mean
-\end{enumerate}
-\end{frame}
-
-\begin{frame}
-    \frametitle{Metropolis Algorithm}
-    For each site, propose a new value and accept change with probability
-        \begin{equation*}
-    P(\mu\rightarrow\nu) = \begin{cases} 
-        e^{-(S_\nu - S_\mu)} & S_\nu > S_\mu \\
-        1 & \mathrm{S_\nu \leq S_\mu} \\
-       \end{cases}
-        \end{equation*} 
-        A \textbf{sweep} consists of perfoming this on every site.
-
-\end{frame}
-
-\begin{frame}
-    \frametitle{Wolff Cluster Algorithm}
-        \textbf{Problem}: Markov Chain gets ``stuck'' in metastable states, i.e. local minima of the action.
-
-        \textbf{Solution}: Identify cluster of similar-values points and flip the sign.
-    \begin{enumerate}
-        \item Recursively grow a cluster with the probability of adding a new site
-        \begin{equation*}
-            P_{add} = 1-e^{-2\phi_a\cdot\phi_b}.
-        \end{equation*}
-        \item Flip sign of sites
-
-    \end{enumerate}
-    %\begin{figure}[h]
-
-      %\centering
-          %\subfigure[before]{\includegraphics[width=0.20\textwidth]{imgs/wolff1}}
-          %\subfigure[after]{\includegraphics[width=0.20\textwidth]{imgs/wolff2}}
-          %\caption{Wolff cluster algorithm applied on $\phi^4$ lattice in $1+1$ dimensions, red corresponding to positive values and green to negative}
-      %\label{fig:flow}
-    %\end{figure}
-\end{frame}
-
-\begin{frame}
-    \frametitle{$\phi^4$ model}
-    \begin{itemize}
-        \item Given by the Euclidean action
-            \begin{equation}
-                S_E = \int d^D x \left[ \partial^\mu \phi \partial_\mu\phi + \frac{1}{2} m_0^2 \phi^2 + \frac{\lambda}{4}\phi^4\right]
-            \end{equation}
-
-        \item Simplest interacting field
-        \item In 1+1 dimensions, exhibits spontaneous symmetry breaking at critical $m_0^2$.
-
-    \end{itemize}
-
-    We measure\ldots
-\begin{itemize}
-    \item Magnetization: $\langle |\bar\phi|\rangle$
-    \item Susceptibility: $\chi = \langle {\bar\phi}^2\rangle - \langle\bar\phi\rangle^2$
-    \item Binder cumulant: $U = 1 - \frac{\langle {\bar\phi}^4\rangle}{3 \langle {\bar\phi}^2\rangle^2}$
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-    \frametitle{Computational Implementation}
-    \begin{enumerate}
-        \item Begin with simple Python $\phi^4$ model with Cython for efficiency
-        \item Transition to C++ code, generalized to nonlinear $\sigma$ model
-        \item Compare results of Python and C++ code.
-    \end{enumerate}
-    \begin{block}{Parallelization}
-        Message Passing Interface: shared RAM between processes.
-    \end{block}
-\end{frame}
-
-\begin{frame}
-\frametitle{Spontaneous Symmetry Breaking in $\phi^4$ model}
-\begin{figure}[h]
               \centering
-      \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
-      \caption{Average $\bar\phi$, Binder cumulant and susceptibility for various $m_0^2$ values.
-$\lambda=0.5$, $L=64$, $1000$ measurements, $1000$ sweep thermalization with measurements taken every $100$ sweeps.}
-  \label{fig:flow}
+              \includegraphics[width=0.9\textwidth]{imgs/divergence.png}
+              \caption{$\tau = 0$}
+          \end{subfigure}%
+          \begin{subfigure}[b]{0.45\paperwidth}
+              \centering
+              \includegraphics[width=0.9\textwidth]{imgs/divergence_flowed.png}
+              \caption{$\tau = 5t_0$}
+          \end{subfigure}
+        \end{center}
+        \caption{Divergent properties with comparison to~\footfullcite{bietenholz2018}, 10,000 measurements}
     \end{figure}
+          \note{Simulation run with 10,000 measurements, once every 50 sweeps, 1,000 sweep thermalization. In the tau=0 case, we have compared our result with the curve fit found in Bietenholz}
 \end{frame}
@@
           \includegraphics[width=0.8\textwidth]{imgs/theta.png}
-      %\caption{Absolute difference between Python and C++ results of average $\bar\phi$, Binder cumulant and susceptibility for various $m_0^2$ values.
-$\lambda=0.5$, $L=64$, $1000$ measurements, $1000$ sweep thermalization with measurements taken every $100$ sweeps.
-}
-
-  \label{fig:flow}
     \end{figure}
@@
 \begin{frame}
-\frametitle{Future Work}
+    \frametitle{Conclusions}
     \begin{itemize}
-    \item Fully implement nonlinear $\sigma$ model.
-    \item Calculate twist-2 operators and vacuum expectation value
-    \item Expand to 3 dimensions
-    \item Profile and optimize C++ code
-    \item Mathematical analysis of results
+        \item  Berg \& L\"uscher~\footfullcite{berg1981} provide three possible reasons
+            \begin{enumerate}
+                \only<1>{\item high-frequency modes cause divergence}
+                \only<2->{\item {\transparent{0.4} high-frequency modes cause divergence}}
+                \item the definition of $Q$ is problematic
+                \item the NLSM has no well-defined continuum limit
+            \end{enumerate}
+
+        \item<3-> Relatively high $\chi^2/DOF$ values indicate errors underestimated.
+        \item<4-> Clearer perspective on the effect $\theta$ helpful for condensed matter systems~\footfullcite{bogli2012}.
     \end{itemize}
-
 \end{frame}
-
-%%------------------------------------------------
-
-%\begin{frame}
-%\frametitle{Multiple Columns}
-%\begin{columns}[c] % The "c" option specifies centered vertical alignment while the "t" option is used for top vertical alignment
-
-%\column{.45\textwidth} % Left column and width
-%\textbf{Heading}
-%\begin{enumerate}
-%\item Statement
-%\item Explanation
-%\item Example
-%\end{enumerate}
-
-%\column{.5\textwidth} % Right column and width
-%Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Integer lectus nisl, ultricies in feugiat rutrum, porttitor sit amet augue.
-Aliquam ut tortor mauris.
-Sed volutpat ante purus, quis accumsan dolor.
-
-%\end{columns}
-%\end{frame}
-%\begin{frame}
-%\frametitle{Blocks of Highlighted Text}
-
-%\begin{block}{Block 2}
-%Pellentesque sed tellus purus. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos. Vestibulum quis magna at risus dictum tempor eu vitae velit.
-%\end{block}
-
-%\begin{block}{Block 3}
-%Suspendisse tincidunt sagittis gravida. Curabitur condimentum, enim sed venenatis rutrum, ipsum neque consectetur orci, sed blandit justo nisi ac lacus.
-%\end{block}
-%\end{frame}
-
-%%------------------------------------------------
-
-%\begin{frame}
-%\frametitle{Multiple Columns}
-%\begin{columns}[c] % The "c" option specifies centered vertical alignment while the "t" option is used for top vertical alignment
-
-%\column{.45\textwidth} % Left column and width
-%\textbf{Heading}
-%\begin{enumerate}
-%\item Statement
-%\item Explanation
-%\item Example
-%\end{enumerate}
-
-%\column{.5\textwidth} % Right column and width
-%Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Integer lectus nisl, ultricies in feugiat rutrum, porttitor sit amet augue.
-Aliquam ut tortor mauris.
-Sed volutpat ante purus, quis accumsan dolor.
-
-%\end{columns}
-%\end{frame}
-
-%%------------------------------------------------
-%\section{Second Section}
-%%------------------------------------------------
-
-%\begin{frame}
-%\frametitle{Table}
-%\begin{table}
-%\begin{tabular}{l l l}
-%\toprule
-%\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2}\\
-%\midrule
-%Treatment 1 & 0.0003262 & 0.562 \\
-%Treatment 2 & 0.0015681 & 0.910 \\
-%Treatment 3 & 0.0009271 & 0.296 \\
-%\bottomrule
-%\end{tabular}
-%\caption{Table caption}
-%\end{table}
-%\end{frame}
-
-%%------------------------------------------------
-
-%\begin{frame}
-%\frametitle{Theorem}
-%\begin{theorem}[Mass--energy equivalence]
-%$E = mc^2$
-%\end{theorem}
-%\end{frame}
-
-%%------------------------------------------------
-
-%\begin{frame}[fragile] % Need to use the fragile option when verbatim is used in the slide
-%\frametitle{Verbatim}
-%\begin{example}[Theorem Slide Code]
-%\begin{verbatim}
-%\begin{frame}
-%\frametitle{Theorem}
-%\begin{theorem}[Mass--energy equivalence]
-%$E = mc^2$
-%\end{theorem}
-%\end{frame}\end{verbatim}
-%\end{example}
-%\end{frame}
-
-%%------------------------------------------------
-
-%\begin{frame}
-%\frametitle{Figure}
-%Uncomment the code on this slide to include your own image from the same directory as the template .TeX file.
-%%\begin{figure}
-%%\includegraphics[width=0.8\linewidth]{test}
-%%\end{figure}
-%\end{frame}
-
-%%------------------------------------------------
-
-%\begin{frame}[fragile] % Need to use the fragile option when verbatim is used in the slide
-%\frametitle{Citation}
-%An example of the \verb|\cite| command to cite within the presentation:\\~
-
-%This statement requires citation \cite{p1}.
-%\end{frame}
-
-%%------------------------------------------------
-
-%\begin{frame}
-%\frametitle{References}
-%\footnotesize{
-%\begin{thebibliography}{99} % Beamer does not support BibTeX so references must be inserted manually as below
-%\bibitem[Smith, 2012]{p1} John Smith (2012)
-%\newblock Title of the publication
-%\newblock \emph{Journal Name} 12(3), 45 -- 678.
-%\end{thebibliography}
-%}
-%\end{frame}
-
-%%------------------------------------------------
-
-%\begin{frame}
-%\Huge{\centerline{The End}}
-%\end{frame}
-
-%%----------------------------------------------------------------------------------------
-
+\begin{frame}
+    \frametitle{Acknowledgements}
+    \begin{itemize}
+        \item<1-> 1693 Scholars program (funding for summer research)
+        \item<2-> HPC Computing Group
+        \item<3-> Prof. Christopher Monahan
+        \item<4-> Friends and family
+    \end{itemize}
+\end{frame}
 \end{document}

### 2021-10-07 abstract.latex/abstract.tex
fix .gitignore issue

--- 2021-04-19 final-report.latex/main.tex
@@
 \usepackage{amsmath}
-\usepackage{tabularx}
+\usepackage{bbm}
 \usepackage{amsfonts}
-\usepackage{hyperref}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta}
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{listings}
+
 \setlength{\oddsidemargin}{0.5in}
@@
 \setlength{\textheight}{8.6in}
-
-\newcommand{\e}{\vec e}
-\newcommand{\citeneeded}{\footnote{\textit{citation needed}}}
-\newcommand{\half}{\frac{1}{2}}
-\newcommand{\codeexample}[1]{
-    \section{\texttt{#1}}
-    \lstinputlisting[language=C++]{../code/cpp_code/#1}
-}
 
@@
 \begin{document}
-\setlength{\baselineskip}{24pt}
-\begin{titlepage}
-\LARGE
-\begin{center}
-    {\bf Topology of the $O(3)$ non-linear sigma model under the gradient flow}\\[2.3cm]
-
-\normalsize
-A thesis submitted in partial fulfillment of the requirement \\
-for the degree of Bachelor of Science with Honors in \\
-Physics from the College of William and Mary in Virginia,\\[0.5cm]
-by\\[0.5cm]
-Stuart Thomas \\[2.5cm]
-\end{center}
-\normalsize
-\begin{flushright}
-%\hfill Accepted for Honors \\[.5cm]
-\hfill \hrulefill \\
-\hfill \hfill Advisor: Prof. Christopher J. Monahan\\[.6cm]
-\hfill \hrulefill \\Prof. Todd Averett \\[.6cm]
-\hfill \hrulefill \\Prof. Andreas Stathopoulos \\[.6cm]
-\end{flushright}
-\begin{center}
-Williamsburg, Virginia\\
-May 2021
-\end{center}
-\end{titlepage}
-
-\setlength{\topmargin}{0.0in}
-
-\pagenumbering{roman}
-\tableofcontents
-\chapter*{Acknowledgments}
-\addcontentsline{toc}{chapter}{Acknowledgments}
-\paragraph
-\indent I would like to thank my advisor, Prof.
-Christopher Monahan.
-His constant helpfulness has made this research possible and enjoyable.
-I would also like to thank the entire faculty of the William \& Mary Physics Department.
-The amount of knowledge and wisdom they have bestowed upon me has been beyond my imagination.
-
-I also acknowledge the help the High Performance Computing group at William \& Mary, whose cluster performed many of these simulations.
-And finally, I would like to acknowledge the financial help of the 1693 Scholars Program which funded the summer portion of this research.
-
-\listoffigures
-\addcontentsline{toc}{chapter}{List of Figures}
-\listoftables
-\addcontentsline{toc}{chapter}{List of Tables}
-
 \begin{abstract}
@@
 \paragraph
-\indent The $O(3)$ non-linear model features topological effects such as instantons but cannot be perturbatively renormalized.
-Other nonperturbative theories such as QCD have motivated the development of lattice quantum field theory, where fields are simulated as discrete arrays and extrapolated to the continuum limit.
-Unfortunately, operators such as the topological susceptibility diverge in this limit.
-The gradient flow is a strong contender to remove these infinites, having prior success in QCD.
-We confirm a previous result\cite{bietenholz2018} that the gradient flow does not make the topological susceptibility infinite.
-We also study the nontrivial field theory by introducing a $\theta$ term into the action and plot the topological charge as a function of $\theta$.
+\indent The $O(3)$ non-linear sigma model (NLSM) is a prototypical field theory for QCD and ferromagnetism, featuring topological qualities.
+Though the topological susceptibility $\chi_t$ should vanish in physical theories, lattice simulations of the NLSM find that $\chi_t$ diverges in the continuum limit.
+We study the effect of the gradient flow on this quantity using a Markov Chain Monte Carlo method, finding that a logarithmic divergence persists.
+This result supports a previous study and indicates that either the definition of topological charge is problematic or the NLSM has no well-defined continuum limit.
+We also introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow.
 
 \end{abstract}
-
-\pagenumbering{arabic}
-\chapter{Introduction}
-\input{intro.tex}
-\chapter{Theory}
-\input{theory.tex}
-\chapter{Methods}
-\label{sec:methods}
-\input{methods.tex}
-\chapter{Results}
-\input{results.tex}
-\chapter{Conclusion \& Outlook}
-\input{conclusion.tex}
-\appendix
-\chapter{C++ NLSM Monte Carlo Program}
-\lstset{language=C++,
-                basicstyle=\scriptsize\ttfamily,
-                keywordstyle=\color{blue}\ttfamily,
-                stringstyle=\color{red}\ttfamily,
-                commentstyle=\color{green}\ttfamily,
-                morecomment=[l][\color{magenta}]{\#},
-                breaklines=true,
-                postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}
-}
-
-\codeexample{sweep.h}
-\codeexample{sweep.cpp}
-\codeexample{lattice.h}
-\codeexample{lattice.cpp}
-\codeexample{phi.h}
-\codeexample{phi.cpp}
-\codeexample{observables.h}
-\codeexample{constants.h}
-
-\bibliographystyle{unsrt}
-\bibliography{library}
-
 \end{document}

### 2021-10-07 midyear-report/midyear-report.tex
fix .gitignore issue

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subfigure}
\setlength{\oddsidemargin}{0.5in}
\setlength{\evensidemargin}{0.5in}
\setlength{\textwidth}{6.0in}
\setlength{\topmargin}{.75in}
\setlength{\textheight}{8.6in}

% Modified by S. Aubin for the 2018 mid-year reports

%
% This is a LaTeX template for an undergraduate mid-year report
%   in physics at William & Mary. Created from
%   several theses from previous students of D.S. Armstrong,
%   pared down to bare bones.
%
%   You will need this present file (MidyearReport_template.tex)
%   as well as the file figure1.pdf  to make the whole document

%  
%   For simplicity, we don't use bibtex here; experts can choose
%   to use that more powerful way of dealing with citations.
%

\begin{document}
\setlength{\baselineskip}{24pt}
\begin{titlepage}
\LARGE
\begin{center}
Mid-Year Report\\[2.3cm]

{\bf Defining condensates in low-dimensional spin models through the gradient flow}\\[2.3cm]

\normalsize

Stuart Thomas\\[2.5cm]
\end{center}
\normalsize
\begin{flushright}
%\hfill Accepted for Honors \\[.5cm]
\hfill \hrulefill \\
\hfill \hfill Advisor: Prof. Christopher Monahan \\[7.0cm]
%\hfill \hrulefill \\Prof. Richard P. Feynman \\[.6cm]
%\hfill \hrulefill \\Prof. Marie Curie \\[.6cm]
%\hfill \hrulefill \\Prof. Ernest J. Rutherford \\[.6cm]
\end{flushright}
\begin{center}
Department of Physics\\
College of William and Mary\\
Williamsburg, Virginia\\
November 2020
\end{center}
\end{titlepage}

\setlength{\topmargin}{0.0in}

\pagenumbering{roman}

\setcounter{page}{5}

\begin{abstract}

\indent
We study the the effect of operator smearing---defined as a local averaging of the field---on the $O(3)$ nonlinear $\sigma$ model, particularly analyzing renormalization properties. Specifically, we analyze the ``gradient flow,'' a gauge-invariant form of smearing which introduces a new dimension known as ``flow time.'' We perform Monte-Carlo simulations with the Metropolis algorithm and the Wolff cluster algorithm to numerically calculate correlation functions in small ($N\sim 32$) and mid-size ($N\sim128$) lattices, in 1+1 and 2+1 dimensions. To verify numerics, we first find the phase transition of scalar $\phi^4$ theory to be at $m_0^2=-0.72$ and $\lambda=0.5$, in agreement with previous literature.

\end{abstract}

\section{Introduction}
\pagenumbering{arabic}
\indent Lattice quantum chromodynamics (QCD) provides a method to explore and understand the strong nuclear force by discretizing spacetime. Specifically, lattice QCD can provide an understanding of nonperturbative theories which cannot be solved with traditional QFT techniques. With this technique, the path integrals that are key to quantum field theory become manageable and numerically calculable. When taking the lattice spacing to zero, which requires renormalizing all observables, we expect operators to remain finite, however this is not always the case. Some observables mix on the lattice, leading to divergencies. As an example, states of definite angular momentum mix when discretized. In our paper, we study power-divergent mixing, arising from the mixing of operators with different mass dimension. \cite{monahan2016}.

\subsection{Gradient Flow}

One technique to resolve this divergence is by ``smearing,'' a local averaging of a gauge field in order to remove diverging fluctuations \cite{solbrig2007}. Specifically, we use a technique known as ``gradient flow'' \cite{monahan2015} which introduces a new half-dimension called ``flow time''. The flow time parameterizes the smearing; an evolution in flow time corresponds to suppressing ultraviolet divergences. In a 2D $\phi^4$ scalar field theory, given by the action
\begin{equation}
    S_\phi [\phi] = \frac{1}{2}\int d^2x\left[(\partial \phi)^2+m^2\phi^2+\frac{\lambda}{2}\right],
\end{equation}
we can define the evolution in flow time as
\begin{equation}
    \frac{\partial \rho(\tau, x)}{\partial \tau} = \partial^2 \rho(\tau,x)
\end{equation}
where $\partial^2$ is the Laplacian in 4-D Euclidean spacetime and $\tau$ is the flow time. Here, $\rho$ is the field evolved into a nonzero flow time, bounded by the condition $\rho(\tau=0,x) = \phi(x)$. When solved exactly, this function forms a Gaussian which forces operators to converge. Therefore, in all orders of pertubation theory, any correlation function in the gauge field that has been evolved in flow time becomes immune to these divergences\cite{makino2015a}.

Generally, we can choose any flow time equation that drives the field towards a classical minimum. Beyond the $\phi^4$ model, we need flow equations that incorporate different types of fields. In the nonlinear $\sigma$ model, it has been shown that an appropriate manifestation of the flow time can resolve divergencies \cite{makino2015a}. Therefore, it is possible to define a different flow equation for this model as well. Our goal is to understand this renormalizability in the nonlinear $\sigma$ model using the gradient flow. Specifically, we seek to define vacuum condensates, or nonpertubative vacuum expectation values.

To explore this technique, we first plan to produce numerical calculations using a nonlinear $\sigma$ model. Using a Monte Carlo method in 1+1 dimensions, we can calculate the correlation functions of quantum field theories. With more computing power, expanding this simulation to 2+1 dimensions would provide an analysis more relevant to quantum chromodynamics.  Following this numerical calculation, we seek to find and understand an implementation of the gradient flow in this model using analytic methods. This understanding can then be verified by the numerical solution.

\subsection{Field theories on the lattice}

The basis for lattice quantum field theory lies in the Feynman path integral formulation of quantum field theory. This formalism predicts the behavior of quantum fields using a ``path integral,'' or a integral of a function over all possible field configurations in space and time. Though nearly impossible to solve analytically, discretizing the field makes numerical solutions clear. Particularly, the correlation functions of the theory can be calculated using the generating functional, given as 
\begin{equation}
    Z[J] = \int \mathcal{D} \phi e^{iS[\phi]}
\end{equation}
where $\int \mathcal{D} \phi$ is a path integral over $\phi$ and $S$ is the action. We can apply a Wick rotation to the lattice to transform the Minkowski spacetime to Euclidean spacetime. This change of basis produces an alternative definition 
\begin{equation}
    Z[J] = \int \mathcal{D}\phi e^{-S_E[\phi]}
\end{equation}
where $S_E$ is the action in Euclidean space.

Theoretically, we could calculate every possible configuration of $\phi$ to obtain an exact value of $Z[J]$, but the computational time makes this method impossible. A more intelligent approach starts with noticing that significant contributions are relegated to the minimum of the action. Therefore, we can accurately determine the generating function, and also most relevant physics, by only considering contributions near the classical limit. However, determining which configurations to consider is nontrivial and is discussed in Section~\ref{sec:mc}.

In order to compare with existing literature, we first study the $\phi^4$ model, described by the action 
\begin{equation}
    S_E = \int d^D x \left[ \partial^\mu \phi \partial_\mu\phi + \frac{1}{2} m_0^2 \phi^2 + \frac{\lambda}{4}\phi^4\right]
\end{equation}
where $m_0^2$ and $\lambda$ are parameters of the theory. This model will provide a stepping stone for the vector-valued nonlinear $\sigma$ model.

\subsection{Non-linear $\sigma$ model}

The nonlinear $\sigma$ model is a prototypical theory for many physical phenomena, including applications in string theory. As a simple nonperturbative model, it provides an ideal starting point for lattice QCD studies. Specifically, the nonlinear $\sigma$ model exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability. Furthermore, it has an exact application to Heisenberg ferromagnets in condensed matter. The theory is defined by the action 
\begin{equation}
    S_E = \frac{1}{2g^2} \int d^Dx \partial^\mu \vec\phi \cdot \partial_\mu \vec\phi
\end{equation}
subject to the constraint that $\vec\phi\cdot\vec\phi = 1$. In this report, we primarily consider $O(3)$, where $\vec\phi$ is a $3$ dimensional vector.

In order to use this theory on the lattice, we discretize the action. We replace the derivatives with a simple difference and impose periodic boundary conditions on the system.

\section{Computational Framework}
\label{sec:methods}
Our study of the gradient flow in the nonlinear $\sigma$ model consists of a computational part and an analytical part. We begin by outlining a numerical Monte Carlo method to simulate the lattice in 2 and 3 dimensions. We verify our program with the well-studied $\phi^4$ scalar field theory. Then, we plan to use this system to calculate twist-2 operators and vacuum expectation values in flow time, supporting a subsequent analytical study of such objects.

We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories.

\subsection{Monte Carlo Simulations}
\label{sec:mc}

We implement a Markov Chain Monte Carlo method following Schaich's thesis \cite{schaich2006}. This implementation utilizes a ``random walk,'' i.e. a set of random steps through phase space, to determine statistical values such as correlation functions across the lattice. By the definition of the Markov chain, the probability of adoption of each state, and therefore its inclusion in the Monte Carlo calculation, depends only on the current state and the proposed state. This probability is denoted as $P(\mu\rightarrow\nu)$ where $\mu$ and $\nu$ are the existing and proposed lattice configurations respectively. We use a combination of the Metropolis and Wolff algorithms to determine this value.

\subsubsection{Metropolis Algorithm}
We primarily use the Metropolis algorithm for the calculation of new Markov chain configurations. We begin with a so-called ``hot start,'' where each field value at each lattice site is randomly selected. Then we propose a new value for a single lattice point, which is accepted with a probability
\begin{equation}
    P(\mu\rightarrow\nu) = \begin{cases} 
        e^{-(S_\nu - S_\mu)} & S_\nu > S_\mu \\
        1 & \mathrm{otherwise} \\
   \end{cases}
\end{equation}
where $S_\nu$ and $S_\mu$ are the lattice actions of configuration before and after the change. This process is performed for each point on the lattice, making up a ``sweep.'' Repeating this sweep many times pushes the lattice toward the action minimum.

\subsubsection{Wolff Cluster Algorithm}

Though the Metropolis algorithm will slowly find the absolute minimum of the theory, the presence of local minima can greatly prolong the convergence. Both the $\phi^4$ model and the nonlinear $\sigma$ model feature ``kinetic'' terms with gradients of $\phi$. Therefore, the presence of large similarly-valued regions in the lattice can lead to a local minimum. The Wolff algorithm helps reduce the presence of these clusters through two steps: identifying a cluster and flipping it along some arbitrary vector. In the case of $\phi^4$ theory, this flipping takes the form of a simple sign change. To identify the cluster, the algorithm uses a recursive algorithm defined by the probability of adding a new site. In the $\phi^4$ model, this takes the form of
\begin{equation}
    P_{add} = 1-e^{-2\phi_a\phi_b}.
\end{equation}

\subsubsection{Checkerboard algorithm}

In order to parallelize the Metropolis algorithm, we use a checkerboard algorithm. Since the Lagrangian density at each site does not depend on any sites diagonal to it, the lattice can be split into ``white'' sites and ``black'' sites, like the tiles on a checkerboard. Each white site is independent of every other white site and likewise with black sites. Therefore, we can split the sites of each color into separate parallel processing nodes and independently run the Metropolis algorithm, ensuring that no site affects the Lagrangian density at any other site. We use this method to parallelize the code through the Message Passing Interface (MPI).

\subsubsection{Autocorrelation times and thermalization}
One important aspect is the correlation between different states of the Markov chain. Ideally, each sample of the lattice would be completely independent, but this is not the case. Therefore, we must sweep the lattice sufficiently to produce an effective Monte Carlo result. We use Wolff's automatic windowing procedure \cite{wolff2007} and the $|\bar\phi|$ operator to estimate the autocorrelation and determine an appropriate number of sweep between measurements for each theory. Furthermore, the initial values of our Markov chain will be highly correlated with the initial random start. Therefore, we also wait a set number of sweeps before taking any measurements so that the lattice can approach a minimum of the action.

\section{Results}
We initially implemented the $\phi^4$ model using the Monte Carlo method described in Sec.~\ref{sec:methods} to verify the results of our system. According to previous studies (\cite{monahan2016}, \cite{schaich2006}), the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ for $\lambda = 0.5$. We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.

\begin{figure}[h]
  \centering
      \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
      \caption{The lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$. $N=32$, $\lambda=0.5$. The lattice was thermalized from a hot start for 1000 sweeps. Afterwards, 100 measurements were taken with 100 sweeps between each. The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
  \label{fig:phi4}
\end{figure}

To improve efficiency, we transitioned from a Python based simulation to a C++ code. The comparison between these results are found in Fig.~\ref{fig:compare}, using the same values from Fig.~\ref{fig:phi4}.

\begin{figure}[h]
  \centering
      \includegraphics[width=0.7\textwidth]{imgs/compare.png}
      \caption{Comparison between Python and C++ code, using the lattice average $|\langle\bar\phi\rangle|$, the lattice variance $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$. $N=32$, $\lambda=0.5$. The lattice was thermalized from a hot start for 1000 sweeps. Afterwards, 100 measurements were taken with 100 sweeps between each. The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
  \label{fig:compare}
\end{figure}

Furthermore, we implemented the gradient flow in this theory. Fig.~\ref{fig:flow} demonstrates the effect of the gradient flow on scalar lattice, specifically the dampening of high frequency modes.
\begin{figure}[h]
  \centering
      \subfigure[$\tau=0$]{\includegraphics[width=0.20\textwidth]{imgs/0.png}}
      \subfigure[$\tau=0.5$]{\includegraphics[width=0.20\textwidth]{imgs/0_5.png}}
      \subfigure[$\tau=2$]{\includegraphics[width=0.20\textwidth]{imgs/2.png}} 
      \subfigure[$\tau=10$]{\includegraphics[width=0.20\textwidth]{imgs/10.png}}
          \caption{Effect of flow time evolution on a random lattice in the symmetric phase. Red and blue indicate positive and negative values of the field in 2D spacetime.}
  \label{fig:flow}
\end{figure}

\section{Future work}
All of our current results apply only to the $\phi^4$ theory. Therefore, our future work primarily includes completing the migration to the nonlinear $\sigma$ model implemented in C/C++. Afterwards, we plan on analyzing the effect of the gradient flow numerically. Finally, we plan to use analytical techniques to understand the numerical results.

Our main numerical result will be the vacuum expectation values of operators, which we will use to define condensates in the nonlinear $\sigma$ model. These condensates are vital for understanding the physical make-up of hadrons and therefore would be valuable to the field of QCD. We also may consider topological states on the lattice.

\begin{flushleft}
\bibliographystyle{unsrt}
\bibliography{midyear-report}
%\begin{thebibliography}{99.}

%\bibitem{Perkins} Perkins, Donald H.  Introduction to High Energy
%Physics.  3rd ed.  Menlo Park: Addison-Wesley, 1987.

%\bibitem{Perkins} Perkins, Donald H.  Introduction to High Energy
%Physics.  3rd ed.  Menlo Park: Addison-Wesley, 1987.

%\bibitem{IsgurKumano} F. Close, N. Isgur and S. Kumano,
%Nucl. Phys. {\bf B389}, 513 (1993).

%\bibitem{UPV} Alex Dubanowitz, Senior project thesis, College of
%William and Mary, 1998. (unpublished).

%\end{thebibliography}
\end{flushleft}

\end{document}

### 2021-10-07 poster.latex/main.tex
fix .gitignore issue

\documentclass[18pt]{beamer}
\mode<presentation>{
    \usetheme{WM}
    \usefonttheme{serif}
    \setbeamertemplate{caption}[numbered]
}
\usepackage{amsmath,amsfonts,amssymb,eulervm,xspace}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\graphicspath{{figures/}}
\usepackage[orientation=landscape,size=custom,width=70,height=40,scale=.6]{beamerposter}
\usepackage[%
    style=phys,%
    articletitle=false,biblabel=brackets,%
    chaptertitle=false,pageranges=false,%
    doi=false,maxnames=2,minnames=1,eprint=false%
  ]
{biblatex}
\addbibresource{library.bib}
\makeatletter
\renewcommand{\@makefnmark}{\makebox{\normalfont[\@thefnmark]}}
\renewcommand\@makefntext[1]{%
    {{\normalfont[\@thefnmark]}\enspace #1}
  }
\makeatother
%\renewcommand{\cite}[1]{[\footfullcite{#1}]}
\renewenvironment{equation}
    {
    \begin{equation*}
    }
    { 
    \end{equation*}
    }

\newcommand{\e}{\vec e}
%-- Header and footer information ----------------------------------
\newcommand{\footleft}{https://scholarworks.wm.edu/honorstheses/1621}
\newcommand{\footright}{snthomas01@email.wm.edu}
\title{Topology of the $O(3)$ non-linear sigma model under the gradient flow}
\author{Stuart Thomas\inst{1} \quad Christopher Monahan\inst{1,2}}
\institute{\inst{1} William \& Mary \quad \inst{2}Jefferson Lab}
%-------------------------------------------------------------------

%-- Main Document --------------------------------------------------
\begin{document}
\begin{frame}{}
  \begin{columns}[t]

    %-- Column 1 ---------------------------------------------------
    \begin{column}{0.32\linewidth}
        \begin{block}{Abstract}
            The $O(3)$ non-linear sigma model (NLSM) is a prototypical field theory for QCD and ferromagnetism, featuring topological qualities. {\bf Though the topological susceptibility $\chi_t$  should vanish in physical theories, lattice simulations of the NLSM find that $\chi_t$ diverges in the continuum limit. \cite{bietenholz2018, berg1981}} We study the effect of the gradient flow on this quantity using a Markov chain Monte Carlo method, finding that a logarithmic divergence persists. This result supports a previous study and indicates that either the definition of topological charge is problematic or the NLSM has no well-defined continuum limit. We introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow for the first time.
        \end{block}
        \begin{block}{Non-Linear $\sigma$ Model}
            We study the $O(3)$ non-linear sigma model (NLSM) in 1+1 dimensions, defined by the Euclidean action
            \begin{equation*}
                \label{eq:nlsm euclidean action}
                S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right]
            \end{equation*}
        where
        \begin{itemize}
            \item $\e$ is 3-component real vector constrained by $|\e\,|=1$.
            \item $\beta$ is the inverse coupling constant.
            %\item<3-> Applications
                %\begin{itemize}
                    %\item Prototypical model for strong nuclear force
                    %\item Models Heisenberg ferromagnets
                    %\item Applications to string theory
                %\end{itemize}
            %\item<4-> Merits 
                %\begin{itemize}
                    %\item mass gap
                    %\item asymptotic freedom
                %\end{itemize}
        \end{itemize}

        \end{block}

        \begin{block}{Topology of 1+1 $O(3)$ NLSM}
            \begin{columns}
            \begin{column}{0.6\textwidth}
            \begin{enumerate}
                \item Since the Lagrangian must disappear as $x \rightarrow \infty$, the field $\e$ becomes uniform.
                \item Therefore, we can envision 1+1 spacetime as a sphere.
                \item The NLSM field $\e$ becomes mapping between two Riemann spheres ($S^2 \rightarrow S^2$).
                \item Every configuration has an associated {\bf topological charge} $Q\in\mathbb{Z}$ (see Fig.~\ref{fig:homotopy}).
                \item The ensemble has a {\bf topological susceptibility} $\chi_t \equiv \left(\langle Q^2 \rangle - \langle Q \rangle^2\right)/L^2$ which should disappear in the continuum limit.
            \end{enumerate}
            \end{column}
            \begin{column}{0.3\textwidth}
            \begin{figure}
                \includegraphics[width=0.9\textwidth]{homotopy.png}
                \caption{\label{fig:homotopy} Visualization of homotopy group of $S^1\rightarrow S^1$}
            \end{figure}
            \end{column}
            \end{columns}
        \end{block}

        \begin{block}{The Gradient Flow}
        \begin{itemize}
            \item Successful in removing $\chi_t$ divergence in QCD by reducing ultraviolet divergences.
            \item Introduces a new dimension, $\tau$, called the ``{\bf flow time},'' which pushes fields towards action minima 
            %\item For NLSM, we use a fourth-order Runge-Kutta to approximate the flow.
        \end{itemize}
        \begin{figure}[h]
          \centering
          \def \Width {0.14\textwidth}
              \begin{subfigure}[b]{\Width}\centering
                \includegraphics[width=0.9\textwidth]{gf0.png}
                \caption{$\tau=0$}
              \end{subfigure}%
              \begin{subfigure}[b]{\Width}\centering
                \includegraphics[width=0.9\textwidth]{gf1.png}
                \caption{$\tau=0.001$}
              \end{subfigure}%
              \begin{subfigure}[b]{\Width}\centering
                \includegraphics[width=0.9\textwidth]{gf2.png}
                \caption{$\tau=0.01$}
              \end{subfigure}%
              \begin{subfigure}[b]{\Width}\centering
                \includegraphics[width=0.9\textwidth]{gf3.png}
                \caption{$\tau=0.1$}
              \end{subfigure}%
              \caption{\label{fig:flow} Visualization of gradient flow in $\phi^4$ model}
        \end{figure}
        \end{block}
    \end{column}%1
    \begin{column}{0.32\linewidth}
        %\begin{block}{Research Question}
            %{\bf Can the gradient flow remove the topological susceptibility divergence in the NLSM?}
        %\end{block}

        \begin{block}{Computational Method}
            We generate lattice field configurations with a Markov Chain Monte Carlo method
            \begin{itemize}
                \item Metropolis algorithm on each site, forming a ``{\bf sweep}'' of the lattice.
                \item Wolff cluster algorithm every five Metropolis sweeps
                \begin{figure}[h]
                  \centering
                      \begin{subfigure}[b]{0.3\textwidth}\centering
                        \includegraphics[width=0.5\textwidth]{wolffa.png}
                        \caption{before cluster flip}
                      \end{subfigure}%
                      \begin{subfigure}[b]{0.3\textwidth}\centering
                        \includegraphics[width=0.5\textwidth]{wolffb.png}
                        \caption{after cluster flip}
                      \end{subfigure}
                      \hfill
                      \caption{\label{fig:wolff} Visualization of Wolff cluster algorithm in $\phi^4$ model}
                \end{figure}
                 \item Thermalize with 1,000 sweeps
                \item Sample every 50 sweeps to reduce autocorrelation
            \end{itemize}
            %\hline

            The ensemble has a topological susceptibility $\chi_t \equiv \left(\langle Q^2 \rangle - \langle Q \rangle^2\right)/L^2$ which should disappear in the continuum limit.
        \end{block}

        \begin{block}{Topological Charge on the Lattice}
            %Following Berg \& L\"usher~\footfullcite{berg1981}, define topological charge density $q(x^*)$ for each plaquette $x^*$ such that
            Following~\cite{berg1981}, we define topological charge density $q(x^*)$ for each plaquette $x^*$ such that
            \begin{equation}
                Q = \sum_{x^*} q(x^*)
            \end{equation}
            \begin{equation}
                q(x^*) = \frac{1}{4\pi} \bigg[A\Big(\e(x_1), \e(x_2), \e(x_3)\Big) + A\Big(\e(x_1), \e(x_3), \e(x_4)\Big) \bigg].
            \end{equation}
            \begin{figure}[H]
                \begin{subfigure}[b]{0.4\textwidth}
                \centering
                \begin{tikzpicture}[scale=0.8, every node/.style={scale=1.2}]
                    \draw[step=4cm,gray,thin] (-1,-1) grid (5,5);
                    \draw[dotted] (0,0) -- (4,4);
                    \draw[thick](0,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_1$} -- 
                                (0,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_2$} -- 
                                (4,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_3$} -- 
                                (4,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_4$} -- cycle ;
                                \draw[] (2,2) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north east]{$x^*$};

                    \draw[-stealth] (0.2,1) -- (0.2,3);
                    \draw[-stealth] (1,3.8) -- (3,3.8);
                    \draw[-stealth] (3,3.5) -- (1,1.5);

                    \draw[-stealth] (1,0.5) -- (3,2.5);
                    \draw[-stealth] (3.8,3) -- (3.8,1);
                    \draw[-stealth] (3,0.2) -- (1,0.2);
                \end{tikzpicture}
                \caption{a plaquette broken up into two triangles}
                %\caption{\label{fig:plaquette} Visualization of plaquette $x^*$. The dotted line separates the plaquette into two signed areas which are used to define the topological charge density $q(x^*)$. Arrows represent order of signed area.}
                \end{subfigure}
                \begin{subfigure}[b]{0.4\textwidth}
                \centering
                \begin{tikzpicture}[scale=0.7, every node/.style={scale=1}]
                    \pgfdeclarelayer{nodelayer}
                    \pgfdeclarelayer{edgelayer}
                    \pgfdeclareradialshading{sphere4}{\pgfpoint{-0.2cm}{0.5cm}}% 
                        {rgb(0cm)=(1,1,1);
                        rgb(1cm)=(0.5,0.5,0.5); rgb(1.05cm)=(1,1,1)}
                        %rgb(0.7cm)=(0.1,0.1,0.1); rgb(1cm)=(0.5,0.05,0); rgb(1.05cm)=(1,1,1)}
                    \pgfsetlayers{nodelayer,edgelayer}
                    \tikzstyle{label}=[fill=none, draw=none, shape=circle]
                    \tikzstyle{point}=[inner sep=0pt, minimum size=0.2cm,fill=black, draw, shape=circle]
                    %\tikzstyle{interior line}=[{Stealth[scale=1.5]}-,dotted,thick]
                    \tikzstyle{interior line}=[dotted,thick]

                    \tikzstyle{triangle}=[thick]
                    \tikzstyle{arrow}=[->, thick]
                    \begin{pgfonlayer}{nodelayer}
                        \node [style=point] (4) at (0.5, 2.5) {};
                        \node [style=point] (5) at (0.55, 0.625) {};
                        \node [style=point] (6) at (2.25, 0.75) {};
                        \node (7) at (0, 0) {};
                        \node (8) at (1.25, 1.375) {};
                        \node (9) at (2.425, 2.525) {};
                    \end{pgfonlayer}
                    \begin{pgfonlayer}{edgelayer}
                        \draw (0,0) circle (3cm);
                        %\shade[inner color=white,outer color=lightgray] (0,0) circle (3cm);
                        \shade[shading=sphere4] (0,0) circle (3cm);
                        \draw [bend left=15,style=triangle] (4.center) to (6.center);
                        \draw [bend left=355,style=triangle] (6.center) to (5.center);
                        \draw [bend right=5,style=triangle] (5.center) to (4.center);
                        \draw [style=interior line] (4.center) to (7.center);
                        \draw [style=interior line] (5.center) to (7.center);
                        \draw [style=interior line] (6.center) to (7.center);
                        \draw [fill=gray!80] (4.center) to [bend left=15] (6.center) to [bend left=355] (5.center) to [bend right=5] cycle;

                        %\draw [style=arrow] (8.center) to (9.center);

                        \node [style=label, anchor=north] at (2.25, 0.75) {\small $\e(x_1)$};
                        \node [style=label, anchor=north] at (0.55, 0.625) {\small $\e(x_2)$};
                        \node [style=label, anchor=east] at (0.5, 2.5) {\small $\e(x_3)$};
                        \node [style=point] at (4){};
                        \node [style=point] at (5){};
                        \node [style=point] at (6){};

                        \node [style=label] at (8) {\small $A$};
                    \end{pgfonlayer}
                \end{tikzpicture}
                \caption{signed area of triangle in target space}
                \end{subfigure}
            \end{figure}

        \vspace{2em}
        {\bf Non-trivial theory}
        \begin{equation*}
            S[\e\,] \rightarrow S[\e\,] - i \theta Q[\e\,].
        \end{equation*}
        A nonzero $\theta$ implies nonzero $\langle Q \rangle$. Furthermore,
        \begin{equation}
            \chi_t \propto \left. \frac{d\,\mathrm{Im}\langle Q \rangle}{d\theta}\right|_{\theta=0}
        \end{equation}
        \end{block}
        \printbibliography
    \end{column}%2
    \begin{column}{0.32\linewidth}

    \begin{block}{$\chi_t$ Divergence}
        \begin{figure}[h!]
            \begin{center}
              \begin{subfigure}[b]{0.45\textwidth}
                  \includegraphics[height=0.7\textwidth]{divergence.png}
                  \caption{$\tau = 0$}
              \end{subfigure}%
              \begin{subfigure}[b]{0.45\textwidth}
                  \centering
                  \includegraphics[height=0.7\textwidth]{divergence_flowed.png}
                  \caption{$\tau = 5t_0$}
              \end{subfigure}
            \end{center}
              \caption{\label{fig:divergence} $\chi_t\xi_2^2$ as a function of $L$. $\xi_2$ is the second moment of the correlation function and $t_0$ is a scale-independent unit of flow time. We fit the data with both a logarithmic and power fit. Simulation run with 10,000 measurements every 50 sweeps, 1,000 sweep thermalization. In the $\tau=0$ case, we have compared our result with the curve fits found in \cite{bietenholz2018}.}
              %\caption{Divergent properties with comparison to~\footfullcite{bietenholz2018}, 10,000 measurements}
        \end{figure}
    \end{block}

    \begin{block}{Effect of $\theta$-term on $\langle Q \rangle$}
        \begin{columns}
            \begin{column}{0.35\textwidth}
                Using the path integral formulation, we show that
                \begin{align*}
                    \langle Q \rangle_\theta &=\int \mathcal{D}\e\:Q[\e\,]e^{-S[\e]+i\theta Q[\e]} \\
                                             &=\int \mathcal{D}\e\:\left( Q[\e\,]e^{i\theta Q[\e]} \right) e^{-S[\e]} \\
                                             &=\langle Q e^{i \theta Q} \rangle_{\theta=0}
                \end{align*}
                Therefore, the trivial Monte Carlo method can calculate topological observables, shown in Fig.~\ref{fig:theta}. The increasing slope at $\theta=0$ as $L\rightarrow\infty$ indicates a nonzero suscpetibility in the continuum limit.
            \end{column}

            \begin{column}{0.55\textwidth}
                \begin{figure}[h]
                  \centering
                  \includegraphics[width=\textwidth]{theta.png}
                  \caption{\label{fig:theta} Nontrivial $\mathrm{Im}\langle Q \rangle$ as a function of $\theta$. Simulation run with 10,000 measurements, every 50 sweeps, 1,000 sweep thermalization. Note the different scaling of the $y$-axis.}
                \end{figure}
            \end{column}
        \end{columns}

    \end{block}

    \begin{block}{Conclusion}
We have studied the topological properties of the 1+1 dimensional NLSM under the gradient flow. Berg \& L\"uscher \cite{berg1981} give three possible sources of divergence:
        {\setlength\parindent{5pt}\begin{enumerate}
                \item ultraviolet modes
                \item the definition of $Q$ on the lattice is problematic
                \item the NLSM has no well-defined continuum limit
        \end{enumerate}}
        Our result supports either options two or three, and requires further work to identify the physical mechanism underlying our results.
    \end{block}

    \end{column}%3

  \end{columns}
\end{frame}
\end{document}

### 2021-10-07 summary.latex/main.tex
fix .gitignore issue

\documentclass[english,11pt]{article}
\usepackage[latin9]{inputenc}
\usepackage[letterpaper]{geometry}
\geometry{verbose}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[english]{babel}
\usepackage{hyperref}
\usepackage[%
    style=phys,%
    articletitle=false,biblabel=brackets,%
    chaptertitle=false,pageranges=false,%
    doi=false,maxnames=2,minnames=1,eprint=false
  ]
{biblatex}
\addbibresource{library.bib}

\makeatletter
\renewcommand{\@makefnmark}{\makebox{\normalfont\@thefnmark}}
\renewcommand\@makefntext[1]{%
    {{\normalfont[\@thefnmark]}\enspace #1}
  }
\makeatother
\renewcommand{\cite}[1]{[\footfullcite{#1}]}

\setlength{\textheight}{8.6in}

\begin{document}
\setlength{\baselineskip}{24pt}
\pagenumbering{gobble}

\title{ Topology of a prototypical field theory on a discretized lattice\vspace{0.5cm} %
\\ \large\sc COLL 400 Summary }
\author{Stuart Thomas \\ Advisor: Christopher Monahan}
\date{April 30, 2021}
\maketitle

\indent Quantum field theory (QFT) is an extraordinarily successful framework that describes particles as packets of energy on all-permeating fields. These fields consist of mathematical objects assigned to each point of space and time, existing in an infinite number of possible states. Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics and describes emergent phenomena such as vibrations in metals. Compared to experiment, QFT is remarkably accurate, famously predicting the electron $g$-factor to eleven significant figures~\cite{odom2006}, arguably the best prediction in all of science.

However this power comes at a cost: the study of quantum fields is rife with infinities where there should not be. A na\"ive treatment of QFT predicts divergent (i.e. infinite) values for physical measurements, a clearly impossible result, and the process of removing these infinites is not simple. One specific technique includes breaking the field into discrete chunks, creating a so-called ``lattice''. Since this imposition is arbitrary, we expect physical quantities to approach fixed values as we decrease the size of the lattice chunks to zero, returning to a continuous field. However, some quantities still diverge.

In this thesis, we consider one specific field: the non-linear sigma model (NLSM). This field is important because it describes solid magnets, has applications in string theory, and shares many attributes with the theory of the strong nuclear force. Based on this last application, the NLSM is known as a ``prototypical'' theory in nuclear physics, meaning that it shares many qualities with the real theory but is simpler.

The NLSM features topologically stable sectors, meaning that a state in one sector cannot continuously change into a state of a different sector. This stability makes topology vitally important to cosmology and possibly key to building fault-tolerant quantum computers~\cite{kitaev1997}. We can quantify the stability of these topological sectors with the ``topological susceptibility''. Though the topological susceptibility should approach zero, indicating perfect stability, numerical simulations find that it diverges in the continuum limit, i.e. when the lattice chunks become infinitesimally small~\cite{berg1981}.

To remedy this divergence, we apply a technique known as the ``gradient flow''. This mathematical operation ``smooths out'' high-frequency waves on the field, thereby reducing some infinities caused by the lattice. We study the effect of the gradient flow on the topological susceptibility using numerical methods and find that a divergence persists, demonstrating the gradient flow is not effective. This result supports a previous study \cite{bietenholz2018}, implying one of two possibilities: (1) there is an issue with the accepted definition of topology in the NLSM or (2), the NLSM is not well-defined in the continuum limit.

We also study a variation of the NLSM which forces the vacuum field (i.e. the field without inserted particles) into a topological state. We analyze the effect of this perturbation on the overall topology and apply the gradient flow. These results support the divergence of the topological susceptibility in the original NLSM and provide new insights into the effect of the topological variation.

\end{document}

### 2021-10-07 code/mcmc/sweep.cpp
remove unwanted files

@@ -1,446 +1,0 @@
-#include <iostream>
-#include "sweep.h"
-#include <algorithm>
-using namespace std;
-
-namespace croutines {
-
-    Phi::Phi() {}; 
-
-    Phi::Phi(array<double, N> phi){
-        this->phi = phi;
-    };
-
-    void Phi::init_as_zero(){
-        for (int i=0; i<N; i++) {
-            phi[i] = 0;
-        }
-    };
-
-
-
-    Phi Phi::operator+ (const Phi & aPhi) const {
-        Phi new_phi;
-        for (int i=0; i<N; i++) {
-            new_phi[i] = phi[i] + aPhi[i];
-        }
-        return new_phi;
-    }
-
-    Phi Phi::operator- (const Phi & aPhi) const {
-        return *this + (-aPhi);    
-    }
-    Phi Phi::operator- () const {
-        Phi new_phi;
-        for (int i=0; i<N; i++) {
-            new_phi[i] = -phi[i];
-        }
-        return new_phi;
-    }
-    double Phi::operator* (const Phi & aPhi) const{
-        //dot product
-        double dot = 0;
-        for (int i=0; i<N; i++) {
-            dot += phi[i] * aPhi[i];
-        }
-        return dot;
-    }
-
-    Phi Phi::operator* (const double & a) const{
-        Phi new_phi;
-        for (int i=0; i<N; i++) {
-            new_phi[i] = phi[i] * a;
-        }
-        return new_phi;
-
-    }
-    void Phi::print(){
-    }
-
-    ostream& operator<<(ostream& os, const Phi & aPhi) 
-    { 
-        os << "(";
-        for (int i=0; i<N; i++) {
-            os << aPhi[i];
-            if (i<N-1) { os<<", "; }
-        }
-        os << ")";
-        return os; 
-    }
-    
-    double Phi::operator[] (int i) const {
-        return phi[i];
-    }
-    double & Phi::operator[] (int i) {
-        return phi[i];
-    }
-
-    bool Phi::operator== (const Phi & aPhi) const {
-        for (int i=0; i<N; i++) {
-            if (phi[i] != aPhi[i]) {
-                return false;
-            }
-        }
-        return true;
-    }
-
-
-    Sweeper::Sweeper (double m02, double lam, MPI_Comm c) : process_Rank(get_rank(c)), size_Of_Cluster(get_size(c)), sites_per_node((DIM*DIM)/2 / get_size(c)) {
[…]

### 2021-10-07 proceedings/draft.tex
update proceedings bibliography
add some text from poster/report to proceedings

@@
 \usepackage{pos}
+\usepackage{amsmath,amsfonts,amssymb,eulervm,xspace}
+\usepackage{amsmath,amsfonts,amssymb}
+\usepackage{graphicx}
+\usepackage{tikz}
+\usepackage{subcaption}
+\newcommand{\e}{\vec e}
+\graphicspath{{../figures/}}
 
@@
 
-            %\item Prototypical model for strong nuclear force
-                    %\item Models Heisenberg ferromagnets
-                    %\item Applications to string theory
-                %\end{itemize}
-            %\item<4-> Merits
-                %\begin{itemize}
-                    %\item mass gap
-                    %\item asymptotic freedom
-                %\end{itemize}
-        \end{itemize}
+Following~\cite{berg1981}, we define topological charge density $q(x^*)$ for each plaquette $x^*$ such that
+\begin{equation}
+    Q = \sum_{x^*} q(x^*)
+\end{equation}
+where
+\begin{equation}
+    q(x^*) = \frac{1}{4\pi} \bigg[A\Big(\e(x_1), \e(x_2), \e(x_3)\Big) + A\Big(\e(x_1), \e(x_3), \e(x_4)\Big) \bigg].
+  \end{equation}
+  The topology of the NLSM is visualized in Fig.~\ref{fig:topology}
+\begin{figure}
+  \centering
+    \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \begin{tikzpicture}[scale=0.4, every node/.style={scale=1.2}]
+        \draw[step=4cm,gray,thin] (-1,-1) grid (5,5);
+        \draw[dotted] (0,0) -- (4,4);
+        \draw[thick](0,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_1$} --
+                    (0,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_2$} --
+                    (4,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_3$} --
+                    (4,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_4$} -- cycle ;
+                    \draw[] (2,2) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north east]{$x^*$};
 
-\begin{thebibliography}{99}
-\bibitem{...}
-....
+        \draw[-stealth] (0.2,1) -- (0.2,3);
+        \draw[-stealth] (1,3.8) -- (3,3.8);
+        \draw[-stealth] (3,3.5) -- (1,1.5);
 
-\end{thebibliography}
+        \draw[-stealth] (1,0.5) -- (3,2.5);
+        \draw[-stealth] (3.8,3) -- (3.8,1);
+        \draw[-stealth] (3,0.2) -- (1,0.2);
+    \end{tikzpicture}
+    \caption{a plaquette broken up into two triangles}
+    %\caption{\label{fig:plaquette} Visualization of plaquette $x^*$.
+The dotted line separates the plaquette into two signed areas which are used to define the topological charge density $q(x^*)$.
+Arrows represent order of signed area.}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \begin{tikzpicture}[scale=0.6, every node/.style={scale=1}]
+        \pgfdeclarelayer{nodelayer}
+        \pgfdeclarelayer{edgelayer}
+        \pgfdeclareradialshading{sphere4}{\pgfpoint{-0.2cm}{0.5cm}}%
+            {rgb(0cm)=(1,1,1);
+            rgb(1cm)=(0.5,0.5,0.5); rgb(1.05cm)=(1,1,1)}
+            %rgb(0.7cm)=(0.1,0.1,0.1); rgb(1cm)=(0.5,0.05,0); rgb(1.05cm)=(1,1,1)}
+        \pgfsetlayers{nodelayer,edgelayer}
+        \tikzstyle{label}=[fill=none, draw=none, shape=circle]
+        \tikzstyle{point}=[inner sep=0pt, minimum size=0.2cm,fill=black, draw, shape=circle]
+        %\tikzstyle{interior line}=[{Stealth[scale=1.5]}-,dotted,thick]
+        \tikzstyle{interior line}=[dotted,thick]
 
+        \tikzstyle{triangle}=[thick]
+        \tikzstyle{arrow}=[->, thick]
+        \begin{pgfonlayer}{nodelayer}
+            \node [style=point] (4) at (0.5, 2.5) {};
+            \node [style=point] (5) at (0.55, 0.625) {};
+            \node [style=point] (6) at (2.25, 0.75) {};
+            \node (7) at (0, 0) {};
+            \node (8) at (1.25, 1.375) {};
+            \node (9) at (2.425, 2.525) {};
+        \end{pgfonlayer}
+        \begin{pgfonlayer}{edgelayer}
+            \draw (0,0) circle (3cm);
+            %\shade[inner color=white,outer color=lightgray] (0,0) circle (3cm);
+            \shade[shading=sphere4] (0,0) circle (3cm);
+            \draw [bend left=15,style=triangle] (4.center) to (6.center);
+            \draw [bend left=355,style=triangle] (6.center) to (5.center);
+            \draw [bend right=5,style=triangle] (5.center) to (4.center);
+            \draw [style=interior line] (4.center) to (7.center);
+            \draw [style=interior line] (5.center) to (7.center);
+            \draw [style=interior line] (6.center) to (7.center);
+            \draw [fill=gray!80] (4.center) to [bend left=15] (6.center) to [bend left=355] (5.center) to [bend right=5] cycle;
+
+            %\draw [style=arrow] (8.center) to (9.center);
+
+            \node [style=label, anchor=north] at (2.25, 0.75) {\small $\e(x_1)$};
+            \node [style=label, anchor=north] at (0.55, 0.625) {\small $\e(x_2)$};
+            \node [style=label, anchor=east] at (0.5, 2.5) {\small $\e(x_3)$};
+            \node [style=point] at (4){};
+            \node [style=point] at (5){};
+            \node [style=point] at (6){};
+
+            \node [style=label] at (8) {\small $A$};
+        \end{pgfonlayer}
+    \end{tikzpicture}
+    \caption{signed area of triangle in target space}
+    \end{subfigure}
+    \caption{\label{fig:topology}}
+\end{figure}
+
+\begin{equation*}
+    S[\e\,] \rightarrow S[\e\,] - i \theta Q[\e\,].
+\end{equation*}
+A nonzero $\theta$ implies nonzero $\langle Q \rangle$. Furthermore,
+\begin{equation}
+    \chi_t \propto \left. \frac{d\,\mathrm{Im}\langle Q \rangle}{d\theta}\right|_{\theta=0}
+\end{equation}
+
+\section{Computational Methods}
+Our study of the gradient flow in the NLSM is based on a computational system that simulates quantum fields numerically.
+We begin by implementing a numerical Monte Carlo method to simulate the lattice in two dimensions and verify our program with the well-studied $\phi^4$ scalar field theory.
+We then generalize our model to a vector field to simulate the NLSM and implement a numerical solution to the gradient flow.
+We use the data produced by this program to study the topological charge and susceptibility under the gradient flow.
+
+We first implement the $\phi^4$ model in Python.
+Afterwards, we transition to the NLSM, using C++ for increased efficiency .
+To compile the C++ simulation, we use the \texttt{gcc} compiler with the highest level of optimization.
+
+\section{Results}
+\begin{figure}[h!]
+    \begin{center}
+        \begin{subfigure}[b]{0.45\textwidth}
+            \includegraphics[height=0.7\textwidth]{divergence.png}
+            \caption{$\tau = 0$}
+        \end{subfigure}%
+        \begin{subfigure}[b]{0.45\textwidth}
+            \centering
+            \includegraphics[height=0.7\textwidth]{divergence_flowed.png}
+            \caption{$\tau = 5t_0$}
+        \end{subfigure}
+    \end{center}
+        \caption{\label{fig:divergence} $\chi_t\xi_2^2$ as a function of $L$.
+$\xi_2$ is the second moment of the correlation function and $t_0$ is a scale-independent unit of flow time.
+We fit the data with both a logarithmic and power fit.
+Simulation run with 10,000 measurements every 50 sweeps, 1,000 sweep thermalization.
+In the $\tau=0$ case, we have compared our result with the curve fits found in \cite{bietenholz2018}.}
+        %\caption{Divergent properties with
+        %comparison to~\footfullcite{bietenholz2018}, 10,000 measurements}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.5\textwidth]{theta.png}
+    \caption{\label{fig:theta} Nontrivial $\mathrm{Im}\langle Q \rangle$ as a function of $\theta$.
+Simulation run with 10,000 measurements, every 50 sweeps, 1,000 sweep thermalization.
+Note the different scaling of the $y$-axis.}
+  \end{figure}
+
+\bibliographystyle{JHEP}
+\bibliography{library}
 \end{document}

### 2021-10-19 proceedings/draft.tex
Added figures folder and first edits of proceedings draft

@@
 \usepackage{pos}
-\usepackage{amsmath,amsfonts,amssymb,eulervm,xspace}
-\usepackage{amsmath,amsfonts,amssymb}
+%\usepackage{amsmath,amsfonts,amssymb,eulervm,xspace} Currently causing a package clash for me.
 \usepackage{graphicx}
@@
 \graphicspath{{../figures/}}
+\newcommand\chris[1]{\textcolor{teal}{CJM: #1}}
 
@@
 \author*[a]{Stuart Thomas}
-\author[a]{Christopher Monahan}
+\author[a,b]{Christopher Monahan}
 
@@
 
-%\affiliation[b]{Department, University,\\
-%Street number, City, Country}
+\affiliation[b]{Thomas Jefferson National Accelerator Facility, Newport News, Virginia, USA}
 
@@
 
+\section{Introduction}
+
+Spin models provide a framework for understanding the physics of strongly-coupled systems, from solid state and condensed matter systems to nuclear and particle physics.
+The non-linear sigma model (NLSM), in particular, has provided a rich arena in which to study nonperturbative effects.
+In solid-state systems, this model describes Heisenberg ferromagnets \cite{callan1985} and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), the gauge theory of the strong nuclear force.
+In general, the NLSM shares key features with non-Abelian gauge theories such as QCD, including a mass gap and asymptotic freedom \cite{polyakov1975}, and has proved a useful model for exploring the effect of these properties in a simpler system.
+
+We consider the $O(3)$ NLSM in 1+1 dimensions (one dimension of space, one dimension of time).
+This theory exhibits topological properties, such as \textit{instantons}, or classical field solutions at local minima of the action in Euclidean space.
+These topologically protected solutions cannot evolve into the vacuum state via local fluctuations.
+This property has become critically important to quantum field theories in cosmology and high energy physics \cite{goddard1986}.
+\chris{Can we provide a more explicit example(s) here?
+Statement is very vague at the moment.} Additionally, topological stability may become a key tool for fault-tolerant quantum computers \cite{kitaev1997}.
+In these devices, topology protects the delicate quantum states necessary for information processing.
+
 \section{The Non-Linear Sigma Model}
-We study the $O(3)$ non-linear sigma model (NLSM) in 1+1 dimensions, defined by the Euclidean action
+
+We study the $O(3)$ NLSM in two dimensions \chris{[In Euclidean space it does not really make sense to distinguish time and space, unless we start in Minkowski spacetime and then Wick rotate.]}, defined by the Euclidean action
 \begin{equation*}
     \label{eq:nlsm euclidean action}
-    S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right]
+    S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right],
 \end{equation*}
-where, $\e$ is 3-component real vector constrained by $|\e\,|=1$ and $\beta$ is the inverse coupling constant.
-In solid-state systems, this model describes Heisenberg ferromagnets \cite{callan1985} and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), the gauge theory that describes the strong nuclear force.
-In general, the NLSM shares key features with non-Abelian gauge theories such as QCD, including a mass gap and asymptotic freedom \cite{polyakov1975}.
-Therefore, the NLSM is a useful model for exploring the effect of these properties in a simpler system.
+where $\e$ is 3-component real vector constrained by $|\e\,|=1$ and $\beta$ is the inverse coupling constant.
 
-Following~\cite{berg1981}, we define topological charge density $q(x^*)$ for each plaquette $x^*$ such that
+Following~\cite{berg1981}, we define the topological charge density, $q(x^*)$, for each plaquette $x^*$ such that the total topological charge is
 \begin{equation}
-    Q = \sum_{x^*} q(x^*)
+    Q = \sum_{x^*} q(x^*),
 \end{equation}
@@
 \end{equation}
-  The topology of the NLSM is visualized in Fig.~\ref{fig:topology}
+Here $A$ is the signed area of the triangle in target space, which we represent in the left-hand figure of Fig.~\ref{fig:topology}.
 \begin{figure}
@@
     \end{tikzpicture}
-    \caption{a plaquette broken up into two triangles}
+    \caption{Visualization of a plaquette $x^*$.
+The dotted line separates the plaquette into two signed areas, used to define the topological charge density $q(x^*)$.
+Arrows represent the order of signed area.}
     %\caption{\label{fig:plaquette} Visualization of plaquette $x^*$.
@@
 Arrows represent order of signed area.}
+    \vspace*{-\baselineskip}
     \end{subfigure}
+    \hspace*{1cm}
     \begin{subfigure}[b]{0.4\textwidth}
@@
     \end{tikzpicture}
-    \caption{signed area of triangle in target space}
+    \caption{The signed area of a triangle in target space.}
     \end{subfigure}
@@
 
+\chris{Need text here to explain the next equations}
 \begin{equation*}
@@
 \section{Computational Methods}
-Our study of the gradient flow in the NLSM is based on a computational system that simulates quantum fields numerically.
-We begin by implementing a numerical Monte Carlo method to simulate the lattice in two dimensions and verify our program with the well-studied $\phi^4$ scalar field theory.
+\chris{Don't need to mention phi-4 model, or python, except to note that the phi-4 model served as a check of your code.}
+We implement a numerical Monte Carlo method to simulate the lattice in two dimensions and verify our program with the well-studied $\phi^4$ scalar field theory.
 We then generalize our model to a vector field to simulate the NLSM and implement a numerical solution to the gradient flow.
@@
         \begin{subfigure}[b]{0.45\textwidth}
-            \includegraphics[height=0.7\textwidth]{divergence.png}
+            %\includegraphics[height=0.7\textwidth]{divergence.png}
             \caption{$\tau = 0$}

### 2021-10-19 proceedings/draft.tex
More info in numerical section

@@
 Here $A$ is the signed area of the triangle in target space, which we represent in the left-hand figure of Fig.~\ref{fig:topology}.
+This value is defined if $A\neq 0, 2\pi$, or in other words, as long as the three points on the sphere are distinct and do not form a hemisphere.
+In numerical calculations, these points can be ignored.
+Therefore, we impose that the signed area is defined on the smallest spherical triangle, or $-2\pi < A < 2\pi$.
 \begin{figure}
@@
 
-\section{Computational Methods}
-\chris{Don't need to mention phi-4 model, or python, except to note that the phi-4 model served as a check of your code.}
-We implement a numerical Monte Carlo method to simulate the lattice in two dimensions and verify our program with the well-studied $\phi^4$ scalar field theory.
-We then generalize our model to a vector field to simulate the NLSM and implement a numerical solution to the gradient flow.
-We use the data produced by this program to study the topological charge and susceptibility under the gradient flow.
+We define the topological susceptibility $\chi_t$
+\begin{equation}
+\chi_t \equiv \frac{1}{L^2} \Big( \langle Q^2 \rangle - \langle Q \rangle^2 \Big).
+\end{equation}
+In the trivial case, $\langle Q \rangle$ is equal to $0$ and on this lattice, this becomes   
+\begin{equation}
+    \chi_t = \frac{1}{L^2} \sum_{x^*,y^*} \langle q(x^*)q(y^*)\rangle  = \frac{1}{L^2} \sum_{x^*} \langle q(x^*)q(0)\rangle,
+\end{equation}
+where in the second equality we have assumed periodic boundary conditions.
+The topological susceptibility diverges in the continuum limit owing solely to the $x^*=0$ term \cite{bietenholz2018}.
+This divergence exists in QCD as well \cite{bruno2014}.
 
-We first implement the $\phi^4$ model in Python.
-Afterwards, we transition to the NLSM, using C++ for increased efficiency .
-To compile the C++ simulation, we use the \texttt{gcc} compiler with the highest level of optimization.
+\subsection{Gradient flow}
+
+\chris{Sentences/paragraph introducing continuous gradient flow here}
+
+\section{Numerical implementation}
+We implement a numerical Monte Carlo method to study the NLSM in two dimensions using the discretized action
+\begin{equation}
+    \label{eq:nlsm discretized action}
+    S_\mathrm{lat}[\e\,] = \sum_i \left[ 2 - \e(x+a\hat{t}\,)\cdot\e(x) - \e(x+a\hat{x})\cdot\e(x) \right].
+\end{equation}
+We generate configurations using the Metropolis \cite{} and Wolff cluster \cite{wolff1989} algorithms, implemented in C++ and parallelized through the Message Passing Interface with a checkerboard algorithm \cite{}.
+
+We thermalize the configurations with 1000 sweeps, with a cluster update every five sweeps, and illustrate a sample Markov chain in Fig.~\ref{fig:therm}, where we plot the action as a function of Metropolis sweeps.
+We use Wolff's automatic windowing procedure \cite{wolff2007} to estimate the autocorrelation times for various observables, such as the magnetic susceptibility $\chi_m$.
+We measure observables every $50$ sweeps for each simulation.
+\begin{figure}[h]
+  \centering
+      \begin{subfigure}[b]{0.5\textwidth}\centering
+        \includegraphics[width=\textwidth]{therm24.png}
+        \caption{$L=24$}
+      \end{subfigure}%
+      \hfill
+      \begin{subfigure}[b]{0.5\textwidth}\centering
+        \includegraphics[width=\textwidth]{therm404.png}
+        \caption{$L=404$}
+      \end{subfigure}
+      \hfill
+      \caption{\label{fig:therm} Plots of the action as a function of Monte Carlo time, starting with a random NLSM lattice.}
+\end{figure}
+
+We apply the gradient flow equation
+\begin{equation}
+    \label{eq:nsm_gradflow_disc}
+    \partial_\tau \e (\tau, x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T\right) \partial^2 \e(\tau,x),
+\end{equation}
+where the Laplacian operator $\partial^2$ is defined as
+\begin{equation*}
+    \partial^2 \e(\tau,x) = \e(\tau, x+a \hat{t}) + \e(\tau,x-a\hat t) + \e(\tau, x+a \hat{x}) + \e(\tau,x-a\hat x) - 4 \e(t,x).
+\end{equation*}
+
+We numerically solve the ordinary differential equation in Eq.~\ref{eq:nsm_gradflow_disc} using a fourth-order Runga-Kutta approximation.
+To increase the efficiency of this algorithm, we implement the step-doubling algorithm to adaptively adjust the step size.
+If the error of a Runge-Kutta step is greater than the tolerance, the same step is repeated with half the step size.
+Alternatively, if the error is less than half of the tolerance, the step size is doubled for the next calculation.
+Finally, if the step size is greater than the distance to the next measurement, that distance is used as the step size, using the normal value afterwards.
+Otherwise, the algorithm proceeds with the consistent step size.
+
+To calculate the error, we compare one lattice $\e_1$ produced using a step of size $2h$ with another lattice $\e_2$ produced via two steps of size $h$.
+The error $\Delta$ can be estimated to up the fifth order of $h$ as \cite{vetterling1992}
+\begin{equation}
+\Delta = \frac{1}{15}\sqrt{\sum_x \left| \e_2(x) - \e_1(x) \right|}
+\end{equation}
+The tolerance used in this work is $\Delta_{max}=0.01$.
 

### 2021-10-20 proceedings/draft.tex
write introduction

@@
 \usepackage{pos}
-%\usepackage{amsmath,amsfonts,amssymb,eulervm,xspace} Currently causing a package clash for me.
+% \usepackage{eulervm}
+% \usepackage{xspace}
+% \usepackage{inputenc}
 \usepackage{graphicx}
@@
 \newcommand\chris[1]{\textcolor{teal}{CJM: #1}}
+\newcommand\stuart[1]{\textcolor{pink}{SNT: #1}}
 
@@
 
-\affiliation[a]{Department of Physics, William \& Mary,  Williamsburg, Virginia 23187, USA}
+\affiliation[a]{Department of Physics, William \& Mary, \\ 300 Ukrop Way, Williamsburg, VA, USA} %23187
+\affiliation[b]{Thomas Jefferson National Accelerator Facility, \\ 12000 Jefferson Avenue, Newport News, VA, USA} %23606
+% These match the PoS format a little better
 
-\affiliation[b]{Thomas Jefferson National Accelerator Facility, Newport News, Virginia, USA}
+% \affiliation[b]{Thomas Jefferson National Accelerator Facility, Newport News, Virginia, USA}
 
@@
 
+The protection of topological instantons in the 1+1 $O(3)$ NLSM relies on a vanishing topological susceptibilty.
+However, the convergence of this quantity is still unknown.
+\cite{bietenholz2018} While the some analytical arguments argue the topological susceptibility should approach zero in the continuum limit, numerical results on the lattice predict infinities \cite{berg1981}.
+In an attempt to shed light on this contradiction, we apply the gradient flow, a local smearing of operators which preserves gauge invariance.
+In quantum chromodynamics, this technique has corroborated a previous analytical result \cite{giusti2004} by removing ultraviolet divergences on the lattice \cite{bruno2014}.
+This success has motivated the gradient flow to calculate the topological susceptibility in the 1+1 $O(3)$ NLSM.
+Despite this intuition, recent studies demonstrate that the observable still diverges in the continuum limit \cite{bietenholz2018}.
+
+We also study a second perspective on the topological susceptibility arising from the introduction of a $\theta$-term into the field Lagrangian.
+This term drives the vacuum state into a topological phase \cite{allessalom2008}.
+Differentiating the field's partition function with respect to $\theta$ yields a value proportional to the topological susceptibility.
+The effect of nonzero $\theta$ on the theory therefore should reflect the divergence in the continuum limit.
+In this work we verify the divergence of the topological susceptibility and develop a clearer picture of how the $\theta$-term affects the topology of the 1+1 $O(3)$ NLSM.
+
+To numerically study the topological qualities of the NLSM, we first implement a Markov chain Monte Carlo simulation using Metropolis and Wolff cluster \cite{wolff1989} algorithms.
+Since the gradient flow has no exact solution in the NLSM we implement a numerical solution using a fourth-order Runge-Kutta approximation with automatic step sizing.
+By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.
+
 \section{The Non-Linear Sigma Model}
@@
     \label{eq:nlsm euclidean action}
-    S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right],
+    % S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right],
+    S_E = \frac{\beta}{2} \int d^2x \; \sum_{i=1}^{2}\left(\partial_i \e\, \right)^2,
 \end{equation*}
 where $\e$ is 3-component real vector constrained by $|\e\,|=1$ and $\beta$ is the inverse coupling constant.
-
 Following~\cite{berg1981}, we define the topological charge density, $q(x^*)$, for each plaquette $x^*$ such that the total topological charge is
@@
 \end{equation}
-Here $A$ is the signed area of the triangle in target space, which we represent in the left-hand figure of Fig.~\ref{fig:topology}.
+Here, $A$ is the signed area of the triangle in target space, which we represent in the left-hand figure of Fig.~\ref{fig:signedarea}.
 This value is defined if $A\neq 0, 2\pi$, or in other words, as long as the three points on the sphere are distinct and do not form a hemisphere.
@@
     \end{tikzpicture}
-    \caption{The signed area of a triangle in target space.}
+    \caption{\label{fig:signedarea} The signed area of a triangle in target space.}
     \end{subfigure}
@@
 \end{figure}
-
-\chris{Need text here to explain the next equations}
-\begin{equation*}
-    S[\e\,] \rightarrow S[\e\,] - i \theta Q[\e\,].
-\end{equation*}
-A nonzero $\theta$ implies nonzero $\langle Q \rangle$. Furthermore,
-\begin{equation}
-    \chi_t \propto \left. \frac{d\,\mathrm{Im}\langle Q \rangle}{d\theta}\right|_{\theta=0}
-\end{equation}
 
@@
 This divergence exists in QCD as well \cite{bruno2014}.
+
+Additionally, we can generalize the NLSM to have a nonzero vacuum expectation value for the topological charge, manifested by the addition of a ``$\theta$-term'':
+\begin{equation*}
+    S[\e\,] \rightarrow S[\e\,] - i \theta Q[\e\,].
+\end{equation*}
+defining the ``nontrivial'' NLSM. We find that the topological susceptibility in the trivial case depends on the charge in the nontrival model:
+\begin{equation}
+    \chi_t \propto \left. \frac{d\,\mathrm{Im}\langle Q \rangle}{d\theta}\right|_{\theta=0}
+\end{equation}
 

### 2021-10-27 proceedings/draft.tex
small edits to draft

@@
 \newcommand{\e}{\vec e}
+\newcommand{\logchi}{7.4 }
+\newcommand{\powchi}{8.8 }
 \graphicspath{{../figures/}}
@@
 \section{Introduction}
-
+\label{sec:introduction}
 Spin models provide a framework for understanding the physics of strongly-coupled systems, from solid state and condensed matter systems to nuclear and particle physics.
@@
 These topologically protected solutions cannot evolve into the vacuum state via local fluctuations.
-This property has become critically important to quantum field theories in cosmology and high energy physics \cite{goddard1986}.
-\chris{Can we provide a more explicit example(s) here?
+This property has become critically important to quantum field theories in cosmology and high energy physics, relating to the existence of magnetic monopoles\cite{goddard1986} and the mass of the $\eta'$ Goldstone boson \cite{witten1979, veneziano1979} \chris{Can we provide a more explicit example(s) here?
 Statement is very vague at the moment.} Additionally, topological stability may become a key tool for fault-tolerant quantum computers \cite{kitaev1997}.
@@
 
-\section{The Non-Linear Sigma Model}
+\section{The non-linear sigma model}
+\label{sec:nlsm}
+\chris{[In Euclidean space it does not really make sense to distinguish time and space, unless we start in Minkowski spacetime and then Wick rotate.]}
 
-We study the $O(3)$ NLSM in two dimensions \chris{[In Euclidean space it does not really make sense to distinguish time and space, unless we start in Minkowski spacetime and then Wick rotate.]}, defined by the Euclidean action
+\stuart{I think I resolved this}
+
+We study the $O(3)$ NLSM in two dimensions, defined by the Euclidean action
 \begin{equation*}
     \label{eq:nlsm euclidean action}
-    % S_E = \frac{\beta}{2} \int d^2x \; \left[ \left(\partial_t \e\, \right)^2+ \left( \partial_x \e\,\right)^2 \right],
     S_E = \frac{\beta}{2} \int d^2x \; \sum_{i=1}^{2}\left(\partial_i \e\, \right)^2,
@@
 \begin{equation}
+  \label{eq:topodensity}
     q(x^*) = \frac{1}{4\pi} \bigg[A\Big(\e(x_1), \e(x_2), \e(x_3)\Big) + A\Big(\e(x_1), \e(x_3), \e(x_4)\Big) \bigg].
 \end{equation}
-Here, $A$ is the signed area of the triangle in target space, which we represent in the left-hand figure of Fig.~\ref{fig:signedarea}.
-This value is defined if $A\neq 0, 2\pi$, or in other words, as long as the three points on the sphere are distinct and do not form a hemisphere.
+Here, $A$ is the signed area of the triangle in target space.
+Visually, the three points in each of the two terms in Eq.~\ref{eq:topodensity} form halves of the plaquette which we represent in Fig.~\ref{fig:plaquette}.
+The resultant signed area is represented in Fig.~\ref{fig:signedarea}.
+The value $A$ is defined if $A\neq 0, 2\pi$, or in other words, as long as the three points on the sphere are distinct and do not form a hemisphere.
 In numerical calculations, these points can be ignored.
@@
     \centering
-    \begin{tikzpicture}[scale=0.4, every node/.style={scale=1.2}]
+    \begin{tikzpicture}[scale=0.4, every node/.style={scale=0.9}]
         \draw[step=4cm,gray,thin] (-1,-1) grid (5,5);
         \draw[dotted] (0,0) -- (4,4);
-        \draw[thick](0,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_1$} --
-                    (0,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_2$} --
-                    (4,4) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_3$} --
-                    (4,0) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north west] {$x_4$} -- cycle ;
-                    \draw[] (2,2) node[circle,fill,inner sep=0pt, minimum size=0.3cm]{} node[anchor=north east]{$x^*$};
+        \draw[thick](0,0) node[circle,fill,inner sep=0pt, minimum size=0.2cm]{} node[anchor=north west] {$x_1$} --
+                    (0,4) node[circle,fill,inner sep=0pt, minimum size=0.2cm]{} node[anchor=north west] {$x_2$} --
+                    (4,4) node[circle,fill,inner sep=0pt, minimum size=0.2cm]{} node[anchor=north west] {$x_3$} --
+                    (4,0) node[circle,fill,inner sep=0pt, minimum size=0.2cm]{} node[anchor=north west] {$x_4$} -- cycle ;
+                    \draw[] (2,2) node[circle,fill,inner sep=0pt, minimum size=0.2cm]{} node[anchor=north west]{$x^*$};
 
@@
 
-We define the topological susceptibility $\chi_t$
+From the definition of $Q$, we define the topological susceptibility as
 \begin{equation}
@@
 \end{equation}
-In the trivial case, $\langle Q \rangle$ is equal to $0$ and on this lattice, this becomes
+Since $\langle Q \rangle=0$ in the NLSM, this quantity becomes
 \begin{equation}
-    \chi_t = \frac{1}{L^2} \sum_{x^*,y^*} \langle q(x^*)q(y^*)\rangle  = \frac{1}{L^2} \sum_{x^*} \langle q(x^*)q(0)\rangle,
+    \chi_t = \frac{1}{L^2} \sum_{x^*} \langle q(x^*)q(0)\rangle,
 \end{equation}
-where in the second equality we have assumed periodic boundary conditions.
-The topological susceptibility diverges in the continuum limit owing solely to the $x^*=0$ term \cite{bietenholz2018}.
-This divergence exists in QCD as well \cite{bruno2014}.
+where we have assumed periodic boundary conditions.
+In the continuum limit, the topological susceptibility diverges owing solely to the $x^*=0$ term \cite{bietenholz2018}, a divergence that exists in QCD as well \cite{bruno2014}.
 
-Additionally, we can generalize the NLSM to have a nonzero vacuum expectation value for the topological charge, manifested by the addition of a ``$\theta$-term'':
+As an extension, we can generalize the NLSM to have a nonzero vacuum expectation value for the topological charge ($\langle Q \rangle\neq0$), manifested by the addition of a ``$\theta$-term'':
 \begin{equation*}
@@
 \end{equation}
+This final relation will allow us to prode the behavior of the topological susceptibility in the nontopological NLSM by considering the topological charge in the nontrivial theory.
 
-\subsection{Gradient flow}
+\section{Gradient flow}
+\label{sec:gradflow}
+Seeking to resolve the ultraviolet divergence in the $\chi_{t}$, we adopt a technique known as ``smearing'', a local averaging of the field \cite{solbrig2008}.
+Specifically, we use a technique known as the gradient flow \cite{monahan2015} which introduces a new half-dimension\footnote{The term ``half-dimension'' indicates that the flow time is exclusively positive.} called ``flow time''.
+The flow time $\tau$ parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences, pushing field configurations toward classical minima of the action.
 
-\chris{Sentences/paragraph introducing continuous gradient flow here}
+We can choose any flow time equation that drives the field towards a classical minimum. Following \cite{bietenholz2018}, we can define the gradient flow for the NLSM via the differential equation
+\begin{equation}
+    \label{eq:nsm_gradflow}
+  \partial_\tau \e (\tau,x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T \right) \partial^2 \e(\tau,x).
+\end{equation}
+We solve this equation numerically using the boundary condition $\e(\tau=0,x) = \e(x)$, a process described in Sec.~\ref{sec:numerics}.
 
 \section{Numerical implementation}
+\label{sec:numerics}
 We implement a numerical Monte Carlo method to study the NLSM in two dimensions using the discretized action
+
 \begin{equation}
     \label{eq:nlsm discretized action}
-    S_\mathrm{lat}[\e\,] = \sum_i \left[ 2 - \e(x+a\hat{t}\,)\cdot\e(x) - \e(x+a\hat{x})\cdot\e(x) \right].
+    S_\mathrm{lat}[\e\,] = \sum_i \left[ 2 - \sum_{\mu=0}^{2}\e(x+a\hat{\mu}\,)\cdot\e(x)  \right]
 \end{equation}
-We generate configurations using the Metropolis \cite{} and Wolff cluster \cite{wolff1989} algorithms, implemented in C++ and parallelized through the Message Passing Interface with a checkerboard algorithm \cite{}.
+where $a$ is a lattice constant and $\hat \mu$ are the Euclidean unit vectors.
+We generate configurations using the Metropolis \stuart{Does this need to be cited?} and Wolff cluster \cite{wolff1989} algorithms.
 
@@
 \begin{equation}
-    \label{eq:nsm_gradflow_disc}
     \partial_\tau \e (\tau, x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T\right) \partial^2 \e(\tau,x),
 \end{equation}
-where the Laplacian operator $\partial^2$ is defined as
+replacing the continuous Laplacian operator $\partial^2$ with a discrete analogue
 \begin{equation*}
@@
 
-We numerically solve the ordinary differential equation in Eq.~\ref{eq:nsm_gradflow_disc} using a fourth-order Runga-Kutta approximation.
+We numerically solve the ordinary differential equation in Eq.~\ref{eq:nsm_gradflow} using a fourth-order Runga-Kutta approximation.
 To increase the efficiency of this algorithm, we implement the step-doubling algorithm to adaptively adjust the step size.
@@
 \section{Results}
+\label{sec:results}
+In order to confirm the accuracy of the model, we compare results with existing literature.
+We first consider the results of Berg \& L\"uscher \cite{berg1981}, specifically the internal energy and magnetic susceptibility.
+Following \cite{berg1981}, we approximate the internal energy in the strong ($\beta<1$) and weak ($\beta>2$) regimes as
+\begin{equation}
+    \label{eq:analytic_energy}
+E \approx \begin{cases}
+    4-4y-8y^3-\frac{48}{5}y^5& \beta<1 \\
+\frac{2}{\beta} + \frac{4}{\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2
+\end{cases}
+\end{equation}
+where
+\begin{equation}
+    y=\mathrm{coth}\beta-\frac{1}{\beta}.
+\end{equation}
+We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{fig:bergluscher}.
+\begin{figure}[h!]
+    \centering
+      \includegraphics[width=0.6\textwidth]{internal_energy.png}
+      \caption{\label{fig:bergluscher}Comparison with \cite{berg1981}.
+First panel: internal energy compared with analytic  energy (Eq.~\ref{eq:analytic_energy}).
+Second panel: magnetic susceptibility compared with literature values.}
+\end{figure}
+These two charts show a high degree of agreement with the current literature however there is a slight discrepancy, prehaps arising from different Monte Carlo methods.
+
+We also seek to confirm the results from Bietenholz et al.
+\cite{bietenholz2018}.
+Specifically, we show the topological susceptibility $\chi_t$ diverges in the continuum limit even at finite flow time.
+Since $\chi_t$ is in units of inverse distance squared, we multiply by $\xi_2^2$, the square of the second moment correlation length, to achieve a scale-invariant value $\chi_t\xi_2^2$.
+Additionally, we use a parameter $t_0$ to scale the flow time such that $t_0\sim L^2$.
+In our Monte Carlo simulation, we utilize the same values as \cite{bietenholz2018} for $\xi_2$, $\beta$ and $t_0$.
+
+To begin the comparison, we plot $\chi_t\xi_2^2$ as a function of flow time $\tau$, shown in Fig.~\ref{fig:bietenholz}.
+\begin{figure}[h!]
+    \centering
+      \includegraphics[width=0.8\textwidth]{bietenholz.png}
+      \caption{\label{fig:bietenholz} $\chi_t L^2$ as a  function of flow time $\tau$. Simulation run with 10,000 measurements every 50 sweeps, 1,000 sweep thermalization.}
+\end{figure}
+We find that the flow time effectively decreases the topological susceptibility by dampening high-momentum modes.
+To analyze the divergence of $\chi_t$ in the continuum limit, we plot $\chi_t \xi_2^2$ as a function of lattice size $L$.
+We perform this simulation at flow times $\tau=0$ and $\tau=5t_0$ (Fig.~\ref{fig:divergence}).
 \begin{figure}[h!]
@@
       \begin{subfigure}[b]{0.45\textwidth}
-            %\includegraphics[height=0.7\textwidth]{divergence.png}
+          \centering
+          \includegraphics[width=\textwidth]{divergence.png}
           \caption{$\tau = 0$}
-        \end{subfigure}%
+      \end{subfigure} %
       \begin{subfigure}[b]{0.45\textwidth}
           \centering
-            \includegraphics[height=0.7\textwidth]{divergence_flowed.png}
+          \includegraphics[width=\textwidth]{divergence_flowed.png}
           \caption{$\tau = 5t_0$}
       \end{subfigure}
+      \caption{\label{fig:divergence} $\chi_t\xi_2^2$ as a function of $L$.
+We fit the data with both a logarithmic and power fit.
+Simulation run with 10,000 measurements, once every 50 sweeps, 1,000 sweep thermalization.
+In the $\tau=0$ case, we have compared our result with the curve fit found in \cite{bietenholz2018}.}
     \end{center}
-        \caption{\label{fig:divergence} $\chi_t\xi_2^2$ as a function of $L$.
-$\xi_2$ is the second moment of the correlation function and $t_0$ is a scale-independent unit of flow time.
-We fit the data with both a logarithmic and power fit.
-Simulation run with 10,000 measurements every 50 sweeps, 1,000 sweep thermalization.
-In the $\tau=0$ case, we have compared our result with the curve fits found in \cite{bietenholz2018}.}
-        %\caption{Divergent properties with
-        %comparison to~\footfullcite{bietenholz2018}, 10,000 measurements}
 \end{figure}
+We fit the data with two options: a log fit
+\begin{equation}
+    \chi_t \xi_2^2 = a \mathrm{log}(b L + c)
+\end{equation}
+and a power law fit
+\begin{equation}
+    \chi_t \xi_2^2 = a L^b + c.
+\end{equation}
+We calculate the parameters to these functions using the \texttt{curve\_fit} tool in the \texttt{scipy} Python package \cite{virtanen2020}.
+When $\tau=5t_0$, the data fits these functions with $\chi^2/DOF$ of \logchi and \powchi respectively, indicating errors were underestimated.
+Both of these functions diverge as $L\rightarrow \infty$, indicating that the topological susceptibility also diverges in the continuum limit.
+Though there is a clear difference between the quantitative fit from \cite{bietenholz2018} and the fit calculated in this work, both demonstrate divergent behavior.
+This result supports the inherent divergence of $\chi_t$ in the continuum limit.
 
-\begin{figure}[h]
+\subsection{Topological charge in nontrivial theory}
+
+Following the method explained in Sec.~\ref{sec:nlsm}, we calculate the imaginary part of $\langle Q \rangle$ for arbitrary $\theta$.
+We perform this calculation for three values of the flow time $\tau$, shown in Fig~\ref{fig:theta}.
+\begin{figure}[h!]
     \centering
-    \includegraphics[width=0.5\textwidth]{theta.png}
-    \caption{\label{fig:theta} Nontrivial $\mathrm{Im}\langle Q \rangle$ as a function of $\theta$.
-Simulation run with 10,000 measurements, every 50 sweeps, 1,000 sweep thermalization.
+      \includegraphics[width=0.6\textwidth]{theta-small.png}
+      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$.
+Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization.
 Note the different scaling of the $y$-axis.}
 \end{figure}
+These plots demonstrate the divergence of the continuum limit in the $\tau=0$ and the flowed regimes.
+In the $\tau=0$ case, the slope increases sharply, reflecting the rapid divergence of $\chi_t$.
+However in the flowed regime, this divergence is much slower, reflecting the decreased values of $\chi_t$ shown in Fig.~\ref{fig:bietenholz}.
+
+\section{Discussion}
+
+Berg \& L\"uscher \cite{berg1981} originally illuminated this discrepancy between the renormalization group hypothesis (that $\chi_t \rightarrow 0$) and the numerical results, providing three possible causes:
+\begin{enumerate}
+    \item The definition of the topological charge does not scale to the continuum.
+    \item There are ultraviolet divergences.
+    \item There is no reasonable continuum limit.
+\end{enumerate}
+Since the gradient flow suppresses ultraviolet fluctuations, the persistence of a divergent topological susceptibility under the gradient flow undermines the second option, thereby supporting the other two.
+
+% Further discussions\footnote{We would like to thank Daniel Nogradi for these conference discussions} \cite{nogradi2012}
 
 \bibliographystyle{JHEP}
-\bibliography{library}
+\bibliography{../library}
 \end{document}

### 2021-10-28 proceedings/draft.tex
fix bibliography issues
fix: small edit

@@
 \newcommand\stuart[1]{\textcolor{pink}{SNT: #1}}
+% \newcommand\chris[1]{}
+% \newcommand\stuart[1]{}
 
@@
     \end{tikzpicture}
-    \caption{Visualization of a plaquette $x^*$.
+    \caption{\label{fig:plaquette} Visualization of a plaquette $x^*$.
 The dotted line separates the plaquette into two signed areas, used to define the topological charge density $q(x^*)$.
@@
 \begin{equation}
-    \label{eq:nsm_gradflow}
+    \label{eq:nlsm_gradflow}
   \partial_\tau \e (\tau,x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T \right) \partial^2 \e(\tau,x).
@@
 
-We apply the gradient flow equation
-\begin{equation}
-    \partial_\tau \e (\tau, x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T\right) \partial^2 \e(\tau,x),
-\end{equation}
-replacing the continuous Laplacian operator $\partial^2$ with a discrete analogue
+We apply the gradient flow equation (Eq.~\ref{eq:nlsm_gradflow}) replacing the continuous Laplacian operator $\partial^2$ with a discrete analogue
 \begin{equation*}
@@
 \end{equation*}
-
-We numerically solve the ordinary differential equation in Eq.~\ref{eq:nsm_gradflow} using a fourth-order Runga-Kutta approximation.
+and numerically solving the differential equation using a fourth-order Runga-Kutta approximation.
 To increase the efficiency of this algorithm, we implement the step-doubling algorithm to adaptively adjust the step size.

### 2021-11-03 proceedings/draft.tex
Updates (up to results section) to draft

@@
 This result supports a previous study and indicates that either the definition of topological charge is problematic or the NLSM has no well-defined continuum limit.
-We also introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow.}
+We also introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow.
+\chris{May need to adjust this to new reference.}}
 
@@
 
+\chris{TO DO:}
+\begin{enumerate}
+ \item \chris{Incorporate new reference from conference.}
+ \item \chris{Double check fits with e.g. lsqfit}
+\end{enumerate}
+
 \section{Introduction}
@@
 In solid-state systems, this model describes Heisenberg ferromagnets \cite{callan1985} and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), the gauge theory of the strong nuclear force.
-In general, the NLSM shares key features with non-Abelian gauge theories such as QCD, including a mass gap and asymptotic freedom \cite{polyakov1975}, and has proved a useful model for exploring the effect of these properties in a simpler system.
+In general, the NLSM shares key features with non-Abelian gauge theories such as QCD, including a mass gap and asymptotic freedom \cite{polyakov1975}, and has proved a useful model for exploring the effect of these properties in simple systems.
 
 We consider the $O(3)$ NLSM in 1+1 dimensions (one dimension of space, one dimension of time).
-This theory exhibits topological properties, such as \textit{instantons}, or classical field solutions at local minima of the action in Euclidean space.
+This theory exhibits topological effects, such as \textit{instantons}, which are classical field solutions at local minima of the action in Euclidean space.
 These topologically protected solutions cannot evolve into the vacuum state via local fluctuations.
-This property has become critically important to quantum field theories in cosmology and high energy physics, relating to the existence of magnetic monopoles\cite{goddard1986} and the mass of the $\eta'$ Goldstone boson \cite{witten1979, veneziano1979} \chris{Can we provide a more explicit example(s) here?
-Statement is very vague at the moment.} Additionally, topological stability may become a key tool for fault-tolerant quantum computers \cite{kitaev1997}.
+This property has become critically important to understanding several applications of quantum field theories in cosmology and high energy physics, such as the existence of magnetic monopoles \cite{goddard1986} and the mass of the $\eta'$ Goldstone boson \cite{witten1979, veneziano1979}.
+Additionally, topological stability may become a key tool for fault-tolerant quantum computers \cite{kitaev1997}.
 In these devices, topology protects the delicate quantum states necessary for information processing.
@@
 The protection of topological instantons in the 1+1 $O(3)$ NLSM relies on a vanishing topological susceptibilty.
-However, the convergence of this quantity is still unknown.
-\cite{bietenholz2018} While the some analytical arguments argue the topological susceptibility should approach zero in the continuum limit, numerical results on the lattice predict infinities \cite{berg1981}.
-In an attempt to shed light on this contradiction, we apply the gradient flow, a local smearing of operators which preserves gauge invariance.
+However, the convergence of this quantity is still unknown \cite{bietenholz2018}.
+Analytical arguments suggest the topological susceptibility should approach zero in the continuum limit, but numerical results on the lattice predict infinities \cite{berg1981}.
+To shed light on this apparent contradiction, we apply the gradient flow, a local smearing of operators which preserves gauge invariance.
 In quantum chromodynamics, this technique has corroborated a previous analytical result \cite{giusti2004} by removing ultraviolet divergences on the lattice \cite{bruno2014}.
@@
 
-To numerically study the topological qualities of the NLSM, we first implement a Markov chain Monte Carlo simulation using Metropolis and Wolff cluster \cite{wolff1989} algorithms.
-Since the gradient flow has no exact solution in the NLSM we implement a numerical solution using a fourth-order Runge-Kutta approximation with automatic step sizing.
-By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.
+To study the topological qualities of the NLSM numerically, we first implement a Markov chain Monte Carlo simulation using Metropolis and Wolff cluster \cite{wolff1989} algorithms.
+The gradient flow has no exact solution in the NLSM, so we implement a numerical solution using a fourth-order Runge-Kutta approximation with automatic step sizing.
+%By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.
 
@@
 \label{sec:nlsm}
-\chris{[In Euclidean space it does not really make sense to distinguish time and space, unless we start in Minkowski spacetime and then Wick rotate.]}
-
-\stuart{I think I resolved this}
 
@@
 Here, $A$ is the signed area of the triangle in target space.
-Visually, the three points in each of the two terms in Eq.~\ref{eq:topodensity} form halves of the plaquette which we represent in Fig.~\ref{fig:plaquette}.
+Visually, the three points in each of the two terms in Eq.~\ref{eq:topodensity} form halves of the plaquette, which we represent in Fig.~\ref{fig:plaquette}.
 The resultant signed area is represented in Fig.~\ref{fig:signedarea}.
@@
 where we have assumed periodic boundary conditions.
-In the continuum limit, the topological susceptibility diverges owing solely to the $x^*=0$ term \cite{bietenholz2018}, a divergence that exists in QCD as well \cite{bruno2014}.
+In the continuum limit, the topological susceptibility diverges owing solely to the $x^*=0$ term \cite{bietenholz2018}, a divergence that also exists in QCD \cite{bruno2014}.
 
@@
 \end{equation}
-This final relation will allow us to prode the behavior of the topological susceptibility in the nontopological NLSM by considering the topological charge in the nontrivial theory.
+This final relation will allow us to probe the behavior of the topological susceptibility in the nontopological NLSM by considering the topological charge in the nontrivial theory.
+\chris{Might want to rephrase this?
+Lots of ``nons'' here.}
 
@@
 \label{sec:gradflow}
-Seeking to resolve the ultraviolet divergence in the $\chi_{t}$, we adopt a technique known as ``smearing'', a local averaging of the field \cite{solbrig2008}.
-Specifically, we use a technique known as the gradient flow \cite{monahan2015} which introduces a new half-dimension\footnote{The term ``half-dimension'' indicates that the flow time is exclusively positive.} called ``flow time''.
+To resolve the ultraviolet divergence in the $\chi_{t}$, we adopt a technique known as ``smearing'', a local averaging of the field \cite{solbrig2008}.
+Specifically, we use the gradient flow \cite{,Narayanan:2006rf,Luscher:2010iy,Luscher:2010we,monahan2015,monahan2016}, which introduces a new half-dimension\footnote{The term ``half-dimension'' indicates that the flow time is exclusively positive.} called ``flow time''.
 The flow time $\tau$ parameterizes the smearing such that an evolution in flow time corresponds to suppressing ultraviolet divergences, pushing field configurations toward classical minima of the action.
 
-We can choose any flow time equation that drives the field towards a classical minimum. Following \cite{bietenholz2018}, we can define the gradient flow for the NLSM via the differential equation
+We can choose any flow time equation that drives the field towards a classical minimum.
+Following \cite{Makino:2014sta,Kikuchi:2014rla,bietenholz2018}, we can define the gradient flow for the NLSM via the differential equation
 \begin{equation}
     \label{eq:nlsm_gradflow}
-  \partial_\tau \e (\tau,x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T \right) \partial^2 \e(\tau,x).
+  \partial_\tau \e (\tau,x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T \right) \partial^2 \e(\tau,x),
 \end{equation}
-We solve this equation numerically using the boundary condition $\e(\tau=0,x) = \e(x)$, a process described in Sec.~\ref{sec:numerics}.
+which we solve numerically with the boundary condition $\e(\tau=0,x) = \e(x)$.
 
@@
 where $a$ is a lattice constant and $\hat \mu$ are the Euclidean unit vectors.
-We generate configurations using the Metropolis \stuart{Does this need to be cited?} and Wolff cluster \cite{wolff1989} algorithms.
+We generate configurations using the Metropolis \stuart{Does this need to be cited?} \chris{Technically probably yes, but in practice no.
+If you want to look up the reference, we can include it.} and Wolff cluster \cite{wolff1989} algorithms.
 

### 2021-11-04 proceedings/draft.tex
Final comments added on v1

@@
 \begin{enumerate}
- \item \chris{Incorporate new reference from conference.}
- \item \chris{Double check fits with e.g. lsqfit}
+ \item \chris{Stuart: Address comments in figures.}
+ \item \chris{Chris: Incorporate new reference from conference.}
+ \item \chris{Chris: Double check fits with e.g. lsqfit}
 \end{enumerate}
@@
 \label{sec:results}
-In order to confirm the accuracy of the model, we compare results with existing literature.
-We first consider the results of Berg \& L\"uscher \cite{berg1981}, specifically the internal energy and magnetic susceptibility.
+
+We check our numerical implementation of the NLSM by comparison to studies of the internal energy and magnetic susceptibility in Refs.~\cite{berg1981} and \cite{bietenholz2018}.
 Following \cite{berg1981}, we approximate the internal energy in the strong ($\beta<1$) and weak ($\beta>2$) regimes as
@@
     4-4y-8y^3-\frac{48}{5}y^5& \beta<1 \\
-\frac{2}{\beta} + \frac{4}{\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2
+\frac{2}{\beta} + \frac{4}{\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2,
 \end{cases}
@@
     \centering
-      \includegraphics[width=0.6\textwidth]{internal_energy.png}
-      \caption{\label{fig:bergluscher}Comparison with \cite{berg1981}.
+      \includegraphics[width=0.45\textwidth]{int_energy-0.png}
+      \includegraphics[width=0.45\textwidth]{int_energy-1.png}
+      \caption{\label{fig:bergluscher} Comparison with \cite{berg1981}.
 First panel: internal energy compared with analytic  energy (Eq.~\ref{eq:analytic_energy}).
-Second panel: magnetic susceptibility compared with literature values.}
+Second panel: magnetic susceptibility compared with literature values.
+\chris{Can you please remake these plots so that they both have x-axis labels?
+Please also make the labels on both axes bigger.}}
 \end{figure}
-These two charts show a high degree of agreement with the current literature however there is a slight discrepancy, prehaps arising from different Monte Carlo methods.
+The slight discrepancy between numerical results in the cross-over region provides an estimate of unquantified systematic uncertainties.
 
@@
 \cite{bietenholz2018}.
-Specifically, we show the topological susceptibility $\chi_t$ diverges in the continuum limit even at finite flow time.
-Since $\chi_t$ is in units of inverse distance squared, we multiply by $\xi_2^2$, the square of the second moment correlation length, to achieve a scale-invariant value $\chi_t\xi_2^2$.
+In addition, we show the topological susceptibility $\chi_t$ diverges in the continuum limit even at finite flow time.
+Since $\chi_t$ is in units of inverse distance squared, we multiply by , to achieve a scale-invariant value $\chi_t\xi_2^2$.
 Additionally, we use a parameter $t_0$ to scale the flow time such that $t_0\sim L^2$.
@@
 
-To begin the comparison, we plot $\chi_t\xi_2^2$ as a function of flow time $\tau$, shown in Fig.~\ref{fig:bietenholz}.
+In Fig.~\ref{fig:bietenholz} we plot the dimensionless variable $\chi_t\xi_2^2$, where $\xi_2^2$ is the square of the second moment correlation length, as a function of flow time $\tau/t_0$, with $t_0 \sim L^2$.
 \begin{figure}[h!]
@@
       \includegraphics[width=0.8\textwidth]{bietenholz.png}
-      \caption{\label{fig:bietenholz} $\chi_t L^2$ as a  function of flow time $\tau$. Simulation run with 10,000 measurements every 50 sweeps, 1,000 sweep thermalization.}
+      \caption{\label{fig:bietenholz} Dimensionless magnetic susceptibility, $\chi_t L^2$, \chris{In the text/y-axis label you say this is $\chi_t\xi_2^2$?} as a  function of flow time $\tau/t_0$.
+Simulation run with 10,000 measurements every 50 sweeps, 1,000 sweep thermalization.
+\chris{Please make axis labels bigger.}}
 \end{figure}
-We find that the flow time effectively decreases the topological susceptibility by dampening high-momentum modes.
-To analyze the divergence of $\chi_t$ in the continuum limit, we plot $\chi_t \xi_2^2$ as a function of lattice size $L$.
-We perform this simulation at flow times $\tau=0$ and $\tau=5t_0$ (Fig.~\ref{fig:divergence}).
+Fig.~\ref{fig:bietenholz} shows that the flow time effectively decreases the topological susceptibility by dampening high-momentum modes.
+To analyze the divergence of $\chi_t$ in the continuum limit, we plot $\chi_t \xi_2^2$ as a function of lattice size $L$, for $\tau=0$ and $\tau=5t_0$, in Fig.~\ref{fig:divergence}.
 \begin{figure}[h!]
@@
 Simulation run with 10,000 measurements, once every 50 sweeps, 1,000 sweep thermalization.
-In the $\tau=0$ case, we have compared our result with the curve fit found in \cite{bietenholz2018}.}
+In the $\tau=0$ case, we have compared our result with the curve fit found in \cite{bietenholz2018}.
+\chris{Please make axis labels bigger.}}
     \end{center}
 \end{figure}
-We fit the data with two options: a log fit
+We fit the data with two fit functions: a log fit
 \begin{equation}
-    \chi_t \xi_2^2 = a \mathrm{log}(b L + c)
+    \chi_t \xi_2^2 = a \mathrm{log}(b L + c);
 \end{equation}
@@
 This result supports the inherent divergence of $\chi_t$ in the continuum limit.
+\chris{I need to think more about how to address this discrepancy.}
 
@@
 
-Following the method explained in Sec.~\ref{sec:nlsm}, we calculate the imaginary part of $\langle Q \rangle$ for arbitrary $\theta$.
-We perform this calculation for three values of the flow time $\tau$, shown in Fig~\ref{fig:theta}.
+We calculate the imaginary part of $\langle Q \rangle$ for arbitrary $\theta$ and plot the results for three values of the flow time $\tau$ in Fig~\ref{fig:theta}.
 \begin{figure}[h!]
@@
 \section{Discussion}
+
+\chris{This discussion will need to be updated to incorporate the new reference from our lattice conference discussions.}
 

### 2021-11-05 proceedings/draft.tex
Adjust scale of figures

@@
     \centering
-      \includegraphics[width=0.45\textwidth]{int_energy-0.png}
-      \includegraphics[width=0.45\textwidth]{int_energy-1.png}
+      % \includegraphics[width=0.45\textwidth]{int_energy-0.png}
+      % \includegraphics[width=0.45\textwidth]{int_energy-1.png}
+      \includegraphics[width=0.9\textwidth]{berg_luscher.png}
       \caption{\label{fig:bergluscher} Comparison with \cite{berg1981}.
 First panel: internal energy compared with analytic  energy (Eq.~\ref{eq:analytic_energy}).
-Second panel: magnetic susceptibility compared with literature values.
-\chris{Can you please remake these plots so that they both have x-axis labels?
-Please also make the labels on both axes bigger.}}
+Second panel: magnetic susceptibility compared with literature values.}
 \end{figure}
@@
       \includegraphics[width=0.8\textwidth]{bietenholz.png}
-      \caption{\label{fig:bietenholz} Dimensionless magnetic susceptibility, $\chi_t L^2$, \chris{In the text/y-axis label you say this is $\chi_t\xi_2^2$?} as a  function of flow time $\tau/t_0$.
+      \caption{\label{fig:bietenholz} Dimensionless magnetic susceptibility, $\chi_t \xi_{2}^2$, as a  function of flow time $\tau/t_0$.
 Simulation run with 10,000 measurements every 50 sweeps, 1,000 sweep thermalization.
-\chris{Please make axis labels bigger.}}
+}
 \end{figure}
@@
           \centering
-          \includegraphics[width=\textwidth]{divergence.png}
+          \includegraphics[height=0.8\textwidth]{divergence.png}
           \caption{$\tau = 0$}
@@
           \centering
-          \includegraphics[width=\textwidth]{divergence_flowed.png}
+          \includegraphics[height=0.8\textwidth]{divergence_flowed.png}
           \caption{$\tau = 5t_0$}
@@
 Simulation run with 10,000 measurements, once every 50 sweeps, 1,000 sweep thermalization.
-In the $\tau=0$ case, we have compared our result with the curve fit found in \cite{bietenholz2018}.
-\chris{Please make axis labels bigger.}}
+In the $\tau=0$ case, we have compared our result with the curve fit found in \cite{bietenholz2018}.}
     \end{center}

### 2021-11-08 proceedings/draft.tex
addressed a few notes

@@
 \end{equation}
-This final relation will allow us to probe the behavior of the topological susceptibility in the nontopological NLSM by considering the topological charge in the nontrivial theory.
-\chris{Might want to rephrase this?
-Lots of ``nons'' here.}
+This final relation will allow us to probe the behavior of the topological susceptibility in the trivial NLSM by considering the topological charge in the nontrivial theory.
 
@@
 \end{equation}
-where $a$ is a lattice constant and $\hat \mu$ are the Euclidean unit vectors.
-We generate configurations using the Metropolis \stuart{Does this need to be cited?} \chris{Technically probably yes, but in practice no.
-If you want to look up the reference, we can include it.} and Wolff cluster \cite{wolff1989} algorithms.
+where $a$ is a lattice constant and $\hat \mu$ are the Euclidean unit vectors. We generate configurations using the Metropolis \cite{metropolis1953} and Wolff cluster \cite{wolff1989} algorithms.
 

### 2021-11-10 proceedings/draft.tex
Incorporated Daniel's paper and removed general comments

@@
 \usepackage{subcaption}
+\usepackage{enumitem}
 \newcommand{\e}{\vec e}
@@
 
-% \affiliation[b]{Thomas Jefferson National Accelerator Facility, Newport News, Virginia, USA}
-
 \emailAdd{snthomas01@email.wm.edu}
@@
 The $O(3)$ non-linear sigma model (NLSM) is a prototypical field theory for QCD and ferromagnetism, featuring topological qualities.
-Though the topological susceptibility $\chi_t$ should vanish in physical theories, lattice simulations of the NLSM find that $\chi_t$ diverges in the continuum limit.
-We study the effect of the gradient flow on this quantity using a Markov Chain Monte Carlo method, finding that a logarithmic divergence persists.
-This result supports a previous study and indicates that either the definition of topological charge is problematic or the NLSM has no well-defined continuum limit.
-We also introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow.
-\chris{May need to adjust this to new reference.}}
+Though the topological susceptibility should vanish in physical theories, lattice simulations of the NLSM find that the topological susceptibility diverges in the continuum limit.
+We introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow.
+Our results show that divergence persists in the presence of the flow, even at non-zero $\theta$.
+}
 
@@
 \begin{enumerate}
- \item \chris{Stuart: Address comments in figures.}
- \item \chris{Chris: Incorporate new reference from conference.}
+ \item \chris{Stuart: check reference highlighted below in introduction}
  \item \chris{Chris: Double check fits with e.g. lsqfit}
@@
 The protection of topological instantons in the 1+1 $O(3)$ NLSM relies on a vanishing topological susceptibilty.
-However, the convergence of this quantity is still unknown \cite{bietenholz2018}.
-Analytical arguments suggest the topological susceptibility should approach zero in the continuum limit, but numerical results on the lattice predict infinities \cite{berg1981}.
-To shed light on this apparent contradiction, we apply the gradient flow, a local smearing of operators which preserves gauge invariance.
+However, the convergence properties of this quantity are still unclear \cite{bietenholz2018}.
+Analytical arguments suggest the topological susceptibility should approach zero in the continuum limit \cite{berg1981} \chris{Can you please confirm this is the correct reference for the origin of these ``analytical arguments''?}, but numerical results on the lattice, summarized in \cite{bietenholz2018}, indicate a logarithmic divergence and support the semi-classical picture of small-size instantons generating the divergence.
+To elucidate this apparent contradiction, we apply the gradient flow, a local smearing of operators which preserves gauge invariance.
 In quantum chromodynamics, this technique has corroborated a previous analytical result \cite{giusti2004} by removing ultraviolet divergences on the lattice \cite{bruno2014}.
-This success has motivated the gradient flow to calculate the topological susceptibility in the 1+1 $O(3)$ NLSM.
-Despite this intuition, recent studies demonstrate that the observable still diverges in the continuum limit \cite{bietenholz2018}.
+This success has motivated the gradient flow to as a tool to define the continuum limit of the topological susceptibility in the 1+1 $O(3)$ NLSM.
+Despite this intuition, recent studies demonstrate that the topological susceptibility still diverges in the continuum limit \cite{bietenholz2018}.
 
 We also study a second perspective on the topological susceptibility arising from the introduction of a $\theta$-term into the field Lagrangian.
-This term drives the vacuum state into a topological phase \cite{allessalom2008}.
-Differentiating the field's partition function with respect to $\theta$ yields a value proportional to the topological susceptibility.
-The effect of nonzero $\theta$ on the theory therefore should reflect the divergence in the continuum limit.
+This term drives the vacuum state into a topological phase \cite{allessalom2008}, and the resulting theory exhibits confining, walking, and conformal behaviors at different values of $\theta$ \cite{Nogradi:2012dj}.
+Differentiating the partition function with respect to $\theta$ yields a value proportional to the topological susceptibility.
+The effect of nonzero $\theta$ on the theory therefore should reflect the divergence of the susceptibility in the continuum limit.
 In this work we verify the divergence of the topological susceptibility and develop a clearer picture of how the $\theta$-term affects the topology of the 1+1 $O(3)$ NLSM.
@@
 In the continuum limit, the topological susceptibility diverges owing solely to the $x^*=0$ term \cite{bietenholz2018}, a divergence that also exists in QCD \cite{bruno2014}.
+In the semiclassical approximation, this divergence arises from small-size instantons \cite{Luscher:1981tq}.
 
@@
       \hfill
-      \caption{\label{fig:therm} Plots of the action as a function of Monte Carlo time, starting with a random NLSM lattice.}
+      \caption{\label{fig:therm} Plots of the action, $S$, as a function of Monte Carlo time, starting with a random NLSM lattice.}
 \end{figure}
@@
 \end{equation}
-where
-\begin{equation}
-    y=\mathrm{coth}\beta-\frac{1}{\beta}.
-\end{equation}
+where $y=\operatorname{coth}\,\beta-1/\beta$.
 We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{fig:bergluscher}.
@@
       % \includegraphics[width=0.45\textwidth]{int_energy-1.png}
-      \includegraphics[width=0.9\textwidth]{berg_luscher.png}
+      \includegraphics[width=0.8\textwidth]{berg_luscher.png}
       \caption{\label{fig:bergluscher} Comparison with \cite{berg1981}.
@@
 
-\section{Discussion}
+\section{Summary}
 
-\chris{This discussion will need to be updated to incorporate the new reference from our lattice conference discussions.}
-
-Berg \& L\"uscher \cite{berg1981} originally illuminated this discrepancy between the renormalization group hypothesis (that $\chi_t \rightarrow 0$) and the numerical results, providing three possible causes:
-\begin{enumerate}
+The topological behavior of the non-linear sigma model has been a topic of debate for several decades.
+Berg and L\"uscher \cite{berg1981} originally highlighted the discrepancy between the renormalization group hypothesis, that the topological susceptibility is zero in the continuum limit, and the numerical results of nonperturbative calculations, supported by a semi-classical instanton picture, which indicate that the topological susceptibility is ill-defined in the continuum limit.
+They suggested three possible causes:
+\begin{enumerate}[itemsep=-0.5em]
     \item The definition of the topological charge does not scale to the continuum.
@@
 \end{enumerate}
-Since the gradient flow suppresses ultraviolet fluctuations, the persistence of a divergent topological susceptibility under the gradient flow undermines the second option, thereby supporting the other two.
+Our results at finite gradient flow time, for both $\theta = 0$ and at non-zero $\theta$,  support the numerical results of \cite{bietenholz2018} at $\theta=0$ and indicate the presence of a divergence in the continuum limit.
+Pre-reflectively, this may suggest that ultraviolet divergences are not the origin of this divergence, because the flow suppresses ultraviolet fluctuations and introduces no new divergences in the renormalized NLSM \cite{makino2015a}.
+The story of this divergence is, however, more subtle and is consistent with divergences at finite flow time.
+The topological susceptibility (and all other even moments of the topological charge) are divergent in the continuum limit, differences between cumulants of the topological charge are finite \cite{Nogradi:2012dj}.
+This indicates that there is a single, non-perturbatively generated divergence that can be removed by introducing a single counter term in the Lagrangian, proportional to the identity.
+In the absence of any such counter-term, the NLSM at zero flow time is not completely renormalized, and introducing the flow cannot remove this divergence.
+One future application of the gradient flow to the study of the topology of the NLSM, could be, however, to define a nonperturbative renormalization scheme for this counter term, and study the corresponding running of its coupling.
 
-% Further discussions\footnote{We would like to thank Daniel Nogradi for these conference discussions} \cite{nogradi2012}
+\section{Acknowledgments}
+
+We thank Daniel Nogradi for enlightening discussions on the NLSM at nonzero $\theta$ during this conference, and for drawing our attention to  \cite{Nogradi:2012dj}.
 

### 2021-11-12 proceedings/draft.tex
Switch curve fit to 'lsqfit'

@@
 \usepackage{enumitem}
+
 \newcommand{\e}{\vec e}
-\newcommand{\logchi}{7.4 }
-\newcommand{\powchi}{8.8 }
+\newcommand{\logchi}{6.2 }
+\newcommand{\powchi}{5.5 }
 \graphicspath{{../figures/}}
@@
 However, the convergence properties of this quantity are still unclear \cite{bietenholz2018}.
-Analytical arguments suggest the topological susceptibility should approach zero in the continuum limit \cite{berg1981} \chris{Can you please confirm this is the correct reference for the origin of these ``analytical arguments''?}, but numerical results on the lattice, summarized in \cite{bietenholz2018}, indicate a logarithmic divergence and support the semi-classical picture of small-size instantons generating the divergence.
+Analytical arguments suggest the topological susceptibility should approach zero in the continuum limit \cite{berg1981}, but numerical results on the lattice, summarized in \cite{bietenholz2018}, indicate a logarithmic divergence and support the semi-classical picture of small-size instantons generating the divergence.
 To elucidate this apparent contradiction, we apply the gradient flow, a local smearing of operators which preserves gauge invariance.
@@
 \end{equation}
-We calculate the parameters to these functions using the \texttt{curve\_fit} tool in the \texttt{scipy} Python package \cite{virtanen2020}.
+We calculate the parameters to these functions using the \texttt{lsqfit} Python package \cite{lsqfit}.
 When $\tau=5t_0$, the data fits these functions with $\chi^2/DOF$ of \logchi and \powchi respectively, indicating errors were underestimated.

### 2021-11-16 proceedings/draft.tex
Final changes

@@
 \newcommand\stuart[1]{\textcolor{pink}{SNT: #1}}
-% \newcommand\chris[1]{}
-% \newcommand\stuart[1]{}
+
+\let\OLDthebibliography\thebibliography
+\renewcommand\thebibliography[1]{
+  \OLDthebibliography{#1}
+  \setlength{\parskip}{0pt}
+  \setlength{\itemsep}{0pt plus 0.3ex}
+}
 
@@
 \maketitle
-
-\chris{TO DO:}
-\begin{enumerate}
- \item \chris{Stuart: check reference highlighted below in introduction}
- \item \chris{Chris: Double check fits with e.g. lsqfit}
-\end{enumerate}
 
@@
 \end{equation}
-We calculate the parameters to these functions using the \texttt{lsqfit} Python package \cite{lsqfit}.
-When $\tau=5t_0$, the data fits these functions with $\chi^2/DOF$ of \logchi and \powchi respectively, indicating errors were underestimated.
-Both of these functions diverge as $L\rightarrow \infty$, indicating that the topological susceptibility also diverges in the continuum limit.
-Though there is a clear difference between the quantitative fit from \cite{bietenholz2018} and the fit calculated in this work, both demonstrate divergent behavior.
-This result supports the inherent divergence of $\chi_t$ in the continuum limit.
-\chris{I need to think more about how to address this discrepancy.}
-
-\subsection{Topological charge in nontrivial theory}
+We calculate the parameters to these functions using two independent fitting methods, the \texttt{lsqfit} package \cite{lsqfit} and the \texttt{scipy} \texttt{curve\_fit} tool \cite{virtanen2020}.
+We quote results from \cite{lsqfit}.
+For flow time $\tau=5t_0$, the $\chi^2/DOF$  for the logarithmic and power fits are \logchi and \powchi respectively, which potentially indicates underestimated systematic uncertainties.
+Both of these functions diverge as $L\rightarrow \infty$, indicating that the topological susceptibility also diverges in the continuum limit, in line with the results of \cite{bietenholz2018}.
+The origin of the discrepancies between our results and \cite{bietenholz2018} will be studied in future work.
 
@@
 The story of this divergence is, however, more subtle and is consistent with divergences at finite flow time.
-The topological susceptibility (and all other even moments of the topological charge) are divergent in the continuum limit, differences between cumulants of the topological charge are finite \cite{Nogradi:2012dj}.
-This indicates that there is a single, non-perturbatively generated divergence that can be removed by introducing a single counter term in the Lagrangian, proportional to the identity.
+The topological susceptibility (and all other even moments of the topological charge) are divergent in the continuum limit, while differences between cumulants of the topological charge are finite \cite{Nogradi:2012dj}.
+This indicates that there is a single, non-perturbatively generated divergence that can be removed by introducing a single counter-term in the Lagrangian, proportional to the identity.
 In the absence of any such counter-term, the NLSM at zero flow time is not completely renormalized, and introducing the flow cannot remove this divergence.
@@
 
+\setlength{\itemsep}{-2pt}
 \bibliographystyle{JHEP}

### 2021-11-22 proceedings/draft.tex
Add Daniel's notes, check spelling

@@
 The $O(3)$ non-linear sigma model (NLSM) is a prototypical field theory for QCD and ferromagnetism, featuring topological qualities.
-Though the topological susceptibility should vanish in physical theories, lattice simulations of the NLSM find that the topological susceptibility diverges in the continuum limit.
+Though the topological susceptibility should approach zero in physical theories, lattice simulations of the NLSM find that the topological susceptibility diverges in the continuum limit.
 We introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow.
@@
 
-The protection of topological instantons in the 1+1 $O(3)$ NLSM relies on a vanishing topological susceptibilty.
+The protection of topological instantons in the 1+1 $O(3)$ NLSM relies on a vanishing topological susceptibility.
 However, the convergence properties of this quantity are still unclear \cite{bietenholz2018}.
@@
 We also study a second perspective on the topological susceptibility arising from the introduction of a $\theta$-term into the field Lagrangian.
-This term drives the vacuum state into a topological phase \cite{allessalom2008}, and the resulting theory exhibits confining, walking, and conformal behaviors at different values of $\theta$ \cite{Nogradi:2012dj}.
+This term drives the vacuum state into a topological phase \cite{allessalom2008}, and the resulting theory may exhibit confining, walking, and conformal behaviors at different values of $\theta$ \cite{Nogradi:2012dj}.
 Differentiating the partition function with respect to $\theta$ yields a value proportional to the topological susceptibility.
@@
 The topological susceptibility (and all other even moments of the topological charge) are divergent in the continuum limit, while differences between cumulants of the topological charge are finite \cite{Nogradi:2012dj}.
-This indicates that there is a single, non-perturbatively generated divergence that can be removed by introducing a single counter-term in the Lagrangian, proportional to the identity.
+This indicates that there is a single, nonperturbatively generated divergence that can be removed by introducing a single counter-term in the Lagrangian, proportional to the identity.
 In the absence of any such counter-term, the NLSM at zero flow time is not completely renormalized, and introducing the flow cannot remove this divergence.

### 2021-11-22 proceedings/draft.tex
Updates, based on Daniel's suggestions

@@
 \abstract{
-The $O(3)$ non-linear sigma model (NLSM) is a prototypical field theory for QCD and ferromagnetism, featuring topological qualities.
-Though the topological susceptibility should approach zero in physical theories, lattice simulations of the NLSM find that the topological susceptibility diverges in the continuum limit.
+The $O(3)$ non-linear sigma model (NLSM) is a prototypical field theory for QCD and ferromagnetism, and provides a simple system in which to study topological effects.
+In lattice QCD, the gradient flow has been demonstrated to remove ultraviolet singularities from the topological susceptibility.
+In contrast, lattice simulations of the NLSM find that the topological susceptibility diverges in the continuum limit, even in the presence of the gradient flow.
 We introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow.
@@
 We consider the $O(3)$ NLSM in 1+1 dimensions (one dimension of space, one dimension of time).
-This theory exhibits topological effects, such as \textit{instantons}, which are classical field solutions at local minima of the action in Euclidean space.
+This theory may exhibit topological effects, such as \textit{instantons}, which are classical field solutions at local minima of the action in Euclidean space.
 These topologically protected solutions cannot evolve into the vacuum state via local fluctuations.
@@
 
-We thank Daniel Nogradi for enlightening discussions on the NLSM at nonzero $\theta$ during this conference, and for drawing our attention to  \cite{Nogradi:2012dj}.
+\noindent We thank Daniel Nogradi for enlightening discussions on the NLSM at nonzero $\theta$ during this conference, for drawing our attention to \cite{Nogradi:2012dj}, and for reading an early version of the manuscript.
 

### 2021-11-29 proceedings/draft.tex
Added acknowledgments
New reference added and discussed

@@
 \usepackage{enumitem}
+\usepackage{verbatim}
 
@@
 \graphicspath{{../figures/}}
-\newcommand\chris[1]{\textcolor{teal}{CJM: #1}}
-\newcommand\stuart[1]{\textcolor{pink}{SNT: #1}}
 
@@
   \setlength{\parskip}{0pt}
-  \setlength{\itemsep}{0pt plus 0.3ex}
+  \setlength{\itemsep}{0pt plus 0.1ex}
 }
@@
 The protection of topological instantons in the 1+1 $O(3)$ NLSM relies on a vanishing topological susceptibility.
-However, the convergence properties of this quantity are still unclear \cite{bietenholz2018}.
+However, the convergence properties of this quantity are still unclear \cite{bietenholz2018,Berni:2020ebn}.
 Analytical arguments suggest the topological susceptibility should approach zero in the continuum limit \cite{berg1981}, but numerical results on the lattice, summarized in \cite{bietenholz2018}, indicate a logarithmic divergence and support the semi-classical picture of small-size instantons generating the divergence.
+Similar, but ultimately inconclusive, results were found for the equivalent CP$^{1}$ model in the presence of an alternative smoothing procedure, ``cooling'', in \cite{Berni:2020ebn}.
 To elucidate this apparent contradiction, we apply the gradient flow, a local smearing of operators which preserves gauge invariance.
@@
 This success has motivated the gradient flow to as a tool to define the continuum limit of the topological susceptibility in the 1+1 $O(3)$ NLSM.
-Despite this intuition, recent studies demonstrate that the topological susceptibility still diverges in the continuum limit \cite{bietenholz2018}.
+Despite this intuition, recent studies demonstrate that the topological susceptibility still diverges in the continuum limit in the presence of the ultraviolet smearing procedures \cite{bietenholz2018,Berni:2020ebn}.
 
@@
 They suggested three possible causes:
+\vspace*{-0.5\baselineskip}
 \begin{enumerate}[itemsep=-0.5em]
@@
 \end{enumerate}
+
 Our results at finite gradient flow time, for both $\theta = 0$ and at non-zero $\theta$,  support the numerical results of \cite{bietenholz2018} at $\theta=0$ and indicate the presence of a divergence in the continuum limit.
@@
 \noindent We thank Daniel Nogradi for enlightening discussions on the NLSM at nonzero $\theta$ during this conference, for drawing our attention to \cite{Nogradi:2012dj}, and for reading an early version of the manuscript.
+CJM is supported in part by USDOE grant No.~DE-AC05-06OR23177, under which Jefferson Science Associates, LLC, manages and operates Jefferson Lab.
 
